### 2005-11-15 inst/doc/MVCClass.Rnw
Added a vignette.
git-svn-id: file:///home/git/hedgehog.fhcrc.org/bioconductor/trunk/madman/Rpacks/MVCClass@14838 bc3139a8-67e5-0310-9ffc-ced21a209358

%
% NOTE -- ONLY EDIT THE .Rnw FILE!!!  The .tex file is
% likely to be overwritten.
%
%\VignetteDepends{RGtk, gtkDevice, graph, Rgraphviz}
%\VignetteKeywords{Model, View, Controller}
%\VignettePackage{MVCClass}
%\VignetteIndexEntry{MVCClass}

\documentclass[11pt]{article}

\usepackage{amsmath,pstricks}
\usepackage[authoryear,round]{natbib}
\usepackage{hyperref}
\usepackage{times}
\usepackage{comment}
\usepackage{epsfig}
\usepackage{graphics}
\usepackage{graphicx}

\textwidth=6.2in
\textheight=8.5in
%\parskip=.3cm
\oddsidemargin=.1in
\evensidemargin=.1in
\headheight=-.3in

\newcommand{\scscst}{\scriptscriptstyle}
\newcommand{\scst}{\scriptstyle}

\newcommand{\Rfunction}[1]{{\texttt{#1}}}
\newcommand{\Robject}[1]{{\texttt{#1}}}
\newcommand{\Rpackage}[1]{{\textit{#1}}}

\bibliographystyle{plainnat}

\title{Model-View-Controller Package}
\author{Elizabeth Whalen}
\date{November 11, 2005}

\begin{document}
\maketitle

\section{Overview}\label{Sec:Overview}

The purpose of the \Rpackage{MVCClass} package is to provide the definitions of
classes and generic functions that will be used by other packages to create 
model-view-controller applications.  Currently, the packages that use the 
\Rpackage{MVCClass} package are \Rpackage{iSPlot} and \Rpackage{iSNetwork}.  
Both packages use the definitions from \Rpackage{MVCClass} to create linked, 
interactive views of data.

\section{MVC Class}\label{Sec:MVC}

The \Robject{MVC} class represents a model-view-controller object.  Basically, 
this class will bind the model with its views and its controller so that 
an object of this class can be created each time there is a new data set 
(model).  Thus, there is a one-to-one relationship between a MVC object and
a model object.  Because of this relationship, the
MVC object and the model object are identified by the same name,
which is stored in the modelName slot of the model object (see Section 
\ref{Sec:Model}).  The figure below 
shows the class definition for \Robject{MVC}.

\begin{figure}[ht]
  \begin{center}
    \scalebox{0.3}{\includegraphics{MVCClass.jpg}}
    \caption{ The MVC Class. }
    \label{Fig:MVCClass}
  \end{center}
\end{figure}

The \Robject{MVC} class has five slots which are model, viewList, controller, 
parentMVC, and childMVCList.  The
model slot stores the model object, the viewList slot is a list of the view
objects for this model, the controller slot is an environment that stores
information for this MVC, the parentMVC slot is the name of the parent model
(MVC), and the childMVCList slot is a list of the children models
(MVCs).  Note that model and MVC objects use the same name so the
parentMVC slot refers to both the name of the model and the name of the
MVC.  Similarly, for the childMVCList slot, the names in the list
refer to both the name of the model and the name of the MVC.

Thus, the \Robject{MVC} class not only binds the model, view, and controller
objects together, but MVC objects can be related to each through a 
parent-child (tree) hierarchy because of the slots, parentMVC and 
childMVCList. 
These two slots also show that a MVC object can have at most one parent MVC and
can have as many children MVCs as it wants.

\section{Model Classes}\label{Sec:Model}

The model class is responsible for storing and updating the data.  These two
functions are reflected in Figure \ref{Fig:Model}, which shows the inheritance
structure for the model classes.  Here the virtual class,
\Robject{gModel}, has the slots modelData, linkData, virtualData, and
modelName.  These four slots are common information that all models will
need.  The modelData slot is the data set for this model, the linkData slot is
a list of two functions, \Rfunction{linkToParent} and \Rfunction{linkToChild},
which link this model to its parent and child models, respectively (see Section
\ref{Sec:MVC}), the virtualData slot is data pertaining to the views
that needs to be stored with the model so that it can be shown in all views of
this model, and the modelName slot is the name of the model.

\begin{figure}[ht]
  \begin{center}
    \scalebox{0.7}{\includegraphics{VigModelClass.jpg}}
    \caption{ Inheritance for Model Objects. }
    \label{Fig:Model}
  \end{center}
\end{figure}

For specific model classes, there are \Robject{dfModel}, \Robject{graphModel},
and \Robject{exprModel}.  The \Robject{dfModel} class represents a model where
the modelData has class data frame (or matrix), the \Robject{graphModel} class
represents a model where the modelData has class graph, and the
\Robject{exprModel} class represents a model where the modelData has class
exprSet.

As mentioned previously, the model classes are also responsible for updating
the data so a generic function \Rfunction{updateModel} has also been 
defined.  The \Rfunction{updateModel} methods will be defined in packages that 
use this package (i.e. the methods will be defined in \Rpackage{iSNetwork} and 
\Rpackage{iSPlot}).

\section{View Classes}\label{Sec:View}

The view classes represent the visual depictions of the model.  All views will
need to store some common information and respond to certain events through
methods.  This consideration led to an object model where the different view
classes inherit from a view virtual class, called \Robject{genView}.  The
object model for the view classes is shown in Figure \ref{Fig:View}.

\begin{figure}[ht]
  \begin{center}
    \scalebox{0.7}{\includegraphics{VigViewClass.jpg}}
    \caption{ Inheritance for View Objects. }
    \label{Fig:View}
  \end{center}
\end{figure}

The \Robject{genView} class has three slots, dataName, win, and winNum.
dataName is the name of the model that the view displays, win is the Gtk
window object that holds the view, and winNum is the number of the window so
that the window can be identified in the GUI that other packages create.  All 
views will need to know these three pieces of information so \Robject{genView} 
will bind the view classes together.

As for specific view classes, the \Robject{spreadView} class will represent a
spreadsheet view of the data.  This view will only make sense for models that
are a two dimensional data structure, such as a matrix or a data frame.  The
information needed for this view is stored in the slot, clist, which is the
Gtk spreadsheet object.

When creating a plot of a data set, there is a general class, called
\Robject{plotView}, that has the slots, plotDevice, plotPar, and drArea, which
store the device number of the plot, the plotting parameters, and the Gtk
drawing area object, respectively.  This class is not intended to have any
objects as it just represents a general plot.  The plot classes that can have
objects are the specific plot classes, \Robject{sPlotView},
\Robject{graphView}, and \Robject{heatmapView}, which inherit from the
\Robject{plotView} class.

The \Robject{sPlotView} class represents a scatterplot view and it has the
extra slots, dfRows, colx, and coly, where dfRows are the row names from the
model that are shown in the plot, colx is the column name from the model that
is shown as the x variable in the plot, and coly is the column name from the
model that is shown as the y variable in the plot.  The \Robject{graphView}
class represents a graph plot view and it contains no extra slots besides
those in the \Robject{plotView} class.  The \Robject{heatmapView} represents a
heatmap view of the data and it contains the slot, ordering, which is a list
that is returned from the \Rfunction{heatmap} function to give information
about the dendrogram ordering.  Note that some of these views are only
applicable for certain model types. For example, the \Robject{graphView} class
will only make sense for a view of a graph object, which would be stored in the
\Robject{graphModel} class.

Several generic functions have been defined that refer to operations 
performed on views.  These generic functions are \Rfunction{motionEvent}, 
\Rfunction{clickEvent}, \Rfunction{updateView}, and \Rfunction{redrawView}.  
Methods for these generic 
functions will be defined in packages that use this package (in 
\Rpackage{iSPlot} and \Rpackage{iSNetwork}).  When these methods are defined, 
the \Rfunction{motionEvent} method will be used to respond to a mouse over 
event on a view, the \Rfunction{clickEvent} method will be used to respond to 
a click event over a view, the \Rfunction{updateView} method will be used to 
update only a portion of the view, and the \Rfunction{redrawView} method will 
be used to completely redraw the view.

\section{Message Classes}\label{Sec:Message}

The message classes are intended to provide communication between different
components of the MVC design, such as when the model changes a message needs to
be sent to the views to let them know that they should be updated.  This
communication between the model, view and controller is crucial for the pieces
to work together and yet, still be independent of each other.

The object model for the message classes was derived from the idea that all
messages must have certain methods in common, such as \Rfunction{initialize}
and \Rfunction{handleMessage}, because all messages must be created and
handled in some manner so that the message is read and acted upon.  Also,
since messages are sent when something has changed, either through addition,
alteration, or deletion, the message classes reflect only certain operations.
These common purposes for the messages allow the message object model to be
constructed, and this model is shown in Figure \ref{Fig:Mess}.

\begin{figure}[ht]
  \begin{center}
    \scalebox{0.7}{\includegraphics{VigMessageClass.jpg}}
    \caption{ Inheritance for Message Objects. }
    \label{Fig:Mess}
  \end{center}
\end{figure}

The above figure shows the inheritance structure for message objects that can 
be passed within one MVC object and a later figure in this section will show 
the inheritance structure for messages objects that can be passed between MVC 
objects.

As with the view classes, the top message class, \Robject{gMessage}, is a
virtual class.  It contains no slots and its purpose is to bind the other
message classes together.

Currently, there are two types of messages for messaging within a model, view,
controller object: an add and an update message.  Both the add message,
\Robject{gAddMessage}, and the update message, \Robject{gUpdateMessage},
contain three slots, dataName, mData, and type.  dataName is the character
string that gives the name of the model, mData is a list of data needed to
perform the addition or alteration operation, and type is a character string
that gives the type of addition or alteration to perform.  Although
\Robject{gAddMessage} and \Robject{gUpdateMessage} are not virtual classes, it
is not intended to create objects of these classes.

Since adding a view requires different information and methods from adding a
model (and similarly with updating views versus models), it makes sense to
have two separate message classes for these operations.  Thus, when messages
are sent between components, they are from one of the following classes:
\Robject{gAddDataMessage}, \Robject{gAddViewMessage},
\Robject{gUpdateDataMessage}, and \Robject{gUpdateViewMessage}.  As shown in
Figure \ref{Fig:Mess}, the \Robject{gAddDataMessage} and
\Robject{gAddViewMessage} classes inherit from the \Robject{gAddMessage}
class, and the \Robject{gUpdateDataMessage} and \Robject{gUpdateViewMessage}
classes inherit from the \Robject{gUpdateMessage} class.

The \Robject{gAddDataMessage} class represents a message to add a model and a
MVC object because whenever a model is added, a MVC object is also added that
contains this model.  The \Robject{gAddViewMessage} class represents a message
to create a view.

The \Robject{gUpdateDataMessage} class represents a message to update
a model.  This message is crucial for linking because it is responsible for
ensuring that once a model has been updated this information must propagate to
any views of the model and this information must also be passed along to any
other MVC object that is related to this model.  Because this information gets
passed to other MVC objects, this message also has one new slot, from, in 
addition to
the slots it inherits from \Robject{gUpdateMessage}.  The from slot is the
name of the model that the update data message came from.  This slot is
necessary because the update data message can come from one of two sources:
the message can occur because of user interaction with a view
of this model or the message may occur because a different model that is
linked to this model has been updated.  So the update data message may come
from within this MVC object or it may come from a different MVC object that
has changed its model and is related to this MVC object.  
Information on passing messages between MVC
objects will be given later in this section.  Finally, the
\Robject{gUpdateViewMessage} class represents a message to update a view.

Message classes have also been defined to pass information between MVC objects.
The two main types of messages, the add and update messages, will also be used 
here.  The inheritance structure for messages passed between MVC objects is 
shown in the figure below.

\clearpage

\begin{figure}[ht]
  \begin{center}
    \scalebox{0.7}{\includegraphics{VigMessageClass2.jpg}}
    \caption{ Inheritance for Message Classes that are Passed Between MVCs. }
    \label{Fig:BetwMess}
  \end{center}
\end{figure}

The first message class that will be discussed is \Robject{gAddChildMessage}.
A \Robject{gAddChildMessage} object will be very similar to a
\Robject{gAddDataMessage} object because they are both adding a new
model and a new MVC object.  The difference with a \Robject{gAddChildMessage}
object is that it must fill in some extra information to tie this new MVC
object to its parent MVC object.  The \Robject{gAddChildMessage} class
inherits from the \Robject{gAddMessage} class, with the same slots of
dataName, mData, and type.  The dataName slot is the name of the new model (and
new MVC), the mData slot is a list that contains the model data and virtual
data to fill the modelData and virtualData slots, respectively, of the new
model, and the type slot gives the type of the model (currently, the options
are ``exprSet'', ``graph'', or ``data.frame'').

The next two message classes pertain to updating MVC objects when a parent MVC
or child MVC object has changed.  These classes are
\Robject{gUpdateParentMessage} and \Robject{gUpdateChildMessage} and they both
inherit from the \Robject{gMessage} class, as shown in Figure
\ref{Fig:BetwMess}.  Even though they are both update messages, they do not
inherit from the \Robject{gUpdateMessage} class because the information they
need to contain is a \Robject{gUpdateDataMessage} object.  Thus, the
\Robject{gUpdateParentMessage} class has one slot, childUpdateDataMessage, and
this slot contains the \Robject{gUpdateDataMessage} object that was used to
update the child model.  Similarly, the \Robject{gUpdateChildMessage} class
has two slots, parentUpdateDataMessage and childName, where the
parentUpdateDataMessage slot contains the \Robject{gUpdateDataMessage} object
that was used to update the parent model and the childName slot contains the
name of the child model that is being updated (because a parent MVC can have
more than one child MVC).

A generic function, \Rfunction{handleMessage}, is defined that will be used to 
read message objects.  Methods for this generic function will be defined in 
packages that use this package (\Rpackage{iSPlot} and \Rpackage{iSNetwork}).

\section{Event-Callback Function Class}\label{Sec:EventFun}

Linking callback functions to events, such as a button click, a mouse movement
or a key press event, is what allows users to have an interactive
environment.  Then when an event happens something will occur in response
depending on the callback function linked to the event.  Now to create a
flexible environment we do not want a callback function that always does the
same thing when an event occurs.  For example, when a left button press event
occurs, we may sometimes want a point to be colored and at other times we may
want a point to be hidden.  To allow this flexibility, a class called
\Robject{gEventFun} was created.  By using this class, users will be able to
change the response to an event.

The \Robject{gEventFun} class stores all of the information about a potential
callback function.  Having the callback function information stored in an 
object will allow a user to connect events to callback functions.  The class 
definition for \Robject{gEventFun} is shown in the figure below.

\begin{figure}[ht]
  \begin{center}
    \scalebox{0.3}{\includegraphics{EventClass.jpg}}
    \caption{ The gEventFun Class. }
    \label{Fig:EventFun}
  \end{center}
\end{figure}

As shown in Fig \ref{Fig:EventFun}, the slots in the \Robject{gEventFun} class
are callFun, shortName, and preprocessFun.  The callFun slot is the name of
the callback function, the shortName slot is a short description of the
function, and the preprocessFun slot is a character vector of all
preprocessing functions that must be called before the callback function.

The callback function that is stored in the callFun slot must take only one
parameter and the class of this parameter will depend on the model class in
the active MVC object.  For example, for a graph model, the
parameter will be of class \Robject{AgNode}, representing a node in the graph.
For a data frame model, the parameter will be of class character, representing
a row in the data frame.

The preprocessing functions that are stored in the preprocessFun slot will
take no parameters.  The purpose of the preprocessing functions is to set some
variables that the callback function will need.  Thus, not all callback
functions will need preprocessing functions and if no preprocessing functions
are needed, then this slot will be NULL.  If the callback function sets the
color of a node in a graph, then a preprocessing function is needed to
determine what color this will be.  So in this example, the preprocessing
function will open a color browser to allow the user to choose what the new
color will be.

\section{Conclusions}\label{Sec:Conc}

The inheritance structure for the classes defined in this package are intended 
to be extensible.  If a user needs a model of a different type that is not 
currently represented, then a new model class can be defined that inherits 
from the \Robject{gModel} virtual class.  Similarly, new view and message 
classes can be created if new views or messages are needed by a user.

As mentioned in Section \ref{Sec:Overview} this package is intended to be a 
behind the scenes package that is used by other packages for its class and 
generic function definitions.  Currently, this package is used by the 
\Rpackage{iSPlot} and \Rpackage{iSNetwork} packages.

\end{document}

### 2005-11-15 inst/paper/paper2.tex
Added my rough draft of paper 2.
git-svn-id: file:///home/git/hedgehog.fhcrc.org/bioconductor/trunk/madman/Rpacks/MVCClass@14843 bc3139a8-67e5-0310-9ffc-ced21a209358

\documentclass{article}[11pt]
\setlength{\topmargin}{-2cm}
\setlength{\textheight}{24cm}
\setlength{\oddsidemargin}{10mm}
\setlength{\evensidemargin}{10mm}
\setlength{\textwidth}{15cm}
\setlength{\parskip}{3mm}
\setlength{\parindent}{0mm}

\usepackage{epsfig}
\usepackage{comment}
\usepackage{graphics}
\usepackage{graphicx}

\newcommand{\Rfunction}[1]{{\texttt{#1}}}
\newcommand{\Robject}[1]{{\texttt{#1}}}
\newcommand{\Rpackage}[1]{{\textit{#1}}}

\begin{document}

\section{Introduction}\label{Sec:Intro}

\subsection{Background}\label{Ssec:Backg}

There are many ways to visualize multivariate data where more than
two or three dimensions are needed to view the data.  Several of these ideas
were discussed in a previous paper, titled ``Creating Linked, Interactive
Plots to Explore Multivariate Data'', so the focus in this paper will be one
particular method of visualizing multivariate data that is particularly useful
for exploratory data analysis.  This paper will discuss creating linked,
interactive views of linked data sets.

First brief definitions of linked and interactive views will be given.  Linked
plots mean that if a point on one plot changes its appearance, such as it
becomes colored red, then the corresponding point on a second plot will also
change its appearance to the color red.  For points to be corresponding, the
data displayed in the points must come from the same entity.  As an example,
suppose that the data consisted of each of the 50 states' murder rate, assault
rate, rape rate and percent of population living in urban areas.  Then if
there were two scatterplots, one with murder rate versus assault rate and the
other with rape rate versus percent urban population, the points on each plot
that referred to the state of Montana would be corresponding points.

Interactivity is an idea that will be familiar from using the Internet.  For
example, clicking on a hyperlink on a web page results in a new web page
opening.  Another example is clicking on the login button after entering a
user name and password to view one's email.  Interactivity means that there is
a response to some user action.

Creating interactive applications requires three
things: some user action or input, which will be referred to as an
event; a response to that action, which will be executed by a callback
function; and a method that connects the event to the callback
function, which will be referred to as the signal handler.  These
three steps are ordered as follows: an event occurs, the event causes a
signal to be emitted that is caught by the signal handler, and then
the signal handler calls the callback function, which results in the
response to the user action.  A flowchart of these steps is shown
below in Figure \ref{Fig:Event}.  In the example of clicking on a
hyperlink mentioned above, the event is the user clicking on the
hyperlink and the callback function causes a new web page to open.
The signal handler, which is the method that connects the event to the
callback function, is not noticed by the user.

\begin{figure}[!h]
  \begin{center}    
    \scalebox{0.8}{\includegraphics{Event.jpg}}
    \caption{ Flow Chart of the Response to an Event. }
    \label{Fig:Event}
  \end{center}
\end{figure}

This method of creating linked, interactive views of data is flexible for
exploring multivariate data because many different views can be created, such
as scatterplots, spreadsheets, and heatmaps, and then these views can be
linked.  Thus, users can decide which views best visually represent their
data.  Also, since the views are interactive, users can change the views while
looking at them to make the views more informative about the underlying data
structure.  This flexibility in linked, interactive views gives users a
powerful tool for visually exploring their data.

\subsection{Design and Implementation of Linked, Interactive
  Views}\label{Ssec:Design}

Focusing on linked, interactive views as a powerful visualization tool, the
next step is to determine how to design a software package that can implement
this method.  A popular design for creating linked views of data is the
model-view-controller (MVC) paradigm.  This design consists of three types
of objects: the controller, which defines what actions occur in response to
user input; the view, which consists of displays of the data; and the model,
which manages the data.  This paradigm is powerful because it decouples the
views from the model.  Thus, only one copy of the data needs to be stored and
then all views refer to this data set.  With all views based on the same data,
any changes to the data will be propagated to all views, which will give the
appearance of linked views.

To link the views to this one data set there is a subscribe/notify procedure
between the views and the model.  The views subscribe to a particular model
and the model must notify the views when a change occurs so that the views
will be updated.  This design allows multiple views of the model that are
linked, but are not aware of each other, as shown in Figure \ref{Fig:ExMVC}.

\begin{figure}[ht]
  \begin{center}
    \scalebox{0.7}{\includegraphics{ExMVC.jpg}}
    \caption{ A Model with its Views. }
    \label{Fig:ExMVC}
  \end{center}
\end{figure}

In the ``Creating Linked, Interactive Plots to Explore Multivariate Data''
paper, the creation of an R package called \Rpackage{iSPlot} was discussed.
The \Rpackage{iSPlot} package implemented the MVC design to satisfy three
goals: to create views that are linked, to create views that are interactive,
and to create a design that is extensible.  This package currently allows
users to generate linked, interactive scatterplots and spreadsheets of
two-dimensional data, such as data frames and matrices.

\subsection{Limitations of the MVC Design}\label{Ssec:Limit}

The MVC design assumes that there is one data set that all views will
reference.  But what if the data consisted of more than one data set?  For
example, consider the data stored in a relational database.  Here there are
often multiple tables of data, where rows in different tables are linked
through a unique identifier called a key.  Although these tables of data
could be combined into one large table, there are several reasons not to do
this.  Keeping the data in separate tables reduces redundancy, particularly
when the rows are related by a one-to-many or a many-to-many relationship.
Also separate data tables lead to more maintainable data.  By storing only one
copy of each piece of information (in a row in one table), it is easier to
ensure that the data are correctly updated when a change occurs.  All other
tables that are connected with this information can then just reference this
row that has been updated.

Consider Figure \ref{Fig:DBTab} shown below.  Here the database consists of
three tables, Students, Classes, and StudentClasses.  The Students table has
four columns, StudentID (the key), FirstName, LastName, and Year.  The Classes
table has three columns, ClassID (the key), ClassName, and Department.  The
StudentClasses table has two columns, StudentID and ClassID.  This last table
relates which classes a student is taking and which students are in
each class.  In this example, the students can be taking more than one class
and each class can have more than one student.  So the relationship between
classes and students is many-to-many.

%\clearpage

\begin{figure}[ht]
  \begin{center}
    \scalebox{0.5}{\includegraphics{databaseTables.jpg}}
    \caption{ Example of Relational Database Tables. }
    \label{Fig:DBTab}
  \end{center}
\end{figure}

Clearly, all of this data could be combined into one large
table, called CombinedStudentClass, as shown in Figure \ref{Fig:OneDBTab}.
The CombinedStudentClass table would have the following columns: StudentID,
FirstName, LastName, Year, ClassID, ClassName, and Department.  Now if a
student's last name changed, all instances of this student would need to be
changed in the new large table.  For example, if the student was taking four
classes, then this student would have four rows in the CombinedStudentClass
table and each row would need to change the value for the student's last name.
If instead only one record was kept per student, as shown in Figure
\ref{Fig:DBTab} above, then this process would be less error prone and would
require less memory to store the data. This example shows that there are
advantages to storing the data in separate structures.

%\clearpage

\begin{figure}[ht]
  \begin{center}
    \scalebox{0.5}{\includegraphics{oneDatabaseTable.jpg}}
    \caption{ Example of a Combined Table. }
    \label{Fig:OneDBTab}
  \end{center}
\end{figure}

In a relational database all of the data are stored in tables, but consider the
situation where there are multiple data sets that are of different data
structures.  For example, suppose that you have experimental microarray data,
which looks at gene expression levels.  This experimental data can be
stored as a two dimensional data structure, such as a matrix, where rows are
genes and columns are samples.  Linked to this
experimental data is meta data that gives information about the genes studied
in the microarray experiment.  This meta data may be a gene ontology
graph that gives the molecular functions of these genes.  These two data sets
(the experimental data and the meta data) are linked through the genes, but
they are different data structures.  The experimental data is a matrix and
the meta data is a graph.

Now for the original MVC design, where one data set is expected, it is not
clear how to combine these two data sets into one data set.  The meta data
graph could be stored as a matrix, but a graph is a data structure that can be
well represented as an object with slots for nodes and edges, especially if
the graph is sparse.  As with the relational database tables, it seems
best in terms of data storage, alteration, and retrieval to keep the two data
sets separate.  Also, keeping the two data sets separate does not force a data
structure on the data that may not fit well.

Currently, linked data sets are a common occurrence that can not be easily
visualized simultaneously.  Any time there is meta data linked to experimental
or study data there is a multiple data set situation where visually exploring
the data would be helpful to understand relationships between variables.  In
this paper, the design of software to create linked views based on multiple
linked data sets will be discussed.  The goals of this software will include
creating linked views of linked data sets, creating interactive
views where the response to an event can be changed, and creating
an extensible design that will allow future users to make additions.

\section{Extending the MVC Design}\label{Sec:Extend}

As discussed in Section \ref{Sec:Intro}, the MVC design expects one data set
with all views referencing this one data set.  However, several examples were
shown where multiple data sets are linked.  Since linking between multiple
data sets would be useful to visualize, the next step will be to determine how
the MVC design can be revised to account for multiple data sets.

\subsection{One Large Data Set}\label{Ssec:OneDS}

One possibility for dealing with multiple data sets within the MVC design is to
combine the multiple data sets into one large data set.  With this new
combined data, the original MVC design will be appropriate for creating linked
views.  Thus, the strength of combining the data sets into one large data set
is that the MVC paradigm will already be appropriately set up to deal with the
one large data set.

Unfortunately, as discussed in Section \ref{Ssec:Limit}, combining
multiple data sets into one large data set is not always appropriate or even
possible.  Even if the multiple data sets have the same data structure,
combining the data sets often results in redundant data and makes updating the
data more complicated.  In the situation where the data sets are not of the
same data structure, it may not be possible to combine the data sets into one
data structure that retains all of the original information.  Thus, another
way of linking views based on multiple data sets needs to be designed.

\subsection{Multiple MVC Objects}\label{Ssec:MMVC}

Because of the difficulty of combining multiple data sets, the MVC paradigm
will need to be altered to fit the data.  It has been mentioned that the MVC
design is intended to deal with one data set.  Now that we have several data
sets, why not have several MVC objects?  One MVC object to handle each data
set.  Thus, within each MVC object, one could create linked views of that
particular data set, which is the same as the original MVC design.  Then
between MVC objects a messaging system would need to be designed so that a
change in one data set would be sent to the other linked data sets.

\small
\begin{tabular}[t]{ | l | r | r | }
  \hline
  State & Life Expectancy & Income \\ \hline
  Pennsylvania & 70.43 & 4449 \\ \hline
  California & 71.71 & 5114 \\ \hline
  Minnesota & 72.96 & 4675 \\ \hline
\end{tabular}
\hspace{10pt}
\begin{tabular}[t]{ | l | r | r | }
  \hline
  City & Precipitation & Temperature \\ \hline
  Philadelphia & 39.9 & 39 \\ \hline
  Pittsburgh & 36.2 & 37 \\ \hline
  Los Angeles & 14.0 & 68 \\ \hline
  San Francisco & 20.7 & 56 \\ \hline
  Sacramento & 17.2 & 55 \\ \hline
  Duluth & 30.2 & 18 \\ \hline
  Minneapolis St. Paul & 25.9 & 22 \\ \hline
\end{tabular}

\begin{table}[h]
  \begin{center}
    \begin{tabular}{ | r | r | }
      \hline
      State & City \\ \hline
      Pennsylvania & Philadelphia \\ \hline
      Pennsylvania & Pittsburgh \\ \hline
      California & Los Angeles \\ \hline
      California & San Francisco \\ \hline
      California & Sacramento \\ \hline
      Minnesota & Duluth \\ \hline
      Minnesota & Minneapolis St. Paul \\ \hline
    \end{tabular}
    \caption{State and City Values.}\label{Tab:CityState}
  \end{center}
\end{table}

\normalsize
As an example, suppose that one has two data sets: one with information at the
state level and one with information at the city level.  These tables are
shown in Table \ref{Tab:CityState}.  The State data includes Life Expectancy
(in years from 1970) and Income (per capita from 1974) data
while the City data includes Precipitation (in inches from 1975) and Average
High Temperature in January (in Fahrenheit from 2005)
data.  These two data sets are linked by the CityState table that links cities
with the states they reside in.  Now for these two data sets there would be
two MVC objects.  One MVC object for the State data and its views and one MVC
object for the City data and its views.

\begin{figure}[ht]
  \begin{center}
    \scalebox{0.7}{\includegraphics{MultipleMVC.jpg}}
    \caption{ Example of Two MVC Objects. }
    \label{Fig:MultMVC}
  \end{center}
\end{figure}

Figure \ref{Fig:MultMVC} shows one view for the State data and one view
for the City data.  You can see how the State and City data are linked based
on the coloring of the points in the plots.  The points referring to the state
of Pennsylvania are red, the points referring to California are blue, and the
points referring to Minnesota are green.  Even though there are two MVC
objects here, with one MVC object for each data set, it is still clear that
the views are linked through the data sets and that is shown in this example
through the coloring.

This solution uses the power and simplicity of the MVC design and expands it
to the multiple data set problem.  Now we can build on the original design,
which has been well tested, instead of starting a new design from scratch.
The strength of this multiple linked MVC design is that it uses MVC
components, which can be individually created and then linked.  Thus, these
components are reusable.  Also this multiple MVC design is structured to fit
with the linking seen in the multiple data sets.  Clearly, the design should
be built around the problem and multiple MVC objects with messaging between
the MVC objects fits the problem of creating linked views of multiple linked
data sets.

Two parts of this design will need some special thought: creating an
appropriate messaging system between MVC objects and deciding how the MVC
objects will be related.  First, there are several questions that need to be
considered when creating the message system.  What information should be
passed between MVC objects in the messages?  Should there be a system to stop
messages from propagating between MVC objects?  How will the messages be
created and handled?  Second, there are also several questions with regard to
how the MVC objects will be related.  Should the MVC objects be related like
the tables in a relational database where the relationships are determined by
linked keys in different tables?  Should the MVC objects be linked through some
type of hierarchy, like a parent-child relationship where one MVC creates
another MVC and thus, the original MVC is the parent?  These decisions will be
based on the data that needs to be visualized and will be discussed in the
following sections as the design for multiple MVC objects is explained.

\section{Design Details of One MVC Object}\label{Sec:OneMVC}

\subsection{Overview}\label{SSec:OneOver}

Implementing a design to create linked, interactive views of multiple linked
data sets will be done in the R language in two packages, called
\Rpackage{MVCClass} and \Rpackage{iSNetwork}.  The \Rpackage{MVCClass} package
will implement the design of the MVC classes and methods.  The
\Rpackage{MVCClass} package will define the individual model, view, and
controller classes as well as the \Robject{MVC} class that binds the
individual model, view, and controller classes together into one unit.  The
package will also define the message classes that will allow the components to
communicate with each other.  The goal of separating these class definitions
and methods into one package is that they will be available for other R
packages that implement linked views of data, such as the \Rpackage{iSPlot}
and the \Rpackage{iSNetwork} packages.

With the class definitions created in the \Rpackage{MVCClass} package, the
\Rpackage{iSNetwork} package will implement the graphical user interface
(GUI) so that users can load data, create views, and interact with the views
through a user interface with menus.  To create the GUI, \Rpackage{iSNetwork}
will use the R packages, \Rpackage{RGtk} and \Rpackage{gtkDevice}.
\Rpackage{RGtk} is an R package that allows one to interact with Gtk functions
from the R interface.  Gtk is an open-source X Window toolkit for creating
user interfaces.  The \Rpackage{gtkDevice} package creates a Gtk device by
turning a Gtk drawing area into a device that acts like an R device, but still
responds to events so that the view can be interactive.  For more information
about these packages, please see the paper, ``Creating Linked, Interactive
Plots to Explore Multivariate Data''.

The next subsections will explain the class definitions and methods that
pertain to a single model-view-controller object.  Thus, the model classes,
the view classes, the controller, the \Robject{MVC} class (which binds the
model, view and controller classes together), and the message classes that pass
information between the three components in one MVC object will all be
discussed in the following subsections.  All of these definitions are
implemented in the \Rpackage{MVCClass} package.  Message class definitions
that pertain to communication between the MVC objects will be given in
Section \ref{Sec:MultMVC}.

\subsection{Model}\label{Ssec:OneModel}

The model class is responsible for storing and updating the data.  These two
functions are reflected in Figure \ref{Fig:Model}, which shows the inheritance
structure for the model classes.  Here the virtual class,
\Robject{gModel}, has the slots modelData, linkData, virtualData, and
modelName.  These four slots are common information that all models will
need.  The modelData slot is the data set for this model, the linkData slot is
a list of two functions, \Rfunction{linkToParent} and \Rfunction{linkToChild},
which link this model to its parent and child models, respectively (see Section
\ref{Ssec:MultLink}), the virtualData slot is data pertaining to the views
that needs to be stored with the model so that it can be shown in all views of
this model, and the modelName slot is the name of the model.

\clearpage

\begin{figure}[ht]
  \begin{center}
    \scalebox{0.7}{\includegraphics{ModelClass.jpg}}
    \caption{ Inheritance for Model Objects. }
    \label{Fig:Model}
  \end{center}
\end{figure}

For specific model classes, there are \Robject{dfModel}, \Robject{graphModel},
and \Robject{exprModel}.  The \Robject{dfModel} class represents a model where
the modelData has class data frame (or matrix), the \Robject{graphModel} class
represents a model where the modelData has class graph, and the
\Robject{exprModel} class represents a model where the modelData has class
exprSet.

As mentioned previously, the model classes are also responsible for updating
the data and thus, you will notice that all three model classes have an
\Rfunction{updateModel} method.  This method is called whenever the data needs
to be updated in response to a user interacting with a view.  Thus, the
\Rfunction{updateModel} method is actually updating the virtualData slot
because it is this slot in the model classes that ensures that all views are
appropriately visually linked.

Although only three model classes are currently implemented, future users are
welcome to make additions to the model classes for new data structures they
would like to represent.  For example, a user may want to add a new model
class called \Robject{tsModel} to represent time series data.

Note that the model classes must also notify views when the data has changed
so that the views can be updated.  The communication between a model and its
views will be discussed in Section \ref{Ssec:OneMess}.

\subsection{Views}\label{Ssec:OneViews}

The view classes represent the visual depictions of the model.  All views will
need to store some common information and respond to certain events through
methods.  This consideration led to an object model where the different view
classes inherit from a view virtual class, called \Robject{genView}.  The
object model for the view classes is shown in Figure \ref{Fig:View}.

\begin{figure}[ht]
  \begin{center}
    \scalebox{0.7}{\includegraphics{ViewClass.jpg}}
    \caption{ Inheritance for View Objects. }
    \label{Fig:View}
  \end{center}
\end{figure}

The \Robject{genView} class has three slots, dataName, win, and winNum.
dataName is the name of the model that the view displays, win is the Gtk
window object that holds the view, and winNum is the number of the window so
that the window can be identified in the Window menu on the control window
(see Section \ref{Ssec:OneCont}).  All views will need to know these three
pieces of information so \Robject{genView} will bind the view classes
together.

As for specific view classes, the \Robject{spreadView} class will represent a
spreadsheet view of the data.  This view will only make sense for models that
are a two dimensional data structure, such as a matrix or a data frame.  The
information needed for this view is stored in the slot, clist, which is the
Gtk spreadsheet object.

When creating a plot of a data set, there is a general class, called
\Robject{plotView}, that has the slots, plotDevice, plotPar, and drArea, which
store the device number of the plot, the plotting parameters, and the Gtk
drawing area object, respectively.  This class is not intended to have any
objects as it just represents a general plot.  The plot classes that can have
objects are the specific plot classes, \Robject{sPlotView},
\Robject{graphView}, and \Robject{heatmapView}, which inherit from the
\Robject{plotView} class.

The \Robject{sPlotView} class represents a scatterplot view and it has the
extra slots, dfRows, colx, and coly, where dfRows are the row names from the
model that are shown in the plot, colx is the column name from the model that
is shown as the x variable in the plot, and coly is the column name from the
model that is shown as the y variable in the plot.  The \Robject{graphView}
class represents a graph plot view and it contains no extra slots besides
those in the \Robject{plotView} class.  The \Robject{heatmapView} represents a
heatmap view of the data and it contains the slot, ordering, which is a list
that is returned from the \Rfunction{heatmap} function to give information
about the dendrogram ordering.  Note that some of these views are only
applicable for certain model types. For example, the \Robject{graphView} class
will only make sense for a view of a graph object, which would be stored in the
\Robject{graphModel} class.

Since all views are expected to be interactive, all
view classes have methods that correspond to user interaction, such as
\Rfunction{clickEvent}.  For \Robject{sPlotView}, the \Rfunction{clickEvent}
method is called when Gtk emits a signal that a mouse button has been pressed
and the \Rfunction{motionEvent} method is called when Gtk emits a signal that
the cursor has been moved over the view.  Both the \Rfunction{clickEvent} and
\Rfunction{motionEvent} methods for a \Robject{sPlotView} object will
determine which point on the plot was clicked or under the cursor,
respectively.  For \Robject{spreadView}, the \Rfunction{clickEvent} method is
called when Gtk emits a signal that a row in the spreadsheet has been selected
or unselected and the \Rfunction{clickEvent} will determine which row has
changed.  For \Robject{graphView}, the \Rfunction{clickEvent} method is called
when Gtk emits a signal that a mouse button has been pressed and the
\Rfunction{motionEvent} is called when Gtk emits a signal that the cursor has
been moved over the view.  Here, though, the \Rfunction{clickEvent} and
\Rfunction{motionEvent} methods will determine which node was clicked or under
the cursor, respectively.  Note that the \Robject{heatmapView} class has
neither a \Rfunction{clickEvent} or a \Rfunction{motionEvent} method because
currently, it is not possible to convert the user coordinates of where the
event occurred into some object on the view (such as the row and column
coordinates of the heatmap).  Thus, the \Robject{heatmapView} is the only view
that is not currently interactive.

As with the model classes, the view classes need an update method and thus, all
views have an \Rfunction{updateView} method.  This method must take the
information that a model has just been updated and appropriately update the
view to reflect that change.  Each method will depend on how the view
represents the data so for a \Robject{spreadView} object the
\Rfunction{updateView} may select or unselect a row on the spreadsheeet while
for a \Robject{sPlotView} object the \Rfunction{updateView} may change the
appearance of a point on the plot.

The inheritance structure of the view classes is intended to be extensible.
For instance, if users wanted to add a new plot view, say a density plot, they
could create a new view class that inherits from \Robject{plotView}, called
\Robject{dPlotView}.  This new class, \Robject{dPlotView}, would have slots
names dfRows and col to store which rows and which column, respectively, were
used to calculate the density shown in the view.

\subsection{Controller}\label{Ssec:OneCont}

The controller must perform several tasks, including storing information,
deciding how a view will respond to user interaction, and allowing users to
load data, to create views, and to quit the program.  These tasks can be
separated into two functions, where the first is that the controller needs to
be a storage location and the second is that the controller must be some type
of interactive interface that allows the user to determine the course of
actions.

For the first function the controller will be an environment that is
stored in the controller slot of a MVC object, which will be discussed in more
detail in Section \ref{Ssec:OneMVC}.  This environment will store data
pertaining to state, such as information about appearance of the user interface
and the views.

For the second function the controller will be a GUI with menus to
allow the user to choose which action will occur next.  This aspect of the
controller, which is implemented in \Rpackage{iSNetwork}, is shown in the
Figure \ref{Fig:ContWin} and will be referred to as the control window.

\begin{figure}[ht]
  \begin{center}
    \scalebox{0.5}{\includegraphics{ControlWindow.jpg}}
    \caption{ The Control Window. }
    \label{Fig:ContWin}
  \end{center}
\end{figure}

As shown in Figure \ref{Fig:ContWin}, the menus on the control window are
File, Edit, Display, Event, and Window.  These menus always appear on the
control window, but the menu items under each menu will change depending on
the type of model that is active.  Because the goal is to have multipled
linked MVC objects, there will be more than one MVC object and thus, more than
one model. However, the control window menus will only perform operations on
one MVC object at a time.  Thus, in the \Rpackage{iSNetwork} package there
will be the concept of the ``active'' MVC object, where the ``active'' MVC
object (and thus, the ``active'' model) means that operations performed
through the control window menus will affect that MVC object.  For example,
the Display menu allows users to create views of the active model, but only
certain views are appropriate for a type of model, as mentioned in Section
\ref{Ssec:OneViews}.  If the active model was a data frame model
(\Robject{dfModel}), then the views that the user could create would be a
scatterplot (\Robject{sPlotView}) and a spreadsheet (\Robject{spreadView}),
but a graph plot (\Robject{graphView}) would not make sense.  Thus, when the
model in the ``active'' MVC is of class \Robject{dfModel}, then the Display
menu will have menu items to create a scatterplot and a spreadsheet.

Because interacting with the views should be driven by the user,
the Event menu allows the user to determine what response will occur when
there is user interaction with a view.  For example, the user may initially
want a point on a plot to become colored when the left button is pressed and
the cursor is over that point.  Then the user may decide that the point
should be hidden when the left button is pressed and the cursor is over that
point.  The Event menu lets the user decide if anything will happen in
response to a left button click, a middle button click, a right button click,
and mouse movement.  More information will be given in Section
\ref{Ssec:OneEvent} about linking an event with a function so that user
interaction with a view is flexible enough to allow for multiple outcomes.

Because one of the goals of creating linked views based on multiple linked
data sets is to create a software package that is flexible and extensible,
users will be able to add menus and menu items to the control window.  As an
example if a future programmer wants to add a new type of view, such as a
histogram, then a new menu item for that view could be added to the Display
menu, as long as that view makes sense for the active MVC's model.  By
allowing users to change the controller, they can add new functionality that
is specific to their problem.

\subsubsection{Linking Events to Functions}\label{Ssec:OneEvent}

Linking callback functions to events, such as a button click, a mouse movement
or a key press event, is what allows users to have an interactive
environment.  Then when an event happens something will occur in response
depending on the callback function linked to the event.  Now to create a
flexible environment we do not want a callback function that always does the
same thing when an event occurs.  For example, when a left button press event
occurs, we may sometimes want a point to be colored and at other times we may
want a point to be hidden.  To allow this flexibility, a class called
\Robject{gEventFun} was created.  By using this class, users will be able to
change the response to an event.

The \Robject{gEventFun} class stores all of the information about a potential
callback function.  When a user wants to connect this callback function to an
event, this can be performed through the Event menu on the control window (see
Section \ref{Ssec:OneCont}).  The controller
environment in each MVC object will store which \Robject{gEventFun} object
(representing the callback function) is linked with each event.  Thus, the
user can decide the response for button clicks and mouse movements, and the
user can change these responses at any time through the control window.

\begin{figure}[ht]
  \begin{center}
    \scalebox{0.6}{\includegraphics{EventClass.jpg}}
    \caption{ The gEventFun Class. }
    \label{Fig:EventFun}
  \end{center}
\end{figure}

As shown in Fig \ref{Fig:EventFun}, the slots in the \Robject{gEventFun} class
are callFun, shortName, and preprocessFun.  The callFun slot is the name of
the callback function, the shortName slot is a short description of the
function, and the preprocessFun slot is a character vector of all
preprocessing functions that must be called before the callback function.

The callback function that is stored in the callFun slot must take only one
parameter and the class of this parameter will depend on the model class in
the active MVC object.  For example, for a graph model, the
parameter will be of class \Robject{AgNode}, representing a node in the graph.
For a data frame model, the parameter will be of class character, representing
a row in the data frame.

The preprocessing functions that are stored in the preprocessFun slot will
take no parameters.  The purpose of the preprocessing functions is to set some
variables that the callback function will need.  Thus, not all callback
functions will need preprocessing functions and if no preprocessing functions
are needed, then this slot will be NULL.  If the callback function sets the
color of a node in a graph, then a preprocessing function is needed to
determine what color this will be.  So in this example, the preprocessing
function will open a color browser to allow the user to choose what the new
color will be.

\subsection{MVC}\label{Ssec:OneMVC}

The \Robject{MVC} class is the class that will bind the model, view, and
controller objects that refer to the same data set into one object.  Thus, the
creation of a MVC object occurs whenever a new data set is added.  A
MVC object can have one model and each model must be in one
MVC object so there is a one-to-one relationship between the
MVC and model objects.  Because of this relationship, the
MVC object and the model object are identified by the same name,
which is stored in the modelName slot of the model object.

As shown in Figure \ref{Fig:MVCClass}, the \Robject{MVC} class has five slots
which are, model, viewList, controller, parentMVC, and childMVCList.  The
model slot stores the model object, the viewList slot is a list of the view
objects for this model, the controller slot is an environment that stores
information for this MVC, the parentMVC slot is the name of the parent model
(MVC), and the childMVCList slot is a list of the children models
(MVCs).  Note that model and MVC objects use the same name so the
parentMVC slot refers to both the name of the model and the name of the
MVC.  Similarly, for the childMVCList slot, the names in the list
refer to both the name of the model and the name of the MVC.

\begin{figure}[ht]
  \begin{center}
    \scalebox{0.6}{\includegraphics{MVCClass.jpg}}
    \caption{ The MVC Class. }
    \label{Fig:MVCClass}
  \end{center}
\end{figure}

Thus, the \Robject{MVC} class not only binds the model, view, and controller
objects together, but it also shows that the MVC objects are related
to each other through a parent-child (tree) hierarchy, as shown in Figure
\ref{Fig:Hier}.  In this picture, `model1' is the name of the original MVC
object (model) and from it, two MVC objects were created, `model2' and
`model3'. Then from `model2', three MVC objects were created, `model4',
`model5', and `model6'.

\clearpage

\begin{figure}[ht]
  \begin{center}
    \scalebox{0.6}{\includegraphics{Hierarchy2.jpg}}
    \caption{ The Parent-Child Relationship Between MVCs. }
    \label{Fig:Hier}
  \end{center}
\end{figure}

A MVC object can have at most one parent MVC, but it can have many child
MVCs.  The relationship between MVC objects shows that child MVC objects are
derived from the parent MVC object (i.e. the model in the child MVC is derived
from the model in the parent MVC).  An easy example of a child model derived
from a parent model is to start with a data frame as the parent model and then
take a subset of this data frame as the child model.

\subsection{Messages}\label{Ssec:OneMess}

The message classes are intended to provide communication between different
components of the MVC design, such as when the model changes a message needs to
be sent to the views to let them know that they should be updated.  This
communication between the model, view and controller is crucial for the pieces
to work together and yet, still be independent of each other.

The object model for the message classes was derived from the idea that all
messages must have certain methods in common, such as \Rfunction{initialize}
and \Rfunction{handleMessage}, because all messages must be created and
handled in some manner so that the message is read and acted upon.  Also,
since messages are sent when something has changed, either through addition,
alteration, or deletion, the message classes reflect only certain operations.
These common purposes for the messages allow the message object model to be
constructed, and this model is shown in Figure \ref{Fig:Mess}.

\begin{figure}[ht]
  \begin{center}
    \scalebox{0.6}{\includegraphics{MessageClass.jpg}}
    \caption{ Inheritance for Message Objects. }
    \label{Fig:Mess}
  \end{center}
\end{figure}

As with the view classes, the top message class, \Robject{gMessage}, is a
virtual class.  It contains no slots and its purpose is to bind the other
message classes together.

Currently, there are two types of messages for messaging within a model, view,
controller object: an add and an update message.  Both the add message,
\Robject{gAddMessage}, and the update message, \Robject{gUpdateMessage},
contain three slots, dataName, mData, and type.  dataName is the character
string that gives the name of the model, mData is a list of data needed to
perform the addition or alteration operation, and type is a character string
that gives the type of addition or alteration to perform.  Although
\Robject{gAddMessage} and \Robject{gUpdateMessage} are not virtual classes, it
is not intended to create objects of these classes.

Since adding a view requires different information and methods from adding a
model (and similarly with updating views versus models), it makes sense to
have two separate message classes for these operations.  Thus, when messages
are sent between components, they are from one of the following classes:
\Robject{gAddDataMessage}, \Robject{gAddViewMessage},
\Robject{gUpdateDataMessage}, and \Robject{gUpdateViewMessage}.  As shown in
Figure \ref{Fig:Mess}, the \Robject{gAddDataMessage} and
\Robject{gAddViewMessage} classes inherit from the \Robject{gAddMessage}
class, and the \Robject{gUpdateDataMessage} and \Robject{gUpdateViewMessage}
classes inherit from the \Robject{gUpdateMessage} class.

The \Robject{gAddDataMessage} class represents a message to add a model and a
MVC object because whenever a model is added, a MVC object is also added that
contains this model.  The \Robject{gAddViewMessage} class represents a message
to create a view.  This message can only be used after a model has been
added.

The \Robject{gUpdateDataMessage} class represents a message to update
a model.  This message is crucial for linking because it is responsible for
ensuring that once a model has been updated this information must propagate to
any views of the model and this information must also be passed along to any
other MVC object that is related to this model.  Because this information gets
passed to other MVC objects, which will be discussed further in Section
\ref{Sec:MultMVC}, this message also has one new slot, from, in addition to
the slots it inherits from \Robject{gUpdateMessage}.  The from slot is the
name of the model that the update data message came from.  This slot is
necessary because the update data message can come from one of two sources:
one the message can occur because of user interaction with a view
of this model or the message may occur because a different model that is
linked to this model has been updated.  So the update data message may come
from within this MVC object or it may come from a different MVC object that
has changed its model and is related to this MVC object.  
Information on passing messages between MVC
objects will be given in Section \ref{Sec:MultMVC}.  Finally, the
\Robject{gUpdateViewMessage} class represents a message to update a view.
This message will be created and handled whenever the model is updated so that
the views accurately reflect the model.

All four classes, \Robject{gAddDataMessage}, \Robject{gAddViewMessage},
\Robject{gUpdateDataMessage}, and \Robject{gUpdateViewMessage}, have
\Rfunction{initialize} and \Rfunction{handleMessage} methods, which perform
the following tasks:

\begin{list}{}{\setlength{\leftmargin}{0.5in}
               \setlength{\itemsep}{0in}}
  \item \Rfunction{initialize} is used to properly fill the slots of the
object when it is created.  
  \item \Rfunction{handleMessage} is necessary to
process the message.  Since a message is a form of communication
between two components, once it is created it must be sent to the
proper recipient and action must be taken in response to its
information, which is the purpose of the \Rfunction{handleMessage} method.
\end{list}

As with the inheritance structures seen in previous sections, the message
inheritance structure is designed to allow for extensibility.  For example, a
user could create a \Robject{gUpdateControlMessage} class that inherited from
the \Robject{gUpdateMessage} class.  This new class,
\Robject{gUpdateControlMessage}, could convey information that told the
control window to be updated in response to a change in the active MVC.
Recall from Section \ref{Ssec:OneCont} that the active MVC is the MVC object
that the menus on the control window refer to so any of the operations chosen
from the control window menus will be applied to this MVC object (the active
MVC).  The goal for the message classes is that the design is flexible enough
to allow a user to make additions or changes.

\section{Details of Multiple MVCs}\label{Sec:MultMVC}

Up until now we have skirted around the fact that multiple MVC objects are
being used in this design for linking views of multiple linked data sets.
Each MVC object will represent one model and its views.  Now for the views of
different data sets to be linked, they must link through the data sets so this
linking information must be stored somewhere.  Also, there must be
message classes to pass information between MVC objects.

As mentioned in Section \ref{Ssec:OneMVC}, the \Robject{MVC} class has slots
for how the current MVC relates to other MVC objects and these slots are
parentMVC and childMVCList.  These slots show that MVC objects are related
through a tree hierarchy where a child model is derived from its parent
model.  Consider Figure \ref{Fig:SmallHier} shown below.  If firstMVC's model
changes, then this information first gets passed to secondMVC, which may make
changes to its model based on this information.  Then secondMVC passes this
information to thirdMVC based on the changes secondMVC made to its model.
Thus, information passing between firstMVC and thirdMVC only occurs through
secondMVC.  So messages can only be sent from parent MVC to child MVC or from
child MVC to parent MVC.

\begin{figure}[ht]
  \begin{center}
    \scalebox{0.6}{\includegraphics{SmallHierarchy.jpg}}
    \caption{ Multiple MVC Objects and Their Relationships. }
    \label{Fig:SmallHier}
  \end{center}
\end{figure}

Next we will consider how the data sets will be linked (and thus, how MVC
objects are linked) followed by how messages will be passed between these
linked MVC objects.

\subsection{Linking Data Sets}\label{Ssec:MultLink}

For messages to pass between MVC objects, there must be a function or a place
that converts information from one MVC into useful information for a different
MVC object.  Thus, there needs to be some way to determine how data from one
MVC object relates to data in a different MVC object.  This conversion is done
by two functions that are stored in the linkData slot of the model objects.
The linkData slot is a list with two elements: a \Rfunction{linkToParent}
function and a \Rfunction{linkToChild} function.

When a child MVC is created from an existing MVC, the child model's linkData
slot will be filled with the two functions, \Rfunction{linkToParent} and
\Rfunction{linkToChild}.  When the model in the child MVC is updated, the
\Rfunction{linkToParent} function in the linkData slot will be used to convert
the child model's data change into useful information for the parent model.

Now suppose that the parent model changed instead.  This information
needs to be passed to the child model.  However, a parent can have multiple
child MVC objects so which \Rfunction{linkToChild} function should be used?
Here, when a parent is updated and needs to pass information to its children,
the \Rfunction{linkToChild} function is stored with each child MVC because a
parent MVC may have multiple child MVC objects that all need to process this
information in a different way.  Thus, when a child MVC object is created and
its model's linkData slot is filled, the \Rfunction{linkToParent} function
will convert information from this child model into useful information for its
parent MVC and the \Rfunction{linkToChild} function will convert a message
from its parent MVC into useful information for this child MVC.

An example of how these functions, \Rfunction{linkToParent} and
\Rfunction{linkToChild}, are used when messages are passed between MVCs will
be given in Section \ref{Ssec:MultMess}.

\subsection{Messages Between MVC Objects}\label{Ssec:MultMess}

In Section \ref{Ssec:OneMess}, where the message classes to pass information
within one MVC object were discussed, it was noted that there are two main
types of messages: an add message and an update message.  These message types
will also be seen in passing messages between different MVC objects, as shown
in Figure \ref{Fig:BetwMess}.

\begin{figure}[ht]
  \begin{center}
    \scalebox{0.6}{\includegraphics{MessageClass2.jpg}}
    \caption{ Inheritance for Message Classes that are Passed Between MVCs. }
    \label{Fig:BetwMess}
  \end{center}
\end{figure}

The first message class that will be discussed is \Robject{gAddChildMessage}.
A \Robject{gAddChildMessage} object will be very similar to a
\Robject{gAddDataMessage} object because they are both adding a new
model and a new MVC object.  The difference with a \Robject{gAddChildMessage}
object is that it must fill in some extra information to tie this new MVC
object to its parent MVC object.  The \Robject{gAddChildMessage} class
inherits from the \Robject{gAddMessage} class, with the same slots of
dataName, mData, and type.  The dataName slot is the name of the new model (and
new MVC), the mData slot is a list that contains the model data and virtual
data to fill the modelData and virtualData slots, respectively, of the new
model, and the type slot gives the type of the model (currently, the options
are ``exprSet'', ``graph'', or ``data.frame'').  As with the previous message
classes, the \Robject{gAddChildMessage} class has two methods:
\Rfunction{initialize} and \Rfunction{handleMessage}.  The
\Rfunction{initialize} method properly fills the slots of the object when it
is created and the \Rfunction{handleMessage} method ensures that the new model
and MVC objects are created (which is what the \Rfunction{handleMessage}
method for the \Robject{gAddDataMessage} class does) and then it fills the
linkData slot of the new child model so that the \Rfunction{linkToParent} and
\Rfunction{linkToChild} functions are available for messaging between MVC
objects and finally it stores information in the child MVC object's controller
that determines if the child model was created from a subset of the parent
model.  Thus, the \Rfunction{handleMessage} method both creates the child MVC
object and relates it to the parent MVC object.

The next two message classes pertain to updating MVC objects when a parent MVC
or child MVC object has changed.  These classes are
\Robject{gUpdateParentMessage} and \Robject{gUpdateChildMessage} and they both
inherit from the \Robject{gMessage} class, as shown in Figure
\ref{Fig:BetwMess}.  Even though they are both update messages, they do not
inherit from the \Robject{gUpdateMessage} class because the information they
need to contain is a \Robject{gUpdateDataMessage} object.  Thus, the
\Robject{gUpdateParentMessage} class has one slot, childUpdateDataMessage, and
this slot contains the \Robject{gUpdateDataMessage} object that was used to
update the child model.  Similarly, the \Robject{gUpdateChildMessage} class
has two slots, parentUpdateDataMessage and childName, where the
parentUpdateDataMessage slot contains the \Robject{gUpdateDataMessage} object
that was used to update the parent model and the childName slot contains the
name of the child model that is being updated (because a parent MVC can have
more than one child MVC).

As with all the other message classes, these two update messages have two
methods: \Rfunction{initialize} and \Rfunction{handleMessage}.  The
\Rfunction{initialize} method properly sets the slots of the new object.  The
\Rfunction{handleMessage} method for a \Robject{gUpdateParentMessage} will
take the \Robject{gUpdateDataMessage} object from the child model and convert
it to a \Robject{gUpdateDataMessage} object for the parent model using the
\Rfunction{linkToParent} function.  Similarly, the \Rfunction{handleMessage}
method for a \Robject{gUpdateChildMessage} will take the
\Robject{gUpdateDataMessage} object from the parent model and convert it to a
\Robject{gUpdateDataMessage} object for the child model using the
\Rfunction{linkToChild} function.

These message classes ensure that when a model is updated, its parent and
child models will be notified of the change and most importantly will be
notified of the change in a way that each model can properly act upon that
information.  Because each MVC object has a different model (and thus a
different data set and potentially a different data structure), information
that pertains to the model for one MVC must be converted to useful information
for the linked model and this conversion is done by the
\Rfunction{linkToParent} and \Rfunction{linkToChild} functions.

\begin{figure}[ht]
  \begin{center}
    \scalebox{0.6}{\includegraphics{MessagePassing.jpg}}
    \caption{ Message Passing Between MVCs. }
    \label{Fig:MessPass}
  \end{center}
\end{figure}

Figure \ref{Fig:MessPass} shows how messages are passed within one MVC as well
as how this message gets passed on to its parent MVC.  In this picture, the
message passing starts when a user interacts with the y vs. x scatterplot
depicting model 2.  After the user interaction with the scatterplot, step 1
(indicated on the figure as a circled 1) is to send an update data message to
model 2.  As soon as model 2 is updated, step 2 is to send an update view
message so that the scatterplot view is updated to reflect the changed data.
In step 3, an update parent message is sent because MVC 2 (which is the MVC
object for model 2) has a parent MVC, which is MVC 1.  For step 4, the
\Rfunction{linkToParent} function converts the child's update data message
into an update data message that the parent will understand.  In step 5, this
new update data message that was created by the \Rfunction{linkToParent}
function will be sent to model 1.  Finally for step 6, after model 1 has been
updated, an update view message will be sent to the view of model 1.  Now all
of the data and the views are in synch.

\section{Conclusions}\label{Sec:Conc}

Several examples of linked data sets were given at the beginning of this
paper, including linked tables from a relational database, and meta data that
is linked to experimental data.  Until now it has not been possible to create
linked views of these data sets.  However, creating linked, interactive views
of underlying linked data sets would be very helpful for exploratory data
analysis.  Thus, this paper discussed the creation of two R packages,
\Rpackage{MVCClass} and \Rpackage{iSNetwork}, which allow users to create
linked, interactive views of linked data sets based on expanding the
model-view-controller paradigm.  Now for visualizing linked data sets, the
model-view-controller design, which is a well known and thus, well tested
software paradigm, was extended to multiple MVC objects where each MVC object
contains one data set and its views.  This solution broke down the problem of
linked, interactive views of linked data sets into components (MVC objects)
that can be reused and linked.

Although one of the goals of the \Rpackage{MVCClass} and
\Rpackage{iSNetwork} packages was to create a design that was extensible,
certain design decisions, such as expecting the data sets to be linked through
a parent-child hierarchy, may not work with some data sets.  As an example,
the linked tables from a relational database are not related through a tree
structure and thus, having parent and child MVC objects may not make sense
with these data sets.  However, the design allows users to add new
types of models, new types of views, new menu items on the control window and
new message classes so there is still flexibility for additions as long as the
data sets are connected through parent-child relationships.

\subsection{Future Work}\label{Ssec:Further}

When a MVC's model is updated, it notifies its parent and child
MVCs that a change has occurred.  Then when a MVC gets a message from a
parent or child MVC, the MVC can either decide to ignore the message or read
the message (which results in this MVC's model being updated).  However, the
MVC object can not decide to only see certain types of messages.  It must
either accept all messages or ignore all messages.  A future goal for message
passing is to allow more selectivity in which messages will be accepted
between MVC objects so that the linking between MVC objects is more
sophisticated.

\end{document}

### 2005-12-13 inst/paper/paper2.tex
Updated paper 2.
git-svn-id: file:///home/git/hedgehog.fhcrc.org/bioconductor/trunk/madman/Rpacks/MVCClass@15362 bc3139a8-67e5-0310-9ffc-ced21a209358

@@
 \newcommand{\Rpackage}[1]{{\textit{#1}}}
+\newcommand{\Rslot}[1]{\textsl{#1}}
 
@@
 
+\bibliographystyle{amsplain}
+\bibliography{paper2}
+
+\begin{titlepage}
+\begin{center}
+\Large
+\vspace*{1in}
+\textbf{Creating Linked, Interactive Views to \linebreak Explore Linked Data} \\
+\vspace{1in}
+\textbf{Dana-Farber Cancer Institute}\\ 
+\textbf{December 9, 2005} \\
+\vspace{1in} 
+\textbf{Elizabeth Whalen} \\
+\end{center}
+\end{titlepage}
+\normalsize
+\newpage
+
+%\renewcommand{\baselinestretch}{1.5}
 \section{Introduction}\label{Sec:Intro}
@@
 were discussed in a previous paper, titled ``Creating Linked, Interactive
-Plots to Explore Multivariate Data'', so the focus in this paper will be one
-particular method of visualizing multivariate data that is particularly useful
-for exploratory data analysis.  This paper will discuss creating linked,
-interactive views of linked data sets.
-
-First brief definitions of linked and interactive views will be given.  Linked
-plots mean that if a point on one plot changes its appearance, such as it
-becomes colored red, then the corresponding point on a second plot will also
-change its appearance to the color red.  For points to be corresponding, the
-data displayed in the points must come from the same entity.  As an example,
-suppose that the data consisted of each of the 50 states' murder rate, assault
-rate, rape rate and percent of population living in urban areas.  Then if
-there were two scatterplots, one with murder rate versus assault rate and the
-other with rape rate versus percent urban population, the points on each plot
-that referred to the state of Montana would be corresponding points.
+Views to Explore Multivariate Data'' \cite{EW05}, so the focus in this paper
+will be one specific method of visualizing multivariate data that is
+particularly useful for exploratory data analysis.  This paper will discuss
+creating linked, interactive views of linked data sets.
+
+First brief definitions of linked views, interactive views and linked data
+sets will be given.  Linked views mean that if a component that represents
+data (such as a point on a plot or a row in a spreadsheet) changes its
+appearance on one view, then the corresponding component on a second view will
+also change its appearance.  For components on different views to be
+corresponding, the data displayed in those components must come from the same
+entity in a model.  As an example, suppose that the model consisted of four
+variables for the fifty states' murder rate, assault rate, rape rate, and
+percent of population living in urban areas.  Now the model could be
+represented as a matrix of four columns by fifty rows with four columns for
+the four variables and fifty rows for the fifty states.  In this model, the
+entities would be the states.  If two scatterplots were created as views of
+this model, where the first plot displayed murder rate versus assault rate and
+the second plot displayed rape rate versus percent urban population, then the
+points on each plot that referred to the state of Montana would be
+corresponding components.
 
@@
 
-Creating interactive applications requires three
-things: some user action or input, which will be referred to as an
-event; a response to that action, which will be executed by a callback
-function; and a method that connects the event to the callback
-function, which will be referred to as the signal handler.  These
-three steps are ordered as follows: an event occurs, the event causes a
-signal to be emitted that is caught by the signal handler, and then
-the signal handler calls the callback function, which results in the
-response to the user action.  A flowchart of these steps is shown
-below in Figure \ref{Fig:Event}.  In the example of clicking on a
+One particular model for creating interactive applications involves three
+components: some user action or input, which will be referred to as an event;
+a response to that action, which will be executed by a function that is called
+a callback function; and a method that connects the event to the callback
+function, which will be referred to as the signal handler.  The events can
+range from the user moving the cursor over a plot to pressing a button on a
+graphical user interface (GUI) to pressing certain keys.  Notice that the
+events almost always occur over some type of interface, though some events,
+such as key press events may occur without an interface.  The callback
+function is a unique function because its signature is determined by the event
+that the callback function is linked to.  The parameters of the callback
+function may include information about the event, such as the location where
+the event occurred, and information about the object that just caused the
+event, such as the button on the GUI.  The signal handler is the piece that
+must look to see if an event has occurred and then when the event has
+occurred, the signal handler calls the appropriate callback function.  The
+strengths of this model of interactivity are that programmers can decide which
+events to notice by deciding which signal handlers to define, programmers
+can change the response to an event by changing the callback function that is
+linked to an event, and the system's event loop can continue working on other
+things until a signal is emitted because an event occurred.
+
+These three pieces in this model of interactivity are ordered as follows:
+an event occurs, the event causes a signal to be emitted that is caught by the
+signal handler, and then the signal handler calls the callback function, which
+results in the response to the user action.  A flowchart of these steps is
+shown below in Figure \ref{Fig:Event}.  In the example of clicking on a
 hyperlink mentioned above, the event is the user clicking on the
@@
 The signal handler, which is the method that connects the event to the
-callback function, is not noticed by the user.
+callback function, is not noticed by the user, but it is handled by the
+system's event loop when a signal is emitted.
+
+\clearpage
 
@@
   \begin{center}    
-    \scalebox{0.8}{\includegraphics{Event.jpg}}
+    \scalebox{0.6}{\includegraphics{Event.jpg}}
     \caption{ Flow Chart of the Response to an Event. }
@@
 
-This method of creating linked, interactive views of data is flexible for
-exploring multivariate data because many different views can be created, such
-as scatterplots, spreadsheets, and heatmaps, and then these views can be
-linked.  Thus, users can decide which views best visually represent their
-data.  Also, since the views are interactive, users can change the views while
-looking at them to make the views more informative about the underlying data
-structure.  This flexibility in linked, interactive views gives users a
-powerful tool for visually exploring their data.
+Linked data sets mean that the data have more than one conceptual grouping,
+but there is a relationship between these different conceptual groupings.
+Linked data sets are an idea that will be familiar to users of relational
+databases.  Data in a relational database are typically stored in many
+linked tables with rows in different tables linked to each other through a
+unique identifier called a key.  In this example, each table would be
+considered a data set and they are linked through the keys.  Another situation
+where linked data sets occur is when there is experimental or study data that
+are linked to meta-data.  An example of this situation is given in Section
+\ref{Ssec:Limit} where the experimental data are microarray data, which looks
+at gene expression levels, and the meta-data are a Gene Ontology graph that
+gives the molecular functions of the genes studied in the microarray
+experiment.
+
+Creating linked, interactive views of data is flexible for exploring
+multivariate data because many different views can be created, such as
+scatterplots, spreadsheets, and heatmaps, and these views are linked because
+they are based on the same underlying model.  Thus, users can decide which
+views best visually represent their data.  Also, since the views are
+interactive, users can change the views while looking at them to make the
+views more informative about the underlying data structure.  This flexibility
+in linked, interactive views gives users a powerful tool for visually
+exploring their data.
+
+In Section \ref{Ssec:Design}, the design of linked, interactive views based on
+the model-view-controller (MVC) paradigm will be discussed.  However, the
+limitation with this design is that it expects one data set, which limits the
+type of data we can visualize.  There are many
+situations where multiple data sets occur, particularly any time there are
+experimental or study data linked to meta-data.  Being able to visualize data
+from multiple data sets would be helpful to understand the relationships
+between variables.  Thus, after discussing the limitations of the MVC design
+in Section \ref{Ssec:Limit}, an extension of the MVC design will be discussed
+in Section \ref{Sec:Extend}.  This extension will allow the creation of linked,
+interactive views based on linked data sets.  The software that is discussed
+in this paper will have three goals: to create linked views of linked data
+sets, to create interactive views where the response to an event can be
+changed, and to create an extensible design that will allow future users to
+make additions.
+
+As a motivating example of linked views based on linked data sets, consider
+Figure \ref{Fig:firstMP}.  In this figure, there are two views, one view of
+data set 1 and one view of data set 2.  Data set 1 and data set 2 are linked.
+When a user initially interacts with the scatterplot view of data set 2, the
+response to this interaction is to tell data set 2 to update itself (this is
+indicated on the figure as a circled 1).  Then in step 2, data set 2 tells the
+scatterplot view that is should update itself in response to the change in the
+data.  Next, in step 3, because data set 2 has changed and because it is
+linked to data set 1, data set 2 sends information to data set 1 about the
+change.  In step 4, the information from data set 2 is converted into
+information that data set 1 can understand.  In step 5, this information is
+received by data set 1 and data set 1 updates its data.  Finally, in step 6,
+data set 1 tells its view to update itself.  Thus, in this example, the views
+are synchronized based on the underlying linked data sets.
+
+\clearpage
+
+\begin{figure}[ht]
+  \begin{center}
+    \scalebox{0.6}{\includegraphics{firstMessagePassing.jpg}}
+    \caption{ Linked Views of Linked Data Sets. }
+    \label{Fig:firstMP}
+  \end{center}
+\end{figure}
 
@@
 then all views refer to this data set.  With all views based on the same data,
-any changes to the data will be propagated to all views, which will give the
-appearance of linked views.
+any changes to the data can be propagated to all views, which will create
+linked views.
 
@@
 will be updated.  This design allows multiple views of the model that are
-linked, but are not aware of each other, as shown in Figure \ref{Fig:ExMVC}.
-
-\begin{figure}[ht]
-  \begin{center}
-    \scalebox{0.7}{\includegraphics{ExMVC.jpg}}
+linked, but are not aware of each other, as shown in Figure \ref{Fig:ExMVC}
+\cite{DesignPatterns}.  Each view depends only on visually representing the
+model and thus, it is not necessary for a view to be aware of other views.  By
+keeping each view ignorant of other views, other views can added and removed
+without affecting a different view.
+
+\begin{figure}[ht]
+  \begin{center}
+    \scalebox{0.6}{\includegraphics{ExMVC.jpg}}
     \caption{ A Model with its Views. }
@@
 
-In the ``Creating Linked, Interactive Plots to Explore Multivariate Data''
-paper, the creation of an R package called \Rpackage{iSPlot} was discussed.
-The \Rpackage{iSPlot} package implemented the MVC design to satisfy three
-goals: to create views that are linked, to create views that are interactive,
-and to create a design that is extensible.  This package currently allows
-users to generate linked, interactive scatterplots and spreadsheets of
-two-dimensional data, such as data frames and matrices.
+Notice that although the components (model, view, and controller) in the MVC
+design are independent of one other, there still needs to be communication
+between them so that information can be shared.  Examples of this
+communication between components include the model notifying views that the
+data have changed, a view asking the controller how it should respond to user
+interaction, and a view telling the model that the data should change in
+response to user interaction with the view.  This message passing between
+components in the MVC paradigm will be discussed in detail in Section
+\ref{Ssec:OneMess}.
+
+The ``Creating Linked, Interactive Views to Explore Multivariate Data''
+\cite{EW05} paper discussed the creation of R software to fulfill three goals:
+to create views that are linked, to create views that are interactive, and to
+create a design that is extensible.  To satisfy the overall goals of the
+software, two R packages, called \Rpackage{MVCClass} and \Rpackage{iSPlot},
+were created.  The \Rpackage{MVCClass} package implemented the class and
+generic function definitions needed for the MVC design and the
+\Rpackage{iSPlot} package created the front end to let the user create linked,
+interactive views of two-dimensional data, such as data frames and matrices,
+through either a GUI or through command line functions.  The reason for
+separating the software into two packages was that it allowed the class
+definitions in the \Rpackage{MVCClass} package to be used by any other R
+package that would like to create linked views.
 
@@
 The MVC design assumes that there is one data set that all views will
-reference.  But what if the data consisted of more than one data set?  For
-example, consider the data stored in a relational database.  Here there are
-often multiple tables of data, where rows in different tables are linked
-through a unique identifier called a key.  Although these tables of data
-could be combined into one large table, there are several reasons not to do
-this.  Keeping the data in separate tables reduces redundancy, particularly
-when the rows are related by a one-to-many or a many-to-many relationship.
-Also separate data tables lead to more maintainable data.  By storing only one
-copy of each piece of information (in a row in one table), it is easier to
-ensure that the data are correctly updated when a change occurs.  All other
-tables that are connected with this information can then just reference this
-row that has been updated.
+reference.  But what if the data consisted of more than one conceptual
+grouping?  For example, consider data stored in a relational database, where
+there are multiple tables of data and rows in different tables are linked
+through a unique identifier called a key.  Although these tables of data could
+be combined into one large table, there are several reasons not to do this.
+Keeping the data in separate tables reduces redundancy, particularly when the
+rows are related by a one-to-many or a many-to-many relationship.  Also
+separate data tables generally lead to more maintainable data.  By storing
+only one copy of each piece of information (in a row in one table), it is
+easier to ensure that the data are correctly updated when a change occurs.
+All other tables that are connected with this information can then just
+reference this row that has been updated.
 
 Consider Figure \ref{Fig:DBTab} shown below.  Here the database consists of
-three tables, Students, Classes, and StudentClasses.  The Students table has
-four columns, StudentID (the key), FirstName, LastName, and Year.  The Classes
-table has three columns, ClassID (the key), ClassName, and Department.  The
-StudentClasses table has two columns, StudentID and ClassID.  This last table
-relates which classes a student is taking and which students are in
-each class.  In this example, the students can be taking more than one class
-and each class can have more than one student.  So the relationship between
-classes and students is many-to-many.
+three tables, State, City and StateCity.  The State table has three columns,
+State (the key), Life Expectancy, and Income.  The City table has three
+columns, City (the key), Precipitation, and Temperature.  The StateCity table
+has two columns, State and City.  This last table relates which cities are in
+which states.  In this example, each city can be in only one state, but each
+state can have more than one city so there is a one-to-many relationship
+between states and cities.
 
@@
   \begin{center}
-    \scalebox{0.5}{\includegraphics{databaseTables.jpg}}
+    \scalebox{0.5}{\includegraphics{newDatabaseTables.jpg}}
     \caption{ Example of Relational Database Tables. }
@@
 
-Clearly, all of this data could be combined into one large
-table, called CombinedStudentClass, as shown in Figure \ref{Fig:OneDBTab}.
-The CombinedStudentClass table would have the following columns: StudentID,
-FirstName, LastName, Year, ClassID, ClassName, and Department.  Now if a
-student's last name changed, all instances of this student would need to be
-changed in the new large table.  For example, if the student was taking four
-classes, then this student would have four rows in the CombinedStudentClass
-table and each row would need to change the value for the student's last name.
-If instead only one record was kept per student, as shown in Figure
-\ref{Fig:DBTab} above, then this process would be less error prone and would
-require less memory to store the data. This example shows that there are
-advantages to storing the data in separate structures.
+Clearly, this information could be combined into one large table, called
+CombinedStateCity, as shown in Figure \ref{Fig:OneDBTab}.  The
+CombinedStateCity table would have the following columns: State, Life
+Expectancy, Income, City, Precipitation, and Temperature.  Now if a state's
+life expectancy changed, all instances of this state would need to be changed
+in the new large table.  For example, if the state had the information for
+three cities recorded, then the state would have three rows in the
+CombinedStateCity table and each row would need to change the value for the
+state's life expectancy.  If instead only one record was kept per state, as
+shown in Figure \ref{Fig:DBTab} above, then this process would be less error
+prone and would require less memory to store the data.  This example shows
+that there are advantages to storing the data in separate structures.
 
@@
   \begin{center}
-    \scalebox{0.5}{\includegraphics{oneDatabaseTable.jpg}}
+    \scalebox{0.5}{\includegraphics{newOneDatabaseTable.jpg}}
     \caption{ Example of a Combined Table. }
@@
 structures.  For example, suppose that you have experimental microarray data,
-which looks at gene expression levels.  This experimental data can be
+which look at gene expression levels.  This experimental data can be
 stored as a two dimensional data structure, such as a matrix, where rows are
-genes and columns are samples.  Linked to this
-experimental data is meta data that gives information about the genes studied
-in the microarray experiment.  This meta data may be a gene ontology
-graph that gives the molecular functions of these genes.  These two data sets
-(the experimental data and the meta data) are linked through the genes, but
-they are different data structures.  The experimental data is a matrix and
-the meta data is a graph.
+genes and columns are samples.  Linked to this experimental data are meta-data
+that give information about the genes studied in the microarray experiment.
+For example, this meta-data may be a Gene Ontology graph that gives the
+molecular functions of these genes.  These two data sets (the experimental
+data and the meta-data) are linked through the genes.  Each gene can have more
+than one molecular function and each molecular function can have more than one
+gene annotated at that function so there is a many-to-many relationship
+between the genes and the molecular functions.  Though these two data sets are
+linked, the microarray data and the Gene Ontology data are different data
+structures.  The experimental microarray data can be represented as a matrix
+and the Gene Ontology meta-data can be represented as a graph.
 
 Now for the original MVC design, where one data set is expected, it is not
-clear how to combine these two data sets into one data set.  The meta data
+clear how to combine these two data sets into one data set.  The meta-data
 graph could be stored as a matrix, but a graph is a data structure that can be
@@
 
-Currently, linked data sets are a common occurrence that can not be easily
-visualized simultaneously.  Any time there is meta data linked to experimental
-or study data there is a multiple data set situation where visually exploring
-the data would be helpful to understand relationships between variables.  In
-this paper, the design of software to create linked views based on multiple
-linked data sets will be discussed.  The goals of this software will include
-creating linked views of linked data sets, creating interactive
-views where the response to an event can be changed, and creating
-an extensible design that will allow future users to make additions.
-
 \section{Extending the MVC Design}\label{Sec:Extend}
 
-As discussed in Section \ref{Sec:Intro}, the MVC design expects one data set
-with all views referencing this one data set.  However, several examples were
-shown where multiple data sets are linked.  Since linking between multiple
-data sets would be useful to visualize, the next step will be to determine how
-the MVC design can be revised to account for multiple data sets.
+As discussed in Section \ref{Sec:Intro}, the MVC design expects a single
+conceptual unit, such as a data frame or a matrix,
+with all views referencing this one data set.  However, in Section
+\ref{Ssec:Limit} several examples were shown where multiple data sets are
+linked.  These examples included linked tables in relational databases and
+experimental data linked to meta-data.  Being able to visualize variables
+simultaneously from different data sets that are linked would be useful for
+exploring relationships between these variables.  Thus, the next step will be
+to determine how the MVC design can be revised to account for multiple data
+sets.
 
@@
 combine the multiple data sets into one large data set.  With this new
-combined data, the original MVC design will be appropriate for creating linked
+combined data, the original MVC design is appropriate for creating linked
 views.  Thus, the strength of combining the data sets into one large data set
-is that the MVC paradigm will already be appropriately set up to deal with the
+is that the MVC paradigm is already appropriately set up to deal with the
 one large data set.
@@
 same data structure, it may not be possible to combine the data sets into one
-data structure that retains all of the original information.  Thus, another
+data structure that retains all of the original information.  Also, combining
+the data into one data structure may give the appearance that some variables
+have a relationship that does not exist.  For example, suppose that in a
+relational database, table 1's rows are linked to table 2's rows and table 3's
+rows are also linked to table 2's rows.  Just because table 1 and table 3 are
+both linked to table 2 does not necessarily mean that there is a meaningful
+relationship between the data in table 1 and table 3.  However, by combining
+the data in these three tables, it would give the appearance that there is a
+relationship between the data in table 1 and table 3.  Thus, another
 way of linking views based on multiple data sets needs to be designed.
@@
 will need to be altered to fit the data.  It has been mentioned that the MVC
-design is intended to deal with one data set.  Now that we have several data
-sets, why not have several MVC objects?  One MVC object to handle each data
-set.  Thus, within each MVC object, one could create linked views of that
-particular data set, which is the same as the original MVC design.  Then
-between MVC objects a messaging system would need to be designed so that a
-change in one data set would be sent to the other linked data sets.
+design is intended to deal with one data set.  A more straightforward approach
+may be to link MVCs.  In linked MVCs, each MVC can be coherent and the
+interactions between MVCs can be encapsulated in a set of messages.  One MVC
+object to handle each data set.  Thus, within each MVC object, one could
+create linked views of that particular data set, which is the same as the
+original MVC design.  Then between MVC objects a messaging system would need
+to be designed so that a change in one data set would be sent to the other
+linked data sets.
 
@@
 state level and one with information at the city level.  These tables are
-shown in Table \ref{Tab:CityState}.  The State data includes Life Expectancy
+shown in Table \ref{Tab:CityState}.  The State data include Life Expectancy
 (in years from 1970) and Income (per capita from 1974) data
-while the City data includes Precipitation (in inches from 1975) and Average
+while the City data include Precipitation (in inches from 1975) and Average
 High Temperature in January (in Fahrenheit from 2005)
@@
 appropriate messaging system between MVC objects and deciding how the MVC
-objects will be related.  First, there are several questions that need to be
-considered when creating the message system.  What information should be
-passed between MVC objects in the messages?  Should there be a system to stop
-messages from propagating between MVC objects?  How will the messages be
-created and handled?  Second, there are also several questions with regard to
-how the MVC objects will be related.  Should the MVC objects be related like
-the tables in a relational database where the relationships are determined by
-linked keys in different tables?  Should the MVC objects be linked through some
-type of hierarchy, like a parent-child relationship where one MVC creates
-another MVC and thus, the original MVC is the parent?  These decisions will be
-based on the data that needs to be visualized and will be discussed in the
-following sections as the design for multiple MVC objects is explained.
+objects will be related.  In designing an appropriate messaging system between
+MVC objects, consideration will be given to what information should pass
+between MVC objects and to how messages will propagate between MVC objects.
+In deciding the relationship between the MVC objects, consideration will be
+given to how the multiple data sets are related.  Clearly, the relationship
+between the MVC objects should reflect the relationship between the data
+sets.  The two examples of linked data sets that have been discussed
+previously were linked tables in a relational database, and experimental data
+that are linked to meta-data.  In this paper, the focus will be on experimental
+data that are linked to meta-data, as those are the data we are most
+interested in visualizing.  In this situation, the experimental data can be
+thought of as determining what meta-data are studied.  In the example of
+microarray experiment data linked to Gene Ontology meta-data, only Gene
+Ontology data that refer to genes in the microarray experiment will be
+included in the meta-data.  Thus, there is a hierarchical relationship between
+these data sets.  Therefore, the design of the linking between MVC objects
+will be based on a hierarchical relationship between the data sets, where one
+MVC object creates another MVC object.  These decisions will be discussed in
+the following sections as the design for multiple MVC objects is explained.
 
@@
 
-Implementing a design to create linked, interactive views of multiple linked
-data sets will be done in the R language in two packages, called
+Implementing an extensible design to create linked, interactive views of
+linked data sets will be done in the R language in two packages, called
 \Rpackage{MVCClass} and \Rpackage{iSNetwork}.  The \Rpackage{MVCClass} package
-will implement the design of the MVC classes and methods.  The
+will implement the design of the MVC classes and generic functions.  The
 \Rpackage{MVCClass} package will define the individual model, view, and
@@
 package will also define the message classes that will allow the components to
-communicate with each other.  The goal of separating these class definitions
-and methods into one package is that they will be available for other R
-packages that implement linked views of data, such as the \Rpackage{iSPlot}
-and the \Rpackage{iSNetwork} packages.
+communicate with each other.  Thus, the \Rpackage{MVCClass} package will
+contain all of the classes necessary for creating linked views.   The goal of
+separating these class and generic function definitions into one package is
+that they will be available for other R packages that implement linked views
+of data, such as the \Rpackage{iSPlot} and the \Rpackage{iSNetwork} packages.
 
 With the class definitions created in the \Rpackage{MVCClass} package, the
-\Rpackage{iSNetwork} package will implement the graphical user interface
-(GUI) so that users can load data, create views, and interact with the views
-through a user interface with menus.  To create the GUI, \Rpackage{iSNetwork}
-will use the R packages, \Rpackage{RGtk} and \Rpackage{gtkDevice}.
-\Rpackage{RGtk} is an R package that allows one to interact with Gtk functions
-from the R interface.  Gtk is an open-source X Window toolkit for creating
-user interfaces.  The \Rpackage{gtkDevice} package creates a Gtk device by
-turning a Gtk drawing area into a device that acts like an R device, but still
-responds to events so that the view can be interactive.  For more information
-about these packages, please see the paper, ``Creating Linked, Interactive
-Plots to Explore Multivariate Data''.
-
-The next subsections will explain the class definitions and methods that
+\Rpackage{iSNetwork} package can decide which classes from \Rpackage{MVCClass}
+to use based on the overall goals of the combined software.  The
+\Rpackage{iSNetwork} package will also determine how objects of these classes
+will behave because the methods for the classes are defined in the
+\Rpackage{iSNetwork} package.  In addition to implementing the behavior of
+objects, the \Rpackage{iSNetwork} package will also create a graphical user
+interface (GUI) so that users can load data, create views, and interact with
+the views through menus on the user interface.  As an alternative to the GUI,
+users will also have access to the same functionality through command line
+functions.
+
+To create the GUI, \Rpackage{iSNetwork} will use the R packages,
+\Rpackage{RGtk} and \Rpackage{gtkDevice}.  \Rpackage{RGtk} is an R package
+that allows one to interact with Gtk functions from the R interface.  Gtk is
+an open-source X Window toolkit \cite{Gtk} for creating user interfaces.  The
+\Rpackage{gtkDevice} package creates a Gtk device by turning a Gtk drawing
+area into a device that acts like an R device, but still responds to events so
+that the view can be interactive.  For more information about these packages,
+please see the paper, ``Creating Linked, Interactive Views to Explore
+Multivariate Data'' \cite{EW05}.
+
+The next subsections will explain the class definitions and the methods that
 pertain to a single model-view-controller object.  Thus, the model classes,
 the view classes, the controller, the \Robject{MVC} class (which binds the
-model, view and controller classes together), and the message classes that pass
-information between the three components in one MVC object will all be
-discussed in the following subsections.  All of these definitions are
-implemented in the \Rpackage{MVCClass} package.  Message class definitions
-that pertain to communication between the MVC objects will be given in
-Section \ref{Sec:MultMVC}.
+model, view and controller classes together), and the message classes that
+pass information between the three components in one MVC object will all be
+discussed in the following subsections.  All of the class definitions are
+implemented in the \Rpackage{MVCClass} package and all of the methods are
+defined in the \Rpackage{iSNetwork} package.  Message class definitions that
+pertain to communication between the MVC objects will be discussed in Section
+\ref{Sec:MultMVC}.
 
@@
 
-The model class is responsible for storing and updating the data.  These two
-functions are reflected in Figure \ref{Fig:Model}, which shows the inheritance
-structure for the model classes.  Here the virtual class,
-\Robject{gModel}, has the slots modelData, linkData, virtualData, and
-modelName.  These four slots are common information that all models will
-need.  The modelData slot is the data set for this model, the linkData slot is
-a list of two functions, \Rfunction{linkToParent} and \Rfunction{linkToChild},
-which link this model to its parent and child models, respectively (see Section
-\ref{Ssec:MultLink}), the virtualData slot is data pertaining to the views
-that needs to be stored with the model so that it can be shown in all views of
-this model, and the modelName slot is the name of the model.
-
-\clearpage
+In the MVC design, the model component is the place where data are stored and
+updated.  Thus, the model is a storage location that can perform certain
+operations.  Information can be sent to the model to add, update, and remove
+data and the model can send out messages that changes have occurred to the
+data that the model stores.  Thus, the object model for the model classes will
+represent both the slots that are needed to store data and the methods that
+are needed to perform the operations of the model.
+
+Figure \ref{Fig:Model} shows the inheritance structure for the model classes.
+Here the virtual class, \Robject{gModel}, has the slots \Rslot{modelData},
+\Rslot{linkData}, \Rslot{virtualData}, and \Rslot{modelName}.  These four
+slots are common information that all models will need to store.  The
+\Rslot{modelData} slot is the data set for this model, the \Rslot{linkData}
+slot is a list of two functions, \Rfunction{toParent} and
+\Rfunction{fromParent}, which let this model send information to its parent
+model and receive information from its parent model, respectively (see Section
+\ref{Ssec:MultLink}), the \Rslot{virtualData} slot is data pertaining to the
+views that needs to be stored with the model so that it can be shown in all
+views of this model, and the \Rslot{modelName} slot is the name of the model.
+
+%\clearpage
 
@@
 
-For specific model classes, there are \Robject{dfModel}, \Robject{graphModel},
+As for specific model classes, the classes that are used in the
+\Rpackage{iSNetwork} package are \Robject{dfModel}, \Robject{graphModel},
 and \Robject{exprModel}.  The \Robject{dfModel} class represents a model where
-the modelData has class data frame (or matrix), the \Robject{graphModel} class
-represents a model where the modelData has class graph, and the
-\Robject{exprModel} class represents a model where the modelData has class
-exprSet.
-
-As mentioned previously, the model classes are also responsible for updating
-the data and thus, you will notice that all three model classes have an
-\Rfunction{updateModel} method.  This method is called whenever the data needs
-to be updated in response to a user interacting with a view.  Thus, the
-\Rfunction{updateModel} method is actually updating the virtualData slot
-because it is this slot in the model classes that ensures that all views are
-appropriately visually linked.
+the \Rslot{modelData} slot has class data frame (or matrix), the
+\Robject{graphModel} class represents a model where the \Rslot{modelData} slot
+has class graph, and the \Robject{exprModel} class represents a model where the
+\Rslot{modelData} slot has class exprSet.
+
+As mentioned previously, the model classes must also perform certain
+operations and these operations are reflected in the methods that are defined
+for the classes.  All three model classes have an \Rfunction{updateModel}
+method that is defined in the \Rpackage{iSNetwork} package.  This method is
+called whenever the data need to be updated in response to a user interacting
+with a view.  Thus, the \Rfunction{updateModel} method is actually updating the
+\Rslot{virtualData} slot because it is this slot in the model classes that
+ensures that all views are appropriately visually linked.
 
@@
 would like to represent.  For example, a user may want to add a new model
-class called \Robject{tsModel} to represent time series data.
-
-Note that the model classes must also notify views when the data has changed
+class called \Robject{tsModel} to represent time series data.  To make this
+addition, a user could extend the \Robject{gModel} class in his or her own
+package.
+
+Note that the model classes must also notify views when the data have changed
 so that the views can be updated.  The communication between a model and its
@@
   \begin{center}
-    \scalebox{0.7}{\includegraphics{ViewClass.jpg}}
+    \scalebox{0.7}{\includegraphics{newViewClass.jpg}}
     \caption{ Inheritance for View Objects. }
@@
 
-The \Robject{genView} class has three slots, dataName, win, and winNum.
-dataName is the name of the model that the view displays, win is the Gtk
-window object that holds the view, and winNum is the number of the window so
-that the window can be identified in the Window menu on the control window
-(see Section \ref{Ssec:OneCont}).  All views will need to know these three
-pieces of information so \Robject{genView} will bind the view classes
-together.
+The \Robject{genView} class has three slots, \Rslot{dataName}, \Rslot{win},
+and \Rslot{winNum}.  \Rslot{dataName} is the name of the model that the view
+displays, \Rslot{win} is the Gtk window object that holds the view, and
+\Rslot{winNum} is the number of the window so that the window can be
+identified in the Window menu on the GUI (see Section \ref{Ssec:OneCont}).
+All views will need to know these three pieces of information so
+\Robject{genView} will bind the view classes together.
 
@@
 are a two dimensional data structure, such as a matrix or a data frame.  The
-information needed for this view is stored in the slot, clist, which is the
-Gtk spreadsheet object.
+information needed for this view is stored in the slot, \Rslot{clist}, which
+is the Gtk spreadsheet object.
 
 When creating a plot of a data set, there is a general class, called
-\Robject{plotView}, that has the slots, plotDevice, plotPar, and drArea, which
-store the device number of the plot, the plotting parameters, and the Gtk
-drawing area object, respectively.  This class is not intended to have any
-objects as it just represents a general plot.  The plot classes that can have
-objects are the specific plot classes, \Robject{sPlotView},
-\Robject{graphView}, and \Robject{heatmapView}, which inherit from the
-\Robject{plotView} class.
+\Robject{plotView}, that has the slots, \Rslot{plotDevice}, \Rslot{plotPar},
+and \Rslot{drArea}, which store the device number of the plot, the plotting
+parameters, and the Gtk drawing area object, respectively.  This class is a
+virtual class as it represents a general plot.  The instantiable plot classes
+are the specific plot classes, \Robject{sPlotView}, \Robject{graphView}, and
+\Robject{heatmapView}, which inherit from the \Robject{plotView} class.
 
 The \Robject{sPlotView} class represents a scatterplot view and it has the
-extra slots, dfRows, colx, and coly, where dfRows are the row names from the
-model that are shown in the plot, colx is the column name from the model that
-is shown as the x variable in the plot, and coly is the column name from the
-model that is shown as the y variable in the plot.  The \Robject{graphView}
-class represents a graph plot view and it contains no extra slots besides
-those in the \Robject{plotView} class.  The \Robject{heatmapView} represents a
-heatmap view of the data and it contains the slot, ordering, which is a list
+extra slots, \Rslot{dfRows}, \Rslot{xvar}, and \Rslot{yvar}, where
+\Rslot{dfRows} are the row names from the model that are shown in the plot,
+\Rslot{xvar} is the variable name from the model that is shown as the x
+variable in the plot, and \Rslot{yvar} is the variable name from the model
+that is shown as the y variable in the plot.  The \Robject{graphView} class
+represents a graph plot view and it contains no extra slots besides those in
+the \Robject{plotView} class.  The \Robject{heatmapView} represents a heatmap
+view of the data and it contains the slot, \Rslot{ordering}, which is a list
 that is returned from the \Rfunction{heatmap} function to give information
-about the dendrogram ordering.  Note that some of these views are only
-applicable for certain model types. For example, the \Robject{graphView} class
-will only make sense for a view of a graph object, which would be stored in the
-\Robject{graphModel} class.
+about the ordering in the row and column dendrograms.  Note that some of these
+views are only applicable for certain model types. For example, the
+\Robject{graphView} class will only make sense for a view of a graph object,
+which would be stored in the \Robject{graphModel} class.
 
@@
 view classes have methods that correspond to user interaction, such as
-\Rfunction{clickEvent}.  For \Robject{sPlotView}, the \Rfunction{clickEvent}
-method is called when Gtk emits a signal that a mouse button has been pressed
-and the \Rfunction{motionEvent} method is called when Gtk emits a signal that
+\Rfunction{clickEvent}.  Recall from Section \ref{Ssec:Backg} that when an
+event occurs Gtk emits a signal that is caught by the signal handler.  The
+signal handler then calls the callback function and it is in the callback
+function that the corresponding method is called.  For example, when the
+``clicked'' event occurs on a scatterplot view, the callback function calls the
+scatterplot's \Rfunction{clickEvent} method.  Using the information that is
+passed to the \Rfunction{clickEvent} method in the \Rslot{where} parameter,
+the method converts the Gtk pixel coordinates that indicate where the event
+occurred into user coordinates and then into an object from the plot, such as
+a point on the plot.
+
+For \Robject{sPlotView}, the \Rfunction{clickEvent} method is called when Gtk
+emits a signal that a mouse button has been pressed and the
+\Rfunction{motionEvent} method is called when Gtk emits a signal that
 the cursor has been moved over the view.  Both the \Rfunction{clickEvent} and
@@
 determine which point on the plot was clicked or under the cursor,
-respectively.  For \Robject{spreadView}, the \Rfunction{clickEvent} method is
-called when Gtk emits a signal that a row in the spreadsheet has been selected
-or unselected and the \Rfunction{clickEvent} will determine which row has
-changed.  For \Robject{graphView}, the \Rfunction{clickEvent} method is called
-when Gtk emits a signal that a mouse button has been pressed and the
+respectively, using the data in the \Rslot{where} parameter.  For
+\Robject{spreadView}, the \Rfunction{clickEvent} method is called when Gtk
+emits a signal that a row in the spreadsheet has been selected or unselected
+and the \Rfunction{clickEvent} method will determine which row was changed.
+For \Robject{graphView}, the \Rfunction{clickEvent} method is called when Gtk
+emits a signal that a mouse button has been pressed and the
 \Rfunction{motionEvent} is called when Gtk emits a signal that the cursor has
@@
 \Rfunction{motionEvent} methods will determine which node was clicked or under
-the cursor, respectively.  Note that the \Robject{heatmapView} class has
-neither a \Rfunction{clickEvent} or a \Rfunction{motionEvent} method because
-currently, it is not possible to convert the user coordinates of where the
-event occurred into some object on the view (such as the row and column
-coordinates of the heatmap).  Thus, the \Robject{heatmapView} is the only view
-that is not currently interactive.
+the cursor, respectively, using the data in the \Rslot{where} parameter.
+
+The \Robject{heatmapView} class is different from the other views in that it
+has neither a \Rfunction{clickEvent} or a \Rfunction{motionEvent} method.  The
+current implementation of the \Rfunction{heatmap} function does not return
+information about where plot objects are placed on the device.  As a result,
+it is not possible to convert the user coordinates of where the event occurred
+into an object (such as the row and column coordinates of the heatmap) on
+the view.  Thus, the \Robject{heatmapView} class is the only view that is not
+currently interactive.
 
 As with the model classes, the view classes need an update method and thus, all
-views have an \Rfunction{updateView} method.  This method must take the
+views have an \Rfunction{updateView} method that is defined in the
+\Rpackage{iSNetwork} package.  This method must take the
 information that a model has just been updated and appropriately update the
@@
 represents the data so for a \Robject{spreadView} object the
-\Rfunction{updateView} may select or unselect a row on the spreadsheeet while
-for a \Robject{sPlotView} object the \Rfunction{updateView} may change the
-appearance of a point on the plot.
+\Rfunction{updateView} method may select or unselect a row on the spreadsheeet
+while for a \Robject{sPlotView} object the \Rfunction{updateView} method may
+change the appearance of a point on the plot.
+
+Note that when updating a view in R we do not have access to objects on a
+view.  Thus, a point on a plot is not an object that allows us to update its
+attributes and then have that point redrawn to reflect its new attributes.
+Instead in R we must currently draw over a portion of the view to update the
+view.  However, if the method of creating R plots changed, the
+\Rfunction{updateView} methods could be modified.
 
@@
 \Robject{dPlotView}.  This new class, \Robject{dPlotView}, would have slots
-names dfRows and col to store which rows and which column, respectively, were
-used to calculate the density shown in the view.
+named \Rslot{dfRows} and \Rslot{col} to store which rows and which column,
+respectively, were used to calculate the density shown in the view.
 
@@
 
-The controller must perform several tasks, including storing information,
-deciding how a view will respond to user interaction, and allowing users to
-load data, to create views, and to quit the program.  These tasks can be
-separated into two functions, where the first is that the controller needs to
-be a storage location and the second is that the controller must be some type
-of interactive interface that allows the user to determine the course of
-actions.
-
-For the first function the controller will be an environment that is
-stored in the controller slot of a MVC object, which will be discussed in more
-detail in Section \ref{Ssec:OneMVC}.  This environment will store data
-pertaining to state, such as information about appearance of the user interface
-and the views.
-
-For the second function the controller will be a GUI with menus to
-allow the user to choose which action will occur next.  This aspect of the
-controller, which is implemented in \Rpackage{iSNetwork}, is shown in the
-Figure \ref{Fig:ContWin} and will be referred to as the control window.
+The controller must perform two tasks: the first task is to store information
+and the second task is to allow the user to determine the course of actions.
+For the first function, which is implemented in the \Rpackage{MVCClass}
+package, the controller is implemented as an environment that is stored in the
+\Rslot{controller} slot of a \Robject{MVC} instance, which will be discussed in
+more detail in Section \ref{Ssec:OneMVC}.  Although other options exist for
+storing data, such as a list or a database, an environment is nice because in
+R an environment is not copied each time a value within the environment
+changes.  The controller in this first role will need to store information
+pertaining to state, such as information about the appearance of the views in
+a particular model-view-controller entity.
+
+For the second function, which is implemented in the \Rpackage{iSNetwork}
+package, the controller gives the user access to the application program
+interface (API), either through the GUI or through command line functions.
+Through the API the user can load data, create views, set the response to user
+interaction with a view, quit the program and perform other functions.  On the
+GUI, there are menus to let users decide which action will be taken next.  If
+the user would rather not use the GUI, there are command line functions that
+provide access to the same functionality that is available through the GUI.
+The GUI is shown in Figure \ref{Fig:ContWin} and will be referred to as the
+control window.
 
@@
 control window, but the menu items under each menu will change depending on
-the type of model that is active.  Because the goal is to have multipled
-linked MVC objects, there will be more than one MVC object and thus, more than
-one model. However, the control window menus will only perform operations on
-one MVC object at a time.  Thus, in the \Rpackage{iSNetwork} package there
-will be the concept of the ``active'' MVC object, where the ``active'' MVC
-object (and thus, the ``active'' model) means that operations performed
-through the control window menus will affect that MVC object.  For example,
-the Display menu allows users to create views of the active model, but only
-certain views are appropriate for a type of model, as mentioned in Section
-\ref{Ssec:OneViews}.  If the active model was a data frame model
+the type of model that is active.  Because the goal is to have multiple
+linked MVC objects, there will be more than one MVC object loaded and thus,
+more than one model. However, the control window menus and the command line
+functions will only perform operations on one MVC object at a time.  Thus, in
+the \Rpackage{iSNetwork} package there will be the concept of the ``active''
+MVC object, where the ``active'' MVC object (and thus, the ``active'' model)
+means that operations performed through the control window menus or through
+the command line functions will affect that MVC object.  For
+example, the Display menu allows users to create views of the active model,
+but only certain views are appropriate for a type of model, as mentioned in
+Section \ref{Ssec:OneViews}.  If the active model was a data frame model
 (\Robject{dfModel}), then the views that the user could create would be a
@@
 model in the ``active'' MVC is of class \Robject{dfModel}, then the Display
-menu will have menu items to create a scatterplot and a spreadsheet.
-
-Because interacting with the views should be driven by the user,
-the Event menu allows the user to determine what response will occur when
-there is user interaction with a view.  For example, the user may initially
-want a point on a plot to become colored when the left button is pressed and
-the cursor is over that point.  Then the user may decide that the point
-should be hidden when the left button is pressed and the cursor is over that
-point.  The Event menu lets the user decide if anything will happen in
-response to a left button click, a middle button click, a right button click,
-and mouse movement.  More information will be given in Section
-\ref{Ssec:OneEvent} about linking an event with a function so that user
-interaction with a view is flexible enough to allow for multiple outcomes.
+menu will have menu items to create a scatterplot and a spreadsheet.  Also, if
+the ``active'' MVC is of class \Robject{dfModel}, then the command line
+functions will only let the user create a scatterplot and a spreadsheet view
+of the active model.
 
@@
 data sets is to create a software package that is flexible and extensible,
-users will be able to add menus and menu items to the control window.  As an
+users will be able to add menus and menu items to the control window through
+the command line to access new functionality that they have created.  As an
 example if a future programmer wants to add a new type of view, such as a
@@
 allowing users to change the controller, they can add new functionality that
-is specific to their problem.
+is specific to their problem.  Note that when users make additions to the API,
+they should add new command line functions in addition to adding menus and
+menu items to access that functionality.
 
@@
 or a key press event, is what allows users to have an interactive
-environment.  Then when an event happens something will occur in response
-depending on the callback function linked to the event.  Now to create a
-flexible environment we do not want a callback function that always does the
-same thing when an event occurs.  For example, when a left button press event
-occurs, we may sometimes want a point to be colored and at other times we may
-want a point to be hidden.  To allow this flexibility, a class called
-\Robject{gEventFun} was created.  By using this class, users will be able to
-change the response to an event.
-
-The \Robject{gEventFun} class stores all of the information about a potential
-callback function.  When a user wants to connect this callback function to an
-event, this can be performed through the Event menu on the control window (see
-Section \ref{Ssec:OneCont}).  The controller
-environment in each MVC object will store which \Robject{gEventFun} object
-(representing the callback function) is linked with each event.  Thus, the
-user can decide the response for button clicks and mouse movements, and the
-user can change these responses at any time through the control window.
+environment.  When an event occurs, the callback function linked to the event
+will determine the response to that event.  To create a flexible environment
+it would be nice to be able to change the response, which is determined by the
+callback function, to an event.  For example, when a left button press event
+occurs, a user may sometimes want a point to be colored and at other times she
+may want a point to be hidden.
+
+To allow this flexibility in the response to an event, a class called
+\Robject{gEventFun} was created.  The \Robject{gEventFun} class stores all of
+the information needed to create a particular response to an event.
+Clearly, the response to an event can be performed by a single function,
+which is called the callback function, as discussed in Section
+\ref{Ssec:Backg}.  However, different responses to an event may require
+functions with different signatures.  Using the example in the previous
+paragraph, a callback function that colored a point would need two parameters,
+the new color to use and the point to color, while a callback function that
+hid a point would need just one parameter, the point to hide.  Since the
+intention is to be able to interchange the responses to an event, the
+signatures of different callback functions should be the same.  These design
+decisions determined the slots in the \Robject{gEventFun} class and will be
+discussed more fully in the following paragraphs.
+
+As shown in Figure \ref{Fig:EventFun}, the slots in the \Robject{gEventFun}
+class are \Rslot{callFun}, \Rslot{shortName}, and \Rslot{preprocessFun}.  The
+\Rslot{callFun} slot is the name of the callback function, the
+\Rslot{shortName} slot is a short description of the function, and the
+\Rslot{preprocessFun} slot is a character vector of all preprocessing
+functions that must be called before the callback function.
 
@@
 
-As shown in Fig \ref{Fig:EventFun}, the slots in the \Robject{gEventFun} class
-are callFun, shortName, and preprocessFun.  The callFun slot is the name of
-the callback function, the shortName slot is a short description of the
-function, and the preprocessFun slot is a character vector of all
-preprocessing functions that must be called before the callback function.
-
-The callback function that is stored in the callFun slot must take only one
-parameter and the class of this parameter will depend on the model class in
-the active MVC object.  For example, for a graph model, the
-parameter will be of class \Robject{AgNode}, representing a node in the graph.
-For a data frame model, the parameter will be of class character, representing
-a row in the data frame.
+The callback function that is stored in the \Rslot{callFun} slot must take
+only one parameter.  As discussed previously, all callback functions must have
+the same signature so that different callback functions can be interchanged in
+the code to allow different responses to an event.  All callback functions
+will need to know what object they should act on so this will be the one
+parameter that the callback functions take.  Any other information a callback
+function may need will be set using the preprocessing functions.  The class of
+the one parameter for the callback function will depend on the model class in
+the active MVC object.  For example, for a graph model, the parameter will be
+of class \Robject{AgNode}, representing a node in the graph.  For a data frame
+model, the parameter will be of class character, representing a row in the
+data frame.
 
@@
 
+When a user wants to link a callback function stored in an \Robject{gEventFun}
+instance to an event, this can be performed either through the GUI or through
+the command line functions.  The controller environment in each MVC object,
+which will be discussed in detail in Section \ref{Ssec:OneMVC},
+will store which \Robject{gEventFun} object (representing the callback
+function) is linked with each event.  Note that not all callback functions
+make sense for each event so the controller environment will also store a list
+for each event of appropriate \Robject{gEventFun} instances that can be linked
+to that event.
+
 \subsection{MVC}\label{Ssec:OneMVC}
 
-The \Robject{MVC} class is the class that will bind the model, view, and
+The \Robject{MVC} class is the class that binds the model, view, and
 controller objects that refer to the same data set into one object.  Thus, the
-creation of a MVC object occurs whenever a new data set is added.  A
+creation of a MVC object occurs whenever a new data set is loaded.  A
 MVC object can have one model and each model must be in one
@@
 MVC object and the model object are identified by the same name,
-which is stored in the modelName slot of the model object.
+which is stored in the \Rslot{modelName} slot of the model object.  When
+loading a new model through \Rpackage{iSNetwork's} API, the name of the model
+will be checked to ensure that it is unique.
 
 As shown in Figure \ref{Fig:MVCClass}, the \Robject{MVC} class has five slots
-which are, model, viewList, controller, parentMVC, and childMVCList.  The
-model slot stores the model object, the viewList slot is a list of the view
-objects for this model, the controller slot is an environment that stores
-information for this MVC, the parentMVC slot is the name of the parent model
-(MVC), and the childMVCList slot is a list of the children models
+which are, \Rslot{model}, \Rslot{viewList}, \Rslot{controller},
+\Rslot{parentMVC}, and \Rslot{childMVCList}.  The \Rslot{model} slot stores
+the model object, the \Rslot{viewList} slot is a list of the view objects for
+this model, the \Rslot{controller} slot is an environment that stores
+information for this MVC, the \Rslot{parentMVC} slot is the name of the parent
+model (MVC), and the \Rslot{childMVCList} slot is a list of the children models
 (MVCs).  Note that model and MVC objects use the same name so the
-parentMVC slot refers to both the name of the model and the name of the
-MVC.  Similarly, for the childMVCList slot, the names in the list
+\Rslot{parentMVC} slot refers to both the name of the model and the name of the
+MVC.  Similarly, for the \Rslot{childMVCList} slot, the names in the list
 refer to both the name of the model and the name of the MVC.
@@
 
-\clearpage
+%\clearpage
 
@@
 
-A MVC object can have at most one parent MVC, but it can have many child
-MVCs.  The relationship between MVC objects shows that child MVC objects are
-derived from the parent MVC object (i.e. the model in the child MVC is derived
-from the model in the parent MVC).  An easy example of a child model derived
-from a parent model is to start with a data frame as the parent model and then
-take a subset of this data frame as the child model.
-
-\subsection{Messages}\label{Ssec:OneMess}
+A design decision was that a new model (and thus, a new MVC object) could
+either be loaded or could be created from an already loaded MVC object.  As
+mentioned in Section \ref{Ssec:MMVC}, the focus in this paper will be on
+experimental data that are linked to meta-data, as the linked data sets that we
+are most interested in visualizing.  In this situation, the experimental data
+(and its MVC object) can be thought of as determining what meta-data are
+studied.  Thus, there is a hierarchical relationship between these data sets
+where one MVC object creates another MVC object.  This focus led to the
+decision that a MVC object can have at most one parent MVC, from which it was
+derived, but it can have many child MVCs.  An easy example of a child model
+derived from a parent model is to start with a data frame as the parent model
+and then take a subset of this data frame as the child model.
+
+\subsection{Messages Within a MVC Object}\label{Ssec:OneMess}
 
@@
 communication between the model, view and controller is crucial for the pieces
-to work together and yet, still be independent of each other.
+to work together and yet, still be independent of each other.  For information
+about messages that are passed between MVC objects, please see Section
+\ref{Ssec:MultMess}.
 
@@
 and \Rfunction{handleMessage}, because all messages must be created and
-handled in some manner so that the message is read and acted upon.  Also,
+handled in some manner so that the message is read and acted upon.  The
+\Rfunction{handleMessage} method can be considered a central dispatch for
+messages because all messages will be sent to the \Rfunction{handleMessage}
+method, which will determine where the information should be sent.   Also,
 since messages are sent when something has changed, either through addition,
@@
 \Robject{gAddMessage}, and the update message, \Robject{gUpdateMessage},
-contain three slots, dataName, mData, and type.  dataName is the character
-string that gives the name of the model, mData is a list of data needed to
-perform the addition or alteration operation, and type is a character string
-that gives the type of addition or alteration to perform.  Although
-\Robject{gAddMessage} and \Robject{gUpdateMessage} are not virtual classes, it
-is not intended to create objects of these classes.
+contain three slots, \Rslot{dataName}, \Rslot{mData}, and \Rslot{type}.
+\Rslot{dataName} is the character string that gives the name of the model,
+\Rslot{mData} is a list of data needed to perform the addition or alteration
+operation, and \Rslot{type} is a character string that gives the type of
+addition or alteration to perform.  Both \Robject{gAddMessage} and
+\Robject{gUpdateMessage} are virtual classes.
 
@@
 a model.  This message is crucial for linking because it is responsible for
-ensuring that once a model has been updated this information must propagate to
-any views of the model and this information must also be passed along to any
+ensuring that once a model has been updated this information propagates to
+any views of the model.  This same information must also be passed along to any
 other MVC object that is related to this model.  Because this information gets
 passed to other MVC objects, which will be discussed further in Section
-\ref{Sec:MultMVC}, this message also has one new slot, from, in addition to
-the slots it inherits from \Robject{gUpdateMessage}.  The from slot is the
-name of the model that the update data message came from.  This slot is
-necessary because the update data message can come from one of two sources:
-one the message can occur because of user interaction with a view
-of this model or the message may occur because a different model that is
+\ref{Sec:MultMVC}, this message also has one new slot, \Rslot{from}, in
+addition to the slots it inherits from \Robject{gUpdateMessage}.  The
+\Rslot{from} slot is the name of the model that the update data message came
+from.  This slot is necessary because the update data message can come from
+one of two sources: the message can occur because of user interaction with a
+view of this model or the message may occur because a different model that is
 linked to this model has been updated.  So the update data message may come
 from within this MVC object or it may come from a different MVC object that
-has changed its model and is related to this MVC object.  
-Information on passing messages between MVC
-objects will be given in Section \ref{Sec:MultMVC}.  Finally, the
-\Robject{gUpdateViewMessage} class represents a message to update a view.
-This message will be created and handled whenever the model is updated so that
-the views accurately reflect the model.
+has changed its model and is related to this MVC object.  Information on
+passing messages between MVC objects will be given in Section
+\ref{Sec:MultMVC}.  Finally, the \Robject{gUpdateViewMessage} class represents
+a message to update a view.  This message will be created and handled whenever
+the model is updated so that the views accurately reflect the model.
 
@@
 \Robject{gUpdateControlMessage}, could convey information that told the
-control window to be updated in response to a change in the active MVC.
-Recall from Section \ref{Ssec:OneCont} that the active MVC is the MVC object
-that the menus on the control window refer to so any of the operations chosen
-from the control window menus will be applied to this MVC object (the active
-MVC).  The goal for the message classes is that the design is flexible enough
-to allow a user to make additions or changes.
+control window to be updated in response to a change in the active view (the
+active view is the view that has the focus).  Currently, the control window
+does not reflect what the active view is, but if a view was created that could
+be rotated, then a user may want the control window to have some type of
+interface to allow the user to control the rotation.  The goal for the message
+classes is that the design is flexible enough to allow a user to make
+additions or changes.
 
@@
 
-Up until now we have skirted around the fact that multiple MVC objects are
-being used in this design for linking views of multiple linked data sets.
-Each MVC object will represent one model and its views.  Now for the views of
-different data sets to be linked, they must link through the data sets so this
-linking information must be stored somewhere.  Also, there must be
-message classes to pass information between MVC objects.
+We now describe the implementation of multiple MVC objects for linking views
+of multiple linked data sets.  Each MVC object will represent one model and
+its views.  Now for the views of different data sets to be linked, they must
+link through the data sets so this linking information must be stored
+somewhere.  Also, there must be message classes to pass information between
+MVC objects.
+
+Thinking of experimental data that are linked to meta-data as an example, there
+will be two MVC objects, one for the experimental data and one for the
+meta-data.  In the example of experimental microarray data that are linked to
+the Gene Ontology molecular function graph that was first mentioned in Section
+\ref{Ssec:Limit}, these two data sets are linked in the following manner: the
+Affymetrix identifiers in the experimental microarray data are linked to Locus
+Link identifiers, which are in turn linked to the Gene Ontology molecular
+function terms.  Thus, to pass information between these two data sets a table
+that links Affymetrix identifiers to Gene Ontology molecular function terms
+must be stored or there must be a function to call that converts from
+Affymetrix identifiers to Gene Ontology molecular function terms and vice
+versa.  Once it is possible to link these data sets, then information must be
+passed between them through messages to keep the views synchronized.
 
@@
 for how the current MVC relates to other MVC objects and these slots are
-parentMVC and childMVCList.  These slots show that MVC objects are related
-through a tree hierarchy where a child model is derived from its parent
-model.  Consider Figure \ref{Fig:SmallHier} shown below.  If firstMVC's model
-changes, then this information first gets passed to secondMVC, which may make
-changes to its model based on this information.  Then secondMVC passes this
-information to thirdMVC based on the changes secondMVC made to its model.
+\Rslot{parentMVC} and \Rslot{childMVCList}.  These slots show that MVC objects
+are related through a tree hierarchy where a child model is derived from its
+parent model.  Consider Figure \ref{Fig:SmallHier} shown below.  If firstMVC's
+model changes, then this information first gets passed to secondMVC, which may
+make changes to its model based on this information.  Then secondMVC passes
+this information to thirdMVC based on the changes secondMVC made to its model.
 Thus, information passing between firstMVC and thirdMVC only occurs through
@@
 MVC object relates to data in a different MVC object.  This conversion is done
-by two functions that are stored in the linkData slot of the model objects.
-The linkData slot is a list with two elements: a \Rfunction{linkToParent}
-function and a \Rfunction{linkToChild} function.
-
-When a child MVC is created from an existing MVC, the child model's linkData
-slot will be filled with the two functions, \Rfunction{linkToParent} and
-\Rfunction{linkToChild}.  When the model in the child MVC is updated, the
-\Rfunction{linkToParent} function in the linkData slot will be used to convert
-the child model's data change into useful information for the parent model.
+by two functions that are stored in the \Rslot{linkData} slot of the model
+objects.  The \Rslot{linkData} slot is a list with two elements: a
+\Rfunction{toParent} function and a \Rfunction{fromParent} function.
+
+When a child MVC is created from an existing MVC, the child model's
+\Rslot{linkData} slot will be filled with the two functions,
+\Rfunction{toParent} and \Rfunction{fromParent}.  When the model in the child
+MVC is updated, the \Rfunction{toParent} function in the \Rslot{linkData} slot
+will be used to convert the child model's data change into useful information
+for the parent model.
 
@@
 needs to be passed to the child model.  However, a parent can have multiple
-child MVC objects so which \Rfunction{linkToChild} function should be used?
-Here, when a parent is updated and needs to pass information to its children,
-the \Rfunction{linkToChild} function is stored with each child MVC because a
-parent MVC may have multiple child MVC objects that all need to process this
-information in a different way.  Thus, when a child MVC object is created and
-its model's linkData slot is filled, the \Rfunction{linkToParent} function
+child MVC objects so the parent must look in each child model's
+\Rslot{linkData} slot for the \Rfunction{fromParent} function.  The
+\Rfunction{fromParent} function will convert the parent model's data change
+into useful information for this particular child model.  Thus, the parent
+model's data change may be converted into different information for different
+child models and this can happen because each child model has its own
+\Rfunction{fromParent} function.  When a child MVC object is created and
+its model's \Rslot{linkData} slot is filled, the \Rfunction{toParent} function
 will convert information from this child model into useful information for its
-parent MVC and the \Rfunction{linkToChild} function will convert a message
+parent MVC and the \Rfunction{fromParent} function will convert a message
 from its parent MVC into useful information for this child MVC.
 
-An example of how these functions, \Rfunction{linkToParent} and
-\Rfunction{linkToChild}, are used when messages are passed between MVCs will
+An example of how these functions, \Rfunction{toParent} and
+\Rfunction{fromParent}, are used when messages are passed between MVCs will
 be given in Section \ref{Ssec:MultMess}.
@@
   \begin{center}
-    \scalebox{0.6}{\includegraphics{MessageClass2.jpg}}
+    \scalebox{0.6}{\includegraphics{newMessageClass2.jpg}}
     \caption{ Inheritance for Message Classes that are Passed Between MVCs. }
@@
 inherits from the \Robject{gAddMessage} class, with the same slots of
-dataName, mData, and type.  The dataName slot is the name of the new model (and
-new MVC), the mData slot is a list that contains the model data and virtual
-data to fill the modelData and virtualData slots, respectively, of the new
-model, and the type slot gives the type of the model (currently, the options
-are ``exprSet'', ``graph'', or ``data.frame'').  As with the previous message
+\Rslot{dataName}, \Rslot{mData}, and \Rslot{type}.  The \Rslot{dataName} slot
+is the name of the new model (and new MVC), the \Rslot{mData} slot is a list
+that contains the model data and virtual data to fill the \Rslot{modelData}
+and \Rslot{virtualData} slots, respectively, of the new model, and the
+\Rslot{type} slot gives the type of the model (currently, the options are
+``exprSet'', ``graph'', or ``data.frame'').  As with the previous message
 classes, the \Robject{gAddChildMessage} class has two methods:
@@
 method for the \Robject{gAddDataMessage} class does) and then it fills the
-linkData slot of the new child model so that the \Rfunction{linkToParent} and
-\Rfunction{linkToChild} functions are available for messaging between MVC
+\Rslot{linkData} slot of the new child model so that the \Rfunction{toParent}
+and \Rfunction{fromParent} functions are available for messaging between MVC
 objects and finally it stores information in the child MVC object's controller
 that determines if the child model was created from a subset of the parent
-model.  Thus, the \Rfunction{handleMessage} method both creates the child MVC
-object and relates it to the parent MVC object.
-
-The next two message classes pertain to updating MVC objects when a parent MVC
-or child MVC object has changed.  These classes are
-\Robject{gUpdateParentMessage} and \Robject{gUpdateChildMessage} and they both
+model.  Thus, the \Rfunction{handleMessage} method for the
+\Robject{gAddChildMessage} class both creates the child MVC object and relates
+it to the parent MVC object.
+
+The next two message classes pertain to sending information to linked MVC
+objects when a parent MVC or child MVC object has changed.  These classes are
+\Robject{gSendParentMessage} and \Robject{gSendChildMessage} and they both
 inherit from the \Robject{gMessage} class, as shown in Figure
-\ref{Fig:BetwMess}.  Even though they are both update messages, they do not
-inherit from the \Robject{gUpdateMessage} class because the information they
-need to contain is a \Robject{gUpdateDataMessage} object.  Thus, the
-\Robject{gUpdateParentMessage} class has one slot, childUpdateDataMessage, and
-this slot contains the \Robject{gUpdateDataMessage} object that was used to
-update the child model.  Similarly, the \Robject{gUpdateChildMessage} class
-has two slots, parentUpdateDataMessage and childName, where the
-parentUpdateDataMessage slot contains the \Robject{gUpdateDataMessage} object
-that was used to update the parent model and the childName slot contains the
-name of the child model that is being updated (because a parent MVC can have
-more than one child MVC).
-
-As with all the other message classes, these two update messages have two
+\ref{Fig:BetwMess}.  Because the messages are sending information to a linked
+MVC object, they both contain a slot for a \Robject{gUpdateDataMessage} object
+that was just used to update a MVC object.  The \Robject{gSendParentMessage}
+class has one slot, \Rslot{childUpdateDataMessage}, and this slot contains the
+\Robject{gUpdateDataMessage} object that was used to update the child model.
+Similarly, the \Robject{gSendChildMessage} class has two slots,
+\Rslot{parentUpdateDataMessage} and \Rslot{childName}, where the
+\Rslot{parentUpdateDataMessage} slot contains the \Robject{gUpdateDataMessage}
+object that was used to update the parent model and the \Rslot{childName} slot
+contains the name of the child model that is being updated because a parent
+MVC can have more than one child MVC.
+
+As with all the other message classes, these two send messages have two
 methods: \Rfunction{initialize} and \Rfunction{handleMessage}.  The
 \Rfunction{initialize} method properly sets the slots of the new object.  The
-\Rfunction{handleMessage} method for a \Robject{gUpdateParentMessage} will
+\Rfunction{handleMessage} method for a \Robject{gSendParentMessage} will
 take the \Robject{gUpdateDataMessage} object from the child model and convert
 it to a \Robject{gUpdateDataMessage} object for the parent model using the
-\Rfunction{linkToParent} function.  Similarly, the \Rfunction{handleMessage}
-method for a \Robject{gUpdateChildMessage} will take the
+\Rfunction{toParent} function.  Similarly, the \Rfunction{handleMessage}
+method for a \Robject{gSendChildMessage} will take the
 \Robject{gUpdateDataMessage} object from the parent model and convert it to a
 \Robject{gUpdateDataMessage} object for the child model using the
-\Rfunction{linkToChild} function.
+\Rfunction{fromParent} function.  An example of how these messages are passed
+and converted is shown in Figure \ref{Fig:MessPass}.
 
@@
 that pertains to the model for one MVC must be converted to useful information
-for the linked model and this conversion is done by the
-\Rfunction{linkToParent} and \Rfunction{linkToChild} functions.
-
-\begin{figure}[ht]
-  \begin{center}
-    \scalebox{0.6}{\includegraphics{MessagePassing.jpg}}
-    \caption{ Message Passing Between MVCs. }
+for the linked model and this conversion is done by the \Rfunction{toParent}
+and \Rfunction{fromParent} functions.
+
+\begin{figure}[ht]
+  \begin{center}
+    \scalebox{0.55}{\includegraphics{newMessagePassing.jpg}}
+    \caption{ Message Passing Between and Within MVCs. }
     \label{Fig:MessPass}
@@
 
-Figure \ref{Fig:MessPass} shows how messages are passed within one MVC as well
-as how this message gets passed on to its parent MVC.  In this picture, the
-message passing starts when a user interacts with the y vs. x scatterplot
-depicting model 2.  After the user interaction with the scatterplot, step 1
-(indicated on the figure as a circled 1) is to send an update data message to
-model 2.  As soon as model 2 is updated, step 2 is to send an update view
-message so that the scatterplot view is updated to reflect the changed data.
-In step 3, an update parent message is sent because MVC 2 (which is the MVC
-object for model 2) has a parent MVC, which is MVC 1.  For step 4, the
-\Rfunction{linkToParent} function converts the child's update data message
-into an update data message that the parent will understand.  In step 5, this
-new update data message that was created by the \Rfunction{linkToParent}
+Figure \ref{Fig:MessPass} is very similar to Figure \ref{Fig:firstMP} shown in
+Section \ref{Ssec:Backg}.  Now, however, Figure \ref{Fig:MessPass} shows how
+MVC objects are defined and how messages are passed within one MVC as well
+as how this information gets passed on to its parent MVC using messages.  In
+this picture, the message passing starts when a user interacts with the y
+vs. x scatterplot depicting model 2.  After the user interaction with the
+scatterplot, step 1 (indicated on the figure as a circled 1) is to send an
+update data message to model 2.  As soon as model 2 is updated, step 2 is to
+send an update view message so that the scatterplot view is updated to reflect
+the changed data.  In step 3, a send parent message is sent because MVC 2
+(which is the MVC object for model 2) has a parent MVC, which is MVC 1.  For
+step 4, the \Rfunction{toParent} function converts the child's update data
+message into an update data message that the parent will understand.  In step
+5, this new update data message that was created by the \Rfunction{toParent}
 function will be sent to model 1.  Finally for step 6, after model 1 has been
 updated, an update view message will be sent to the view of model 1.  Now all
-of the data and the views are in synch.
+of the data and the views are synchronized.
 
@@
 paper, including linked tables from a relational database, and meta data that
-is linked to experimental data.  Until now it has not been possible to create
-linked views of these data sets.  However, creating linked, interactive views
-of underlying linked data sets would be very helpful for exploratory data
-analysis.  Thus, this paper discussed the creation of two R packages,
-\Rpackage{MVCClass} and \Rpackage{iSNetwork}, which allow users to create
-linked, interactive views of linked data sets based on expanding the
-model-view-controller paradigm.  Now for visualizing linked data sets, the
-model-view-controller design, which is a well known and thus, well tested
-software paradigm, was extended to multiple MVC objects where each MVC object
-contains one data set and its views.  This solution broke down the problem of
-linked, interactive views of linked data sets into components (MVC objects)
-that can be reused and linked.
-
-Although one of the goals of the \Rpackage{MVCClass} and
-\Rpackage{iSNetwork} packages was to create a design that was extensible,
-certain design decisions, such as expecting the data sets to be linked through
+are linked to experimental data.  The model-view-controller (MVC) paradigm was
+discussed as a useful design for creating linked views of a data set.
+Starting with the MVC paradigm, this paper provided a design for implementing
+linked, interactive views of linked data sets based on expanding the MVC
+paradigm.  For visualizing linked data sets, the MVC design, which is a well
+known and thus, well tested software paradigm, was extended to multiple MVC
+objects where each MVC object contains one data set and its views.  This
+solution broke down the problem of linked, interactive views of linked data
+sets into components (MVC objects) that can be reused, extended and linked.
+
+Once the design focused on MVC objects as the component of interest for
+creating linked views of linked data sets, the next step was to determine how
+these MVC objects would be related.  Because the focus in this paper was on
+experimental data that are linked to meta-data, a hierarchical relationship
+between MVC objects was built into the MVC class definition.  The experimental
+data can be thought of as determining what meta-data are studied and thus, the
+meta-data are derived from the experimental data, which shows a parent-child
+relationship.
+
+Another consideration in the design was creating a method for communicating
+between MVC objects.  Already communication within a MVC object was sent using
+instances of the message classes.  Thus, it seemed a natural extension to have
+message classes that pertained to passing information between MVC objects.
+The goal of the message classes was to allow communication between independent
+components in the design.
+
+The goals for this paper were to create linked views of linked data sets, to
+create interactive views where the response to an event can be changed, and to
+create an extensible design that will allow future users to make additions.
+The first goal was accomplished by implementing MVC objects as components that
+communicate with each other through messages.  The second goal of creating
+interactive views was met by using the \Rpackage{RGtk} and
+\Rpackage{gtkDevice} packages to create devices that could respond to events.
+In addition the \Robject{gEventFun} class was designed so that users could
+alter the response to an event.  Finally, the last goal of creating an
+extensible design was achieved through the inheritance structure for classes,
+which allows for extensions by the user, and through command line functions,
+which let the user create new menu items to add to the GUI and also let the
+user create new callback functions that can be linked to events.
+
+Although extensibility was a goal, certain design decisions, such as expecting
+the data sets to be linked through 
 a parent-child hierarchy, may not work with some data sets.  As an example,
@@
 with these data sets.  However, the design allows users to add new
-types of models, new types of views, new menu items on the control window and
-new message classes so there is still flexibility for additions as long as the
-data sets are connected through parent-child relationships.
+types of models, new types of views, and new message classes so there is still
+flexibility for additions as long as the data sets are connected through
+parent-child relationships.  These new class definitions could be created by
+the user in a new package that extends the \Rpackage{MVCClass} package that
+was discussed in this paper.  Other additions that the user may want to
+perform, such as adding new menus to the GUI and adding new responses to an
+event can be performed through command line functions that are defined in the
+\Rpackage{iSNetwork} package.
 
@@
 parent or child MVC, the MVC can either decide to ignore the message or read
-the message (which results in this MVC's model being updated).  However, the
+the message, which results in this MVC's model being updated.  However, the
 MVC object can not decide to only see certain types of messages.  It must
-either accept all messages or ignore all messages.  A future goal for message
-passing is to allow more selectivity in which messages will be accepted
-between MVC objects so that the linking between MVC objects is more
-sophisticated.
+either accept all messages and update its model based on the messages, or
+ignore all messages.  A future goal for message passing is to allow more
+selectivity in which messages will be accepted between MVC objects so that the
+linking between MVC objects is more sophisticated.
 

### 2005-12-13 inst/doc/MVCClass.Rnw
Changed the vignette to incorporate the new classes.
git-svn-id: file:///home/git/hedgehog.fhcrc.org/bioconductor/trunk/madman/Rpacks/MVCClass@15363 bc3139a8-67e5-0310-9ffc-ced21a209358

@@
 need.  The modelData slot is the data set for this model, the linkData slot is
-a list of two functions, \Rfunction{linkToParent} and \Rfunction{linkToChild},
+a list of two functions, \Rfunction{toParent} and \Rfunction{fromParent},
 which link this model to its parent and child models, respectively (see Section
@@
   \begin{center}
-    \scalebox{0.7}{\includegraphics{VigMessageClass2.jpg}}
+    \scalebox{0.7}{\includegraphics{newVigMessageClass2.jpg}}
     \caption{ Inheritance for Message Classes that are Passed Between MVCs. }
@@
 or child MVC object has changed.  These classes are
-\Robject{gUpdateParentMessage} and \Robject{gUpdateChildMessage} and they both
+\Robject{gSendParentMessage} and \Robject{gSendChildMessage} and they both
 inherit from the \Robject{gMessage} class, as shown in Figure
-\ref{Fig:BetwMess}.  Even though they are both update messages, they do not
-inherit from the \Robject{gUpdateMessage} class because the information they
+\ref{Fig:BetwMess}.  The information these two classes
 need to contain is a \Robject{gUpdateDataMessage} object.  Thus, the
-\Robject{gUpdateParentMessage} class has one slot, childUpdateDataMessage, and
+\Robject{gSendParentMessage} class has one slot, childUpdateDataMessage, and
 this slot contains the \Robject{gUpdateDataMessage} object that was used to
-update the child model.  Similarly, the \Robject{gUpdateChildMessage} class
+update the child model.  Similarly, the \Robject{gSendChildMessage} class
 has two slots, parentUpdateDataMessage and childName, where the

### 2005-12-20 inst/paper/paper2.tex
Added new pictures and the bibliography file and updated the paper.
git-svn-id: file:///home/git/hedgehog.fhcrc.org/bioconductor/trunk/madman/Rpacks/MVCClass@15482 bc3139a8-67e5-0310-9ffc-ced21a209358

@@
 \textbf{Dana-Farber Cancer Institute}\\ 
-\textbf{December 9, 2005} \\
+\textbf{December 19, 2005} \\
 \vspace{1in} 
@@
 Views to Explore Multivariate Data'' \cite{EW05}, so the focus in this paper
-will be one specific method of visualizing multivariate data that is
-particularly useful for exploratory data analysis.  This paper will discuss
-creating linked, interactive views of linked data sets.
+is on one specific method of visualizing multivariate data that is
+particularly useful for exploratory data analysis; creating linked,
+interactive views of linked data sets.
 
 First brief definitions of linked views, interactive views and linked data
-sets will be given.  Linked views mean that if a component that represents
+sets are given.  Linked views mean that if a component that represents
 data (such as a point on a plot or a row in a spreadsheet) changes its
-appearance on one view, then the corresponding component on a second view will
-also change its appearance.  For components on different views to be
+appearance on one view, then the corresponding component on a second view
+also changes its appearance.  For components on different views to be
 corresponding, the data displayed in those components must come from the same
@@
 the four variables and fifty rows for the fifty states.  In this model, the
-entities would be the states.  If two scatterplots were created as views of
+entities are the states.  If two scatterplots were created as views of
 this model, where the first plot displayed murder rate versus assault rate and
 the second plot displayed rape rate versus percent urban population, then the
-points on each plot that referred to the state of Montana would be
-corresponding components.
-
-Interactivity is an idea that will be familiar from using the Internet.  For
+points on each plot that referred to the state of Montana are
+corresponding components.   Linked views, which have a long history in
+statistics, are now considered a standard feature of interactive data
+visualization software \cite{GGobi}.  Linked views have been implemented in
+several pieces of statistical software, such as XLisp-Stat and GGobi.
+Implementing linked views is based on the model-view-controller (MVC) design
+\cite{DesignPatterns}, which is a widely used and well understood paradigm.
+The MVC design is discussed in more detail in Section \ref{Ssec:Design}.
+
+Interactivity is an idea that is familiar from using the Internet.  For
 example, clicking on a hyperlink on a web page results in a new web page
@@
 user name and password to view one's email.  Interactivity means that there is
-a response to some user action.
+a response to some user action.  It is a powerful paradigm and allows users to
+quickly navigate and explore complex documents or complex data sets.  As an
+example, suppose that a user has two views: one view is a graph of molecular
+functions for genes and the other view is a heatmap of gene expression data.
+If the user clicks on a node that represents a particular molecular function,
+then the response to this action could be to highlit all genes in the heatmap
+view that express this molecular function.  This interactivity allows the
+user to quickly see if genes with the same molecular function have similar
+expression levels.
 
 One particular model for creating interactive applications involves three
-components: some user action or input, which will be referred to as an event;
-a response to that action, which will be executed by a function that is called
+components: some user action or input, which is referred to as an event;
+a response to that action, which is executed by a function that is called
 a callback function; and a method that connects the event to the callback
-function, which will be referred to as the signal handler.  The events can
-range from the user moving the cursor over a plot to pressing a button on a
-graphical user interface (GUI) to pressing certain keys.  Notice that the
-events almost always occur over some type of interface, though some events,
-such as key press events may occur without an interface.  The callback
-function is a unique function because its signature is determined by the event
-that the callback function is linked to.  The parameters of the callback
-function may include information about the event, such as the location where
-the event occurred, and information about the object that just caused the
-event, such as the button on the GUI.  The signal handler is the piece that
-must look to see if an event has occurred and then when the event has
-occurred, the signal handler calls the appropriate callback function.  The
-strengths of this model of interactivity are that programmers can decide which
-events to notice by deciding which signal handlers to define, programmers
-can change the response to an event by changing the callback function that is
-linked to an event, and the system's event loop can continue working on other
-things until a signal is emitted because an event occurred.
+function, which is referred to as the signal handler.  The events can
+range from the user moving the cursor over a plot, to pressing a button on a
+graphical user interface (GUI), to pressing certain keys.  Events are
+interpreted based on which window has focus.  For example, in the XWindows
+system, if a user types C on the keyboard and the window that has focus is a
+text editor, then the letter C is printed on the text editor's screen.
+Any other open windows that do not have
+focus do not respond to this event.  The callback function is a unique
+function because its signature is determined by the event that the callback
+function is linked to.  The parameters of the callback function may include
+information about the event, such as the on screen location of the event, and
+information about the object that had focus when the event occurred, such as a
+button on the GUI.  The signal handler is the piece that must look to see if
+an event has occurred and then when the event has occurred, the signal handler
+calls the appropriate callback function.
+
+The strengths of this model of interactivity are that programmers can decide
+which events to react to by deciding which signal handlers to implement,
+programmers can change the response to an event by changing the callback
+function that is linked to an event, and the system's event loop can continue
+working on other things until a signal is emitted when an event occurs.  A
+system's event loop controls how work is processed.  The event loop checks for
+work to do and handles that work.  In the model of interactivity described
+above when a signal handler is implemented by the programmer, this results in
+the event loop periodically checking to see if an event has occurred.  Thus,
+the event loop can perform other work and then check back to see if an event
+has occurred.
 
@@
 The signal handler, which is the method that connects the event to the
-callback function, is not noticed by the user, but it is handled by the
-system's event loop when a signal is emitted.
-
-\clearpage
+callback function, is handled by the system's event loop when a signal is
+emitted.
+
+%\clearpage
 
@@
 but there is a relationship between these different conceptual groupings.
-Linked data sets are an idea that will be familiar to users of relational
+Linked data sets are an idea that is familiar to users of relational
 databases.  Data in a relational database are typically stored in many
 linked tables with rows in different tables linked to each other through a
-unique identifier called a key.  In this example, each table would be
+unique identifier called a key.  In this example, each table is
 considered a data set and they are linked through the keys.  Another situation
@@
 scatterplots, spreadsheets, and heatmaps, and these views are linked because
-they are based on the same underlying model.  Thus, users can decide which
+they are based on the same underlying data.  Thus, users can decide which
 views best visually represent their data.  Also, since the views are
 interactive, users can change the views while looking at them to make the
-views more informative about the underlying data structure.  This flexibility
-in linked, interactive views gives users a powerful tool for visually
-exploring their data.
-
-In Section \ref{Ssec:Design}, the design of linked, interactive views based on
-the model-view-controller (MVC) paradigm will be discussed.  However, the
-limitation with this design is that it expects one data set, which limits the
-type of data we can visualize.  There are many
-situations where multiple data sets occur, particularly any time there are
-experimental or study data linked to meta-data.  Being able to visualize data
-from multiple data sets would be helpful to understand the relationships
-between variables.  Thus, after discussing the limitations of the MVC design
-in Section \ref{Ssec:Limit}, an extension of the MVC design will be discussed
-in Section \ref{Sec:Extend}.  This extension will allow the creation of linked,
-interactive views based on linked data sets.  The software that is discussed
-in this paper will have three goals: to create linked views of linked data
-sets, to create interactive views where the response to an event can be
-changed, and to create an extensible design that will allow future users to
-make additions.
+views more informative about the underlying structure in the data.  This
+flexibility in linked, interactive views gives users a powerful tool for
+visually exploring their data.  However, linked views can not solve all
+problems.  Clearly, only a few linked views can be viewed simultaneously by
+the user because only a certain number of views can be shown on a computer
+screen.  Thus, there is a limit to the number of dimensions that can be
+represented by linked views.
 
@@
 received by data set 1 and data set 1 updates its data.  Finally, in step 6,
-data set 1 tells its view to update itself.  Thus, in this example, the views
-are synchronized based on the underlying linked data sets.
-
-\clearpage
+data set 1 tells its view to update itself.  This example shows the components
+and the communication between the components that are necessary to create
+linked views of underlying linked data sets.
+
+In the design presented in this paper, the components consist of MVC objects
+and communication within and between MVC objects is performed by message
+objects.  This solution uses the power and simplicity of the MVC design and
+expands it to the multiple data set problem.  Now we can build on the original
+MVC design, which has been well tested, instead of starting a new design from
+scratch.  The strength of this multiple linked MVC design is that it uses MVC
+components, which can be individually created and then linked.  Thus, these
+components are reusable.  These design decisions are discussed further in
+Sections \ref{Sec:OneMVC} and \ref{Sec:MultMVC}.
+
+%\clearpage
 
@@
 
+In Section \ref{Ssec:Design}, the design of linked, interactive views based on
+the MVC paradigm is discussed.  However, the limitation with this design is
+that it expects one data set, which limits the type of data we can visualize.
+There are many situations where multiple data sets occur, particularly any
+time there are experimental or study data linked to meta-data.  Being able to
+visualize data from multiple data sets is helpful to understand the
+relationships between variables.  Thus, after discussing the limitations of
+the MVC design in Section \ref{Ssec:Limit}, an extension of the MVC design is
+discussed in Section \ref{Sec:Extend}.  This extension supports the creation of
+linked MVC objects.  The design of individual MVC objects is discussed in
+Section \ref{Sec:OneMVC} while the implementation of linked MVC objects is
+discussed in Section \ref{Sec:MultMVC}.  The goals of the software discussed
+in this paper are to create linked views of linked data sets, to create
+interactive views where the response to an event can be changed, and to create
+an extensible design that allows future users to make additions.
+
 \subsection{Design and Implementation of Linked, Interactive
@@
 this method.  A popular design for creating linked views of data is the
-model-view-controller (MVC) paradigm.  This design consists of three types
-of objects: the controller, which defines what actions occur in response to
-user input; the view, which consists of displays of the data; and the model,
-which manages the data.  This paradigm is powerful because it decouples the
-views from the model.  Thus, only one copy of the data needs to be stored and
-then all views refer to this data set.  With all views based on the same data,
-any changes to the data can be propagated to all views, which will create
-linked views.
+MVC paradigm.  This design consists of three types of objects: the controller,
+which defines what actions occur in response to user input; the view, which
+consists of displays of the data; and the model, which manages the data.  This
+paradigm is powerful because it decouples the views from the model.  Thus,
+only one copy of the data needs to be stored and then all views refer to this
+data set.  With all views based on the same data, any changes to the data can
+be propagated to all views, which creates linked views.
 
@@
 and the model must notify the views when a change occurs so that the views
-will be updated.  This design allows multiple views of the model that are
-linked, but are not aware of each other, as shown in Figure \ref{Fig:ExMVC}
-\cite{DesignPatterns}.  Each view depends only on visually representing the
-model and thus, it is not necessary for a view to be aware of other views.  By
-keeping each view ignorant of other views, other views can added and removed
-without affecting a different view.
+are updated.  This design allows multiple views of the model that are
+linked, but are not aware of each other, as shown in Figure \ref{Fig:ExMVC}.
+Each view is responsible for visually representing some aspect of the model
+and thus, it is not necessary for a view to be aware of other views.  By
+keeping each view ignorant of other views, views can added and removed without
+affecting each other.
 
@@
 Notice that although the components (model, view, and controller) in the MVC
-design are independent of one other, there still needs to be communication
+design are independent of one other, there needs to be communication
 between them so that information can be shared.  Examples of this
@@
 response to user interaction with the view.  This message passing between
-components in the MVC paradigm will be discussed in detail in Section
+components in the MVC paradigm is discussed in detail in Section
 \ref{Ssec:OneMess}.
 
-The ``Creating Linked, Interactive Views to Explore Multivariate Data''
-\cite{EW05} paper discussed the creation of R software to fulfill three goals:
+In \cite{EW05}, we discuss the creation of R software to fulfill three goals:
 to create views that are linked, to create views that are interactive, and to
 create a design that is extensible.  To satisfy the overall goals of the
-software, two R packages, called \Rpackage{MVCClass} and \Rpackage{iSPlot},
-were created.  The \Rpackage{MVCClass} package implemented the class and
+design, two R packages, called \Rpackage{MVCClass} and \Rpackage{iSPlot},
+were created.  The \Rpackage{MVCClass} package implements the class and
 generic function definitions needed for the MVC design and the
-\Rpackage{iSPlot} package created the front end to let the user create linked,
+\Rpackage{iSPlot} package creates the front end to let the user create linked,
 interactive views of two-dimensional data, such as data frames and matrices,
-through either a GUI or through command line functions.  The reason for
+through a GUI or through command line functions.  The reason for
 separating the software into two packages was that it allowed the class
 definitions in the \Rpackage{MVCClass} package to be used by any other R
-package that would like to create linked views.
+package that wanted to create linked views.
 
@@
 
-The MVC design assumes that there is one data set that all views will
-reference.  But what if the data consisted of more than one conceptual
-grouping?  For example, consider data stored in a relational database, where
-there are multiple tables of data and rows in different tables are linked
-through a unique identifier called a key.  Although these tables of data could
-be combined into one large table, there are several reasons not to do this.
-Keeping the data in separate tables reduces redundancy, particularly when the
-rows are related by a one-to-many or a many-to-many relationship.  Also
-separate data tables generally lead to more maintainable data.  By storing
-only one copy of each piece of information (in a row in one table), it is
-easier to ensure that the data are correctly updated when a change occurs.
-All other tables that are connected with this information can then just
-reference this row that has been updated.
+The MVC design assumes that there is one data set, which is called the model,
+that all views reference.  But problems arise when the data consist of more
+than one conceptual grouping.  For example, consider data stored in a
+relational database, where there are multiple tables of data and rows in
+different tables are linked through a unique identifier called a key.
+Although these tables of data could be combined into one large table, there
+are several reasons not to do this.  Keeping the data in separate tables
+reduces redundancy, particularly when the rows are related by a one-to-many or
+a many-to-many relationship.  Also separate data tables generally lead to more
+maintainable data.  By storing only one copy of each piece of information (in
+a row in one table), it is easier to ensure that the data are correctly
+updated when a change occurs.  All other tables that are connected with this
+information can then just reference this row that has been updated.
 
@@
 CombinedStateCity, as shown in Figure \ref{Fig:OneDBTab}.  The
-CombinedStateCity table would have the following columns: State, Life
+CombinedStateCity table has the following columns: State, Life
 Expectancy, Income, City, Precipitation, and Temperature.  Now if a state's
-life expectancy changed, all instances of this state would need to be changed
+life expectancy changes, all instances of this state need to change
 in the new large table.  For example, if the state had the information for
-three cities recorded, then the state would have three rows in the
-CombinedStateCity table and each row would need to change the value for the
+three cities recorded, then the state has three rows in the
+CombinedStateCity table and each row needs to change the value for the
 state's life expectancy.  If instead only one record was kept per state, as
-shown in Figure \ref{Fig:DBTab} above, then this process would be less error
-prone and would require less memory to store the data.  This example shows
+shown in Figure \ref{Fig:DBTab} above, then this process is less error
+prone and requires less memory to store the data.  Another consideration in
+combining data sets is that it may give the appearance of a relationship
+between variables that does not exist.  For example, it may not make sense to
+compare the Income data stored in the State table to the Precipitation data
+stored in the City table.  Combining the data tables assumes that there is a
+relationship between the variables in all of the tables.  This example shows
 that there are advantages to storing the data in separate structures.
@@
 that give information about the genes studied in the microarray experiment.
-For example, this meta-data may be a Gene Ontology graph that gives the
-molecular functions of these genes.  These two data sets (the experimental
-data and the meta-data) are linked through the genes.  Each gene can have more
-than one molecular function and each molecular function can have more than one
-gene annotated at that function so there is a many-to-many relationship
-between the genes and the molecular functions.  Though these two data sets are
-linked, the microarray data and the Gene Ontology data are different data
-structures.  The experimental microarray data can be represented as a matrix
-and the Gene Ontology meta-data can be represented as a graph.
+For example, this meta-data may be terms from Gene Ontology, which form a
+directed acyclic graph, describing the molecular functions carried out by
+these genes.  These two data sets (the experimental data and the meta-data)
+are linked through the genes.  Each gene can have more than one molecular
+function and each molecular function can have more than one gene annotated at
+that function so there is a many-to-many relationship between the genes and
+the molecular functions.  Though these two data sets are linked, the
+microarray data and the Gene Ontology data are different data structures.
 
@@
 well represented as an object with slots for nodes and edges, especially if
-the graph is sparse.  As with the relational database tables, it seems
-best in terms of data storage, alteration, and retrieval to keep the two data
-sets separate.  Also, keeping the two data sets separate does not force a data
-structure on the data that may not fit well.
+the graph is sparse.  As with relational database tables, combining the data
+sets complicates data storage, alteration, and retrieval.  For example, by
+combining the data sets a structure may be forced on the data that may not fit
+well, especially if the original data sets are instances of different classes.
+In contrast, keeping the data sets stored separately in their original
+structures simplifies the processes of storing, updating, and accessing the
+data.
 
@@
 
-As discussed in Section \ref{Sec:Intro}, the MVC design expects a single
-conceptual unit, such as a data frame or a matrix,
-with all views referencing this one data set.  However, in Section
-\ref{Ssec:Limit} several examples were shown where multiple data sets are
-linked.  These examples included linked tables in relational databases and
-experimental data linked to meta-data.  Being able to visualize variables
-simultaneously from different data sets that are linked would be useful for
-exploring relationships between these variables.  Thus, the next step will be
-to determine how the MVC design can be revised to account for multiple data
-sets.
+As discussed in Section \ref{Sec:Intro}, the MVC design typically uses a single
+conceptual unit, such as a data frame or a matrix, for the model.  In Section
+\ref{Ssec:Limit} we provided several examples where it is advantageous to link
+multiple data sets.  These examples included linked tables in relational
+databases and experimental data linked to meta-data.  Being able to visualize
+variables simultaneously from different data sets that are linked is useful for
+exploring relationships between these variables.  We next describe how the MVC
+design can be revised and extended to accommodate multiple data sets.
 
@@
 combine the multiple data sets into one large data set.  With this new
-combined data, the original MVC design is appropriate for creating linked
-views.  Thus, the strength of combining the data sets into one large data set
-is that the MVC paradigm is already appropriately set up to deal with the
-one large data set.
+combined data, the original MVC design can be used to create linked views.
 
 Unfortunately, as discussed in Section \ref{Ssec:Limit}, combining
-multiple data sets into one large data set is not always appropriate or even
-possible.  Even if the multiple data sets have the same data structure,
-combining the data sets often results in redundant data and makes updating the
-data more complicated.  In the situation where the data sets are not of the
-same data structure, it may not be possible to combine the data sets into one
-data structure that retains all of the original information.  Also, combining
-the data into one data structure may give the appearance that some variables
-have a relationship that does not exist.  For example, suppose that in a
-relational database, table 1's rows are linked to table 2's rows and table 3's
-rows are also linked to table 2's rows.  Just because table 1 and table 3 are
-both linked to table 2 does not necessarily mean that there is a meaningful
-relationship between the data in table 1 and table 3.  However, by combining
-the data in these three tables, it would give the appearance that there is a
-relationship between the data in table 1 and table 3.  Thus, another
-way of linking views based on multiple data sets needs to be designed.
+multiple data sets into one large data set can be problematic.  Even if the
+multiple data sets have the same data structure, combining the data sets often
+results in redundant data and makes updating and representing the
+data more complicated.  In the situation where the data sets do not have the
+same structure, it may be necessary to define a new class to appropriately
+represent the combined data.  Also, combining the data into one data structure
+may give the appearance that some variables have a relationship that does not
+exist.  Just because two data sets are linked does not necessarily mean that
+there is a meaningful relationship between their data.  Thus, another way
+of linking views based on multiple data sets needs to be designed.
 
@@
 
-Because of the difficulty of combining multiple data sets, the MVC paradigm
-will need to be altered to fit the data.  It has been mentioned that the MVC
-design is intended to deal with one data set.  A more straightforward approach
-may be to link MVCs.  In linked MVCs, each MVC can be coherent and the
-interactions between MVCs can be encapsulated in a set of messages.  One MVC
-object to handle each data set.  Thus, within each MVC object, one could
-create linked views of that particular data set, which is the same as the
-original MVC design.  Then between MVC objects a messaging system would need
-to be designed so that a change in one data set would be sent to the other
-linked data sets.
+A more straightforward approach may be to link MVCs.  Each MVC is a coherent
+object and the interactions between MVCs are encapsulated in a set of
+messages.  We use one MVC object to handle each data set.  Thus, within a MVC
+object, one can create linked views of that particular data set, using the
+standard MVC design.  But between MVC objects we design a second messaging
+system so that a change in one data set is sent to the other linked data sets.
 
@@
 High Temperature in January (in Fahrenheit from 2005)
-data.  These two data sets are linked by the CityState table that links cities
-with the states they reside in.  Now for these two data sets there would be
+data.  These two data sets are linked by the StateCity table that links cities
+with the states they reside in.  Now for these two data sets there are
 two MVC objects.  One MVC object for the State data and its views and one MVC
@@
 
-This solution uses the power and simplicity of the MVC design and expands it
-to the multiple data set problem.  Now we can build on the original design,
-which has been well tested, instead of starting a new design from scratch.
-The strength of this multiple linked MVC design is that it uses MVC
-components, which can be individually created and then linked.  Thus, these
-components are reusable.  Also this multiple MVC design is structured to fit
-with the linking seen in the multiple data sets.  Clearly, the design should
-be built around the problem and multiple MVC objects with messaging between
-the MVC objects fits the problem of creating linked views of multiple linked
-data sets.
-
-Two parts of this design will need some special thought: creating an
-appropriate messaging system between MVC objects and deciding how the MVC
-objects will be related.  In designing an appropriate messaging system between
-MVC objects, consideration will be given to what information should pass
-between MVC objects and to how messages will propagate between MVC objects.
-In deciding the relationship between the MVC objects, consideration will be
-given to how the multiple data sets are related.  Clearly, the relationship
-between the MVC objects should reflect the relationship between the data
-sets.  The two examples of linked data sets that have been discussed
-previously were linked tables in a relational database, and experimental data
-that are linked to meta-data.  In this paper, the focus will be on experimental
-data that are linked to meta-data, as those are the data we are most
-interested in visualizing.  In this situation, the experimental data can be
-thought of as determining what meta-data are studied.  In the example of
-microarray experiment data linked to Gene Ontology meta-data, only Gene
-Ontology data that refer to genes in the microarray experiment will be
-included in the meta-data.  Thus, there is a hierarchical relationship between
-these data sets.  Therefore, the design of the linking between MVC objects
-will be based on a hierarchical relationship between the data sets, where one
-MVC object creates another MVC object.  These decisions will be discussed in
-the following sections as the design for multiple MVC objects is explained.
+Not only does this solution use the power and simplicity of the MVC design, but
+this multiple MVC design is structured to fit the linking seen in the multiple
+data sets.  Clearly, the design should be built around the problem and
+multiple MVC objects with messaging between the MVC objects fits the problem
+of creating linked views of multiple linked data sets.  In designing an
+appropriate messaging system between MVC objects, consideration is given to
+what information should pass between MVC objects and to how messages propagate
+between MVC objects.  As for the relationship between the MVC objects, it
+should reflect the relationship between the data sets.
+
+Although two examples of linked data sets have been discussed thus far,
+this paper focuses on experimental data that are linked to meta-data, as
+the data we are most interested in visualizing.  In this situation, the
+experimental data can be thought of as determining what meta-data are studied.
+In the example of microarray experiment data linked to Gene Ontology
+meta-data, only Gene Ontology data that refer to genes in the microarray
+experiment are included in the meta-data.  Thus, there is a hierarchical
+relationship between these data sets and the MVC objects that contain them,
+where one MVC object creates another MVC object.  The original MVC
+object is the parent MVC and the new MVC object that is created from the
+parent MVC is the child MVC.  These decisions are discussed in the following
+sections as the design for multiple MVC objects is explained.
 
@@
 
-\subsection{Overview}\label{SSec:OneOver}
-
-Implementing an extensible design to create linked, interactive views of
-linked data sets will be done in the R language in two packages, called
-\Rpackage{MVCClass} and \Rpackage{iSNetwork}.  The \Rpackage{MVCClass} package
-will implement the design of the MVC classes and generic functions.  The
-\Rpackage{MVCClass} package will define the individual model, view, and
+\subsection{Implementation Details}\label{SSec:OneOver}
+
+Implementing linked, interactive views of linked data sets is done using an
+object-oriented programming design for extensibility.  The \Rpackage{MVCClass}
+package implements the design of the MVC classes and generic functions.  The
+\Rpackage{MVCClass} package defines the individual model, view, and
 controller classes as well as the \Robject{MVC} class that binds the
 individual model, view, and controller classes together into one unit.  The
-package will also define the message classes that will allow the components to
-communicate with each other.  Thus, the \Rpackage{MVCClass} package will
-contain all of the classes necessary for creating linked views.   The goal of
-separating these class and generic function definitions into one package is
-that they will be available for other R packages that implement linked views
+package also defines the message classes that allow the components to
+communicate with each other.  Thus, the \Rpackage{MVCClass} package
+contains the classes necessary for creating linked views.   The goal of
+placing these class and generic function definitions into one package is
+that they are available for other R packages that implement linked views
 of data, such as the \Rpackage{iSPlot} and the \Rpackage{iSNetwork} packages.
@@
 to use based on the overall goals of the combined software.  The
-\Rpackage{iSNetwork} package will also determine how objects of these classes
-will behave because the methods for the classes are defined in the
+\Rpackage{iSNetwork} package also determines how instances of these classes
+behave because the methods for the classes are defined in the
 \Rpackage{iSNetwork} package.  In addition to implementing the behavior of
-objects, the \Rpackage{iSNetwork} package will also create a graphical user
+objects, the \Rpackage{iSNetwork} package also creates a graphical user
 interface (GUI) so that users can load data, create views, and interact with
 the views through menus on the user interface.  As an alternative to the GUI,
-users will also have access to the same functionality through command line
+users also have access to the same functionality through command line
 functions.
 
-To create the GUI, \Rpackage{iSNetwork} will use the R packages,
+To create the GUI, \Rpackage{iSNetwork} uses the R packages,
 \Rpackage{RGtk} and \Rpackage{gtkDevice}.  \Rpackage{RGtk} is an R package
@@
 that the view can be interactive.  For more information about these packages,
-please see the paper, ``Creating Linked, Interactive Views to Explore
-Multivariate Data'' \cite{EW05}.
-
-The next subsections will explain the class definitions and the methods that
-pertain to a single model-view-controller object.  Thus, the model classes,
+see \cite{EW05}.
+
+The next subsections explain the class definitions and the methods that
+pertain to a single MVC object.  Thus, the model classes,
 the view classes, the controller, the \Robject{MVC} class (which binds the
 model, view and controller classes together), and the message classes that
-pass information between the three components in one MVC object will all be
+pass information between the three components in one MVC object are
 discussed in the following subsections.  All of the class definitions are
@@
 defined in the \Rpackage{iSNetwork} package.  Message class definitions that
-pertain to communication between the MVC objects will be discussed in Section
+pertain to communication between the MVC objects are discussed in Section
 \ref{Sec:MultMVC}.
@@
 data and the model can send out messages that changes have occurred to the
-data that the model stores.  Thus, the object model for the model classes will
-represent both the slots that are needed to store data and the methods that
+data that the model stores.  Thus, the object model for the model classes
+represents both the slots that are needed to store data and the methods that
 are needed to perform the operations of the model.
@@
 \Rslot{linkData}, \Rslot{virtualData}, and \Rslot{modelName}.  These four
-slots are common information that all models will need to store.  The
+slots are common information that all models need to store.  The
 \Rslot{modelData} slot is the data set for this model, the \Rslot{linkData}
@@
 \Rfunction{fromParent}, which let this model send information to its parent
-model and receive information from its parent model, respectively (see Section
-\ref{Ssec:MultLink}), the \Rslot{virtualData} slot is data pertaining to the
-views that needs to be stored with the model so that it can be shown in all
-views of this model, and the \Rslot{modelName} slot is the name of the model.
+model and receive information from its parent model, respectively (see
+Sections \ref{Ssec:OneMVC} and \ref{Ssec:MultLink} for more information on how
+parent and child models are linked), the \Rslot{virtualData} slot is data
+pertaining to the views that needs to be stored with the model so that it can
+be shown in all views of this model, and the \Rslot{modelName} slot is the
+name of the model.
 
@@
 
-As for specific model classes, the classes that are used in the
-\Rpackage{iSNetwork} package are \Robject{dfModel}, \Robject{graphModel},
-and \Robject{exprModel}.  The \Robject{dfModel} class represents a model where
-the \Rslot{modelData} slot has class data frame (or matrix), the
-\Robject{graphModel} class represents a model where the \Rslot{modelData} slot
-has class graph, and the \Robject{exprModel} class represents a model where the
-\Rslot{modelData} slot has class exprSet.
+We encapsulate the particular data model in the class definition, which allows
+for efficient dispatch, and accessor functions that are appropriate for the
+underlying implementation.  The specific model classes used in the
+\Rpackage{iSNetwork} package to represent data are \Robject{dfModel},
+\Robject{graphModel}, and \Robject{exprModel}.  The \Robject{dfModel} class
+represents a model where the \Rslot{modelData} slot has class data frame (or
+matrix), the \Robject{graphModel} class represents a model where the
+\Rslot{modelData} slot has class graph, and the \Robject{exprModel} class
+represents a model where the \Rslot{modelData} slot has class exprSet, a class
+used to represent gene expression data \cite{BioC}.
 
@@
 welcome to make additions to the model classes for new data structures they
-would like to represent.  For example, a user may want to add a new model
+want to represent.  For example, a user may want to add a new model
 class called \Robject{tsModel} to represent time series data.  To make this
-addition, a user could extend the \Robject{gModel} class in his or her own
+addition, the user extends the \Robject{gModel} class in his or her own
 package.
 
-Note that the model classes must also notify views when the data have changed
+The model classes must also notify views when the data have changed
 so that the views can be updated.  The communication between a model and its
-views will be discussed in Section \ref{Ssec:OneMess}.
+views are discussed in Section \ref{Ssec:OneMess}.
 
@@
 
-The view classes represent the visual depictions of the model.  All views will
+The view classes represent the visual depictions of the model.  All views
 need to store some common information and respond to certain events through
-methods.  This consideration led to an object model where the different view
+methods.  This consideration leads to an object model where the different view
 classes inherit from a view virtual class, called \Robject{genView}.  The
@@
 identified in the Window menu on the GUI (see Section \ref{Ssec:OneCont}).
-All views will need to know these three pieces of information so
-\Robject{genView} will bind the view classes together.
-
-As for specific view classes, the \Robject{spreadView} class will represent a
-spreadsheet view of the data.  This view will only make sense for models that
+All views need to know these three pieces of information so
+\Robject{genView} binds the view classes together.
+
+As for specific view classes, the \Robject{spreadView} class represents a
+spreadsheet view of the data.  This view only makes sense for models that
 are a two dimensional data structure, such as a matrix or a data frame.  The
@@
 views are only applicable for certain model types.  For example, the
-\Robject{graphView} class will only make sense for a view of a graph object,
-which would be stored in the \Robject{graphModel} class.
-
-Since all views are expected to be interactive, all
-view classes have methods that correspond to user interaction, such as
-\Rfunction{clickEvent}.  Recall from Section \ref{Ssec:Backg} that when an
-event occurs Gtk emits a signal that is caught by the signal handler.  The
-signal handler then calls the callback function and it is in the callback
-function that the corresponding method is called.  For example, when the
-``clicked'' event occurs on a scatterplot view, the callback function calls the
-scatterplot's \Rfunction{clickEvent} method.  Using the information that is
-passed to the \Rfunction{clickEvent} method in the \Rslot{where} parameter,
-the method converts the Gtk pixel coordinates that indicate where the event
-occurred into user coordinates and then into an object from the plot, such as
-a point on the plot.
+\Robject{graphView} class only makes sense for a view of a graph object,
+which is stored in the \Robject{graphModel} class.
+
+Since all views are expected to be interactive, all view classes have methods
+that correspond to user interaction.  Interactive views mean a response
+occurs when the view has focus and an event occurs.  The events we are most
+interested in responding to include key press
+events, where a certain configuration of keys are pressed; button click
+events, where one of the mouse buttons is clicked over the view; and mouse
+movement events, where the cursor moves over the view.  Recall from Section
+\ref{Ssec:Backg} that when an event occurs Gtk emits a signal that is caught
+by the signal handler.  The signal handler then calls the callback function
+and it is in the callback function that the corresponding method is called.
+For example, when the button click event occurs on a scatterplot view, the
+callback function calls the scatterplot's \Rfunction{clickEvent} method.
+Using the information that is passed to the \Rfunction{clickEvent} method in
+the \Rslot{where} parameter, the method converts the Gtk pixel coordinates
+that indicate where the event occurred into user coordinates and then into an
+object from the plot, such as a point on the plot.
 
@@
 the cursor has been moved over the view.  Both the \Rfunction{clickEvent} and
-\Rfunction{motionEvent} methods for a \Robject{sPlotView} object will
+\Rfunction{motionEvent} methods for a \Robject{sPlotView} object
 determine which point on the plot was clicked or under the cursor,
@@
 emits a signal that a row in the spreadsheet has been selected or unselected
-and the \Rfunction{clickEvent} method will determine which row was changed.
+and the \Rfunction{clickEvent} method determines which row was changed.
 For \Robject{graphView}, the \Rfunction{clickEvent} method is called when Gtk
@@
 been moved over the view.  Here, though, the \Rfunction{clickEvent} and
-\Rfunction{motionEvent} methods will determine which node was clicked or under
+\Rfunction{motionEvent} methods determine which node was clicked or under
 the cursor, respectively, using the data in the \Rslot{where} parameter.
@@
 information that a model has just been updated and appropriately update the
-view to reflect that change.  Each method will depend on how the view
+view to reflect that change.  Each method depends on how the view
 represents the data so for a \Robject{spreadView} object the
@@
 could create a new view class that inherits from \Robject{plotView}, called
-\Robject{dPlotView}.  This new class, \Robject{dPlotView}, would have slots
+\Robject{dPlotView}.  This new class, \Robject{dPlotView}, has slots
 named \Rslot{dfRows} and \Rslot{col} to store which rows and which column,
@@
 package, the controller is implemented as an environment that is stored in the
-\Rslot{controller} slot of a \Robject{MVC} instance, which will be discussed in
+\Rslot{controller} slot of a \Robject{MVC} instance, which is discussed in
 more detail in Section \ref{Ssec:OneMVC}.  Although other options exist for
@@
 R an environment is not copied each time a value within the environment
-changes.  The controller in this first role will need to store information
+changes.  The controller in this first role needs to store information
 pertaining to state, such as information about the appearance of the views in
@@
 interaction with a view, quit the program and perform other functions.  On the
-GUI, there are menus to let users decide which action will be taken next.  If
+GUI, there are menus to let users decide which action is taken next.  If
 the user would rather not use the GUI, there are command line functions that
 provide access to the same functionality that is available through the GUI.
-The GUI is shown in Figure \ref{Fig:ContWin} and will be referred to as the
+The GUI is shown in Figure \ref{Fig:ContWin} and is referred to as the
 control window.
@@
 File, Edit, Display, Event, and Window.  These menus always appear on the
-control window, but the menu items under each menu will change depending on
+control window, but the menu items under each menu change depending on
 the type of model that is active.  Because the goal is to have multiple
-linked MVC objects, there will be more than one MVC object loaded and thus,
+linked MVC objects, more than one MVC object is loaded and thus,
 more than one model. However, the control window menus and the command line
-functions will only perform operations on one MVC object at a time.  Thus, in
-the \Rpackage{iSNetwork} package there will be the concept of the ``active''
+functions only perform operations on one MVC object at a time.  Thus, in
+the \Rpackage{iSNetwork} package there is the concept of the ``active''
 MVC object, where the ``active'' MVC object (and thus, the ``active'' model)
 means that operations performed through the control window menus or through
-the command line functions will affect that MVC object.  For
+the command line functions affect that MVC object.  For
 example, the Display menu allows users to create views of the active model,
@@
 Section \ref{Ssec:OneViews}.  If the active model was a data frame model
-(\Robject{dfModel}), then the views that the user could create would be a
+(\Robject{dfModel}), then the views that the user could create are a
 scatterplot (\Robject{sPlotView}) and a spreadsheet (\Robject{spreadView}),
-but a graph plot (\Robject{graphView}) would not make sense.  Thus, when the
+but a graph plot (\Robject{graphView}) does not make sense.  Thus, when the
 model in the ``active'' MVC is of class \Robject{dfModel}, then the Display
-menu will have menu items to create a scatterplot and a spreadsheet.  Also, if
+menu has menu items to create a scatterplot and a spreadsheet.  Also, if
 the ``active'' MVC is of class \Robject{dfModel}, then the command line
-functions will only let the user create a scatterplot and a spreadsheet view
+functions only let the user create a scatterplot and a spreadsheet view
 of the active model.
@@
 data sets is to create a software package that is flexible and extensible,
-users will be able to add menus and menu items to the control window through
+users are able to add menus and menu items to the control window through
 the command line to access new functionality that they have created.  As an
@@
 environment.  When an event occurs, the callback function linked to the event
-will determine the response to that event.  To create a flexible environment
-it would be nice to be able to change the response, which is determined by the
+determines the response to that event.  To create a flexible environment
+it is nice to be able to change the response, which is determined by the
 callback function, to an event.  For example, when a left button press event
@@
 functions with different signatures.  Using the example in the previous
-paragraph, a callback function that colored a point would need two parameters,
+paragraph, a callback function that colored a point needs two parameters,
 the new color to use and the point to color, while a callback function that
-hid a point would need just one parameter, the point to hide.  Since the
+hid a point needs just one parameter, the point to hide.  Since the
 intention is to be able to interchange the responses to an event, the
 signatures of different callback functions should be the same.  These design
-decisions determined the slots in the \Robject{gEventFun} class and will be
+decisions determined the slots in the \Robject{gEventFun} class and are
 discussed more fully in the following paragraphs.
@@
 the code to allow different responses to an event.  All callback functions
-will need to know what object they should act on so this will be the one
+need to know what object they should act on so this is the one
 parameter that the callback functions take.  Any other information a callback
-function may need will be set using the preprocessing functions.  The class of
-the one parameter for the callback function will depend on the model class in
-the active MVC object.  For example, for a graph model, the parameter will be
+function may need is set using the preprocessing functions.  The class of
+the one parameter for the callback function depends on the model class in
+the active MVC object.  For example, for a graph model, the parameter is
 of class \Robject{AgNode}, representing a node in the graph.  For a data frame
-model, the parameter will be of class character, representing a row in the
+model, the parameter is of class character, representing a row in the
 data frame.
 
-The preprocessing functions that are stored in the preprocessFun slot will
-take no parameters.  The purpose of the preprocessing functions is to set some
-variables that the callback function will need.  Thus, not all callback
-functions will need preprocessing functions and if no preprocessing functions
-are needed, then this slot will be NULL.  If the callback function sets the
+The preprocessing functions that are stored in the preprocessFun slot
+take no parameters.  The purpose of the preprocessing functions is to set
+variables that the callback function needs.  Thus, not all callback
+functions need preprocessing functions and if no preprocessing functions
+are needed, then this slot is NULL.  If the callback function sets the
 color of a node in a graph, then a preprocessing function is needed to
-determine what color this will be.  So in this example, the preprocessing
-function will open a color browser to allow the user to choose what the new
-color will be.
+determine what color is used.  So in this example, the preprocessing
+function opens a color browser to allow the user to choose what the new
+color is.
 
@@
 the command line functions.  The controller environment in each MVC object,
-which will be discussed in detail in Section \ref{Ssec:OneMVC},
-will store which \Robject{gEventFun} object (representing the callback
+which is discussed in detail in Section \ref{Ssec:OneMVC},
+stores which \Robject{gEventFun} object (representing the callback
 function) is linked with each event.  Note that not all callback functions
-make sense for each event so the controller environment will also store a list
+make sense for each event so the controller environment also stores a list
 for each event of appropriate \Robject{gEventFun} instances that can be linked
@@
 loading a new model through \Rpackage{iSNetwork's} API, the name of the model
-will be checked to ensure that it is unique.
+is checked to ensure that it is unique.
 
@@
 either be loaded or could be created from an already loaded MVC object.  As
-mentioned in Section \ref{Ssec:MMVC}, the focus in this paper will be on
+mentioned in Section \ref{Ssec:MMVC}, the focus in this paper is on
 experimental data that are linked to meta-data, as the linked data sets that we
@@
 \Rfunction{handleMessage} method can be considered a central dispatch for
-messages because all messages will be sent to the \Rfunction{handleMessage}
-method, which will determine where the information should be sent.   Also,
+messages because all messages are sent to the \Rfunction{handleMessage}
+method, which determines where the information should be sent.   Also,
 since messages are sent when something has changed, either through addition,
@@
 \Robject{gAddDataMessage}, \Robject{gAddViewMessage},
-\Robject{gUpdateDataMessage}, and \Robject{gUpdateViewMessage}.  As shown in
-Figure \ref{Fig:Mess}, the \Robject{gAddDataMessage} and
-\Robject{gAddViewMessage} classes inherit from the \Robject{gAddMessage}
-class, and the \Robject{gUpdateDataMessage} and \Robject{gUpdateViewMessage}
-classes inherit from the \Robject{gUpdateMessage} class.
+\Robject{gUpdateDataMessage}, and \Robject{gUpdateViewMessage}.
 
@@
 other MVC object that is related to this model.  Because this information gets
-passed to other MVC objects, which will be discussed further in Section
+passed to other MVC objects, which is discussed further in Section
 \ref{Sec:MultMVC}, this message also has one new slot, \Rslot{from}, in
@@
 linked to this model has been updated.  So the update data message may come
-from within this MVC object or it may come from a different MVC object that
-has changed its model and is related to this MVC object.  Information on
-passing messages between MVC objects will be given in Section
-\ref{Sec:MultMVC}.  Finally, the \Robject{gUpdateViewMessage} class represents
-a message to update a view.  This message will be created and handled whenever
-the model is updated so that the views accurately reflect the model.
+from within this MVC object or it may come from a MVC object that
+has changed its model and is related to this MVC object, either as this MVC
+object's parent MVC or one of this MVC object's child MVCs.  Information on
+passing messages between MVC objects is given in Section \ref{Sec:MultMVC}.
+Finally, the \Robject{gUpdateViewMessage} class represents a message to update
+a view.  This message is created and handled whenever the model is updated so
+that the views accurately reflect the model.
+
+An example of message passing within a MVC object is shown in Figure
+\ref{Fig:MPwithin}.  In this example, the response to a user performing a
+button click event over a point in the scatterplot is to color the point red.
+Setting the response to an event was discussed in Section
+\ref{Ssec:OneEvent}.  Here the user first clicks on a point in the
+scatterplot, which is shown as step 1 (the circled 1 on the figure).  In the
+figure, the point that has been clicked has an arrow next to it.  Then, in
+step 2, an update data message is sent to the model, which updates the
+\Rslot{virtualData} slot of the model object.  After the model has been
+updated, the views must be notified that a change has occurred so in step 3,
+an update view message is created and sent to all views of this model.  In
+response to this message, the view updates itself in step 4 and now the point
+is colored red on the view to reflect the change in the model.  In this
+example, the \Rslot{from} slot in the \Robject{gUpdateDataMessage} object is
+the name of this MVC object since the message came from within the MVC.
+
+\begin{figure}[ht]
+  \begin{center}
+    \scalebox{0.6}{\includegraphics{newMPwithin.jpg}}
+    \caption{ Message Passing within a MVC object. }
+    \label{Fig:MPwithin}
+  \end{center}
+\end{figure}
 
@@
 \Rfunction{initialize} and \Rfunction{handleMessage} methods, which perform
-the following tasks:
-
-\begin{list}{}{\setlength{\leftmargin}{0.5in}
-               \setlength{\itemsep}{0in}}
-  \item \Rfunction{initialize} is used to properly fill the slots of the
-object when it is created.  
-  \item \Rfunction{handleMessage} is necessary to
-process the message.  Since a message is a form of communication
-between two components, once it is created it must be sent to the
-proper recipient and action must be taken in response to its
-information, which is the purpose of the \Rfunction{handleMessage} method.
-\end{list}
+the following tasks.  The \Rfunction{initialize} method is used to properly
+fill the slots of the object when it is created.  The
+\Rfunction{handleMessage} method is necessary to process the message.  Since a
+message is a form of communication between two components, once it is created
+it must be sent to the proper recipient and action must be taken in response
+to its information, which is the purpose of the \Rfunction{handleMessage}
+method.
 
@@
 We now describe the implementation of multiple MVC objects for linking views
-of multiple linked data sets.  Each MVC object will represent one model and
+of multiple linked data sets.  Each MVC object represents one model and
 its views.  Now for the views of different data sets to be linked, they must
@@
 Thinking of experimental data that are linked to meta-data as an example, there
-will be two MVC objects, one for the experimental data and one for the
+are two MVC objects, one for the experimental data and one for the
 meta-data.  In the example of experimental microarray data that are linked to
@@
 are related through a tree hierarchy where a child model is derived from its
-parent model.  Consider Figure \ref{Fig:SmallHier} shown below.  If firstMVC's
-model changes, then this information first gets passed to secondMVC, which may
-make changes to its model based on this information.  Then secondMVC passes
-this information to thirdMVC based on the changes secondMVC made to its model.
-Thus, information passing between firstMVC and thirdMVC only occurs through
-secondMVC.  So messages can only be sent from parent MVC to child MVC or from
-child MVC to parent MVC.
-
-\begin{figure}[ht]
-  \begin{center}
-    \scalebox{0.6}{\includegraphics{SmallHierarchy.jpg}}
-    \caption{ Multiple MVC Objects and Their Relationships. }
-    \label{Fig:SmallHier}
-  \end{center}
-\end{figure}
-
-Next we will consider how the data sets will be linked (and thus, how MVC
-objects are linked) followed by how messages will be passed between these
+parent model.  Consider Figure \ref{Fig:Hier}, which was shown previously.  
+If model1's model changes, then this information first gets passed to model2
+and model3 and these two MVC's may make changes to their models based on this
+information.  Then model2 passes this information to model4, model5, and
+model6 based on the changes that model2 made to its model.  Thus, information
+passing between model1 and model4, for instance, only occurs through model2.
+So messages can only be sent from parent MVC to child MVC or from child MVC to
+parent MVC.
+
+Next we consider how the data sets are linked (and thus, how MVC
+objects are linked) followed by how messages are passed between these
 linked MVC objects.
@@
 
-For messages to pass between MVC objects, there must be a function or a place
-that converts information from one MVC into useful information for a different
-MVC object.  Thus, there needs to be some way to determine how data from one
-MVC object relates to data in a different MVC object.  This conversion is done
-by two functions that are stored in the \Rslot{linkData} slot of the model
-objects.  The \Rslot{linkData} slot is a list with two elements: a
-\Rfunction{toParent} function and a \Rfunction{fromParent} function.
+For messages to pass between MVC objects, the information from one MVC must be
+converted into useful information for a different MVC object.  Thus, there
+needs to be some way to determine how data from one MVC object relates to data
+in a different MVC object.  This conversion is done by two functions that are
+stored in the \Rslot{linkData} slot of the model objects.  The
+\Rslot{linkData} slot is a list with two elements: a \Rfunction{toParent}
+function and a \Rfunction{fromParent} function.
 
 When a child MVC is created from an existing MVC, the child model's
-\Rslot{linkData} slot will be filled with the two functions,
+\Rslot{linkData} slot is filled with the two functions,
 \Rfunction{toParent} and \Rfunction{fromParent}.  When the model in the child
 MVC is updated, the \Rfunction{toParent} function in the \Rslot{linkData} slot
-will be used to convert the child model's data change into useful information
+is used to convert the child model's data change into useful information
 for the parent model.
@@
 needs to be passed to the child model.  However, a parent can have multiple
-child MVC objects so the parent must look in each child model's
-\Rslot{linkData} slot for the \Rfunction{fromParent} function.  The
-\Rfunction{fromParent} function will convert the parent model's data change
-into useful information for this particular child model.  Thus, the parent
-model's data change may be converted into different information for different
-child models and this can happen because each child model has its own
-\Rfunction{fromParent} function.  When a child MVC object is created and
-its model's \Rslot{linkData} slot is filled, the \Rfunction{toParent} function
-will convert information from this child model into useful information for its
-parent MVC and the \Rfunction{fromParent} function will convert a message
-from its parent MVC into useful information for this child MVC.
+child MVC objects so the parent sends a message to each child MVC to notify
+the child of the change and then each child MVC handles the message by using
+its \Rfunction{fromParent} function.  The \Rfunction{fromParent} function
+converts the parent model's data change into useful information for this
+particular child model.  Thus, the parent model's data change may be converted
+into different information for different child models and this can happen
+because each child model has its own \Rfunction{fromParent} function.  When a
+child MVC object is created and its model's \Rslot{linkData} slot is filled,
+the \Rfunction{toParent} function converts information from this child model
+into useful information for its parent MVC and the \Rfunction{fromParent}
+function converts a message from its parent MVC into useful information for
+this child MVC.
 
 An example of how these functions, \Rfunction{toParent} and
-\Rfunction{fromParent}, are used when messages are passed between MVCs will
-be given in Section \ref{Ssec:MultMess}.
+\Rfunction{fromParent}, are used when messages are passed between MVCs is
+given in Section \ref{Ssec:MultMess} where Figure \ref{Fig:MessPass} shows the
+\Rfunction{toParent} function being used to convert information from a child
+MVC object into useful information for the parent MVC object.
 
@@
 types of messages: an add message and an update message.  These message types
-will also be seen in passing messages between different MVC objects, as shown
+also are seen in passing messages between different MVC objects, as shown
 in Figure \ref{Fig:BetwMess}.
@@
 
-The first message class that will be discussed is \Robject{gAddChildMessage}.
-A \Robject{gAddChildMessage} object will be very similar to a
+The first message class that is discussed is \Robject{gAddChildMessage}.
+A \Robject{gAddChildMessage} object is very similar to a
 \Robject{gAddDataMessage} object because they are both adding a new
@@
 \Rfunction{initialize} method properly sets the slots of the new object.  The
-\Rfunction{handleMessage} method for a \Robject{gSendParentMessage} will
-take the \Robject{gUpdateDataMessage} object from the child model and convert
+\Rfunction{handleMessage} method for a \Robject{gSendParentMessage}
+takes the \Robject{gUpdateDataMessage} object from the child model and converts
 it to a \Robject{gUpdateDataMessage} object for the parent model using the
 \Rfunction{toParent} function.  Similarly, the \Rfunction{handleMessage}
-method for a \Robject{gSendChildMessage} will take the
-\Robject{gUpdateDataMessage} object from the parent model and convert it to a
+method for a \Robject{gSendChildMessage} takes the
+\Robject{gUpdateDataMessage} object from the parent model and converts it to a
 \Robject{gUpdateDataMessage} object for the child model using the
@@
 These message classes ensure that when a model is updated, its parent and
-child models will be notified of the change and most importantly will be
+child models are notified of the change and most importantly are
 notified of the change in a way that each model can properly act upon that
@@
   \begin{center}
-    \scalebox{0.55}{\includegraphics{newMessagePassing.jpg}}
+    \scalebox{0.55}{\includegraphics{newMP.jpg}}
     \caption{ Message Passing Between and Within MVCs. }
@@
 MVC objects are defined and how messages are passed within one MVC as well
-as how this information gets passed on to its parent MVC using messages.  In
-this picture, the message passing starts when a user interacts with the y
-vs. x scatterplot depicting model 2.  After the user interaction with the
-scatterplot, step 1 (indicated on the figure as a circled 1) is to send an
-update data message to model 2.  As soon as model 2 is updated, step 2 is to
-send an update view message so that the scatterplot view is updated to reflect
-the changed data.  In step 3, a send parent message is sent because MVC 2
-(which is the MVC object for model 2) has a parent MVC, which is MVC 1.  For
-step 4, the \Rfunction{toParent} function converts the child's update data
-message into an update data message that the parent will understand.  In step
-5, this new update data message that was created by the \Rfunction{toParent}
-function will be sent to model 1.  Finally for step 6, after model 1 has been
-updated, an update view message will be sent to the view of model 1.  Now all
-of the data and the views are synchronized.
+as how this information gets passed on to its parent MVC using messages.
+Suppose for this example that when the user clicks a mouse button over the
+scatterplot, the response to this event is to color the point under the cursor
+red.  In this picture, the message passing starts when a user clicks a point,
+that is indicated by the arrow, on the scatterplot.  An update data message is
+sent to Model 2 in response to this user interaction.  As soon as Model 2 is
+updated, it sends an update view message to all its views so that the views
+are updated to reflect the changed data.  Now the scatterplot point that was
+initially clicked is red.  Then a send parent message is sent because MVC
+2, which is the MVC object for Model 2, has a parent MVC, which is MVC 1.  Next
+the \Rfunction{toParent} function converts the child's update data message
+into an update data message that the parent MVC understands.  Then the
+update data message that was created by the \Rfunction{toParent} function is
+sent to Model 1.  Finally, after Model 1 has been updated, an update view
+message is sent to all of Model 1's views.  It is in this last step that the
+view of Model 1 has a point colored red to indicate which data from Model 1 is
+linked to data from Model 2.  All of the data and the views are
+synchronized.
 
@@
 paper, including linked tables from a relational database, and meta data that
-are linked to experimental data.  The model-view-controller (MVC) paradigm was
+are linked to experimental data.  The MVC paradigm was
 discussed as a useful design for creating linked views of a data set.
@@
 creating linked views of linked data sets, the next step was to determine how
-these MVC objects would be related.  Because the focus in this paper was on
+these MVC objects were related.  Because the focus in this paper was on
 experimental data that are linked to meta-data, a hierarchical relationship
@@
 
+For message passing between MVC objects, when a MVC's model is updated, it
+notifies its parent and child MVCs that a change has occurred.  Then when a
+MVC gets a message from a parent or child MVC, the MVC can either decide to
+ignore the message or read the message, which results in this MVC's model
+being updated.  However, the MVC object can not decide to only see certain
+types of messages.  It must either accept all messages and update its model
+based on the messages, or ignore all messages.  A future goal for message
+passing is to allow more selectivity in which messages are accepted between
+MVC objects so that the linking between MVC objects is more sophisticated.
+
 The goals for this paper were to create linked views of linked data sets, to
 create interactive views where the response to an event can be changed, and to
-create an extensible design that will allow future users to make additions.
+create an extensible design that allows future users to make additions.
 The first goal was accomplished by implementing MVC objects as components that
@@
 Although extensibility was a goal, certain design decisions, such as expecting
-the data sets to be linked through 
-a parent-child hierarchy, may not work with some data sets.  As an example,
-the linked tables from a relational database are not related through a tree
-structure and thus, having parent and child MVC objects may not make sense
-with these data sets.  However, the design allows users to add new
-types of models, new types of views, and new message classes so there is still
-flexibility for additions as long as the data sets are connected through
-parent-child relationships.  These new class definitions could be created by
-the user in a new package that extends the \Rpackage{MVCClass} package that
-was discussed in this paper.  Other additions that the user may want to
-perform, such as adding new menus to the GUI and adding new responses to an
-event can be performed through command line functions that are defined in the
-\Rpackage{iSNetwork} package.
-
-\subsection{Future Work}\label{Ssec:Further}
-
-When a MVC's model is updated, it notifies its parent and child
-MVCs that a change has occurred.  Then when a MVC gets a message from a
-parent or child MVC, the MVC can either decide to ignore the message or read
-the message, which results in this MVC's model being updated.  However, the
-MVC object can not decide to only see certain types of messages.  It must
-either accept all messages and update its model based on the messages, or
-ignore all messages.  A future goal for message passing is to allow more
-selectivity in which messages will be accepted between MVC objects so that the
-linking between MVC objects is more sophisticated.
+the data sets to be linked through a parent-child hierarchy, may not work with
+some data sets.  As an example, the linked tables from a relational database
+are not related through a tree structure and thus, having parent and child MVC
+objects may not make sense with these data sets.  However, the design allows
+users to add new types of models, new types of views, and new message classes
+so there is still flexibility for additions as long as the data sets are
+connected through parent-child relationships.  These new class definitions
+could be created by the user in a new package that extends the
+\Rpackage{MVCClass} package that was discussed in this paper.  Other additions
+that the user may want to perform, such as adding new menus to the GUI and
+adding new responses to an event can be performed through command line
+functions that are defined in the \Rpackage{iSNetwork} package.
+
+The functionality that is defined in the \Rpackage{iSNetwork} package is
+discussed in detail in the Vignette for this R package.  Users can load data,
+create views of the data, interact with these views and change the response to
+the interaction (these interactions include mouse button click events and
+mouse movement events), and users can create child MVC objects through the GUI
+or through command line functions.  Using the functionality in the
+\Rpackage{iSNetwork} package, the user can create views of linked data sets
+and can explore visually the relationships between variables in different data
+sets.  This functionality was implemented by extending the MVC paradigm in a
+very natural way to encompass a much richer class of models and user
+interactions.
 

### 2006-01-26 inst/doc/MVCClass.Rnw
Added a new version of paper 2.  Must have also made a few small changes to the code.
git-svn-id: file:///home/git/hedgehog.fhcrc.org/bioconductor/trunk/madman/Rpacks/MVCClass@15900 bc3139a8-67e5-0310-9ffc-ced21a209358

@@
 store the device number of the plot, the plotting parameters, and the Gtk
-drawing area object, respectively.  This class is not intended to have any
-objects as it just represents a general plot.  The plot classes that can have
+drawing area object, respectively.  This class is a virtual class because it
+is not intended to have any objects as it just represents a general plot.  The
+plot classes that can have
 objects are the specific plot classes, \Robject{sPlotView},
@@
 perform the addition or alteration operation, and type is a character string
-that gives the type of addition or alteration to perform.  Although
-\Robject{gAddMessage} and \Robject{gUpdateMessage} are not virtual classes, it
-is not intended to create objects of these classes.
+that gives the type of addition or alteration to perform.  Both
+\Robject{gAddMessage} and \Robject{gUpdateMessage} are virtual classes.
 

### 2006-01-26 inst/paper/paper2.tex
Added a new version of paper 2.  Must have also made a few small changes to the code.
git-svn-id: file:///home/git/hedgehog.fhcrc.org/bioconductor/trunk/madman/Rpacks/MVCClass@15900 bc3139a8-67e5-0310-9ffc-ced21a209358

@@
 
-\newcommand{\Rfunction}[1]{{\texttt{#1}}}
+\newcommand{\Rfunction}[1]{{\textbf{#1}}}
 \newcommand{\Robject}[1]{{\texttt{#1}}}
@@
 \newcommand{\Rslot}[1]{\textsl{#1}}
+\newcommand{\Rclass}[1]{\textsf{#1}}
 
@@
 \textbf{Dana-Farber Cancer Institute}\\ 
-\textbf{December 19, 2005} \\
+\textbf{January 25, 2006} \\
 \vspace{1in} 
@@
 two or three dimensions are needed to view the data.  Several of these ideas
-were discussed in a previous paper, titled ``Creating Linked, Interactive
-Views to Explore Multivariate Data'' \cite{EW05}, so the focus in this paper
+were discussed in previous work \cite{EW05}, so the focus in this paper
 is on one specific method of visualizing multivariate data that is
@@
 
-First brief definitions of linked views, interactive views and linked data
-sets are given.  Linked views mean that if a component that represents
-data (such as a point on a plot or a row in a spreadsheet) changes its
-appearance on one view, then the corresponding component on a second view
-also changes its appearance.  For components on different views to be
-corresponding, the data displayed in those components must come from the same
-entity in a model.  As an example, suppose that the model consisted of four
-variables for the fifty states' murder rate, assault rate, rape rate, and
-percent of population living in urban areas.  Now the model could be
-represented as a matrix of four columns by fifty rows with four columns for
-the four variables and fifty rows for the fifty states.  In this model, the
-entities are the states.  If two scatterplots were created as views of
-this model, where the first plot displayed murder rate versus assault rate and
-the second plot displayed rape rate versus percent urban population, then the
-points on each plot that referred to the state of Montana are
-corresponding components.   Linked views, which have a long history in
-statistics, are now considered a standard feature of interactive data
-visualization software \cite{GGobi}.  Linked views have been implemented in
-several pieces of statistical software, such as XLisp-Stat and GGobi.
-Implementing linked views is based on the model-view-controller (MVC) design
-\cite{DesignPatterns}, which is a widely used and well understood paradigm.
-The MVC design is discussed in more detail in Section \ref{Ssec:Design}.
-
-Interactivity is an idea that is familiar from using the Internet.  For
-example, clicking on a hyperlink on a web page results in a new web page
+First definitions of linked views, interactive views and linked data
+sets are given in the following paragraphs.  Linked views mean that if a
+component that represents data (such as a point on a plot or a row in a
+spreadsheet) changes its appearance on one view, then the corresponding
+component on a second view also changes its appearance.  For components on
+different views to be corresponding, the data displayed in those components
+must come from the same entity in a model.  As an example, suppose that the
+model consists of the gene expression levels from four samples.  If the same
+one hundred genes are studied in the four samples, then the model could be
+represented as a matrix of four columns by one hundred rows with four columns
+for the four samples and one hundred rows for the one hundred genes.  In this
+model, the entities are the genes.  If two scatterplots are created as views
+of this model, where the first plot displays sample one versus sample two and
+the second plot displays sample three versus sample four, then the points on
+each plot that referred to the same gene are corresponding components.  Linked
+views, which have a long history in statistics, are now considered a standard
+feature of interactive data visualization software \cite{GGobi}.  Linked views
+have been implemented in several pieces of statistical software, such as
+XLisp-Stat \cite{Lisp} and GGobi \cite{GGobiMan}.  Implementing linked views
+is based on the model-view-controller (MVC) design \cite{DesignPatterns},
+which is a widely used and well understood paradigm.  The MVC design is
+discussed in more detail in Section \ref{Ssec:Design}.
+
+Interactivity is an idea that is familiar from using an Internet browser.
+Clicking on a hyperlink on a web page results in a new web page
 opening.  Another example is clicking on the login button after entering a
@@
 quickly navigate and explore complex documents or complex data sets.  As an
-example, suppose that a user has two views: one view is a graph of molecular
-functions for genes and the other view is a heatmap of gene expression data.
-If the user clicks on a node that represents a particular molecular function,
-then the response to this action could be to highlit all genes in the heatmap
-view that express this molecular function.  This interactivity allows the
-user to quickly see if genes with the same molecular function have similar
-expression levels.
+example, suppose that a user has two views: one view is a graph representing
+the molecular function of genes and the other view is a heatmap of gene
+expression data.  If the user clicks on a node that represents a particular
+molecular function, then the response to this action could be to highlight all
+genes in the heatmap view that correspond to this molecular function.  This
+interactivity allows the user to quickly see if genes with the same molecular
+function have similar expression levels.
 
@@
 text editor, then the letter C is printed on the text editor's screen.
-Any other open windows that do not have
-focus do not respond to this event.  The callback function is a unique
-function because its signature is determined by the event that the callback
-function is linked to.  The parameters of the callback function may include
-information about the event, such as the on screen location of the event, and
-information about the object that had focus when the event occurred, such as a
-button on the GUI.  The signal handler is the piece that must look to see if
-an event has occurred and then when the event has occurred, the signal handler
-calls the appropriate callback function.
+Any other open windows that do not have focus do not respond to this event.
+The callback function is a unique function because its signature is determined
+by the event that the callback function is linked to.  The parameters passed
+to the callback function may include information about the event, such as the
+on screen location of the event, and information about the object that had
+focus when the event occurred, such as a button on the GUI.  The signal
+handler is the piece that must look to see if an event has occurred and then
+when the event has occurred, the signal handler calls the appropriate callback
+function.
 
@@
   \begin{center}    
-    \scalebox{0.6}{\includegraphics{Event.jpg}}
+    \scalebox{0.55}{\includegraphics{Event.jpg}}
     \caption{ Flow Chart of the Response to an Event. }
@@
 response to this interaction is to tell data set 2 to update itself (this is
-indicated on the figure as a circled 1).  Then in step 2, data set 2 tells the
+indicated on the figure as a circled 1).  Then in Step 2, data set 2 tells the
 scatterplot view that is should update itself in response to the change in the
-data.  Next, in step 3, because data set 2 has changed and because it is
+data.  Next, in Step 3, because data set 2 has changed and because it is
 linked to data set 1, data set 2 sends information to data set 1 about the
-change.  In step 4, the information from data set 2 is converted into
-information that data set 1 can understand.  In step 5, this information is
-received by data set 1 and data set 1 updates its data.  Finally, in step 6,
+change.  In Step 4, the information from data set 2 is converted into
+information that data set 1 can understand.  In Step 5, this information is
+received by data set 1 and data set 1 updates its data.  Finally, in Step 6,
 data set 1 tells its view to update itself.  This example shows the components
@@
 
-In the design presented in this paper, the components consist of MVC objects
-and communication within and between MVC objects is performed by message
-objects.  This solution uses the power and simplicity of the MVC design and
-expands it to the multiple data set problem.  Now we can build on the original
-MVC design, which has been well tested, instead of starting a new design from
-scratch.  The strength of this multiple linked MVC design is that it uses MVC
-components, which can be individually created and then linked.  Thus, these
-components are reusable.  These design decisions are discussed further in
-Sections \ref{Sec:OneMVC} and \ref{Sec:MultMVC}.
+In the design presented in this paper, the components consist of \Robject{MVC}
+objects and communication within and between \Robject{MVC} objects is
+performed by message objects.  This solution uses the power and simplicity of
+the MVC design and expands it to the multiple data set problem.  Now we can
+build on the original MVC design, which has been well tested, instead of
+starting a new design from scratch.  The strength of this multiple linked MVC
+design is that it uses MVC components, which can be individually created and
+then linked.  Thus, these components are reusable.  These design decisions are
+discussed further in Sections \ref{Sec:OneMVC} and \ref{Sec:MultMVC}.
 
@@
   \begin{center}
-    \scalebox{0.6}{\includegraphics{firstMessagePassing.jpg}}
-    \caption{ Linked Views of Linked Data Sets. }
+    \scalebox{0.55}{\includegraphics{firstMessagePassing.jpg}}
+    \caption{ Linked views of linked Data Sets.  When a user interacts with
+      the y versus x scatterplot, the first step, indicated by a circled one,
+      is to tell data set 2 to update.  Then all views of data set 2 must be
+      updated.  Next any data that are linked to data set 2 must be notified
+      that data set 2 changed.  This information from data set 2 must be
+      altered so that it can be understood by linked data sets, which is shown
+      in the fourth step.  Finally, data set 1 is notified that it should
+      update in response to the change in data set 2 and then all views of
+      data set 1 are updated. }
     \label{Fig:firstMP}
@@
 discussed in Section \ref{Sec:Extend}.  This extension supports the creation of
-linked MVC objects.  The design of individual MVC objects is discussed in
-Section \ref{Sec:OneMVC} while the implementation of linked MVC objects is
-discussed in Section \ref{Sec:MultMVC}.  The goals of the software discussed
-in this paper are to create linked views of linked data sets, to create
-interactive views where the response to an event can be changed, and to create
-an extensible design that allows future users to make additions.
+linked \Robject{MVC} objects.  The design of individual \Robject{MVC} objects
+is discussed in 
+Section \ref{Sec:OneMVC} while the implementation of linked \Robject{MVC}
+objects is discussed in Section \ref{Sec:MultMVC}.  The goals of the software
+discussed in this paper are to create linked views of linked data sets, to
+create interactive views where the response to an event can be changed, and to
+create an extensible design that allows future users to make additions.
 
@@
   \begin{center}
-    \scalebox{0.6}{\includegraphics{ExMVC.jpg}}
+    \scalebox{0.5}{\includegraphics{ExMVC.jpg}}
     \caption{ A Model with its Views. }
@@
 
-A more straightforward approach may be to link MVCs.  Each MVC is a coherent
-object and the interactions between MVCs are encapsulated in a set of
-messages.  We use one MVC object to handle each data set.  Thus, within a MVC
-object, one can create linked views of that particular data set, using the
-standard MVC design.  But between MVC objects we design a second messaging
-system so that a change in one data set is sent to the other linked data sets.
+A more straightforward approach may be to link \Robject{MVCs}.  Each
+\Robject{MVC} is a coherent object and the interactions between \Robject{MVCs}
+are encapsulated in a set of messages.  We use one \Robject{MVC} object to
+handle each data set.  Thus, within a \Robject{MVC} object, one can create
+linked views of that particular data set, using the standard MVC design.  But
+between \Robject{MVC} objects we design a second messaging system so that a
+change in one data set is sent to the other linked data sets.
 
@@
 with the states they reside in.  Now for these two data sets there are
-two MVC objects.  One MVC object for the State data and its views and one MVC
-object for the City data and its views.
+two \Robject{MVC} objects.  One \Robject{MVC} object for the State data and
+its views and one \Robject{MVC} object for the City data and its views.
 
@@
     \scalebox{0.7}{\includegraphics{MultipleMVC.jpg}}
-    \caption{ Example of Two MVC Objects. }
+    \caption{ Example of Two \Robject{MVC} Objects.  These two \Robject{MVC}
+      objects are linked by the StateCity table and the color of the points
+    indicates the linking between the data.  The points referring to the state
+    of Pennsylvania are red, the points referring to the state of California
+    are blue, and the points referring to Minnesota are green. }
     \label{Fig:MultMVC}
@@
 of Pennsylvania are red, the points referring to California are blue, and the
-points referring to Minnesota are green.  Even though there are two MVC
-objects here, with one MVC object for each data set, it is still clear that
-the views are linked through the data sets and that is shown in this example
-through the coloring.
+points referring to Minnesota are green.  Even though there are two
+\Robject{MVC} objects here, with one \Robject{MVC} object for each data set,
+it is still clear that the views are linked through the data sets and that is
+shown in this example through the coloring.
 
@@
 data sets.  Clearly, the design should be built around the problem and
-multiple MVC objects with messaging between the MVC objects fits the problem
-of creating linked views of multiple linked data sets.  In designing an
-appropriate messaging system between MVC objects, consideration is given to
-what information should pass between MVC objects and to how messages propagate
-between MVC objects.  As for the relationship between the MVC objects, it
-should reflect the relationship between the data sets.
+multiple \Robject{MVC} objects with messaging between the \Robject{MVC}
+objects fits the problem of creating linked views of multiple linked data
+sets.  In designing an appropriate messaging system between \Robject{MVC}
+objects, consideration is given to what information should pass between
+\Robject{MVC} objects and to how messages propagate between \Robject{MVC}
+objects.  As for the relationship between the \Robject{MVC} objects, it should
+reflect the relationship between the data sets.
 
@@
 experiment are included in the meta-data.  Thus, there is a hierarchical
-relationship between these data sets and the MVC objects that contain them,
-where one MVC object creates another MVC object.  The original MVC
-object is the parent MVC and the new MVC object that is created from the
-parent MVC is the child MVC.  These decisions are discussed in the following
-sections as the design for multiple MVC objects is explained.
+relationship between these data sets and the \Robject{MVC} objects that
+contain them, where one \Robject{MVC} object creates another \Robject{MVC}
+object.  The original \Robject{MVC} object is the parent \Robject{MVC} and the
+new \Robject{MVC} object that is created from the parent \Robject{MVC} is the
+child \Robject{MVC}.  These decisions are discussed in the following sections
+as the design for multiple \Robject{MVC} objects is explained.
 
@@
 package implements the design of the MVC classes and generic functions.  The
-\Rpackage{MVCClass} package defines the individual model, view, and
-controller classes as well as the \Robject{MVC} class that binds the
-individual model, view, and controller classes together into one unit.  The
-package also defines the message classes that allow the components to
-communicate with each other.  Thus, the \Rpackage{MVCClass} package
-contains the classes necessary for creating linked views.   The goal of
-placing these class and generic function definitions into one package is
-that they are available for other R packages that implement linked views
-of data, such as the \Rpackage{iSPlot} and the \Rpackage{iSNetwork} packages.
+\Rpackage{MVCClass} package defines the individual model,
+view, and controller classes as well as the \Rclass{MVC}
+class that binds the individual model, view, and
+controller classes together into one unit.  The package also defines
+the message classes that allow the components to communicate with
+each other.  Thus, the \Rpackage{MVCClass} package contains the classes
+necessary for creating linked views.   The goal of placing these class and
+generic function definitions into one package is that they are available for
+other R packages that implement linked views of data, such as the
+\Rpackage{iSPlot} and the \Rpackage{iSNetwork} packages.
 
@@
 The next subsections explain the class definitions and the methods that
-pertain to a single MVC object.  Thus, the model classes,
-the view classes, the controller, the \Robject{MVC} class (which binds the
-model, view and controller classes together), and the message classes that
-pass information between the three components in one MVC object are
-discussed in the following subsections.  All of the class definitions are
-implemented in the \Rpackage{MVCClass} package and all of the methods are
-defined in the \Rpackage{iSNetwork} package.  Message class definitions that
-pertain to communication between the MVC objects are discussed in Section
+pertain to a single \Robject{MVC} object.  Thus, the model classes,
+the view classes, the controller, the \Rclass{MVC} class
+(which binds the model, view and controller classes
+together), and the message classes that pass information between the
+three components in one \Robject{MVC} object are discussed in the following
+subsections.  All of the class definitions are implemented in the
+\Rpackage{MVCClass} package and all of the methods are defined in the
+\Rpackage{iSNetwork} package.  Message class definitions that pertain
+to communication between the \Robject{MVC} objects are discussed in Section
 \ref{Sec:MultMVC}.
@@
 data and the model can send out messages that changes have occurred to the
-data that the model stores.  Thus, the object model for the model classes
-represents both the slots that are needed to store data and the methods that
-are needed to perform the operations of the model.
-
-Figure \ref{Fig:Model} shows the inheritance structure for the model classes.
-Here the virtual class, \Robject{gModel}, has the slots \Rslot{modelData},
-\Rslot{linkData}, \Rslot{virtualData}, and \Rslot{modelName}.  These four
-slots are common information that all models need to store.  The
-\Rslot{modelData} slot is the data set for this model, the \Rslot{linkData}
-slot is a list of two functions, \Rfunction{toParent} and
+data that the model stores.  Thus, the object model for the model
+classes represents both the slots that are needed to store data and the
+methods that are needed to perform the operations of the model.
+
+Figure \ref{Fig:Model} shows the inheritance structure for the model
+classes. Here the virtual class, \Rclass{gModel}, has the slots
+\Rslot{modelData}, \Rslot{linkData}, \Rslot{virtualData}, and
+\Rslot{modelName}.  These four slots are common information that all models
+need to store.  The \Rslot{modelData} slot is the data set for this model, the
+\Rslot{linkData} slot is a list of two functions, \Rfunction{toParent} and
 \Rfunction{fromParent}, which let this model send information to its parent
@@
 underlying implementation.  The specific model classes used in the
-\Rpackage{iSNetwork} package to represent data are \Robject{dfModel},
-\Robject{graphModel}, and \Robject{exprModel}.  The \Robject{dfModel} class
+\Rpackage{iSNetwork} package to represent data are \Rclass{dfModel},
+\Rclass{graphModel}, and \Rclass{exprModel}.  The \Rclass{dfModel} class
 represents a model where the \Rslot{modelData} slot has class data frame (or
-matrix), the \Robject{graphModel} class represents a model where the
-\Rslot{modelData} slot has class graph, and the \Robject{exprModel} class
+matrix), the \Rclass{graphModel} class represents a model where the
+\Rslot{modelData} slot has class graph, and the \Rclass{exprModel} class
 represents a model where the \Rslot{modelData} slot has class exprSet, a class
@@
 with a view.  Thus, the \Rfunction{updateModel} method is actually updating the
-\Rslot{virtualData} slot because it is this slot in the model classes that
-ensures that all views are appropriately visually linked.
-
-Although only three model classes are currently implemented, future users are
-welcome to make additions to the model classes for new data structures they
-want to represent.  For example, a user may want to add a new model
-class called \Robject{tsModel} to represent time series data.  To make this
-addition, the user extends the \Robject{gModel} class in his or her own
-package.
+\Rslot{virtualData} slot because it is this slot in the model classes
+that ensures that all views are appropriately visually linked.
+
+Although only three model classes are currently implemented, future
+users are welcome to make additions to the model classes for new data
+structures they want to represent.  For example, a user may want to add a new
+model class called \Rclass{tsModel} to represent time series data.
+To make this addition, the user extends the \Rclass{gModel} class in his or
+her own package.
 
@@
 
-The view classes represent the visual depictions of the model.  All views
-need to store some common information and respond to certain events through
-methods.  This consideration leads to an object model where the different view
-classes inherit from a view virtual class, called \Robject{genView}.  The
-object model for the view classes is shown in Figure \ref{Fig:View}.
+The view classes represent the visual depictions of the model.  All
+views need to store some common information and respond to certain events
+through methods.  This consideration leads to an object model where the
+different view classes inherit from a view virtual class,
+called \Rclass{genView}.  The object model for the view classes is
+shown in Figure \ref{Fig:View}.
 
@@
 
-The \Robject{genView} class has three slots, \Rslot{dataName}, \Rslot{win},
+The \Rclass{genView} class has three slots, \Rslot{dataName}, \Rslot{win},
 and \Rslot{winNum}.  \Rslot{dataName} is the name of the model that the view
@@
 All views need to know these three pieces of information so
-\Robject{genView} binds the view classes together.
-
-As for specific view classes, the \Robject{spreadView} class represents a
-spreadsheet view of the data.  This view only makes sense for models that
-are a two dimensional data structure, such as a matrix or a data frame.  The
-information needed for this view is stored in the slot, \Rslot{clist}, which
-is the Gtk spreadsheet object.
+\Rclass{genView} binds the view classes together.
+
+As for specific view classes, the \Rclass{spreadView} class
+represents a spreadsheet view of the data.  This view only makes sense for
+models that are a two dimensional data structure, such as a matrix or a data
+frame.  The information needed for this view is stored in the slot,
+\Rslot{clist}, which is the Gtk spreadsheet object.
 
 When creating a plot of a data set, there is a general class, called
-\Robject{plotView}, that has the slots, \Rslot{plotDevice}, \Rslot{plotPar},
+\Rclass{plotView}, that has the slots, \Rslot{plotDevice}, \Rslot{plotPar},
 and \Rslot{drArea}, which store the device number of the plot, the plotting
@@
 virtual class as it represents a general plot.  The instantiable plot classes
-are the specific plot classes, \Robject{sPlotView}, \Robject{graphView}, and
-\Robject{heatmapView}, which inherit from the \Robject{plotView} class.
-
-The \Robject{sPlotView} class represents a scatterplot view and it has the
+are the specific plot classes, \Rclass{sPlotView}, \Rclass{graphView}, and
+\Rclass{heatmapView}, which inherit from the \Rclass{plotView} class.
+
+The \Rclass{sPlotView} class represents a scatterplot view and it has the
 extra slots, \Rslot{dfRows}, \Rslot{xvar}, and \Rslot{yvar}, where
@@
 variable in the plot, and \Rslot{yvar} is the variable name from the model
-that is shown as the y variable in the plot.  The \Robject{graphView} class
+that is shown as the y variable in the plot.  The \Rclass{graphView} class
 represents a graph plot view and it contains no extra slots besides those in
-the \Robject{plotView} class.  The \Robject{heatmapView} represents a heatmap
+the \Rclass{plotView} class.  The \Rclass{heatmapView} represents a heatmap
 view of the data and it contains the slot, \Rslot{ordering}, which is a list
@@
 views are only applicable for certain model types.  For example, the
-\Robject{graphView} class only makes sense for a view of a graph object,
-which is stored in the \Robject{graphModel} class.
+\Rclass{graphView} class only makes sense for a view of a graph object,
+which is stored in the \Rclass{graphModel} class.
 
@@
 
-For \Robject{sPlotView}, the \Rfunction{clickEvent} method is called when Gtk
+For \Rclass{sPlotView}, the \Rfunction{clickEvent} method is called when Gtk
 emits a signal that a mouse button has been pressed and the
@@
 respectively, using the data in the \Rslot{where} parameter.  For
-\Robject{spreadView}, the \Rfunction{clickEvent} method is called when Gtk
+\Rclass{spreadView}, the \Rfunction{clickEvent} method is called when Gtk
 emits a signal that a row in the spreadsheet has been selected or unselected
 and the \Rfunction{clickEvent} method determines which row was changed.
-For \Robject{graphView}, the \Rfunction{clickEvent} method is called when Gtk
+For \Rclass{graphView}, the \Rfunction{clickEvent} method is called when Gtk
 emits a signal that a mouse button has been pressed and the
@@
 
-The \Robject{heatmapView} class is different from the other views in that it
+The \Rclass{heatmapView} class is different from the other views in that it
 has neither a \Rfunction{clickEvent} or a \Rfunction{motionEvent} method.  The
@@
 into an object (such as the row and column coordinates of the heatmap) on
-the view.  Thus, the \Robject{heatmapView} class is the only view that is not
+the view.  Thus, the \Rclass{heatmapView} class is the only view that is not
 currently interactive.
 
-As with the model classes, the view classes need an update method and thus, all
-views have an \Rfunction{updateView} method that is defined in the
-\Rpackage{iSNetwork} package.  This method must take the
+As with the model classes, the view classes need an update
+method and thus, all views have an \Rfunction{updateView} method that is
+defined in the \Rpackage{iSNetwork} package.  This method must take the
 information that a model has just been updated and appropriately update the
@@
 
-The inheritance structure of the view classes is intended to be extensible.
-For instance, if users wanted to add a new plot view, say a density plot, they
-could create a new view class that inherits from \Robject{plotView}, called
-\Robject{dPlotView}.  This new class, \Robject{dPlotView}, has slots
-named \Rslot{dfRows} and \Rslot{col} to store which rows and which column,
-respectively, were used to calculate the density shown in the view.
+The inheritance structure of the view classes is intended to be
+extensible.  For instance, if users wanted to add a new plot view, say a
+density plot, they could create a new view class that inherits from
+\Rclass{plotView}, called \Rclass{dPlotView}.  This new class,
+\Rclass{dPlotView}, has slots named \Rslot{dfRows} and \Rslot{col} to store
+which rows and which column, respectively, were used to calculate the density
+shown in the view.
 
@@
   \begin{center}
-    \scalebox{0.5}{\includegraphics{ControlWindow.jpg}}
+    \scalebox{0.35}{\includegraphics{ControlWindow.jpg}}
     \caption{ The Control Window. }
@@
 the type of model that is active.  Because the goal is to have multiple
-linked MVC objects, more than one MVC object is loaded and thus,
-more than one model. However, the control window menus and the command line
-functions only perform operations on one MVC object at a time.  Thus, in
-the \Rpackage{iSNetwork} package there is the concept of the ``active''
-MVC object, where the ``active'' MVC object (and thus, the ``active'' model)
-means that operations performed through the control window menus or through
-the command line functions affect that MVC object.  For
-example, the Display menu allows users to create views of the active model,
-but only certain views are appropriate for a type of model, as mentioned in
+linked \Robject{MVC} objects, more than one \Robject{MVC} object is loaded and
+thus, more than one model. However, the control window menus and the command
+line functions only perform operations on one \Robject{MVC} object at a time.
+Thus, in the \Rpackage{iSNetwork} package there is the concept of the
+``active'' \Robject{MVC} object, where the ``active'' \Robject{MVC} object
+(and thus, the ``active'' model) means that operations performed through the
+control window menus or through the command line functions affect that
+\Robject{MVC} object.  For example, the Display menu allows users to create
+views of the active model, but only certain views are appropriate for a type
+of model, as mentioned in 
 Section \ref{Ssec:OneViews}.  If the active model was a data frame model
-(\Robject{dfModel}), then the views that the user could create are a
-scatterplot (\Robject{sPlotView}) and a spreadsheet (\Robject{spreadView}),
-but a graph plot (\Robject{graphView}) does not make sense.  Thus, when the
-model in the ``active'' MVC is of class \Robject{dfModel}, then the Display
+(\Rclass{dfModel}), then the views that the user could create are a
+scatterplot (\Rclass{sPlotView}) and a spreadsheet (\Rclass{spreadView}),
+but a graph plot (\Rclass{graphView}) does not make sense.  Thus, when the
+model in the ``active'' \Robject{MVC} is of class \Rclass{dfModel}, then the
+Display 
 menu has menu items to create a scatterplot and a spreadsheet.  Also, if
-the ``active'' MVC is of class \Robject{dfModel}, then the command line
-functions only let the user create a scatterplot and a spreadsheet view
+the ``active'' \Robject{MVC} is of class \Rclass{dfModel}, then the command
+line functions only let the user create a scatterplot and a spreadsheet view
 of the active model.
@@
 histogram, then a new menu item for that view could be added to the Display
-menu, as long as that view makes sense for the active MVC's model.  By
-allowing users to change the controller, they can add new functionality that
+menu, as long as that view makes sense for the active \Robject{MVC's} model.
+By allowing users to change the controller, they can add new functionality that
 is specific to their problem.  Note that when users make additions to the API,
@@
 To allow this flexibility in the response to an event, a class called
-\Robject{gEventFun} was created.  The \Robject{gEventFun} class stores all of
+\Rclass{gEventFun} was created.  The \Rclass{gEventFun} class stores all of
 the information needed to create a particular response to an event.
@@
 signatures of different callback functions should be the same.  These design
-decisions determined the slots in the \Robject{gEventFun} class and are
+decisions determined the slots in the \Rclass{gEventFun} class and are
 discussed more fully in the following paragraphs.
 
-As shown in Figure \ref{Fig:EventFun}, the slots in the \Robject{gEventFun}
+As shown in Figure \ref{Fig:EventFun}, the slots in the \Rclass{gEventFun}
 class are \Rslot{callFun}, \Rslot{shortName}, and \Rslot{preprocessFun}.  The
@@
 the one parameter for the callback function depends on the model class in
-the active MVC object.  For example, for a graph model, the parameter is
-of class \Robject{AgNode}, representing a node in the graph.  For a data frame
-model, the parameter is of class character, representing a row in the
-data frame.
+the active \Robject{MVC} object.  For example, for a graph model, the
+parameter is of class \Rclass{AgNode}, representing a node in the graph.  For
+a data frame model, the parameter is of class \Rclass{character}, representing
+a row in the data frame.
 
@@
 instance to an event, this can be performed either through the GUI or through
-the command line functions.  The controller environment in each MVC object,
-which is discussed in detail in Section \ref{Ssec:OneMVC},
+the command line functions.  The controller environment in each \Robject{MVC}
+object, which is discussed in detail in Section \ref{Ssec:OneMVC},
 stores which \Robject{gEventFun} object (representing the callback
@@
 
-The \Robject{MVC} class is the class that binds the model, view, and
-controller objects that refer to the same data set into one object.  Thus, the
-creation of a MVC object occurs whenever a new data set is loaded.  A
-MVC object can have one model and each model must be in one
-MVC object so there is a one-to-one relationship between the
-MVC and model objects.  Because of this relationship, the
-MVC object and the model object are identified by the same name,
-which is stored in the \Rslot{modelName} slot of the model object.  When
-loading a new model through \Rpackage{iSNetwork's} API, the name of the model
-is checked to ensure that it is unique.
-
-As shown in Figure \ref{Fig:MVCClass}, the \Robject{MVC} class has five slots
+The \Rclass{MVC} class binds the model, view, and
+controller objects into one object.  Thus, whenever a new data set
+is loaded, a \Robject{MVC} object is created.  A \Robject{MVC} object can have
+one model and each model must be in one \Robject{MVC}
+object so there is a one-to-one relationship between the \Robject{MVC} and
+model objects.  Because of this relationship, the \Robject{MVC}
+object and the model object are identified by the same name, which
+is stored in the \Rslot{modelName} slot of the model object.  When
+loading a new model, the name of the model is checked to
+ensure that it is unique.
+
+As shown in Figure \ref{Fig:MVCClass}, the \Rclass{MVC} class has five slots
 which are, \Rslot{model}, \Rslot{viewList}, \Rslot{controller},
 \Rslot{parentMVC}, and \Rslot{childMVCList}.  The \Rslot{model} slot stores
-the model object, the \Rslot{viewList} slot is a list of the view objects for
-this model, the \Rslot{controller} slot is an environment that stores
-information for this MVC, the \Rslot{parentMVC} slot is the name of the parent
-model (MVC), and the \Rslot{childMVCList} slot is a list of the children models
-(MVCs).  Note that model and MVC objects use the same name so the
-\Rslot{parentMVC} slot refers to both the name of the model and the name of the
-MVC.  Similarly, for the \Rslot{childMVCList} slot, the names in the list
-refer to both the name of the model and the name of the MVC.
+the model object, the \Rslot{viewList} slot is a list of the
+view objects for this model, the \Rslot{controller} slot is an
+environment that stores information for this \Robject{MVC}, the
+\Rslot{parentMVC} slot is the name of the parent model
+(\Robject{MVC}), and the \Rslot{childMVCList} slot is a list of the children
+models (\Robject{MVCs}).  Note that model and
+\Robject{MVC} objects use the same name so the \Rslot{parentMVC} slot refers
+to both the name of the model and the name of the \Robject{MVC}.
+Similarly, for the \Rslot{childMVCList} slot, the names in the list refer to
+both the name of the model and the name of the \Robject{MVC}.
 
@@
 
-Thus, the \Robject{MVC} class not only binds the model, view, and controller
-objects together, but it also shows that the MVC objects are related
-to each other through a parent-child (tree) hierarchy, as shown in Figure
-\ref{Fig:Hier}.  In this picture, `model1' is the name of the original MVC
-object (model) and from it, two MVC objects were created, `model2' and
-`model3'. Then from `model2', three MVC objects were created, `model4',
-`model5', and `model6'.
+Thus, the \Rclass{MVC} class not only binds the model,
+view, and controller objects together, but it also shows
+that the \Robject{MVC} objects are related to each other through a
+parent-child (tree) hierarchy, as shown in Figure \ref{Fig:Hier}.  In this
+picture, `model1' is the name of the original \Robject{MVC} object (model) and
+from it, two \Robject{MVC} objects were created, `model2.1' and
+`model2.2'. Then from `model2.1', three \Robject{MVC} objects were created,
+`model3.1', `model3.2', and `model3.3'.  As an example, model1 could be gene
+expression data, model2.1 could be gene expression data that is a subset of
+model1 (based on filtering), model2.2 could be derived by performing
+multidimensional scaling (MDS) on model1, model 3.1 could be the gene ontology
+(GO) graph representing the molecular functions of the genes in model2.1,
+model 3.2 could be the KEGG pathways that the genes in model2.1 are involved
+in, and model 3.3 could be the chromosome locations of the genes in model2.1.
 
@@
     \scalebox{0.6}{\includegraphics{Hierarchy2.jpg}}
-    \caption{ The Parent-Child Relationship Between MVCs. }
+    \caption{ The Parent-Child Relationship Between \Robject{MVCs}. }
     \label{Fig:Hier}
@@
 
-A design decision was that a new model (and thus, a new MVC object) could
-either be loaded or could be created from an already loaded MVC object.  As
-mentioned in Section \ref{Ssec:MMVC}, the focus in this paper is on
-experimental data that are linked to meta-data, as the linked data sets that we
-are most interested in visualizing.  In this situation, the experimental data
-(and its MVC object) can be thought of as determining what meta-data are
-studied.  Thus, there is a hierarchical relationship between these data sets
-where one MVC object creates another MVC object.  This focus led to the
-decision that a MVC object can have at most one parent MVC, from which it was
-derived, but it can have many child MVCs.  An easy example of a child model
-derived from a parent model is to start with a data frame as the parent model
-and then take a subset of this data frame as the child model.
+A design decision was that a new model (and thus, a new
+\Robject{MVC} object) could either be loaded or could be created from an
+already loaded \Robject{MVC} object.  As mentioned in Section \ref{Ssec:MMVC},
+the focus in this paper is on experimental data that are linked to meta-data,
+as the linked data sets that we are most interested in visualizing.  In this
+situation, the experimental data (and its \Robject{MVC} object) can be thought
+of as determining what meta-data are studied.  Thus, there is a hierarchical
+relationship between these data sets where one \Robject{MVC} object creates
+another \Robject{MVC} object.  This focus led to the decision that a
+\Robject{MVC} object can have at most one parent \Robject{MVC}, from which it
+was derived, but it can have many child \Robject{MVCs}.  An easy example of a
+child model derived from a parent model is to start with a
+data frame as the parent model and then take a subset of this data
+frame as the child model.
 
@@
 
-The message classes are intended to provide communication between different
-components of the MVC design, such as when the model changes a message needs to
-be sent to the views to let them know that they should be updated.  This
+The message classes provide communication between different
+components of the MVC design, such as when the model changes a message needs
+to be sent to the views to let them know that they should be updated.  This
 communication between the model, view and controller is crucial for the pieces
 to work together and yet, still be independent of each other.  For information
-about messages that are passed between MVC objects, please see Section
-\ref{Ssec:MultMess}.
-
-The object model for the message classes was derived from the idea that all
-messages must have certain methods in common, such as \Rfunction{initialize}
-and \Rfunction{handleMessage}, because all messages must be created and
-handled in some manner so that the message is read and acted upon.  The
-\Rfunction{handleMessage} method can be considered a central dispatch for
-messages because all messages are sent to the \Rfunction{handleMessage}
-method, which determines where the information should be sent.   Also,
-since messages are sent when something has changed, either through addition,
-alteration, or deletion, the message classes reflect only certain operations.
-These common purposes for the messages allow the message object model to be
-constructed, and this model is shown in Figure \ref{Fig:Mess}.
+about messages that are passed between \Robject{MVC} objects, please see
+Section\ref{Ssec:MultMess}.
+
+The object model for the message classes was derived from the idea
+that all messages must have certain methods in common, such as
+\Rfunction{initialize} and \Rfunction{handleMessage}, because all
+messages must be created and handled in some manner so that the
+message is read and acted upon.  All messages are sent to
+the \Rfunction{handleMessage} method, which determines where the information
+should be sent and hence is the dispatcher.   Also, since messages
+are sent when something has changed, either through addition, alteration, or
+deletion, the message classes reflect only certain operations.  These
+common purposes for the messages allow the message object
+model to be constructed, and this model is shown in Figure \ref{Fig:Mess}.
 
@@
 
-As with the view classes, the top message class, \Robject{gMessage}, is a
-virtual class.  It contains no slots and its purpose is to bind the other
-message classes together.
-
-Currently, there are two types of messages for messaging within a model, view,
-controller object: an add and an update message.  Both the add message,
-\Robject{gAddMessage}, and the update message, \Robject{gUpdateMessage},
+As with the view classes, the top message class,
+\Rclass{gMessage}, is a virtual class.  It contains no slots and its purpose
+is to bind the other message classes together.
+
+Currently, there are two types of messages for messaging within a
+model, view, controller object: an add and an update message.  Both the add
+message, \Rclass{gAddMessage}, and the update message, \Rclass{gUpdateMessage},
 contain three slots, \Rslot{dataName}, \Rslot{mData}, and \Rslot{type}.
@@
 operation, and \Rslot{type} is a character string that gives the type of
-addition or alteration to perform.  Both \Robject{gAddMessage} and
-\Robject{gUpdateMessage} are virtual classes.
+addition or alteration to perform.  Both \Rclass{gAddMessage} and
+\Rclass{gUpdateMessage} are virtual classes.
 
@@
 model (and similarly with updating views versus models), it makes sense to
-have two separate message classes for these operations.  Thus, when messages
-are sent between components, they are from one of the following classes:
-\Robject{gAddDataMessage}, \Robject{gAddViewMessage},
-\Robject{gUpdateDataMessage}, and \Robject{gUpdateViewMessage}.
-
-The \Robject{gAddDataMessage} class represents a message to add a model and a
-MVC object because whenever a model is added, a MVC object is also added that
-contains this model.  The \Robject{gAddViewMessage} class represents a message
-to create a view.  This message can only be used after a model has been
-added.
-
-The \Robject{gUpdateDataMessage} class represents a message to update
-a model.  This message is crucial for linking because it is responsible for
-ensuring that once a model has been updated this information propagates to
-any views of the model.  This same information must also be passed along to any
-other MVC object that is related to this model.  Because this information gets
-passed to other MVC objects, which is discussed further in Section
-\ref{Sec:MultMVC}, this message also has one new slot, \Rslot{from}, in
-addition to the slots it inherits from \Robject{gUpdateMessage}.  The
-\Rslot{from} slot is the name of the model that the update data message came
-from.  This slot is necessary because the update data message can come from
-one of two sources: the message can occur because of user interaction with a
-view of this model or the message may occur because a different model that is
-linked to this model has been updated.  So the update data message may come
-from within this MVC object or it may come from a MVC object that
-has changed its model and is related to this MVC object, either as this MVC
-object's parent MVC or one of this MVC object's child MVCs.  Information on
-passing messages between MVC objects is given in Section \ref{Sec:MultMVC}.
-Finally, the \Robject{gUpdateViewMessage} class represents a message to update
-a view.  This message is created and handled whenever the model is updated so
-that the views accurately reflect the model.
-
-An example of message passing within a MVC object is shown in Figure
+have two separate message classes for these operations.  Messages
+sent between components are from one of the following classes:
+\Rclass{gAddDataMessage}, \Rclass{gAddViewMessage},
+\Rclass{gUpdateDataMessage}, and \Rclass{gUpdateViewMessage}.  The
+\Rclass{gAddDataMessage} class represents a message to add a model
+and a \Robject{MVC} object because whenever a model is added, a
+\Robject{MVC} object is created that contains this model.  The
+\Rclass{gAddViewMessage} class represents a message to create a view.  This
+message can only be used after a model has been added.
+
+The \Rclass{gUpdateDataMessage} class represents a message to update
+a model.  This message is crucial for linking because it is
+responsible for ensuring that once a model has been updated this
+information propagates to all views of the model.  This
+same information must also be passed along to any other \Robject{MVC} object
+that is related to this model.  Because this
+information gets passed to other \Robject{MVC} objects (see Section
+\ref{Sec:MultMVC}), this message also has one new slot, \Rslot{from}, in
+addition to the slots it inherits from \Rclass{gUpdateMessage}.  The
+\Rslot{from} slot is the name of the model that the update data
+message came from.  This slot is necessary because the update data message can
+come from one of two sources: the message can occur because of user
+interaction with a view of this model or the message may
+occur because a different model that is
+linked to this model has been updated.  Thus, the update data
+message may come from within this \Robject{MVC} object or it may come from
+either the parent \Robject{MVC} or a child \Robject{MVC}.  Finally, the
+\Rclass{gUpdateViewMessage} class represents a message to update a
+view.  This message is created and handled whenever the
+model is updated so that the views accurately reflect the model.
+
+%\subsubsection*{Example of Message Passing}
+
+An example of message passing within a \Robject{MVC} object is shown in Figure
 \ref{Fig:MPwithin}.  In this example, the response to a user performing a
@@
 \ref{Ssec:OneEvent}.  Here the user first clicks on a point in the
-scatterplot, which is shown as step 1 (the circled 1 on the figure).  In the
+scatterplot, which is shown as Step 1 (the circled 1 on the figure).  In the
 figure, the point that has been clicked has an arrow next to it.  Then, in
-step 2, an update data message is sent to the model, which updates the
-\Rslot{virtualData} slot of the model object.  After the model has been
-updated, the views must be notified that a change has occurred so in step 3,
-an update view message is created and sent to all views of this model.  In
-response to this message, the view updates itself in step 4 and now the point
-is colored red on the view to reflect the change in the model.  In this
-example, the \Rslot{from} slot in the \Robject{gUpdateDataMessage} object is
-the name of this MVC object since the message came from within the MVC.
+Step 2, an update data message is sent to the model, which updates
+the \Rslot{virtualData} slot of the model object.  After the
+model has been updated, the views must be notified that a
+change has occurred so in Step 3, an update view message is created and sent
+to all views of this model.  In response to this
+message, the view updates itself in Step 4 and now the
+point is colored red on the view to reflect the change in the
+model.  In this example, the \Rslot{from} slot in the
+\Robject{gUpdateDataMessage} object is the name of this \Robject{MVC} object
+since the message came from within the \Robject{MVC}.
 
@@
     \scalebox{0.6}{\includegraphics{newMPwithin.jpg}}
-    \caption{ Message Passing within a MVC object. }
+    \caption{ Message Passing within a \Robject{MVC} object.  The first step,
+      indicated by a circled one, occurs when a user interacts with the y
+      versus x scatterplot.  Then an update data message is sent to the model
+      in the second step, which alters the model.  Next in the third step all
+      views of this model are sent an update view message, which causes the
+      view to be updated in the fourth step. }
     \label{Fig:MPwithin}
@@
 
-All four classes, \Robject{gAddDataMessage}, \Robject{gAddViewMessage},
-\Robject{gUpdateDataMessage}, and \Robject{gUpdateViewMessage}, have
+All four classes, \Rclass{gAddDataMessage}, \Rclass{gAddViewMessage},
+\Rclass{gUpdateDataMessage}, and \Rclass{gUpdateViewMessage}, have
 \Rfunction{initialize} and \Rfunction{handleMessage} methods, which perform
@@
 inheritance structure is designed to allow for extensibility.  For example, a
-user could create a \Robject{gUpdateControlMessage} class that inherited from
-the \Robject{gUpdateMessage} class.  This new class,
-\Robject{gUpdateControlMessage}, could convey information that told the
+user could create a \Rclass{gUpdateControlMessage} class that inherited from
+the \Rclass{gUpdateMessage} class.  This new class,
+\Rclass{gUpdateControlMessage}, could convey information that told the
 control window to be updated in response to a change in the active view (the
@@
 
-We now describe the implementation of multiple MVC objects for linking views
-of multiple linked data sets.  Each MVC object represents one model and
-its views.  Now for the views of different data sets to be linked, they must
-link through the data sets so this linking information must be stored
-somewhere.  Also, there must be message classes to pass information between
-MVC objects.
+We now describe the implementation of multiple \Robject{MVC} objects for
+linking views of multiple linked data sets.  Each \Robject{MVC} object
+represents one model and its views.  Now for the
+views of different data sets to be linked, they must link through
+the data sets so this linking information must be stored somewhere.  Also,
+there must be message classes to pass information between
+\Robject{MVC} objects.
 
 Thinking of experimental data that are linked to meta-data as an example, there
-are two MVC objects, one for the experimental data and one for the
+are two \Robject{MVC} objects, one for the experimental data and one for the
 meta-data.  In the example of experimental microarray data that are linked to
@@
 
-As mentioned in Section \ref{Ssec:OneMVC}, the \Robject{MVC} class has slots
-for how the current MVC relates to other MVC objects and these slots are
-\Rslot{parentMVC} and \Rslot{childMVCList}.  These slots show that MVC objects
-are related through a tree hierarchy where a child model is derived from its
-parent model.  Consider Figure \ref{Fig:Hier}, which was shown previously.  
-If model1's model changes, then this information first gets passed to model2
-and model3 and these two MVC's may make changes to their models based on this
-information.  Then model2 passes this information to model4, model5, and
-model6 based on the changes that model2 made to its model.  Thus, information
-passing between model1 and model4, for instance, only occurs through model2.
-So messages can only be sent from parent MVC to child MVC or from child MVC to
-parent MVC.
-
-Next we consider how the data sets are linked (and thus, how MVC
+As mentioned in Section \ref{Ssec:OneMVC}, the \Rclass{MVC} class has slots
+for how the current \Robject{MVC} relates to other \Robject{MVC} objects and
+these slots are \Rslot{parentMVC} and \Rslot{childMVCList}.  These slots show
+that \Robject{MVC} objects are related through a tree hierarchy where a child
+model is derived from its parent model.  Consider Figure
+\ref{Fig:Hier}, which was shown previously.  If model1's model changes, then
+this information first gets passed to model2.1 and model2.2 and these two
+\Robject{MVCs} may make changes to their models based on this
+information.  Then model2.1 passes this information to model3.1, model3.2, and
+model3.3 based on the changes that model2.1 made to its model.  Thus,
+information passing between model1 and model3.1, for instance, only occurs
+through model2.1.  So messages can only be sent from parent \Robject{MVC} to
+child \Robject{MVC} or from child \Robject{MVC} to parent \Robject{MVC}.
+
+Next we consider how the data sets are linked (and thus, how \Robject{MVC}
 objects are linked) followed by how messages are passed between these
-linked MVC objects.
+linked \Robject{MVC} objects.
 
@@
 
-For messages to pass between MVC objects, the information from one MVC must be
-converted into useful information for a different MVC object.  Thus, there
-needs to be some way to determine how data from one MVC object relates to data
-in a different MVC object.  This conversion is done by two functions that are
-stored in the \Rslot{linkData} slot of the model objects.  The
-\Rslot{linkData} slot is a list with two elements: a \Rfunction{toParent}
-function and a \Rfunction{fromParent} function.
-
-When a child MVC is created from an existing MVC, the child model's
-\Rslot{linkData} slot is filled with the two functions,
-\Rfunction{toParent} and \Rfunction{fromParent}.  When the model in the child
-MVC is updated, the \Rfunction{toParent} function in the \Rslot{linkData} slot
-is used to convert the child model's data change into useful information
-for the parent model.
+For messages to pass between \Robject{MVC} objects, the information from one
+\Robject{MVC} must be converted into useful information for a different
+\Robject{MVC} object.  Thus, there needs to be some way to determine how data
+from one \Robject{MVC} object relates to data in a different \Robject{MVC}
+object.  This conversion is done by two functions that are stored in the
+\Rslot{linkData} slot of the model objects.  The \Rslot{linkData}
+slot is a list with two elements: a \Rfunction{toParent} function and a
+\Rfunction{fromParent} function.
+
+When a child \Robject{MVC} is created from an existing \Robject{MVC}, the
+child model's \Rslot{linkData} slot is filled with the two functions,
+\Rfunction{toParent} and \Rfunction{fromParent}.  When the model in
+the child \Robject{MVC} is updated, the \Rfunction{toParent} function in the
+\Rslot{linkData} slot is used to convert the child model's data
+change into useful information for the parent model.
 
 Now suppose that the parent model changed instead.  This information
-needs to be passed to the child model.  However, a parent can have multiple
-child MVC objects so the parent sends a message to each child MVC to notify
-the child of the change and then each child MVC handles the message by using
-its \Rfunction{fromParent} function.  The \Rfunction{fromParent} function
-converts the parent model's data change into useful information for this
-particular child model.  Thus, the parent model's data change may be converted
-into different information for different child models and this can happen
-because each child model has its own \Rfunction{fromParent} function.  When a
-child MVC object is created and its model's \Rslot{linkData} slot is filled,
-the \Rfunction{toParent} function converts information from this child model
-into useful information for its parent MVC and the \Rfunction{fromParent}
-function converts a message from its parent MVC into useful information for
-this child MVC.
+needs to be passed to the child model.  However, a parent can have
+multiple child \Robject{MVC} objects so the parent sends a message to each
+child \Robject{MVC} to notify the child of the change and then each child
+\Robject{MVC} handles the message by using its \Rfunction{fromParent}
+function.  The \Rfunction{fromParent} function converts the parent
+model's data change into useful information for this
+particular child model.  Thus, the parent model's data
+change may be converted into different information for different child models
+and this can happen because each child model has
+its own \Rfunction{fromParent} function.  When a child \Robject{MVC} object is
+created and its model's \Rslot{linkData} slot is filled, the
+\Rfunction{toParent} function converts information from this child
+model into useful information for its parent \Robject{MVC} and the
+\Rfunction{fromParent} function converts a message from its parent
+\Robject{MVC} into useful information for this child \Robject{MVC}.
 
 An example of how these functions, \Rfunction{toParent} and
-\Rfunction{fromParent}, are used when messages are passed between MVCs is
-given in Section \ref{Ssec:MultMess} where Figure \ref{Fig:MessPass} shows the
-\Rfunction{toParent} function being used to convert information from a child
-MVC object into useful information for the parent MVC object.
+\Rfunction{fromParent}, are used when messages are passed between
+\Robject{MVCs} is given in Section \ref{Ssec:MultMess} where Figure
+\ref{Fig:MessPass} shows the \Rfunction{toParent} function being used to
+convert information from a child \Robject{MVC} object into useful information
+for the parent \Robject{MVC} object.
 
@@
 
-In Section \ref{Ssec:OneMess}, where the message classes to pass information
-within one MVC object were discussed, it was noted that there are two main
-types of messages: an add message and an update message.  These message types
-also are seen in passing messages between different MVC objects, as shown
-in Figure \ref{Fig:BetwMess}.
+In Section \ref{Ssec:OneMess}, where the message classes to pass
+information within one \Robject{MVC} object were discussed, it was noted that
+there are two main types of messages: an add message and an update message.
+These message types also are seen in passing messages between different
+\Robject{MVC} objects, as shown in Figure \ref{Fig:BetwMess}.
 
@@
     \scalebox{0.6}{\includegraphics{newMessageClass2.jpg}}
-    \caption{ Inheritance for Message Classes that are Passed Between MVCs. }
+    \caption{ Inheritance for Message Classes that are Passed Between
+      \Robject{MVCs}. }
     \label{Fig:BetwMess}
@@
 
-The first message class that is discussed is \Robject{gAddChildMessage}.
-A \Robject{gAddChildMessage} object is very similar to a
-\Robject{gAddDataMessage} object because they are both adding a new
-model and a new MVC object.  The difference with a \Robject{gAddChildMessage}
-object is that it must fill in some extra information to tie this new MVC
-object to its parent MVC object.  The \Robject{gAddChildMessage} class
-inherits from the \Robject{gAddMessage} class, with the same slots of
-\Rslot{dataName}, \Rslot{mData}, and \Rslot{type}.  The \Rslot{dataName} slot
-is the name of the new model (and new MVC), the \Rslot{mData} slot is a list
-that contains the model data and virtual data to fill the \Rslot{modelData}
-and \Rslot{virtualData} slots, respectively, of the new model, and the
+The first message class that is discussed is
+\Rclass{gAddChildMessage}.  A \Robject{gAddChildMessage} object is very
+similar to a \Robject{gAddDataMessage} object because they are both adding a
+new model and a new \Robject{MVC} object.  The difference with a
+\Robject{gAddChildMessage} object is that it must fill in some extra
+information to tie this new \Robject{MVC} object to its parent \Robject{MVC}
+object.  The \Rclass{gAddChildMessage} class inherits from the
+\Rclass{gAddMessage} class, with the same slots of \Rslot{dataName},
+\Rslot{mData}, and \Rslot{type}.  The \Rslot{dataName} slot is the name of the
+new model (and new \Robject{MVC}), the \Rslot{mData} slot is a list that
+contains the
+model data and virtual data to fill the \Rslot{modelData} and
+\Rslot{virtualData} slots, respectively, of the new model, and the
 \Rslot{type} slot gives the type of the model (currently, the options are
-``exprSet'', ``graph'', or ``data.frame'').  As with the previous message
-classes, the \Robject{gAddChildMessage} class has two methods:
+``exprSet'', ``graph'', or ``data.frame'').  As with the previous
+message classes, the \Rclass{gAddChildMessage} class has two methods:
 \Rfunction{initialize} and \Rfunction{handleMessage}.  The
 \Rfunction{initialize} method properly fills the slots of the object when it
-is created and the \Rfunction{handleMessage} method ensures that the new model
-and MVC objects are created (which is what the \Rfunction{handleMessage}
-method for the \Robject{gAddDataMessage} class does) and then it fills the
-\Rslot{linkData} slot of the new child model so that the \Rfunction{toParent}
-and \Rfunction{fromParent} functions are available for messaging between MVC
-objects and finally it stores information in the child MVC object's controller
-that determines if the child model was created from a subset of the parent
+is created and the \Rfunction{handleMessage} method ensures that the new
+model and \Robject{MVC} objects are created (which is what the
+\Rfunction{handleMessage} method for the \Rclass{gAddDataMessage} class does)
+and then it fills the \Rslot{linkData} slot of the new child model
+so that the \Rfunction{toParent} and \Rfunction{fromParent} functions are
+available for messaging between \Robject{MVC} objects and finally it stores
+information in the child \Robject{MVC} object's controller that determines if
+the child model was created from a subset of the parent
 model.  Thus, the \Rfunction{handleMessage} method for the
-\Robject{gAddChildMessage} class both creates the child MVC object and relates
-it to the parent MVC object.
-
-The next two message classes pertain to sending information to linked MVC
-objects when a parent MVC or child MVC object has changed.  These classes are
-\Robject{gSendParentMessage} and \Robject{gSendChildMessage} and they both
-inherit from the \Robject{gMessage} class, as shown in Figure
-\ref{Fig:BetwMess}.  Because the messages are sending information to a linked
-MVC object, they both contain a slot for a \Robject{gUpdateDataMessage} object
-that was just used to update a MVC object.  The \Robject{gSendParentMessage}
-class has one slot, \Rslot{childUpdateDataMessage}, and this slot contains the
-\Robject{gUpdateDataMessage} object that was used to update the child model.
-Similarly, the \Robject{gSendChildMessage} class has two slots,
-\Rslot{parentUpdateDataMessage} and \Rslot{childName}, where the
+\Rclass{gAddChildMessage} class both creates the child \Robject{MVC} object
+and relates it to the parent \Robject{MVC} object.
+
+The next two message classes pertain to sending information to linked
+\Robject{MVC} objects when a parent \Robject{MVC} or child \Robject{MVC}
+object has changed.  These classes are \Rclass{gSendParentMessage} and
+\Rclass{gSendChildMessage} and they both inherit from the \Rclass{gMessage}
+class, as shown in Figure \ref{Fig:BetwMess}.  Because the messages are
+sending information to a linked \Robject{MVC} object, they both contain a slot
+for a \Robject{gUpdateDataMessage} object that was just used to update a
+\Robject{MVC} object.  The \Rclass{gSendParentMessage} class has one slot,
+\Rslot{childUpdateDataMessage}, and this slot contains the
+\Robject{gUpdateDataMessage} object that was used to update the child
+model.  Similarly, the \Rclass{gSendChildMessage} class has two
+slots, \Rslot{parentUpdateDataMessage} and \Rslot{childName}, where the
 \Rslot{parentUpdateDataMessage} slot contains the \Robject{gUpdateDataMessage}
-object that was used to update the parent model and the \Rslot{childName} slot
-contains the name of the child model that is being updated because a parent
-MVC can have more than one child MVC.
-
-As with all the other message classes, these two send messages have two
-methods: \Rfunction{initialize} and \Rfunction{handleMessage}.  The
+object that was used to update the parent model and the
+\Rslot{childName} slot contains the name of the child model that is
+being updated because a parent \Robject{MVC} can have more than one child
+\Robject{MVC}.
+
+As with all the other message classes, these two send messages have
+two methods: \Rfunction{initialize} and \Rfunction{handleMessage}.  The
 \Rfunction{initialize} method properly sets the slots of the new object.  The
 \Rfunction{handleMessage} method for a \Robject{gSendParentMessage}
-takes the \Robject{gUpdateDataMessage} object from the child model and converts
-it to a \Robject{gUpdateDataMessage} object for the parent model using the
-\Rfunction{toParent} function.  Similarly, the \Rfunction{handleMessage}
-method for a \Robject{gSendChildMessage} takes the
-\Robject{gUpdateDataMessage} object from the parent model and converts it to a
-\Robject{gUpdateDataMessage} object for the child model using the
-\Rfunction{fromParent} function.  An example of how these messages are passed
-and converted is shown in Figure \ref{Fig:MessPass}.
+takes the \Robject{gUpdateDataMessage} object from the child model
+and converts it to a \Robject{gUpdateDataMessage} object for the parent
+model using the \Rfunction{toParent} function.  Similarly, the
+\Rfunction{handleMessage} method for a \Robject{gSendChildMessage} takes the
+\Robject{gUpdateDataMessage} object from the parent model and
+converts it to a \Robject{gUpdateDataMessage} object for the child
+model using the \Rfunction{fromParent} function.  An example of how
+these messages are passed and converted is shown in Figure \ref{Fig:MessPass}.
 
@@
 notified of the change in a way that each model can properly act upon that
-information.  Because each MVC object has a different model (and thus a
-different data set and potentially a different data structure), information
-that pertains to the model for one MVC must be converted to useful information
-for the linked model and this conversion is done by the \Rfunction{toParent}
-and \Rfunction{fromParent} functions.
+information.  Because each \Robject{MVC} object has a different model (and
+thus a different data set and potentially a different data structure),
+information that pertains to the model for one \Robject{MVC} must be converted
+to useful information for the linked model and this conversion is done by the
+\Rfunction{toParent} and \Rfunction{fromParent} functions.
 
@@
     \scalebox{0.55}{\includegraphics{newMP.jpg}}
-    \caption{ Message Passing Between and Within MVCs. }
+    \caption{ Message Passing Between and Within \Robject{MVCs}.  When a user
+      interacts with the y versus x scatterplot, the first step, indicated by
+      a circled one, is to send an update data message to model 2.  After
+      model 2 has been updated, all views of this model must be notified, which
+      is shown in the second step when an update view message is sent.  Next
+      any data that are linked to model 2 must be notified that model 2
+      changed, which is shown in the third step as a send parent message.
+      This message is converted into useful information for model 1 in
+      the fourth step by the \Rfunction{toParent} function, and this
+      information is sent to model 1 in the fifth step when an update data
+      message is sent.  Finally, after model 1 is updated, an update view
+      message is sent in the sixth step to update all views of model 1. }
     \label{Fig:MessPass}
@@
 Section \ref{Ssec:Backg}.  Now, however, Figure \ref{Fig:MessPass} shows how
-MVC objects are defined and how messages are passed within one MVC as well
-as how this information gets passed on to its parent MVC using messages.
-Suppose for this example that when the user clicks a mouse button over the
-scatterplot, the response to this event is to color the point under the cursor
+\Robject{MVC} objects are defined and how messages are passed within one
+\Robject{MVC} as well as how this information gets passed on to its parent
+\Robject{MVC} using messages.  Suppose for this example that when the user
+clicks a mouse button over the scatterplot, the response to this event is to
+color the point under the cursor 
 red.  In this picture, the message passing starts when a user clicks a point,
@@
 initially clicked is red.  Then a send parent message is sent because MVC
-2, which is the MVC object for Model 2, has a parent MVC, which is MVC 1.  Next
-the \Rfunction{toParent} function converts the child's update data message
-into an update data message that the parent MVC understands.  Then the
-update data message that was created by the \Rfunction{toParent} function is
-sent to Model 1.  Finally, after Model 1 has been updated, an update view
-message is sent to all of Model 1's views.  It is in this last step that the
-view of Model 1 has a point colored red to indicate which data from Model 1 is
-linked to data from Model 2.  All of the data and the views are
-synchronized.
+2, which is the \Robject{MVC} object for Model 2, has a parent \Robject{MVC},
+which is MVC 1.  Next the \Rfunction{toParent} function converts the child's
+update data message into an update data message that the parent \Robject{MVC}
+understands.  Then the update data message that was created by the
+\Rfunction{toParent} function is sent to Model 1.  Finally, after Model 1 has
+been updated, an update view message is sent to all of Model 1's views.  It is
+in this last step that the view of Model 1 has a point colored red to indicate
+which data from Model 1 is linked to data from Model 2.  All of the data and
+the views are synchronized.
 
@@
 paradigm.  For visualizing linked data sets, the MVC design, which is a well
-known and thus, well tested software paradigm, was extended to multiple MVC
-objects where each MVC object contains one data set and its views.  This
-solution broke down the problem of linked, interactive views of linked data
-sets into components (MVC objects) that can be reused, extended and linked.
-
-Once the design focused on MVC objects as the component of interest for
-creating linked views of linked data sets, the next step was to determine how
-these MVC objects were related.  Because the focus in this paper was on
-experimental data that are linked to meta-data, a hierarchical relationship
-between MVC objects was built into the MVC class definition.  The experimental
-data can be thought of as determining what meta-data are studied and thus, the
-meta-data are derived from the experimental data, which shows a parent-child
-relationship.
+known and thus, well tested software paradigm, was extended to multiple
+\Robject{MVC} objects where each \Robject{MVC} object contains one data set
+and its views.  This solution broke down the problem of linked, interactive
+views of linked data sets into components (\Robject{MVC} objects) that can be
+reused, extended and linked.
+
+Once the design settled on \Robject{MVC} objects as the component of interest
+for creating linked views of linked data sets, the next step was to determine
+how these \Robject{MVC} objects were related.  Because the focus in this paper
+was on experimental data that are linked to meta-data, a hierarchical
+relationship between \Robject{MVC} objects was built into the \Rclass{MVC}
+class definition.  The experimental data can be thought of as determining what
+meta-data are studied and thus, the meta-data are derived from the
+experimental data, which shows a parent-child relationship.
 
 Another consideration in the design was creating a method for communicating
-between MVC objects.  Already communication within a MVC object was sent using
-instances of the message classes.  Thus, it seemed a natural extension to have
-message classes that pertained to passing information between MVC objects.
-The goal of the message classes was to allow communication between independent
-components in the design.
-
-For message passing between MVC objects, when a MVC's model is updated, it
-notifies its parent and child MVCs that a change has occurred.  Then when a
-MVC gets a message from a parent or child MVC, the MVC can either decide to
-ignore the message or read the message, which results in this MVC's model
-being updated.  However, the MVC object can not decide to only see certain
+within and between \Robject{MVC} objects.  This method must pass information
+between the components while keeping the components independent.
+Communication within a \Robject{MVC} object was handled using instances of the
+message classes, which performed one of four functions: adding a model, adding
+a view, updating a model, or updating a view.  Starting with this design, it
+seemed a natural extension to have message classes that passed information
+between \Robject{MVC} objects.  When a \Robject{MVC's} model is updated, it
+notifies its parent and child \Robject{MVCs} that a change has occurred using
+message objects.  Then when a \Robject{MVC} gets a message from a parent or
+child \Robject{MVC}, the \Robject{MVC} can either decide to ignore the message
+or read the message, which results in this \Robject{MVC's} model being
+updated.  However, the \Robject{MVC} object can not decide to only see certain
 types of messages.  It must either accept all messages and update its model
@@
 passing is to allow more selectivity in which messages are accepted between
-MVC objects so that the linking between MVC objects is more sophisticated.
+\Robject{MVC} objects so that the linking between \Robject{MVC} objects is
+more sophisticated.
 
@@
 create an extensible design that allows future users to make additions.
-The first goal was accomplished by implementing MVC objects as components that
-communicate with each other through messages.  The second goal of creating
-interactive views was met by using the \Rpackage{RGtk} and
+The first goal was accomplished by implementing \Robject{MVC} objects as
+components that communicate with each other through messages.  The second goal
+of creating interactive views was met by using the \Rpackage{RGtk} and
 \Rpackage{gtkDevice} packages to create devices that could respond to events.
-In addition the \Robject{gEventFun} class was designed so that users could
+In addition the \Rclass{gEventFun} class was designed so that users could
 alter the response to an event.  Finally, the last goal of creating an
@@
 some data sets.  As an example, the linked tables from a relational database
-are not related through a tree structure and thus, having parent and child MVC
-objects may not make sense with these data sets.  However, the design allows
-users to add new types of models, new types of views, and new message classes
-so there is still flexibility for additions as long as the data sets are
-connected through parent-child relationships.  These new class definitions
-could be created by the user in a new package that extends the
+are not related through a tree structure and thus, having parent and child
+\Robject{MVC} objects may not make sense with these data sets.  However, the
+design allows users to add new types of models, new types of views, and new
+message classes so there is still flexibility for additions as long as the
+data sets are connected through parent-child relationships.  These new class
+definitions could be created by the user in a new package that extends the
 \Rpackage{MVCClass} package that was discussed in this paper.  Other additions
@@
 the interaction (these interactions include mouse button click events and
-mouse movement events), and users can create child MVC objects through the GUI
-or through command line functions.  Using the functionality in the
-\Rpackage{iSNetwork} package, the user can create views of linked data sets
+mouse movement events), and users can create child \Robject{MVC} objects
+through the GUI or through command line functions.  Using the functionality in
+the \Rpackage{iSNetwork} package, the user can create views of linked data sets
 and can explore visually the relationships between variables in different data

### 2006-01-26 R/MVC.R

@@ -10,7 +10,7 @@ setOldClass("GtkCList")
 # a virtual model class that all model classes will be derived from
 # modelData will be the actual model data, linkData will be any data
 # that links the model to its parent model (a list of 2 functions:
-# linkToParent and linkToChild), virtualData is any data that
+# toParent and fromParent), virtualData is any data that
 # is needed for views of the model, and modelName is the name of the model
 # (a way for the user to refer to the data)
 setClass("gModel", representation(modelData="ANY", linkData="list", 
@@ -119,7 +119,7 @@ setClass("genView", representation(dataName="character", win="GtkWindow",
          winNum="numeric"), contains=("VIRTUAL"))
 
 setClass("plotView", representation(plotDevice="numeric", plotPar="list",
-         drArea="GtkDrawingArea"), contains="genView")
+         drArea="GtkDrawingArea"), contains=c("genView", "VIRTUAL"))
 
 # store the row names and column names rather than their indices
 setClass("sPlotView", representation(dfRows="character", colx="character",

### 2006-01-26 R/messages.R

@@ -5,7 +5,7 @@
 setClass("gMessage")
 
 setClass("gUpdateMessage", representation(type="character", mData="list",
-          dataName="character"), contains="gMessage")
+          dataName="character"), contains=c("gMessage", "VIRTUAL"))
 
 setClass("gUpdateViewMessage", contains="gUpdateMessage")
 
@@ -16,7 +16,7 @@ setClass("gUpdateDataMessage", representation(from="character"),
           contains="gUpdateMessage")
 
 setClass("gAddMessage", representation(dataName="character", mData="list",
-         type="character"), contains="gMessage")
+         type="character"), contains=c("gMessage", "VIRTUAL"))
 setClass("gAddViewMessage", contains="gAddMessage")
 setClass("gAddDataMessage", contains="gAddMessage")
 
@@ -26,13 +26,13 @@ setClass("gAddChildMessage", contains="gAddMessage")
 # added 7/14/05 to update a parent model
 # the slot for this class tells the parent what the child update data
 # message was
-setClass("gUpdateParentMessage",
+setClass("gSendParentMessage",
          representation(childUpdateDataMessage="gUpdateDataMessage"), 
          contains="gMessage")
 # added 7/19/05 to update a child model
 # the slot for this class tells the child what the parent update data
 # message was
-setClass("gUpdateChildMessage",
+setClass("gSendChildMessage",
          representation(parentUpdateDataMessage="gUpdateDataMessage",
          childName="character"),
          contains="gMessage")
@@ -49,19 +49,19 @@ setMethod("from", "gUpdateDataMessage", function(object)
 if (is.null(getGeneric("childUpdateDataMessage")))
   setGeneric("childUpdateDataMessage", function(object)
             standardGeneric("childUpdateDataMessage"))
-setMethod("childUpdateDataMessage", "gUpdateParentMessage", function(object)
+setMethod("childUpdateDataMessage", "gSendParentMessage", function(object)
          object@childUpdateDataMessage)
 
 if (is.null(getGeneric("parentUpdateDataMessage")))
   setGeneric("parentUpdateDataMessage", function(object)
             standardGeneric("parentUpdateDataMessage"))
-setMethod("parentUpdateDataMessage", "gUpdateChildMessage", function(object)
+setMethod("parentUpdateDataMessage", "gSendChildMessage", function(object)
          object@parentUpdateDataMessage)
 
 if (is.null(getGeneric("childName")))
   setGeneric("childName", function(object)
[…]

### 2006-02-02 inst/paper/paper2.tex
Made a few changes to the paper.
git-svn-id: file:///home/git/hedgehog.fhcrc.org/bioconductor/trunk/madman/Rpacks/MVCClass@16065 bc3139a8-67e5-0310-9ffc-ced21a209358

@@
 
-\newcommand{\Rfunction}[1]{{\textbf{#1}}}
+\newcommand{\Rfunction}[1]{{\textsf{#1}}}
 \newcommand{\Robject}[1]{{\texttt{#1}}}
@@
 \newcommand{\Rslot}[1]{\textsl{#1}}
-\newcommand{\Rclass}[1]{\textsf{#1}}
+\newcommand{\Rclass}[1]{\texttt{#1}}
 
@@
 \textbf{Dana-Farber Cancer Institute}\\ 
-\textbf{January 25, 2006} \\
+\textbf{February 2, 2006} \\
 \vspace{1in} 
@@
 for the four samples and one hundred rows for the one hundred genes.  In this
-model, the entities are the genes.  If two scatterplots are created as views
+model, the entities are the genes.  If two scatter plots are created as views
 of this model, where the first plot displays sample one versus sample two and
@@
 multivariate data because many different views can be created, such as
-scatterplots, spreadsheets, and heatmaps, and these views are linked because
+scatter plots, spreadsheets, and heatmaps, and these views are linked because
 they are based on the same underlying data.  Thus, users can decide which
@@
 data set 1 and one view of data set 2.  Data set 1 and data set 2 are linked.
-When a user initially interacts with the scatterplot view of data set 2, the
+When a user initially interacts with the scatter plot view of data set 2, the
 response to this interaction is to tell data set 2 to update itself (this is
 indicated on the figure as a circled 1).  Then in Step 2, data set 2 tells the
-scatterplot view that is should update itself in response to the change in the
+scatter plot view that is should update itself in response to the change in the
 data.  Next, in Step 3, because data set 2 has changed and because it is
@@
     \caption{ Linked views of linked Data Sets.  When a user interacts with
-      the y versus x scatterplot, the first step, indicated by a circled one,
+      the y versus x scatter plot, the first step, indicated by a circled one,
       is to tell data set 2 to update.  Then all views of data set 2 must be
@@
 
-The \Rclass{sPlotView} class represents a scatterplot view and it has the
+The \Rclass{sPlotView} class represents a scatter plot view and it has the
 extra slots, \Rslot{dfRows}, \Rslot{xvar}, and \Rslot{yvar}, where
@@
 and it is in the callback function that the corresponding method is called.
-For example, when the button click event occurs on a scatterplot view, the
-callback function calls the scatterplot's \Rfunction{clickEvent} method.
+For example, when the button click event occurs on a scatter plot view, the
+callback function calls the scatter plot's \Rfunction{clickEvent} method.
 Using the information that is passed to the \Rfunction{clickEvent} method in
@@
 (\Rclass{dfModel}), then the views that the user could create are a
-scatterplot (\Rclass{sPlotView}) and a spreadsheet (\Rclass{spreadView}),
+scatter plot (\Rclass{sPlotView}) and a spreadsheet (\Rclass{spreadView}),
 but a graph plot (\Rclass{graphView}) does not make sense.  Thus, when the
@@
 Display 
-menu has menu items to create a scatterplot and a spreadsheet.  Also, if
+menu has menu items to create a scatter plot and a spreadsheet.  Also, if
 the ``active'' \Robject{MVC} is of class \Rclass{dfModel}, then the command
-line functions only let the user create a scatterplot and a spreadsheet view
+line functions only let the user create a scatter plot and a spreadsheet view
 of the active model.
@@
 \ref{Fig:MPwithin}.  In this example, the response to a user performing a
-button click event over a point in the scatterplot is to color the point red.
+button click event over a point in the scatter plot is to color the point red.
 Setting the response to an event was discussed in Section
 \ref{Ssec:OneEvent}.  Here the user first clicks on a point in the
-scatterplot, which is shown as Step 1 (the circled 1 on the figure).  In the
+scatter plot, which is shown as Step 1 (the circled 1 on the figure).  In the
 figure, the point that has been clicked has an arrow next to it.  Then, in
@@
       indicated by a circled one, occurs when a user interacts with the y
-      versus x scatterplot.  Then an update data message is sent to the model
+      versus x scatter plot.  Then an update data message is sent to the model
       in the second step, which alters the model.  Next in the third step all
@@
     \caption{ Message Passing Between and Within \Robject{MVCs}.  When a user
-      interacts with the y versus x scatterplot, the first step, indicated by
+      interacts with the y versus x scatter plot, the first step, indicated by
       a circled one, is to send an update data message to model 2.  After
@@
 \Robject{MVC} using messages.  Suppose for this example that when the user
-clicks a mouse button over the scatterplot, the response to this event is to
+clicks a mouse button over the scatter plot, the response to this event is to
 color the point under the cursor 
 red.  In this picture, the message passing starts when a user clicks a point,
-that is indicated by the arrow, on the scatterplot.  An update data message is
+that is indicated by the arrow, on the scatter plot.  An update data message is
 sent to Model 2 in response to this user interaction.  As soon as Model 2 is
 updated, it sends an update view message to all its views so that the views
-are updated to reflect the changed data.  Now the scatterplot point that was
+are updated to reflect the changed data.  Now the scatter plot point that was
 initially clicked is red.  Then a send parent message is sent because MVC

### 2006-02-16 inst/doc/MVCClass.Rnw
Added a slot to the graphView class and then had to update all the documentation.
git-svn-id: file:///home/git/hedgehog.fhcrc.org/bioconductor/trunk/madman/Rpacks/MVCClass@16287 bc3139a8-67e5-0310-9ffc-ced21a209358

@@
 model that is shown as the y variable in the plot.  The \Robject{graphView}
-class represents a graph plot view and it contains no extra slots besides
-those in the \Robject{plotView} class.  The \Robject{heatmapView} represents a
-heatmap view of the data and it contains the slot, ordering, which is a list
-that is returned from the \Rfunction{heatmap} function to give information
-about the dendrogram ordering.  Note that some of these views are only
-applicable for certain model types. For example, the \Robject{graphView} class
-will only make sense for a view of a graph object, which would be stored in the
-\Robject{graphModel} class.
+class represents a graph plot view and it has the extra slot, grLayout, which
+holds a \Robject{Ragraph} object that represents the layout of the graph plot.
+The \Robject{heatmapView} represents a heatmap view of the data and it contains
+the slot, ordering, which is a list that is returned from the
+\Rfunction{heatmap} function to give information about the dendrogram
+ordering.  Note that some of these views are only applicable for certain model
+types.  For example, the \Robject{graphView} class will only make sense for a
+view of a graph object, which would be stored in the \Robject{graphModel}
+class.
 

### 2006-02-16 inst/paper/paper2.tex
Added a slot to the graphView class and then had to update all the documentation.
git-svn-id: file:///home/git/hedgehog.fhcrc.org/bioconductor/trunk/madman/Rpacks/MVCClass@16287 bc3139a8-67e5-0310-9ffc-ced21a209358

@@
 that is shown as the y variable in the plot.  The \Rclass{graphView} class
-represents a graph plot view and it contains no extra slots besides those in
-the \Rclass{plotView} class.  The \Rclass{heatmapView} represents a heatmap
+represents a graph plot view and it contains the slot, \Rslot{grLayout}, which
+holds a \Robject{Ragraph} object that represents the layout of the graph plot.
+The \Rclass{heatmapView} represents a heatmap
 view of the data and it contains the slot, \Rslot{ordering}, which is a list

### 2006-02-16 R/MVC.R

@@ -129,7 +129,7 @@ setClass("spreadView", representation(clist="GtkCList"),
          contains="genView")
 
 # 7/28/05 put the graphLayout info in the graphModel object
-setClass("graphView", contains="plotView")
+setClass("graphView", representation(grLayout="Ragraph"), contains="plotView")
 
 # 9/1/05 not sure if I need to store anything else about a heatmap
 # decided to store the list of row and column reorderings returned from the
@@ -204,6 +204,12 @@ if (is.null(getGeneric("ordering")))
             standardGeneric("ordering"))
 setMethod("ordering", "heatmapView", function(object)
          object@ordering)
+
+if (is.null(getGeneric("grLayout")))
+  setGeneric("grLayout", function(object)
+            standardGeneric("grLayout"))
+setMethod("grLayout", "graphView", function(object)
+         object@grLayout)
 
 #####
 # setting the slots
@@ -314,6 +320,16 @@ if (is.null(getGeneric("ordering<-")))
 setReplaceMethod("ordering", "heatmapView", function(object, value)
          {
            object@ordering<-value
+           object
+         }
+)
+
+if (is.null(getGeneric("grLayout<-")))
+  setGeneric("grLayout<-", function(object, value)
+            standardGeneric("grLayout<-"))
+setReplaceMethod("grLayout", "graphView", function(object, value)
+         {
+           object@grLayout<-value
            object
          }
 )

### 2006-04-20 inst/doc/MVCClass.Rnw
Changed two slots in the sPlotView class.  They used to be colx and coly and now they are xvar and yvar.  Now the code and documentation should match.
git-svn-id: file:///home/git/hedgehog.fhcrc.org/bioconductor/trunk/madman/Rpacks/MVCClass@17400 bc3139a8-67e5-0310-9ffc-ced21a209358

@@
 \author{Elizabeth Whalen}
-\date{November 11, 2005}
+\date{April 20, 2006}
 
@@
 The \Robject{sPlotView} class represents a scatterplot view and it has the
-extra slots, dfRows, colx, and coly, where dfRows are the row names from the
-model that are shown in the plot, colx is the column name from the model that
-is shown as the x variable in the plot, and coly is the column name from the
+extra slots, dfRows, xvar, and yvar, where dfRows are the row names from the
+model that are shown in the plot, xvar is the variable name (the column name
+if the model is a data frame) from the model that
+is shown as the x variable in the plot, and yvar is the variable name (the
+column name if the model is a data frame) from the
 model that is shown as the y variable in the plot.  The \Robject{graphView}

### 2006-04-20 R/MVC.R

@@ -122,8 +122,8 @@ setClass("plotView", representation(plotDevice="numeric", plotPar="list",
          drArea="GtkDrawingArea"), contains=c("genView", "VIRTUAL"))
 
 # store the row names and column names rather than their indices
-setClass("sPlotView", representation(dfRows="character", colx="character",
-         coly="character"), contains="plotView")
+setClass("sPlotView", representation(dfRows="character", xvar="character",
+         yvar="character"), contains="plotView")
 
 setClass("spreadView", representation(clist="GtkCList"), 
          contains="genView")
@@ -181,17 +181,17 @@ if (is.null(getGeneric("dfRows")))
 setMethod("dfRows", "sPlotView", function(object)
          object@dfRows)
 
-if (is.null(getGeneric("colx")))
-  setGeneric("colx", function(object)
-            standardGeneric("colx"))
-setMethod("colx", "sPlotView", function(object)
-         object@colx)
+if (is.null(getGeneric("xvar")))
+  setGeneric("xvar", function(object)
+            standardGeneric("xvar"))
+setMethod("xvar", "sPlotView", function(object)
+         object@xvar)
 
-if (is.null(getGeneric("coly")))
-  setGeneric("coly", function(object)
-            standardGeneric("coly"))
-setMethod("coly", "sPlotView", function(object)
-         object@coly)
+if (is.null(getGeneric("yvar")))
+  setGeneric("yvar", function(object)
+            standardGeneric("yvar"))
+setMethod("yvar", "sPlotView", function(object)
+         object@yvar)
 
 if (is.null(getGeneric("clist")))
   setGeneric("clist", function(object)
@@ -284,22 +284,22 @@ setReplaceMethod("dfRows","sPlotView",function(object, value)
          }
 )
 
-if (is.null(getGeneric("colx<-")))
-  setGeneric("colx<-",function(object,value)
-            standardGeneric("colx<-"))
-setReplaceMethod("colx","sPlotView",function(object,value)
+if (is.null(getGeneric("xvar<-")))
+  setGeneric("xvar<-",function(object, value)
+            standardGeneric("xvar<-"))
+setReplaceMethod("xvar","sPlotView",function(object, value)
          {
-           object@colx<-value
+           object@xvar<-value
            object
          }
 )
 
-if (is.null(getGeneric("coly<-")))
-  setGeneric("coly<-",function(object, value)
-            standardGeneric("coly<-"))
[…]

### 2006-04-26 inst/doc/MVCClass.Rnw
Added a new method: identifyView.
git-svn-id: file:///home/git/hedgehog.fhcrc.org/bioconductor/trunk/madman/Rpacks/MVCClass@17721 bc3139a8-67e5-0310-9ffc-ced21a209358

@@
 performed on views.  These generic functions are \Rfunction{motionEvent}, 
-\Rfunction{clickEvent}, \Rfunction{updateView}, and \Rfunction{redrawView}.  
-Methods for these generic 
+\Rfunction{clickEvent}, \Rfunction{updateView}, \Rfunction{redrawView} and
+\Rfunction{identifyView}.  Methods for these generic 
 functions will be defined in packages that use this package (in 
@@
 a click event over a view, the \Rfunction{updateView} method will be used to 
-update only a portion of the view, and the \Rfunction{redrawView} method will 
-be used to completely redraw the view.
+update only a portion of the view, the \Rfunction{redrawView} method will 
+be used to completely redraw the view, and the \Rfunction{identifyView} method
+will be used to identify an object (such as a point, a node, etc.) on a view.
 

### 2006-04-26 R/MVC.R

@@ -346,6 +346,14 @@ if (is.null(getGeneric("clickEvent")))
             standardGeneric("clickEvent"))
 
 #########
+# added 4/23/06
+# to identify an object on a view (starting from user coordinates)
+#########
+if (is.null(getGeneric("identifyView")))
+  setGeneric("identifyView", function(object, location, ...)
+            standardGeneric("identifyView"))
+
+#########
 # added 6/5/05
 # make method to update a view depending on the view object
 # vData is the view data needed to update the view
@@ -353,8 +361,6 @@ if (is.null(getGeneric("clickEvent")))
 if (is.null(getGeneric("updateView")))
   setGeneric("updateView", function(object, vData)
             standardGeneric("updateView"))
-
-
 
 ########
 # added 6/5/05

### 2006-05-05 inst/doc/MVCClass.Rnw
Added some new classes and changed a few existing classes.  Plus had to change all the documentation.
git-svn-id: file:///home/git/hedgehog.fhcrc.org/bioconductor/trunk/madman/Rpacks/MVCClass@17877 bc3139a8-67e5-0310-9ffc-ced21a209358

@@
 \author{Elizabeth Whalen}
-\date{April 20, 2006}
+\date{May 5, 2006}
 
@@
 which is stored in the modelName slot of the model object (see Section 
-\ref{Sec:Model}).  The figure below 
-shows the class definition for \Robject{MVC}.
+\ref{Sec:Model}).  Figure \ref{Fig:MVCClass} 
+shows the inheritance structure for the \Robject{MVC} classes.
 
@@
   \begin{center}
-    \scalebox{0.3}{\includegraphics{MVCClass.jpg}}
-    \caption{ The MVC Class. }
+    \scalebox{0.2}{\includegraphics{fullMVCClass.jpg}}
+    \caption{ Inheritance for MVC Objects. }
     \label{Fig:MVCClass}
@@
 
-The \Robject{MVC} class has five slots which are model, viewList, controller, 
-parentMVC, and childMVCList.  The
-model slot stores the model object, the viewList slot is a list of the view
-objects for this model, the controller slot is an environment that stores
-information for this MVC, the parentMVC slot is the name of the parent model
-(MVC), and the childMVCList slot is a list of the children models
-(MVCs).  Note that model and MVC objects use the same name so the
-parentMVC slot refers to both the name of the model and the name of the
-MVC.  Similarly, for the childMVCList slot, the names in the list
-refer to both the name of the model and the name of the MVC.
+The \Robject{MVC} class is a virtual class that binds the other \Robject{MVC}
+classes together.  Every \Robject{MVC} object will need the three slots:
+model, viewList and controller.  The model slot stores the model object, the
+viewList slot is a list of the view objects for this model, and the controller
+slot is an environment that stores information for this MVC.  Inheriting from
+the \Robject{MVC} class is the \Robject{singleModelMVC} class that has no new
+slots.  The \Robject{singleModelMVC} class represents a \Robject{MVC} object
+that is not linked to any other \Robject{MVC} objects.
 
-Thus, the \Robject{MVC} class not only binds the model, view, and controller
-objects together, but MVC objects can be related to each through a 
-parent-child (tree) hierarchy because of the slots, parentMVC and 
-childMVCList. 
-These two slots also show that a MVC object can have at most one parent MVC and
-can have as many children MVCs as it wants.
+Inheriting from the \Robject{singleModelMVC} class is the
+\Robject{linkedModelMVC} that contains the new slots: parentMVC and
+childMVCList.  The parentMVC slot is the name of the parent model (MVC) and
+the childMVCList slot is a list of the children models (MVCs).  Note that
+model and MVC objects use the same name so the parentMVC slot refers to both
+the name of the model and the name of the MVC.  Similarly, for the
+childMVCList slot, the names in the list refer to both the name of the model
+and the name of the MVC.
+
+Thus, the \Robject{singleModelMVC} class binds the model, view, and controller
+objects together and the \Robject{linkedModelMVC} class allows the MVC objects
+to be related to each through a parent-child (tree) hierarchy because of the
+slots, parentMVC and childMVCList.  These two slots also show that a MVC
+object can have at most one parent MVC and can have as many children MVCs as
+it wants.
 
@@
 
-The above figure shows the inheritance structure for message objects that can 
+The figure shows the inheritance structure for message objects that can 
 be passed within one MVC object and a later figure in this section will show 
@@
 Currently, there are two types of messages for messaging within a model, view,
-controller object: an add and an update message.  Both the add message,
+controller object: an add and an update message, and both of these messages
+are performing some type of modification.  Thus, both the add message,
 \Robject{gAddMessage}, and the update message, \Robject{gUpdateMessage},
-contain three slots, dataName, mData, and type.  dataName is the character
+inherit from the virtual class, \Robject{gModifyMessage}, which has 
+three slots, dataName, mData, and type.  dataName is the character
 string that gives the name of the model, mData is a list of data needed to
@@
 model (and similarly with updating views versus models), it makes sense to
-have two separate message classes for these operations.  Thus, when messages
+have two separate message classes for these elements.  Thus, when messages
 are sent between components, they are from one of the following classes:
@@
 passed to other MVC objects, this message also has one new slot, from, in 
-addition to
-the slots it inherits from \Robject{gUpdateMessage}.  The from slot is the
-name of the model that the update data message came from.  This slot is
-necessary because the update data message can come from one of two sources:
-the message can occur because of user interaction with a view
+addition to the slots it inherits from \Robject{gUpdateMessage}.  The from
+slot is the name of the model that the update data message came from.  This
+slot is necessary because the update data message can come from one of two
+sources: the message can occur because of user interaction with a view
 of this model or the message may occur because a different model that is
@@
   \begin{center}
-    \scalebox{0.3}{\includegraphics{EventClass.jpg}}
+    \scalebox{0.2}{\includegraphics{EventClass.jpg}}
     \caption{ The gEventFun Class. }
@@
 As shown in Fig \ref{Fig:EventFun}, the slots in the \Robject{gEventFun} class
-are callFun, shortName, and preprocessFun.  The callFun slot is the name of
-the callback function, the shortName slot is a short description of the
-function, and the preprocessFun slot is a character vector of all
+are callFun, shortName, and preprocessFun.  The callFun slot is the callback
+function, the shortName slot is a short description of the 
+function, and the preprocessFun slot is a list of all
 preprocessing functions that must be called before the callback function.
@@
 variables that the callback function will need.  Thus, not all callback
-functions will need preprocessing functions and if no preprocessing functions
-are needed, then this slot will be NULL.  If the callback function sets the
+functions need preprocessing functions and if no preprocessing functions
+are needed, then this slot is set to NULL.  If the callback function sets the
 color of a node in a graph, then a preprocessing function is needed to
 determine what color this will be.  So in this example, the preprocessing
-function will open a color browser to allow the user to choose what the new
-color will be.
+function opens a color browser to allow the user to choose what the new
+color is.
 

### 2006-05-05 R/MVC.R

@@ -380,8 +380,12 @@ if (is.null(getGeneric("redrawView")))
 #############
 
 setClass("MVC", representation(model="gModel", viewList="list", 
-                  controller="environment", parentMVC="character", 
-                  childMVCList="list"))
+                  controller="environment"), contains=("VIRTUAL"))
+
+setClass("singleModelMVC", contains="MVC")
+
+setClass("linkedModelMVC", representation(parentMVC="character",
+                           childMVCList="list"), contains="singleModelMVC")
 
 ########
 # get the slots
@@ -408,13 +412,13 @@ setMethod("controller", "MVC", function(object)
 if (is.null(getGeneric("parentMVC")))
   setGeneric("parentMVC", function(object)
             standardGeneric("parentMVC"))
-setMethod("parentMVC", "MVC", function(object)
+setMethod("parentMVC", "linkedModelMVC", function(object)
          object@parentMVC)
 
 if (is.null(getGeneric("childMVCList")))
   setGeneric("childMVCList", function(object)
             standardGeneric("childMVCList"))
-setMethod("childMVCList", "MVC", function(object)
+setMethod("childMVCList", "linkedModelMVC", function(object)
          object@childMVCList)
 
 ########
@@ -454,7 +458,7 @@ setReplaceMethod("controller", "MVC", function(object,value)
 if (is.null(getGeneric("parentMVC<-")))
   setGeneric("parentMVC<-", function(object,value)
             standardGeneric("parentMVC<-"))
-setReplaceMethod("parentMVC", "MVC", function(object,value)
+setReplaceMethod("parentMVC", "linkedModelMVC", function(object,value)
          {
            object@parentMVC<-value
            object
@@ -464,7 +468,7 @@ setReplaceMethod("parentMVC", "MVC", function(object,value)
 if (is.null(getGeneric("childMVCList<-")))
   setGeneric("childMVCList<-", function(object,value)
             standardGeneric("childMVCList<-"))
-setReplaceMethod("childMVCList", "MVC", function(object,value)
+setReplaceMethod("childMVCList", "linkedModelMVC", function(object,value)
          {
            object@childMVCList<-value
            object

### 2006-05-05 R/eventFun.R

@@ -14,11 +14,11 @@
 #  preprocessing functions must not take any parameters
 ##############
 # there may be no preprocessing functions
-setClassUnion("characterNULL", members=c("character", "NULL"))
+setClassUnion("listNULL", members=c("list", "NULL"))
 
-setClass("gEventFun", representation(callFun="character", 
+setClass("gEventFun", representation(callFun="function", 
                                   shortName="character", 
-                                  preprocessFun="characterNULL"))
+                                  preprocessFun="listNULL"))
 
 if (is.null(getGeneric("callFun")))
   setGeneric("callFun", function(object)

### 2006-05-05 R/messages.R

@@ -4,8 +4,10 @@
 # a virtual class for message
 setClass("gMessage")
 
-setClass("gUpdateMessage", representation(type="character", mData="list",
+setClass("gModifyMessage", representation(type="character", mData="list",
           dataName="character"), contains=c("gMessage", "VIRTUAL"))
+
+setClass("gUpdateMessage", contains=c("gModifyMessage", "VIRTUAL"))
 
 setClass("gUpdateViewMessage", contains="gUpdateMessage")
 
@@ -15,8 +17,7 @@ setClass("gUpdateViewMessage", contains="gUpdateMessage")
 setClass("gUpdateDataMessage", representation(from="character"), 
           contains="gUpdateMessage")
 
-setClass("gAddMessage", representation(dataName="character", mData="list",
-         type="character"), contains=c("gMessage", "VIRTUAL"))
+setClass("gAddMessage", contains=c("gModifyMessage", "VIRTUAL"))
 setClass("gAddViewMessage", contains="gAddMessage")
 setClass("gAddDataMessage", contains="gAddMessage")
 
@@ -67,25 +68,19 @@ setMethod("childName", "gSendChildMessage", function(object)
 if (is.null(getGeneric("type")))
   setGeneric("type", function(object)
             standardGeneric("type"))
-setMethod("type", "gUpdateMessage", function(object)
-         object@type)
-setMethod("type", "gAddMessage", function(object)
+setMethod("type", "gModifyMessage", function(object)
          object@type)
 
 if (is.null(getGeneric("mData")))
   setGeneric("mData", function(object)
             standardGeneric("mData"))
-setMethod("mData", "gUpdateMessage", function(object)
-         object@mData)
-setMethod("mData", "gAddMessage", function(object)
+setMethod("mData", "gModifyMessage", function(object)
          object@mData)
 
 if (is.null(getGeneric("dataName")))
   setGeneric("dataName", function(object)
             standardGeneric("dataName"))
-setMethod("dataName", "gAddMessage", function(object)
-         object@dataName)
-setMethod("dataName", "gUpdateMessage", function(object)
+setMethod("dataName", "gModifyMessage", function(object)
          object@dataName)
 
 #####
@@ -134,13 +129,7 @@ setReplaceMethod("childName", "gSendChildMessage", function(object, value)
 if (is.null(getGeneric("type<-")))
   setGeneric("type<-", function(object, value)
             standardGeneric("type<-"))
-setReplaceMethod("type", "gUpdateMessage", function(object, value)
-         {
-           object@type<-value
[…]

### 2006-05-06 inst/paper/paper2.tex
Changed paper 2 to match the new classes.  Also updated some of the man pages.
git-svn-id: file:///home/git/hedgehog.fhcrc.org/bioconductor/trunk/madman/Rpacks/MVCClass@17887 bc3139a8-67e5-0310-9ffc-ced21a209358

@@
 \bibliographystyle{amsplain}
-\bibliography{paper2}
 
@@
 \textbf{Dana-Farber Cancer Institute}\\ 
-\textbf{February 2, 2006} \\
+\textbf{May 6, 2006} \\
 \vspace{1in} 
@@
 class are \Rslot{callFun}, \Rslot{shortName}, and \Rslot{preprocessFun}.  The
-\Rslot{callFun} slot is the name of the callback function, the
-\Rslot{shortName} slot is a short description of the function, and the
-\Rslot{preprocessFun} slot is a character vector of all preprocessing
+\Rslot{callFun} slot is the callback function, the
+\Rslot{shortName} slot is a short description of the function that
+identifies the callback function, and the
+\Rslot{preprocessFun} slot is a list of all preprocessing
 functions that must be called before the callback function.
@@
 
-As shown in Figure \ref{Fig:MVCClass}, the \Rclass{MVC} class has five slots
-which are, \Rslot{model}, \Rslot{viewList}, \Rslot{controller},
-\Rslot{parentMVC}, and \Rslot{childMVCList}.  The \Rslot{model} slot stores
-the model object, the \Rslot{viewList} slot is a list of the
-view objects for this model, the \Rslot{controller} slot is an
-environment that stores information for this \Robject{MVC}, the
-\Rslot{parentMVC} slot is the name of the parent model
-(\Robject{MVC}), and the \Rslot{childMVCList} slot is a list of the children
-models (\Robject{MVCs}).  Note that model and
+As shown in Figure \ref{Fig:MVCClass}, there is a virtual class, \Rclass{MVC},
+that binds the MVC inheritance structure together.  The virtual class,
+\Rclass{MVC}, has three slots, which are \Rslot{model}, \Rslot{viewList}, and
+\Rslot{controller}.  The \Rslot{model} slot stores the model object, the
+\Rslot{viewList} slot is a list of the view objects for this model, and the
+\Rslot{controller} slot is an environment that stores information for this
+\Robject{MVC}.  These three pieces of information are needed to bind the
+model, view, and controller into one object.  To create instances of a single
+MVC, the \Rclass{singleModelMVC} class, which inherits from the virtual
+\Rclass{MVC} class, can be used.
+
+Inheriting from the \Rclass{singleModelMVC} class is the
+\Rclass{linkedModelMVC} class, which represents a \Robject{MVC} object that
+can be linked to other \Robject{MVC} objects.  The \Rclass{linkedModelMVC}
+class has two additional slots: \Rslot{parentMVC}, which is the name of the
+parent model (\Robject{MVC}), and \Rslot{childMVCList}, which is a list of the
+child models (\Robject{MVCs}).  Note that model and
 \Robject{MVC} objects use the same name so the \Rslot{parentMVC} slot refers
@@
 Similarly, for the \Rslot{childMVCList} slot, the names in the list refer to
-both the name of the model and the name of the \Robject{MVC}.
+both the name of the model and the name of the \Robject{MVC}.  In this paper,
+we are assuming that models are linked so all \Robject{MVC} objects are of
+class \Rclass{linkedModelMVC}.
 
@@
   \begin{center}
-    \scalebox{0.6}{\includegraphics{MVCClass.jpg}}
+    \scalebox{0.6}{\includegraphics{fullMVCClass.jpg}}
     \caption{ The MVC Class. }
@@
 
-Thus, the \Rclass{MVC} class not only binds the model,
+Thus, the \Rclass{linkedModelMVC} class not only binds the model,
 view, and controller objects together, but it also shows
-that the \Robject{MVC} objects are related to each other through a
+that the \Robject{linkedModelMVC} objects are related to each other through a
 parent-child (tree) hierarchy, as shown in Figure \ref{Fig:Hier}.  In this
-picture, `model1' is the name of the original \Robject{MVC} object (model) and
-from it, two \Robject{MVC} objects were created, `model2.1' and
-`model2.2'. Then from `model2.1', three \Robject{MVC} objects were created,
-`model3.1', `model3.2', and `model3.3'.  As an example, model1 could be gene
+picture, `model1' is the name of the original \Robject{MVC} object
+(model) and from it, two \Robject{MVC} objects were created,
+`model2.1' and `model2.2'. Then from `model2.1', three
+\Robject{MVC} objects were created, `model3.1', `model3.2', and
+`model3.3'.  As an example, model1 could be gene 
 expression data, model2.1 could be gene expression data that is a subset of
@@
 \Robject{MVC} object) could either be loaded or could be created from an
-already loaded \Robject{MVC} object.  As mentioned in Section \ref{Ssec:MMVC},
-the focus in this paper is on experimental data that are linked to meta-data,
-as the linked data sets that we are most interested in visualizing.  In this
-situation, the experimental data (and its \Robject{MVC} object) can be thought
-of as determining what meta-data are studied.  Thus, there is a hierarchical
-relationship between these data sets where one \Robject{MVC} object creates
-another \Robject{MVC} object.  This focus led to the decision that a
-\Robject{MVC} object can have at most one parent \Robject{MVC}, from which it
-was derived, but it can have many child \Robject{MVCs}.  An easy example of a
-child model derived from a parent model is to start with a
-data frame as the parent model and then take a subset of this data
-frame as the child model.
+already loaded \Robject{MVC} object.  As mentioned in Section
+\ref{Ssec:MMVC}, the focus in this paper is on experimental data that are
+linked to meta-data, as the linked data sets that we are most interested in
+visualizing.  In this situation, the experimental data (and its
+\Robject{MVC} object) can be thought of as determining what
+meta-data are studied.  Thus, there is a hierarchical relationship between
+these data sets where one \Robject{MVC} object creates another
+\Robject{MVC} object.  This focus led to the decision that a
+\Robject{MVC} object can have at most one parent
+\Robject{MVC}, from which it was derived, but it can have many
+child \Robject{MVCs}.  An easy example of a child model derived
+from a parent model is to start with a data frame as the parent model and then
+take a subset of this data frame as the child model.
 
@@
   \begin{center}
-    \scalebox{0.6}{\includegraphics{MessageClass.jpg}}
+    \scalebox{0.6}{\includegraphics{newMessageClass.jpg}}
     \caption{ Inheritance for Message Objects. }
@@
 Currently, there are two types of messages for messaging within a
-model, view, controller object: an add and an update message.  Both the add
-message, \Rclass{gAddMessage}, and the update message, \Rclass{gUpdateMessage},
-contain three slots, \Rslot{dataName}, \Rslot{mData}, and \Rslot{type}.
-\Rslot{dataName} is the character string that gives the name of the model,
+model, view, controller object: an add and an update message.  These two types
+of messages are both performing some type of modification and thus, both the
+add message, \Rclass{gAddMessage}, and the update message,
+\Rclass{gUpdateMessage}, inherit from the virtual class,
+\Rclass{gModifyMessage}, which contains three slots, \Rslot{dataName},
+\Rslot{mData}, and \Rslot{type}.  \Rslot{dataName} is the character string
+that gives the name of the model, 
 \Rslot{mData} is a list of data needed to perform the addition or alteration
@@
 model (and similarly with updating views versus models), it makes sense to
-have two separate message classes for these operations.  Messages
+have two separate message classes for these elements.  Messages
 sent between components are from one of the following classes:
@@
 
+\bibliography{paper2}
+
 \end{document}

### 2006-06-05 R/MVC.R
Moved some of the class definitions into the BioMVCClass package so that the MVCClass package no longer depends on Biobase, graph, or Rgraphviz.
git-svn-id: file:///home/git/hedgehog.fhcrc.org/bioconductor/trunk/madman/Rpacks/MVCClass@18314 bc3139a8-67e5-0310-9ffc-ced21a209358

@@ -16,20 +16,13 @@ setOldClass("GtkCList")
 setClass("gModel", representation(modelData="ANY", linkData="list", 
          virtualData="ANY", modelName="character"), contains=("VIRTUAL"))
 
-# couldn't get RagraphNULL to pass the installation - didn't recognize
-# Ragraph class
-#setClassUnion("RagraphNULL", members=c("Ragraph", "NULL"))
+## for a model that has graph data
+#setClass("graphModel", representation(modelData="graph"), contains="gModel")
 
-# for a model that has graph data
-# initially the virtualData slot for graphModel will be NULL (until a 
-# view is created)
-#setClass("graphModel", representation(modelData="graph", 
-#                       virtualData="RagraphNULL"), contains="gModel")
-setClass("graphModel", representation(modelData="graph"), contains="gModel")
+## for a model that has expression data - 
+## the data list should include an exprSet object and maybe a vector of LL ids
+#setClass("exprModel", representation(modelData="exprSet"), contains="gModel")
 
-# for a model that has expression data - 
-# the data list should include an exprSet object and maybe a vector of LL ids
-setClass("exprModel", representation(modelData="exprSet"), contains="gModel")
 # for a model that has data frame data
 setClass("dfModel", representation(modelData="data.frame", 
                     virtualData="data.frame"), contains="gModel")
@@ -128,13 +121,13 @@ setClass("sPlotView", representation(dfRows="character", xvar="character",
 setClass("spreadView", representation(clist="GtkCList"), 
          contains="genView")
 
-# 7/28/05 put the graphLayout info in the graphModel object
-setClass("graphView", representation(grLayout="Ragraph"), contains="plotView")
-
-# 9/1/05 not sure if I need to store anything else about a heatmap
-# decided to store the list of row and column reorderings returned from the
-# heatmap function (just in case I need it later)
-setClass("heatmapView", representation(ordering="list"), contains="plotView")
+## 7/28/05 put the graphLayout info in the graphModel object
+#setClass("graphView", representation(grLayout="Ragraph"), contains="plotView")
+#
+## 9/1/05 not sure if I need to store anything else about a heatmap
+## decided to store the list of row and column reorderings returned from the
+## heatmap function (just in case I need it later)
+#setClass("heatmapView", representation(ordering="list"), contains="plotView")
 
 #####
[…]

### 2006-06-05 R/zzz.R

@@ -1,6 +1,2 @@
-setOldClass("GtkWindow")
-setOldClass("GtkDrawingArea")
-setOldClass("GtkCList")
-
-.onLoad <- function(lib, pkg, where) {
+.onLoad <- function(libname, pkgname) {
 }

### 2006-07-03 inst/paper/MVC.tex
added and committed
git-svn-id: file:///home/git/hedgehog.fhcrc.org/bioconductor/trunk/madman/Rpacks/MVCClass@18682 bc3139a8-67e5-0310-9ffc-ced21a209358

\documentclass{article}[11pt]
\setlength{\topmargin}{-2cm}
\setlength{\textheight}{24cm}
\setlength{\oddsidemargin}{10mm}
\setlength{\evensidemargin}{10mm}
\setlength{\textwidth}{15cm}
\setlength{\parskip}{3mm}
\setlength{\parindent}{0mm}

\usepackage{epsfig}
\usepackage{comment}
\usepackage{graphics}
\usepackage{graphicx}
\usepackage{hyperref}
\usepackage[authoryear,round]{natbib}

\newcommand{\Rfunction}[1]{{\textsf{#1}}}
\newcommand{\Robject}[1]{{\texttt{#1}}}
\newcommand{\Rpackage}[1]{{\textit{#1}}}
\newcommand{\Rslot}[1]{\textsl{#1}}
\newcommand{\Rclass}[1]{\texttt{#1}}

\begin{document}

\bibliographystyle{amsplain}

\author{Elizabeth Whalen\\Harvard School of Public Health \\ 
        Robert Gentleman\\Fred Hutchinson Cancer Research Center}
\title{Generalizing the Model View Controller Paradigm}

\maketitle

\abstract{ In this paper we consider a generalization of the model
  view controller (MVC) paradigm to allow for more flexible handling
  of submodels. Instead of using a single MVC we develop a tree
  structure of related MVCs, where each MVC has a single parent, but a
  parent can spawn many different child MVCs. Communication between
  parent and child MVCs allows for very flexible linking of the
  MVCs. We provide an implementation of our work in the R package
\Rpackage{MVC} available at \url{http://www.bioconductor.org}. }

\section{Introduction}
\label{Sec:Intro}

%\subsection{Background}
%\label{Ssec:Backg}

There are many ways to visualize multivariate data when more than
two or three dimensions are needed to view the data. Many authors
(citations) have suggested that linked interactive views of the data
provide a valuable and important set of tools for comprehending such
data. The model view controller (MVC) paradigm is now a standard
mechanism for implementing such systems. The paradigm has proven very
effective and allows for clear separation of different components.
In this paper we consider an extension of the MVC paradigm to allow
for, what we believe to be, easy extension to linked data sets.

The sorts of problems we are interested in, are those where a user
might begin with a large data set, identify some interesting subset
and want to create interactive views of that subset. In some cases the
natural data representation of the subset is in a form other than that
of the initial data set. This leads to complicatins in the usual MVC
paradigm, where the data (i.e. the model) are often presumed to be in
a standard format. We found that adding new views under such
conditions could be quite problematic and often leads to very complex
code. Hence we considered an approach that created new instances of
the MVC that were linked to the parent that spawned them.

In particular, our implementation consists of a hierarchy of MVCs,
stored as a tree. Each parent can have multiple children, but the
children can have a single parent. Communication between parent and
child is bidirectional, and a variety of different filters can be used
to control flow. Thus, we can devise a number of purpose built MVC
classes, each with a model appropriate to the specific data of
interest, that are linked in a natural way. Interacting with the view
in a child MVC, first sends a message to the child MVC, which will
update all of its views, but the message will also propagate to the
parent MVC (and further up to the root). The parent MVC can choose to
react, or not, to messages from its children -- all control being in
the hands of the user, should they want it.

\subsection*{A motivating example}
\label{sec:micro}

Our development of this model was driven primarily by an interest in
analyzing high throughput biological data. Such data tend to be large
and very complex, and hence ideally suited to interactive
visualization. In order to provide the reader with some idea of the
sorts of interactions we would like to support we here provide a short
description of a type of experiment and some of the likely
interactions that would be of interest to a data analyst.

A gene expression experiment provides data on the expression of mRNA
abundance on thousands to tens of thousands of genes for some number
of samples. Usually there are many fewer samples than genes. There is
much that is known about the genes, such as what chromosome they are
located on, what sorts of activities they are engaged in and in many
cases we want to restrict attention to a specific subset of genes, or
to alter the view so that rather than being gene centric it becomes
centered on some biological aspect, such as chromosome location.

One particularly rich data source is the Gene Ontology (GO), \cite{GO}
which provides three controlled vocabularies that are represented as
directed acyclic graphs, with a root node. Terms typically have a
child-parent relationship, with the child terms being more specific
and the parent terms more general. The root node represents the most
general term. Genes are then annotated at specific nodes in these
graphs where the node is determined, by a domain expert, to describe a
property of the gene. A gene is then also associated with all parents,
since if the specific term applies so must all more general ones.

%% FIXME: perhaps we could have a heatmap and a few of the other views
%% here?
There is often substantial interest in relating gene expression data
to such a graph based model, or to chromosomal locations, or to
activity in known biological pathways. Data mappings for this latter
aspect can be obtained from different sources such as \cite{KEGG}.
Further, users typically want to begin with a heatmap view. Then from
that derive a GO graph view, and additionally a chromosomal location
view. In order to interact with the GO graph, the primary unit is the
node in that graph (and sometimes one expects with the edges), while
for the heatmap view the basic element is generally the gene.

\subsection*{Overview}

We found that constructing a single MVC that would encompass all
anticipated views was very cumbersome. It would also be inefficient
since a MVC model for the GO graph, or for a chromosome view would
also be of substantial interest, regardless of whether there was an
initial heatmap view or not. Thus, we looked for a solution that was
more based on component software constructed within an object oriented
programming environment.

We had several objectives, or design criteria. They were that the
software should be extensible in a high level language. We chose R as
the language for our implementation and wanted to allow users that
knew no other languages to be able to extend and enhance the
system. Therefore, we built our software on top the \Rpackage{RGtk} of
D. Temple Lang, since it provides such an interface. We also vastly
preferred Gtk over Tcl/Tk since its widgets are more sophisticated
which we found results in a much shorter development time.

We wanted to provide a system that was usable and extensible by
relatively naive end users and hence have provided a graphical user
interface (GUI) which they can use to both carry out an analysis, but
also supports a variety of extensions, such as changing the actions
performed in response to events. We also wanted to support a more
sophisticate developer who might want to take our basic infrastructure
and create an interactive system, say for the interactive analysis of
time series data, using our infrastructure. Hence, all commands are
available both through the GUI and through command line functions.

\section{Methods}

We first give our definitions of linked views, interactive views and
linked data sets are given. Our model is typically a statistical data
matrix which represents data on a number of variables, for each of a
number of samples.  By \textit{linked view} we mean that if a
component that represents data (such as a point on a plot or a row in
a spreadsheet) changes its appearance on one view, then the
corresponding component in all other views also changes appearance.
In our example, correspondence would be measured through the sample,
but there are other possibilities, as we shall explain later.  As an
example, suppose that the model consists of the gene expression levels
from say 50 samples and 10,000 genes.  The model can be represented as
a matrix of 50 columns and ten thousand rows.  We have two choices for
linking, either the entities can be the samples, or they can be the
genes. If the entities are the genes then when two scatter plots are
created as views, where on displays sample one versus sample two and
the second displays sample three versus sample four, then the points
on each plot that referred to the same gene are corresponding
components.  If instead we had chosen to link on samples, then we
could produce pairwise plots of gene expression values and would link
them through the sample IDs.

Linked views, which have a long history in statistics, are now
considered a standard feature of interactive data visualization
software \cite{GGobi}.  Linked views have been implemented in several
pieces of statistical software, such as XLisp-Stat \cite{Lisp} and
GGobi \cite{GGobiMan}.  Implementing linked views is based on the
model-view-controller (MVC) design \cite{DesignPatterns}, which is a
widely used and well understood paradigm.

We take a fairly general view of interactivity. Users can interact via
a pointing device, such as a mouse, or via key strokes, or any of many
other types of input devices.  Interactivity means that there is a
response to some user action.  It is a powerful paradigm and allows
users to quickly navigate and explore complex documents or complex
data sets.  As an example, suppose that a user has two views: one view
is a graph where each node represents a particular set of genes while
the other view is a heatmap of gene expression data.  If the user
clicks on a node then the response to this action could be to
highlight all genes in the heatmap view that correspond to the
selected node.

By \textit{inked data sets} we mean that the data have more than one
conceptual grouping and that there is a relationship between these
different conceptual groupings.  Linked data sets are 
familiar to users of relational databases.  Data in a relational
database are typically stored in many linked tables with rows in
different tables linked to each other through a unique identifier
called a key.  In this example, each table can be considered a data set
and they are linked through the keys.  But the idea is more general
and we shall discuss other situations where the linked data sets do
not easily correspond to tables, however, we must have a map that goes
from the entities in one data set to those in the other as well as the
inverse map, in order to propagate signals between the MVCs.

Consider Figure \ref{Fig:firstMP}, which provides a cartoon
description of the steps that occur when a user interacts with a
view. In the cartoon there are only two views, one of data set A and the
other of data set B, which is linked to data set A.  When a user
interacts with the scatter plot view of data set B, the response to
this interaction is for data set B to update itself (this is indicated
on the figure as a circled 1).  In Step 2, the scatter plot view is
told to update itself.  Next, in Step 3, because data set B is linked
to data set A, data set B sends a message to data set A.  In Step 4,
the information from data set B is converted into information that
data set A understands.  In Step 5, this information is received by
data set A and data set A updates itself.  Finally, in Step 6, data
set 1 tells its view to update itself.

% In the design presented in this paper, the components consist of
% \Robject{MVC} objects and communication within and between
% \Robject{MVC} objects is performed by message objects.  This solution
% uses the power and simplicity of the MVC design and expands it to the
% multiple data set problem.  Now we can build on the original MVC
% design, which has been well tested, instead of starting a new design
% from scratch.  The strength of this multiple linked MVC design is that
% it uses MVC components, which can be individually created and then
% linked.  Thus, these components are reusable.  These design decisions
% are discussed further in Sections \ref{Sec:OneMVC} and
% \ref{Sec:MultMVC}.

%\clearpage

%%FIXME: please update the figure so that it has A and B, not 1 and 2
%%for the data sets - it was too confusing
\begin{figure}[ht]
  \begin{center}
    \scalebox{0.55}{\includegraphics{firstMessagePassing.jpg}}
    \caption{ Linked views of linked Data Sets.  When a user interacts with
      the y versus x scatter plot, the first step, indicated by a circled one,
      is to tell data set B to update.  Then all views of data set B must be
      updated.  Next any data that are linked to data set B must be notified
      that data set B changed.  This information from data set B must be
      altered so that it can be understood by linked data sets, which is shown
      in the fourth step.  Finally, data set A is notified that it should
      update in response to the change in data set B and then all views of
      data set A are updated. }
    \label{Fig:firstMP}
  \end{center}
\end{figure}

% In Section \ref{Ssec:Design}, the design of linked, interactive views based on
% the MVC paradigm is discussed.  However, the limitation with this design is
% that it expects one data set, which limits the type of data we can visualize.
% There are many situations where multiple data sets occur, particularly any
% time there are experimental or study data linked to meta-data.  Being able to
% visualize data from multiple data sets is helpful to understand the
% relationships between variables.  Thus, after discussing the limitations of
% the MVC design in Section \ref{Ssec:Limit}, an extension of the MVC design is
% discussed in Section \ref{Sec:Extend}.  This extension supports the creation of
% linked \Robject{MVC} objects.  The design of individual \Robject{MVC} objects
% is discussed in 
% Section \ref{Sec:OneMVC} while the implementation of linked \Robject{MVC}
% objects is discussed in Section \ref{Sec:MultMVC}.  The goals of the software
% discussed in this paper are to create linked views of linked data sets, to
% create interactive views where the response to an event can be changed, and to
% create an extensible design that allows future users to make additions.

% \subsection{Design and Implementation of Linked, Interactive
%   Views}\label{Ssec:Design}

% Focusing on linked, interactive views as a powerful visualization tool, the
% next step is to determine how to design a software package that can implement
% this method.  A popular design for creating linked views of data is the
% MVC paradigm.  This design consists of three types of objects: the controller,
% which defines what actions occur in response to user input; the view, which
% consists of displays of the data; and the model, which manages the data.  This
% paradigm is powerful because it decouples the views from the model.  Thus,
% only one copy of the data needs to be stored and then all views refer to this
% data set.  With all views based on the same data, any changes to the data can
% be propagated to all views, which creates linked views.

% To link the views to this one data set there is a subscribe/notify procedure
% between the views and the model.  The views subscribe to a particular model
% and the model must notify the views when a change occurs so that the views
% are updated.  This design allows multiple views of the model that are
% linked, but are not aware of each other, as shown in Figure \ref{Fig:ExMVC}.
% Each view is responsible for visually representing some aspect of the model
% and thus, it is not necessary for a view to be aware of other views.  By
% keeping each view ignorant of other views, views can added and removed without
% affecting each other.

% \begin{figure}[ht]
%   \begin{center}
%     \scalebox{0.5}{\includegraphics{ExMVC.jpg}}
%     \caption{ A Model with its Views. }
%     \label{Fig:ExMVC}
%   \end{center}
% \end{figure}

% Notice that although the components (model, view, and controller) in the MVC
% design are independent of one other, there needs to be communication
% between them so that information can be shared.  Examples of this
% communication between components include the model notifying views that the
% data have changed, a view asking the controller how it should respond to user
% interaction, and a view telling the model that the data should change in
% response to user interaction with the view.  This message passing between
% components in the MVC paradigm is discussed in detail in Section
% \ref{Ssec:OneMess}.

% In \cite{EW05}, we discuss the creation of R software to fulfill three goals:
% to create views that are linked, to create views that are interactive, and to
% create a design that is extensible.  To satisfy the overall goals of the
% design, two R packages, called \Rpackage{MVCClass} and \Rpackage{iSPlot},
% were created.  The \Rpackage{MVCClass} package implements the class and
% generic function definitions needed for the MVC design and the
% \Rpackage{iSPlot} package creates the front end to let the user create linked,
% interactive views of two-dimensional data, such as data frames and matrices,
% through a GUI or through command line functions.  The reason for
% separating the software into two packages was that it allowed the class
% definitions in the \Rpackage{MVCClass} package to be used by any other R
% package that wanted to create linked views.

\subsection{Limitations of the MVC Design}\label{Ssec:Limit}

In essence the MVC design assumes that there is one data set, which is
called the model, that all views reference.  But problems arise when
the data consist of more than one conceptual grouping.  For example,
consider data stored in a relational database, where there are
multiple tables of data and rows in different tables are linked
through a unique identifier called a key.  Although these tables of
data could be combined into one large table, there are several reasons
not to do this.  Keeping the data in separate tables reduces
redundancy, particularly when the rows are related by a one-to-many or
a many-to-many relationship.

%  Also separate data tables generally lead
% to more maintainable data.  By storing only one copy of each piece of
% information (in a row in one table), it is easier to ensure that the
% data are correctly updated when a change occurs.  All other tables
% that are connected with this information can then just reference this
% row that has been updated.

Consider Figure \ref{Fig:DBTab} shown below.  Here the database
consists of three tables, State, City and StateCity.  The State table
has three columns, State (the key), Life Expectancy, and Income.  The
City table has three columns, City (the key), Precipitation, and
Temperature.  The StateCity table has two columns, State and City.
This last table relates which cities are in which states.  In this
example, each city can be in only one state, but each state can have
more than one city so there is a one-to-many relationship between
states and cities.  In order to support interactive views of all data
tables, in a standard MVC paradigm, it seems that either all data must
be combined into a single large table (Figure~\ref{Fig:OneDBTab}), or
the processing and accessing of the model data must become much more
complex.

We felt that neither approach had much appeal. A single large table
suggests that all data are similar, and still leaves issues about how
to deal with ensuring that when, for example, a state is selected all
cities within it are highlit. The second approach requires writing a
large amount of code, but is likely to provide little
extensibility. We should also mention that our main area of
application is a setting where one representation of the data is as a
graph (with nodes and edges) while the other is as a matrix. We can
envisage lots of different MVCs that operate on graph based data, and
equally MVCs that operate on matrix based data. Specifying a mechanism
by which these can interact allows us to create two visible classes
that others can use, but still provide interaction between
appropriately linked instances of the two classes. We also argue that
there are surely other specific types of models that represent
specific types of data (e.g. time series) and it would make sense to
implement these as yet another separate, specialized MVC, that could
be linked to any one of a variety of other MVCs.

\begin{figure}[ht]
  \begin{center}
    \scalebox{0.5}{\includegraphics{newDatabaseTables.jpg}}
    \caption{ Example of Relational Database Tables. }
    \label{Fig:DBTab}
  \end{center}
\end{figure}

% Clearly, this information could be combined into one large table, called
% CombinedStateCity, as shown in Figure \ref{Fig:OneDBTab}.  The
% CombinedStateCity table has the following columns: State, Life
% Expectancy, Income, City, Precipitation, and Temperature.  Now if a state's
% life expectancy changes, all instances of this state need to change
% in the new large table.  For example, if the state had the information for
% three cities recorded, then the state has three rows in the
% CombinedStateCity table and each row needs to change the value for the
% state's life expectancy.  If instead only one record was kept per state, as
% shown in Figure \ref{Fig:DBTab} above, then this process is less error
% prone and requires less memory to store the data.  Another consideration in
% combining data sets is that it may give the appearance of a relationship
% between variables that does not exist.  For example, it may not make sense to
% compare the Income data stored in the State table to the Precipitation data
% stored in the City table.  Combining the data tables assumes that there is a
% relationship between the variables in all of the tables.  This example shows
% that there are advantages to storing the data in separate structures.

%\clearpage

\begin{figure}[ht]
  \begin{center}
    \scalebox{0.5}{\includegraphics{newOneDatabaseTable.jpg}}
    \caption{ Example of a Combined Table. }
    \label{Fig:OneDBTab}
  \end{center}
\end{figure}

% In a relational database all of the data are stored in tables, but consider the
% situation where there are multiple data sets that are of different data
% structures.  For example, suppose that you have experimental microarray data,
% which look at gene expression levels.  This experimental data can be
% stored as a two dimensional data structure, such as a matrix, where rows are
% genes and columns are samples.  Linked to this experimental data are meta-data
% that give information about the genes studied in the microarray experiment.
% For example, this meta-data may be terms from Gene Ontology, which form a
% directed acyclic graph, describing the molecular functions carried out by
% these genes.  These two data sets (the experimental data and the meta-data)
% are linked through the genes.  Each gene can have more than one molecular
% function and each molecular function can have more than one gene annotated at
% that function so there is a many-to-many relationship between the genes and
% the molecular functions.  Though these two data sets are linked, the
% microarray data and the Gene Ontology data are different data structures.

% Now for the original MVC design, where one data set is expected, it is not
% clear how to combine these two data sets into one data set.  The meta-data
% graph could be stored as a matrix, but a graph is a data structure that can be
% well represented as an object with slots for nodes and edges, especially if
% the graph is sparse.  As with relational database tables, combining the data
% sets complicates data storage, alteration, and retrieval.  For example, by
% combining the data sets a structure may be forced on the data that may not fit
% well, especially if the original data sets are instances of different classes.
% In contrast, keeping the data sets stored separately in their original
% structures simplifies the processes of storing, updating, and accessing the
% data.

\section{Extending the MVC Design}
\label{Sec:Extend}

As discussed in Section \ref{Sec:Intro}, the MVC design typically uses a single
conceptual unit, such as a data frame or a matrix, for the model.  And
we believe that such an approach is a good one. MVCs should be
specialized to certain types of data, as this allows them to provide
better and more natural interactions. However, we additionally argue
that there is great benefit to considering linking these different
MVCs when the underlying data themselves are derived from the same source.

Each \Robject{MVC} is a coherent object and the interactions between
\Robject{MVCs} are encapsulated in a set of messages.  We use one
\Robject{MVC} object to handle each data set.  Thus, within a
\Robject{MVC} object, one can create linked views of that particular
data set, using the standard MVC design.  Between \Robject{MVC}
objects we propose a second messaging system that allows messages to
be sent between linked data sets.

% \small
% \begin{tabular}[t]{ | l | r | r | }
%   \hline
%   State & Life Expectancy & Income \\ \hline
%   Pennsylvania & 70.43 & 4449 \\ \hline
%   California & 71.71 & 5114 \\ \hline
%   Minnesota & 72.96 & 4675 \\ \hline
% \end{tabular}
% \hspace{10pt}
% \begin{tabular}[t]{ | l | r | r | }
%   \hline
%   City & Precipitation & Temperature \\ \hline
%   Philadelphia & 39.9 & 39 \\ \hline
%   Pittsburgh & 36.2 & 37 \\ \hline
%   Los Angeles & 14.0 & 68 \\ \hline
%   San Francisco & 20.7 & 56 \\ \hline
%   Sacramento & 17.2 & 55 \\ \hline
%   Duluth & 30.2 & 18 \\ \hline
%   Minneapolis St. Paul & 25.9 & 22 \\ \hline
% \end{tabular}

% \begin{table}[h]
%   \begin{center}
%     \begin{tabular}{ | r | r | }
%       \hline
%       State & City \\ \hline
%       Pennsylvania & Philadelphia \\ \hline
%       Pennsylvania & Pittsburgh \\ \hline
%       California & Los Angeles \\ \hline
%       California & San Francisco \\ \hline
%       California & Sacramento \\ \hline
%       Minnesota & Duluth \\ \hline
%       Minnesota & Minneapolis St. Paul \\ \hline
%     \end{tabular}
%     \caption{State and City Values.}\label{Tab:CityState}
%   \end{center}
% \end{table}

% \normalsize
% As an example, suppose that one has two data sets: one with information at the
% state level and one with information at the city level.  These tables are
% shown in Table \ref{Tab:CityState}.  The State data include Life Expectancy
% (in years from 1970) and Income (per capita from 1974) data
% while the City data include Precipitation (in inches from 1975) and Average
% High Temperature in January (in Fahrenheit from 2005)
% data.  These two data sets are linked by the StateCity table that links cities
% with the states they reside in.  Now for these two data sets there are
% two \Robject{MVC} objects.  One \Robject{MVC} object for the State data and
% its views and one \Robject{MVC} object for the City data and its views.

\begin{figure}[ht]
  \begin{center}
    \scalebox{0.7}{\includegraphics{MultipleMVC.jpg}}
    \caption{ Example of Two \Robject{MVC} Objects.  These two \Robject{MVC}
      objects are linked by the StateCity table and the color of the points
    indicates the linking between the data.  The points referring to the state
    of Pennsylvania are red, the points referring to the state of California
    are blue, and the points referring to Minnesota are green. }
    \label{Fig:MultMVC}
  \end{center}
\end{figure}

In Figure \ref{Fig:MultMVC} we show one view for state-level data and
one view for the city-level data.  The state and city data are linked
based on the color.  Even though there are two \Robject{MVC} objects
here, one \Robject{MVC} for each data set, the views are linked.

% Not only does this solution use the power and simplicity of the MVC design, but
% this multiple MVC design is structured to fit the linking seen in the multiple
% data sets.  Clearly, the design should be built around the problem and
% multiple \Robject{MVC} objects with messaging between the \Robject{MVC}
% objects fits the problem of creating linked views of multiple linked data
% sets.  In designing an appropriate messaging system between \Robject{MVC}
% objects, consideration is given to what information should pass between
% \Robject{MVC} objects and to how messages propagate between \Robject{MVC}
% objects.  As for the relationship between the \Robject{MVC} objects, it should
% reflect the relationship between the data sets.

We believe that there will often be a hierarchical relationship
between the data sets and the MVCs that contain them. We have found
that allowing the user to create new MVCs based on a current MVC is an
important and useful paradigm.  The original MVC is the parent
MVC and the new MVC the child.

\section{Design Details An MVC}
\label{Sec:OneMVC}

\subsection{Implementation Details}
\label{SSec:OneOver}

We have implemented linked, interactive views of linked data sets
using an object-oriented programming design which allows for
extensibility.  We provide this implementation in the
\Rpackage{MVCClass} package, written in R, which implements the basic
classes and generic functions.  The \Rpackage{MVCClass} package
defines the individual model, view, and controller classes as well as
the \Rclass{MVC} class that binds the individual model, view, and
controller classes together into one unit.  The package also defines
the message classes that allow the components to communicate with each
other.  Thus, the \Rpackage{MVCClass} package contains the classes
necessary for creating linked views.

\subsection{Model}
\label{Ssec:OneModel}

In the MVC design, the model stores the data and among other things it
supports all updating and other mechanisms that might be needed.
Information can be sent to the model to add, update, and remove data
and after every change the model is responsible for sending messages
to the views indicating which changes have occurred.  In our
implementation we created a single virtual class that represents an
abstract model and that has storage locations for data that will be
needed by all models. Then, we created a number of subclasses, one for
each type of model. In Figure~\ref{Fig:Model} we show the inheritance
structure for the three types of models we have implemented, on based
on data frames, one based on graphs and one based on the expression
set class of data used by the Bioconductor project \citep{BIOC}.
%%FIXME: we need to send the complete paper2 to the BioC tech report
%%series and then cite it here, saying something like: Complete
%%implementation details are provided in XXX.

% Figure \ref{Fig:Model} shows the inheritance structure for the model
% classes. Here the virtual class, \Rclass{gModel}, has the slots
% \Rslot{modelData}, \Rslot{linkData}, \Rslot{virtualData}, and
% \Rslot{modelName}.  These four slots are common information that all models
% need to store.  The \Rslot{modelData} slot is the data set for this model, the
% \Rslot{linkData} slot is a list of two functions, \Rfunction{toParent} and
% \Rfunction{fromParent}, which let this model send information to its parent
% model and receive information from its parent model, respectively (see
% Sections \ref{Ssec:OneMVC} and \ref{Ssec:MultLink} for more information on how
% parent and child models are linked), the \Rslot{virtualData} slot is data
% pertaining to the views that needs to be stored with the model so that it can
% be shown in all views of this model, and the \Rslot{modelName} slot is the
% name of the model.

\begin{figure}[ht]
  \begin{center}
    \scalebox{0.7}{\includegraphics{ModelClass.jpg}}
    \caption{ Inheritance for Model Objects. }
    \label{Fig:Model}
  \end{center}
\end{figure}

% We encapsulate the particular data model in the class definition, which allows
% for efficient dispatch, and accessor functions that are appropriate for the
% underlying implementation.  The specific model classes used in the
% \Rpackage{iSNetwork} package to represent data are \Rclass{dfModel},
% \Rclass{graphModel}, and \Rclass{exprModel}.  The \Rclass{dfModel} class
% represents a model where the \Rslot{modelData} slot has class data frame (or
% matrix), the \Rclass{graphModel} class represents a model where the
% \Rslot{modelData} slot has class graph, and the \Rclass{exprModel} class
% represents a model where the \Rslot{modelData} slot has class exprSet, a class
% used to represent gene expression data \cite{BioC}.

As mentioned previously, the model classes must also perform certain
operations and these operations are reflected in the methods that are
defined for the classes.  All three model classes have an
\Rfunction{updateModel} method.  This method is called whenever the
data need to be updated in response to a user interacting with a view.

Although only three model classes are currently implemented, future
users are welcome to make additions to the model classes for new data
structures they want to represent.  For example, a user may want to add a new
model class called \Rclass{tsModel} to represent time series data.
To make this addition, the user extends the \Rclass{gModel} class in his or
her own package.

\subsection{Views}\label{Ssec:OneViews}

The view classes represent the visual depictions, or renderings, of
the model.  There are some data that are required by all views and all
must respond to a certain subset of events so we have implemented this
class by first creating a single, virtual class which the specialized
views extend.  As for our MVC classes, this allows us to define
methods and data storage once, for the virtual class, and only extend
it for views that specifically require additional data storage, or
specialized handling.  The object model for the view classes is shown
in Figure \ref{Fig:View}.

\begin{figure}[ht]
  \begin{center}
    \scalebox{0.7}{\includegraphics{newViewClass.jpg}}
    \caption{ Inheritance for View Objects. }
    \label{Fig:View}
  \end{center}
\end{figure}

As for specific view classes, the \Rclass{spreadView} class represents
a spreadsheet view of the data.  This view only makes sense for models
that are a two dimensional data structure, such as a matrix or a data
frame. Other views we have implemented include: scatterplot
(\Rclass{sPlotView}), graph (\Rclass{graphView}), and heatmaps
(\Rclass{heatmapView}).

Since all views are expected to be interactive they have methods that
correspond to the different supported user interactions.  The events
we are most interested in responding to include key press events,
where a certain configuration of keys are pressed; button click
events, where one of the mouse buttons is clicked over the view; and
mouse movement events, where the cursor moves over the view.  When an
event occurs it creates a signal, that is caught by the signal
handler.  The signal handler then calls the callback function and it
is in the callback function that the corresponding method is called.
For example, when the user clicks the mouse button and the cursor is
over a scatter plot view, it is the scatter plot's
\Rfunction{clickEvent} method that is invoked.  Using the information
that is passed to it, the method converts the pixel coordinates into
user coordinates and then finds the closest object in the plot.

All views need an update method so they can alter themselves when the
model sends them a message that the data have been altered. The
actions taken depend on the view, for example for a spread sheet view
the method may select or unselect a row on the spreadsheeet while for
a scatterplot view the method may change the appearance of a point on
the plot.  The inheritance structure of the view classes is
extensible.  For instance, if a user wanted to add a new plot view,
say a density plot, they can create a new view class that inherits
from \Rclass{plotView}, perhaps called \Rclass{dPlotView}.

\subsection{Controller}
\label{Ssec:OneCont}

The controller performs two tasks: the first task is to store
information and the second task is to allow the user to determine the
course of actions.  The information that must be stored primarily
pertains to state. In order for the user to be able to alter the
effects of different interactions, such as changing the response to a
key-click event, or the appearance of the views, the controller must
retain information as to the state.

For the second task, the controller gives the user restricted access
to the state information through an application program interface
(API). In our implementation users can interact either through a GUI
or through command line functions.  Through the API the user can load
data, create views, set the response to user interaction with a view,
quit the program and perform other functions.  On the GUI, there are
menus to let users decide which action is taken next.  If the user
would rather not use the GUI, there are command line functions that
provide access to the same functionality that is available through the
GUI.

% The GUI is shown in Figure \ref{Fig:ContWin} and is referred to
% as the control window.

% \begin{figure}[ht]
%   \begin{center}
%     \scalebox{0.35}{\includegraphics{ControlWindow.jpg}}
%     \caption{ The Control Window. }
%     \label{Fig:ContWin}
%   \end{center}
% \end{figure}

Because we support simultaneous linked \Robject{MVC}
objects, more than one \Robject{MVC} object can be loaded and hence
more than one model can be present. However, the control window menus
and the command line functions only perform operations on one
\Robject{MVC} object at a time.  Thus, we make use of the concept of
an \textit{active} \Robject{MVC} object, where the active
\Robject{MVC} object (and thus, the active model) means that
operations performed through the control window menus or through the
command line functions affect that \Robject{MVC} object. Users can
switch between active models, but since the interactions could be
quite different, there must always be an active MVC.

In our implementation we provide a control window, for users that want
to interact via the GUI.  It has menu items corresponding to
\texttt{File}, \texttt{Edit}, \texttt{Display}, \texttt{Event}, and
\texttt{Window}.  These menus always appear on the control window, but
the items under each menu may change depending on the type of model
that is active.  For example, the \texttt{Display} menu allows users
to create views of the active model, but only certain views are
appropriate for a type of model, as mentioned in
Section~\ref{Ssec:OneViews}.  If the active model was a data frame
model (\Rclass{dfModel}), then the views that the user could create
are a scatter plot (\Rclass{sPlotView}) and a spreadsheet
(\Rclass{spreadView}), but a graph plot (\Rclass{graphView}) does not
make sense.  Thus, when the model in the ``active'' \Robject{MVC} is
of class \Rclass{dfModel}, then the Display menu has menu items to
create a scatter plot and a spreadsheet.  Also, if the ``active''
\Robject{MVC} is of class \Rclass{dfModel}, then the command line
functions only allow the user create a scatterplot or a spreadsheet
view of the active model, since these are the only ones we have
implemented.

Because one of our goals is to create a software package that is
flexible and extensible, users are able to add menus and menu items to
the control window through the command line to access new
functionality that they have created.  As an example if a future
programmer wants to add a new type of view, such as a histogram, then
a new menu item for that view could be added to the \texttt{Display}
menu.  By allowing users to change the controller, they can add new
functionality that is specific to their problem.  Note that when users
make additions to the API, they should add new command line functions
in addition to adding menus and menu items to access that
functionality.

\subsubsection{Linking Events to Functions}
\label{Ssec:OneEvent}

Linking callback functions to events, such as a button click, a mouse
movement or a key press event, is what allows users to have an
interactive environment.  When an event occurs, the callback function
linked to the event determines the response to that event.  To create
a flexible environment it is important to be able to change the
response, to an event.  For example, when a left button press event
occurs, a user may sometimes want a point to be colored and at other
times she may want a point to be hidden.  However, different responses
to an event may require functions with different signatures.  In our
example coloring a point requires two parameters, the new color to use
and the point to color, while hiding a point
requires just one parameter, the point to hide.  Since the intention is
to be able to interchange the responses to an event, the signatures of
different callback functions should be the same.

% As shown in Figure \ref{Fig:EventFun}, the slots in the \Rclass{gEventFun}
% class are \Rslot{callFun}, \Rslot{shortName}, and \Rslot{preprocessFun}.  The
% \Rslot{callFun} slot is the callback function, the
% \Rslot{shortName} slot is a short description of the function that
% identifies the callback function, and the
% \Rslot{preprocessFun} slot is a list of all preprocessing
% functions that must be called before the callback function.

% \begin{figure}[ht]
%   \begin{center}
%     \scalebox{0.6}{\includegraphics{EventClass.jpg}}
%     \caption{ The gEventFun Class. }
%     \label{Fig:EventFun}
%   \end{center}
% \end{figure}

%%FIXME: I have tried to reword this part, but am not sure I have it
%%right, nor that I am following the design decision here.
In our design a callback function takes only one parameter. Further,
within a specific MVC all callback functions must have the same
signature so that different callback functions can be interchanged to
allow different responses to an event.  The class of that parameter
depends on the type of model represented by the \Robject{MVC}.  For
example, for a graph model, the parameter corresponds to a node in the
graph, while for a data frame model, the parameter is of class
\Rclass{character}, representing a case.

%%FIXME: I thought we talked about this, and found it to be a little
%%problematic, as the way this sounds every click will ask me to set a
%%color. Somehow, some of these are essentially global options that
%%need to be set once, and then used.
Any other information a callback function may need is set using the
preprocessing functions.  The preprocessing functions take no
parameters.  The purpose of the preprocessing functions is to set
variables that the callback function needs.  Thus, not all callback
functions need preprocessing functions and if no preprocessing
functions are needed, then this slot is NULL.  If the callback
function sets the color of a node in a graph, then a preprocessing
function is needed to determine what color is used.  So in this
example, the preprocessing function opens a color browser to allow the
user to choose what the new color is.

%%FIXME: any chance you could elucidate here - what events get
%%excluded, and why are we wanting to be the "event police?"
Users can link actions to events either through the GUI or via
commandline functions.  Note that not all callback functions make
sense for each event so the controller environment also stores a list,
for each even,t of appropriate instances that can be linked to that
event.

\subsection{MVC}
\label{Ssec:OneMVC}

The \Rclass{MVC} class binds the model, view, and controller objects
into one object.  Thus, whenever a new data set is loaded, an
appropriate \Robject{MVC} object is created.  There is a one-to-one
relationship between the \Robject{MVC} instances and model objects.
Because of this relationship, the \Robject{MVC} object and the model
object are identified by the same name.

Our implementation is shown in Figure \ref{Fig:MVCClass}. It consists
of a virtual class, \Rclass{MVC}, with three slots, which are
\Rslot{model}, \Rslot{viewList}, and \Rslot{controller}. The three
slots correspond to the three pieces of information that are needed to
bind the model, view, and controller into a single unit. This class is
sufficient for all single MVC activities. To support linked data sets
we extend this class with the \Rclass{linkedModelMVC} class, which has
two additional slots: \Rslot{parentMVC}, which is the name of the
parent model and \Rslot{childMVCList}, which is a
list of the child models.

\begin{figure}[ht]
  \begin{center}
    \scalebox{0.6}{\includegraphics{fullMVCClass.jpg}}
    \caption{ The MVC Class. }
    \label{Fig:MVCClass}
  \end{center}
\end{figure}

%%FIXME: at this point we could perhaps even include a real example
%%with pictures of what is described below?
Thus, the \Rclass{linkedModelMVC} class not only binds the model,
view, and controller objects together, but it also shows
that the \Robject{linkedModelMVC} objects are related to each other through a
parent-child (tree) hierarchy, as shown in Figure \ref{Fig:Hier}.  In this
picture, `model1' is the name of the original \Robject{MVC} object
(model) and from it, two \Robject{MVC} objects were created,
`model2.1' and `model2.2'. Then from `model2.1', three
\Robject{MVC} objects were created, `model3.1', `model3.2', and
`model3.3'.  As an example, model1 could be gene 
expression data, model2.1 could be gene expression data that is a subset of
model1 (based on filtering), model2.2 could be derived by performing
multidimensional scaling (MDS) on model1, model 3.1 could be the gene ontology
(GO) graph representing the molecular functions of the genes in model2.1,
model 3.2 could be the KEGG pathways that the genes in model2.1 are involved
in, and model 3.3 could be the chromosome locations of the genes in model2.1.

%\clearpage

\begin{figure}[ht]
  \begin{center}
    \scalebox{0.6}{\includegraphics{Hierarchy2.jpg}}
    \caption{ The Parent-Child Relationship Between \Robject{MVCs}. }
    \label{Fig:Hier}
  \end{center}
\end{figure}

New models, and thus new \Robject{MVC} objects, can either be loaded
or they can be created from an existing \Robject{MVC}. As described in
Section~\ref{sec:micro} we are interested in data analytic settings
where experimental data are linked to meta-data and that different
views of the data based on the meta-data are of substantial interest.
In this situation, the experimental data can be thought of as
determining what meta-data are studied.  Thus, there is a hierarchical
relationship between these data sets where one \Robject{MVC} object
creates another \Robject{MVC} object.  This focus led to the decision
that a \Robject{MVC} object can have at most one parent \Robject{MVC},
from which it was derived, but it can have many child \Robject{MVCs}.

\subsection{Messages Within a MVC Object}
\label{Ssec:OneMess}

In this section we discuss the messages that are sent between the
components of a single MVC object and later we will discuss the
support needed to link MVCs.  All communication is carried out via
message passing. We have implemented a set of message classes to
provide communication between different components of the MVC design,
such as when the model changes a message needs to be sent to the views
to let them know that they should be updated.  This communication
between the model, view and controller is crucial for the pieces to
work together and yet, still be independent of each other.

%%FIXME: lets talk about what we need to do, what sorts of messages
%%need to be sent, and less about how we implemented it, for this
%%paper. As noted elsewhere we can send the whole tech report to the
%%BioC Report series.

% The object model for the message classes was derived from the idea
% that all messages must have certain methods in common, such as
% \Rfunction{initialize} and \Rfunction{handleMessage}, because all
% messages must be created and handled in some manner so that the
% message is read and acted upon.  All messages are sent to
% the \Rfunction{handleMessage} method, which determines where the information
% should be sent and hence is the dispatcher.   Also, since messages
% are sent when something has changed, either through addition, alteration, or
% deletion, the message classes reflect only certain operations.  These
% common purposes for the messages allow the message object
% model to be constructed, and this model is shown in Figure \ref{Fig:Mess}.

% \begin{figure}[ht]
%   \begin{center}
%     \scalebox{0.6}{\includegraphics{newMessageClass.jpg}}
%     \caption{ Inheritance for Message Objects. }
%     \label{Fig:Mess}
%   \end{center}
% \end{figure}

% As with the view classes, the top message class,
% \Rclass{gMessage}, is a virtual class.  It contains no slots and its purpose
% is to bind the other message classes together.

%%FIXME: this needs a few more words. What do these messages do, and
%%why are they the only ones we care about?
Currently, there are two types of messages within an MVC: an add
message and an update message.  Both messages request that some
modification be carried out.  Since adding a view requires different
information and methods from adding a model (and similarly with
updating views versus models), it makes sense to have two separate
message classes for these elements.  These messages can be sent to
either a model, or a view and their effect is quite different.

%%FIXME: you use numbers, 1, 2, 3 and so on, for too many things. Use
%%letters to label objects, numbers are good for steps, as the numbers
%%mean something.
An example of message passing within a \Robject{MVC} object is shown
in Figure~\ref{Fig:MPwithin}.  In this example, the response to a user
performing a button click event over a point in the scatter plot is to
color the point red.  Setting the response to an event was discussed
in Section~\ref{Ssec:OneEvent}.  Here the user first clicks on a point
in the scatter plot, which is shown as Step 1 (the circled 1 on the
figure).  In the figure, the point that has been clicked has an arrow
next to it.  Then, in Step 2, an update data message is sent to the
model, which updates itself accordingly, ie. by changing the color
associated with the corresponding element.  After the model has been
updated, the views must be notified that a change has occurred so in
Step 3, an update view message is created and sent to all views of
this model.  In response to this message, the view updates itself in
Step 4 and now the point is colored red on the view to reflect the
change in the model. Any other view that contains this point should
also be altered to reflect the new color.

\begin{figure}[ht]
  \begin{center}
    \scalebox{0.6}{\includegraphics{newMPwithin.jpg}}
    \caption{ Message Passing within a \Robject{MVC} object.  The first step,
      indicated by a circled one, occurs when a user interacts with the y
      versus x scatter plot.  Then an update data message is sent to the model
      in the second step, which alters the model.  Next in the third step all
      views of this model are sent an update view message, which causes the
      view to be updated in the fourth step. }
    \label{Fig:MPwithin}
  \end{center}
\end{figure}

%%FIXME: say a bit about why and how the proper recipient is
%%identified, but don't get into the slots or class names, just
%%describe, succinctly, what needs to happen, and if possible how.

% All four classes, \Rclass{gAddDataMessage}, \Rclass{gAddViewMessage},
% \Rclass{gUpdateDataMessage}, and \Rclass{gUpdateViewMessage}, have
% \Rfunction{initialize} and \Rfunction{handleMessage} methods, which perform
% the following tasks.  The \Rfunction{initialize} method is used to properly
% fill the slots of the object when it is created.  The
% \Rfunction{handleMessage} method is necessary to process the message.  Since a
% message is a form of communication between two components, once it is created
% it must be sent to the proper recipient and action must be taken in response
% to its information, which is the purpose of the \Rfunction{handleMessage}
% method.

%%FIXME: say a few words about extension of existing methods, with a
%%more convincing example. Say, what if we allow users to drag a point
%%in a scatter plot. What happens then? It might be good to have a
%%single example, such as this, or the density plot, that you can talk
%%about in several different places. With the density plot, as we saw
%%in our discussion with DTL, we need to resolve where to put the
%%value of the slider, or density smooth.

% As with the inheritance structures seen in previous sections, the message
% inheritance structure is designed to allow for extensibility.  For example, a
% user could create a \Rclass{gUpdateControlMessage} class that inherited from
% the \Rclass{gUpdateMessage} class.  This new class,
% \Rclass{gUpdateControlMessage}, could convey information that told the
% control window to be updated in response to a change in the active view (the
% active view is the view that has the focus).  Currently, the control window
% does not reflect what the active view is, but if a view was created that could
% be rotated, then a user may want the control window to have some type of
% interface to allow the user to control the rotation.  The goal for the message
% classes is that the design is flexible enough to allow a user to make
% additions or changes.

\section{Details of Multiple MVCs}
\label{Sec:MultMVC}

%%In one discussion I had with an interested party, they suggested
%%that it should be possible to subscribe to a MVC. So a new MVC is
%%created and it can insert itself as a child (possibly as a parent),
%%or it can simply ask to get messages from a particular MVC. This
%%would be one way to deal with letting parent child discussions be
%%optional, we have that structure, but when sending messages we look
%%at the subscribers slot, and send the message to all subscribers.

We now describe the support needed for multiple linked data sets.
Each \Robject{MVC} object represents one model and its views.  In our
paradigm data sets are linked through some additional meta-data. In
the case of gene expression data the GO view is linked to the heatmap
view through the mapping of genes to GO categories. In the city and
state example, the cities are linked via the states in which they are
contained. If, in the state MVC, a state is selected that information
can be propagated to the city MVC, but some piece of software must
link the state to the cities. The message is interpreted by the model
which changes some data and then sends messages to all views that
there has been an update.

Thinking of experimental data that are linked to meta-data as an
example, there are two \Robject{MVC} objects, one for the experimental
data and one for the meta-data.  In the example of experimental
microarray data that are linked to the Gene Ontology molecular
function graph (Section~\ref{sec:micro}, these two data sets are
linked in the following manner. The gene identifiers in the
experimental microarray data are linked to the GO terms.  Thus, to
pass information between these two data sets requires that these
mappings be available.

%%FIXME: if it is not too much work, lets give the models names, not
%%numbers.
Consider Figure~\ref{Fig:Hier}, which was shown previously.  If
model1's model changes, then this information first gets passed to
model2.1 and model2.2 and these two \Robject{MVCs} may make changes to
their models based on this information.  Then model2.1 passes this
information to model3.1, model3.2, and model3.3 based on the changes
that model2.1 made to its model.  Thus, information passing between
model1 and model3.1, for instance, only occurs through model2.1.  So
messages can only be sent from parent \Robject{MVC} to child
\Robject{MVC} or from child \Robject{MVC} to parent \Robject{MVC}.

\subsection{Linking Data Sets}
\label{Ssec:MultLink}

%%FIXME: but we said this above, so let's drop this section and move
%%this up, integrate with the discussion there. Say we use two
%%functions, toParent and fromParent. The rest has been said.
For messages to pass between \Robject{MVC} objects, the information from one
\Robject{MVC} must be converted into useful information for a different
\Robject{MVC} object.  Thus, there needs to be some way to determine how data
from one \Robject{MVC} object relates to data in a different \Robject{MVC}
object.  This conversion is done by two functions that are stored in the
\Rslot{linkData} slot of the model objects.  The \Rslot{linkData}
slot is a list with two elements: a \Rfunction{toParent} function and a
\Rfunction{fromParent} function.

When a child \Robject{MVC} is created from an existing \Robject{MVC}, the
child model's \Rslot{linkData} slot is filled with the two functions,
\Rfunction{toParent} and \Rfunction{fromParent}.  When the model in
the child \Robject{MVC} is updated, the \Rfunction{toParent} function in the
\Rslot{linkData} slot is used to convert the child model's data
change into useful information for the parent model.

Now suppose that the parent model changed instead.  This information
needs to be passed to the child model.  However, a parent can have
multiple child \Robject{MVC} objects so the parent sends a message to
each child \Robject{MVC} to notify the child of the change and then
each child \Robject{MVC} handles the message by using its
\Rfunction{fromParent} function.  The \Rfunction{fromParent} function
converts the parent model's data change into useful information for
this particular child model.  Thus, the message sent may be specific
to the child.

In Section~\ref{Ssec:OneMess} we was noted that
there are two main types of messages: an add message and an update message.
These message types also are seen in passing messages between different
\Robject{MVC} objects, as shown in Figure \ref{Fig:BetwMess}.

\begin{figure}[ht]
  \begin{center}
    \scalebox{0.6}{\includegraphics{newMessageClass2.jpg}}
    \caption{ Inheritance for Message Classes that are Passed Between
      \Robject{MVCs}. }
    \label{Fig:BetwMess}
  \end{center}
\end{figure}

% The first message class that is discussed is
% \Rclass{gAddChildMessage}.  A \Robject{gAddChildMessage} object is very
% similar to a \Robject{gAddDataMessage} object because they are both adding a
% new model and a new \Robject{MVC} object.  The difference with a
% \Robject{gAddChildMessage} object is that it must fill in some extra
% information to tie this new \Robject{MVC} object to its parent \Robject{MVC}
% object.  The \Rclass{gAddChildMessage} class inherits from the
% \Rclass{gAddMessage} class, with the same slots of \Rslot{dataName},
% \Rslot{mData}, and \Rslot{type}.  The \Rslot{dataName} slot is the name of the
% new model (and new \Robject{MVC}), the \Rslot{mData} slot is a list that
% contains the
% model data and virtual data to fill the \Rslot{modelData} and
% \Rslot{virtualData} slots, respectively, of the new model, and the
% \Rslot{type} slot gives the type of the model (currently, the options are
% ``exprSet'', ``graph'', or ``data.frame'').  As with the previous
% message classes, the \Rclass{gAddChildMessage} class has two methods:
% \Rfunction{initialize} and \Rfunction{handleMessage}.  The
% \Rfunction{initialize} method properly fills the slots of the object when it
% is created and the \Rfunction{handleMessage} method ensures that the new
% model and \Robject{MVC} objects are created (which is what the
% \Rfunction{handleMessage} method for the \Rclass{gAddDataMessage} class does)
% and then it fills the \Rslot{linkData} slot of the new child model
% so that the \Rfunction{toParent} and \Rfunction{fromParent} functions are
% available for messaging between \Robject{MVC} objects and finally it stores
% information in the child \Robject{MVC} object's controller that determines if
% the child model was created from a subset of the parent
% model.  Thus, the \Rfunction{handleMessage} method for the
% \Rclass{gAddChildMessage} class both creates the child \Robject{MVC} object
% and relates it to the parent \Robject{MVC} object.

% The next two message classes pertain to sending information to linked
% \Robject{MVC} objects when a parent \Robject{MVC} or child \Robject{MVC}
% object has changed.  These classes are \Rclass{gSendParentMessage} and
% \Rclass{gSendChildMessage} and they both inherit from the \Rclass{gMessage}
% class, as shown in Figure \ref{Fig:BetwMess}.  Because the messages are
% sending information to a linked \Robject{MVC} object, they both contain a slot
% for a \Robject{gUpdateDataMessage} object that was just used to update a
% \Robject{MVC} object.  The \Rclass{gSendParentMessage} class has one slot,
% \Rslot{childUpdateDataMessage}, and this slot contains the
% \Robject{gUpdateDataMessage} object that was used to update the child
% model.  Similarly, the \Rclass{gSendChildMessage} class has two
% slots, \Rslot{parentUpdateDataMessage} and \Rslot{childName}, where the
% \Rslot{parentUpdateDataMessage} slot contains the \Robject{gUpdateDataMessage}
% object that was used to update the parent model and the
% \Rslot{childName} slot contains the name of the child model that is
% being updated because a parent \Robject{MVC} can have more than one child
% \Robject{MVC}.

% As with all the other message classes, these two send messages have
% two methods: \Rfunction{initialize} and \Rfunction{handleMessage}.  The
% \Rfunction{initialize} method properly sets the slots of the new object.  The
% \Rfunction{handleMessage} method for a \Robject{gSendParentMessage}
% takes the \Robject{gUpdateDataMessage} object from the child model
% and converts it to a \Robject{gUpdateDataMessage} object for the parent
% model using the \Rfunction{toParent} function.  Similarly, the
% \Rfunction{handleMessage} method for a \Robject{gSendChildMessage} takes the
% \Robject{gUpdateDataMessage} object from the parent model and
% converts it to a \Robject{gUpdateDataMessage} object for the child
% model using the \Rfunction{fromParent} function.

An example of how these messages are passed and converted is shown in
Figure \ref{Fig:MessPass}.  These message classes ensure that when a
model is updated, its parent and child models are notified of the
change and most importantly are notified of the change in a way that
each model can properly act upon that information.  Because each
\Robject{MVC} object has a different model (and thus a different data
set and potentially a different data structure), information that
pertains to the model for one \Robject{MVC} must be converted to
useful information for the linked model and this conversion is done by
the \Rfunction{toParent} and \Rfunction{fromParent} functions.

\begin{figure}[ht]
  \begin{center}
    \scalebox{0.55}{\includegraphics{newMP.jpg}}
    \caption{ Message Passing Between and Within \Robject{MVCs}.  When a user
      interacts with the y versus x scatter plot, the first step, indicated by
      a circled one, is to send an update data message to model 2.  After
      model 2 has been updated, all views of this model must be notified, which
      is shown in the second step when an update view message is sent.  Next
      any data that are linked to model 2 must be notified that model 2
      changed, which is shown in the third step as a send parent message.
      This message is converted into useful information for model 1 in
      the fourth step by the \Rfunction{toParent} function, and this
      information is sent to model 1 in the fifth step when an update data
      message is sent.  Finally, after model 1 is updated, an update view
      message is sent in the sixth step to update all views of model 1. }
    \label{Fig:MessPass}
  \end{center}
\end{figure}

%%FIXME: please give models names, not numbers

Figure \ref{Fig:MessPass} is very similar to Figure \ref{Fig:firstMP} shown in
Section \ref{Sec:Intro}.  Now, however, Figure \ref{Fig:MessPass} shows how
\Robject{MVC} objects are defined and how messages are passed within one
\Robject{MVC} as well as how this information gets passed on to its parent
\Robject{MVC} using messages.  Suppose for this example that when the user
clicks a mouse button over the scatter plot, the response to this event is to
color the point under the cursor 
red.  In this picture, the message passing starts when a user clicks a point,
that is indicated by the arrow, on the scatter plot.  An update data message is
sent to Model 2 in response to this user interaction.  As soon as Model 2 is
updated, it sends an update view message to all its views so that the views
are updated to reflect the changed data.  Now the scatter plot point that was
initially clicked is red.  Then a send parent message is sent because MVC
2, which is the \Robject{MVC} object for Model 2, has a parent \Robject{MVC},
which is MVC 1.  Next the \Rfunction{toParent} function converts the child's
update data message into an update data message that the parent \Robject{MVC}
understands.  Then the update data message that was created by the
\Rfunction{toParent} function is sent to Model 1.  Finally, after Model 1 has
been updated, an update view message is sent to all of Model 1's views.  It is
in this last step that the view of Model 1 has a point colored red to indicate
which data from Model 1 is linked to data from Model 2.  All of the data and
the views are synchronized.

\section{Conclusions}\label{Sec:Conc}

Several examples of linked data sets were given at the beginning of
this paper, including linked tables from a relational database, and
meta data that are linked to experimental data.  The MVC paradigm was
discussed as a useful design for creating linked views of a data set.
Starting with the MVC paradigm, this paper provided a design for
implementing linked, interactive views of linked data sets based on
expanding the MVC paradigm.  For visualizing linked data sets, the MVC
design, which is a well known and thus, well tested software paradigm,
was extended to multiple \Robject{MVC} objects where each
\Robject{MVC} object contains one data set and its views.  This
solution broke down the problem of linked, interactive views of linked
data sets into components (\Robject{MVC} objects) that can be reused,
extended and linked.

Once the design settled on \Robject{MVC} objects as the component of interest
for creating linked views of linked data sets, the next step was to determine
how these \Robject{MVC} objects were related.  Because the focus in this paper
was on experimental data that are linked to meta-data, a hierarchical
relationship between \Robject{MVC} objects was built into the \Rclass{MVC}
class definition.  The experimental data can be thought of as determining what
meta-data are studied and thus, the meta-data are derived from the
experimental data, which shows a parent-child relationship.

Another consideration in the design was creating a method for communicating
within and between \Robject{MVC} objects.  This method must pass information
between the components while keeping the components independent.
Communication within a \Robject{MVC} object was handled using instances of the
message classes, which performed one of four functions: adding a model, adding
a view, updating a model, or updating a view.  Starting with this design, it
seemed a natural extension to have message classes that passed information
between \Robject{MVC} objects.  When a \Robject{MVC's} model is updated, it
notifies its parent and child \Robject{MVCs} that a change has occurred using
message objects.  Then when a \Robject{MVC} gets a message from a parent or
child \Robject{MVC}, the \Robject{MVC} can either decide to ignore the message
or read the message, which results in this \Robject{MVC's} model being
updated.  However, the \Robject{MVC} object can not decide to only see certain
types of messages.  It must either accept all messages and update its model
based on the messages, or ignore all messages.  A future goal for message
passing is to allow more selectivity in which messages are accepted between
\Robject{MVC} objects so that the linking between \Robject{MVC} objects is
more sophisticated.

%%FIXME: this belongs in the introductin. Here you should say
%%something more like: RGtk worked well. We could construct the tools
%%we wanted, relatively easily. You should also say what worked well -
%%ie that we got lots of specialized MVCs that can themselves be
%%implemented and extended - mostly because they are simple, they are
%%easier to understand, an unintended consequence.
%%
%% what about message passing? How did it work? Did you need more or
%% fewer types than you thought? What about the between data set
%% links? How did they turn out? Hard or easy? How much can be
%% automated and how much needs to be done by hand. Can naive users
%% get linked MVCs (which might be a better way to describe it than
%% linked data sets), or do you need to know a lot to get the link
%% function right?

The goals for this paper were to create linked views of linked data sets, to
create interactive views where the response to an event can be changed, and to
create an extensible design that allows future users to make additions.
The first goal was accomplished by implementing \Robject{MVC} objects as
components that communicate with each other through messages.  The second goal
of creating interactive views was met by using the \Rpackage{RGtk} and
\Rpackage{gtkDevice} packages to create devices that could respond to events.
In addition the \Rclass{gEventFun} class was designed so that users could
alter the response to an event.  Finally, the last goal of creating an
extensible design was achieved through the inheritance structure for classes,
which allows for extensions by the user, and through command line functions,
which let the user create new menu items to add to the GUI and also let the
user create new callback functions that can be linked to events.

Although extensibility was a goal, certain design decisions, such as expecting
the data sets to be linked through a parent-child hierarchy, may not work with
some data sets.  As an example, the linked tables from a relational database
are not related through a tree structure and thus, having parent and child
\Robject{MVC} objects may not make sense with these data sets.  However, the
design allows users to add new types of models, new types of views, and new
message classes so there is still flexibility for additions as long as the
data sets are connected through parent-child relationships.  These new class
definitions could be created by the user in a new package that extends the
\Rpackage{MVCClass} package that was discussed in this paper.  Other additions
that the user may want to perform, such as adding new menus to the GUI and
adding new responses to an event can be performed through command line
functions that are defined in the \Rpackage{iSNetwork} package.

The functionality that is defined in the \Rpackage{iSNetwork} package is
discussed in detail in the Vignette for this R package.  Users can load data,
create views of the data, interact with these views and change the response to
the interaction (these interactions include mouse button click events and
mouse movement events), and users can create child \Robject{MVC} objects
through the GUI or through command line functions.  Using the functionality in
the \Rpackage{iSNetwork} package, the user can create views of linked data sets
and can explore visually the relationships between variables in different data
sets.  This functionality was implemented by extending the MVC paradigm in a
very natural way to encompass a much richer class of models and user
interactions.

\bibliography{paper2}

\end{document}

### 2006-07-10 inst/paper/MVC.tex
Made changes to some of the figures and changed some of the text.
git-svn-id: file:///home/git/hedgehog.fhcrc.org/bioconductor/trunk/madman/Rpacks/MVCClass@18784 bc3139a8-67e5-0310-9ffc-ced21a209358

@@
   MVCs. We provide an implementation of our work in the R package
-\Rpackage{MVC} available at \url{http://www.bioconductor.org}. }
+\Rpackage{MVCClass} available at \url{http://www.bioconductor.org}. }
 
@@
 two or three dimensions are needed to view the data. Many authors
-(citations) have suggested that linked interactive views of the data
+(citations) have suggested that linked, interactive views of the data
 provide a valuable and important set of tools for comprehending such
@@
 In this paper we consider an extension of the MVC paradigm to allow
-for, what we believe to be, easy extension to linked data sets.
+for, what we believe to be, an easy expansion to linked data sets.
 
@@
 natural data representation of the subset is in a form other than that
-of the initial data set. This leads to complicatins in the usual MVC
+of the initial data set. This leads to complications in the usual MVC
 paradigm, where the data (i.e. the model) are often presumed to be in
@@
 react, or not, to messages from its children -- all control being in
-the hands of the user, should they want it.
+the hands of the users, should they want it.
 
@@
 visualization. In order to provide the reader with some idea of the
-sorts of interactions we would like to support we here provide a short
+sorts of interactions we would like to support, we provide here a short
 description of a type of experiment and some of the likely
@@
 much that is known about the genes, such as what chromosome they are
-located on, what sorts of activities they are engaged in and in many
+located on and what sorts of activities they are engaged in, and in many
 cases we want to restrict attention to a specific subset of genes, or
@@
 %% here?
+%% EAW: put them below
 There is often substantial interest in relating gene expression data
@@
 node in that graph (and sometimes one expects with the edges), while
-for the heatmap view the basic element is generally the gene.
+for the heatmap view the basic element is generally the gene.  In Figures
+\ref{Fig:heatmap} and \ref{Fig:GOgraph}, a heatmap view of gene expression
+from Chronic Lymphocytic Leukemia (CLL) patients and the corresponding GO
+graph are shown, respectively.  By interacting with the heatmap to color one
+gene blue (shown as gene PDE8A on the heatmap), which causes the corresponding
+GO nodes to be colored blue, one can easily see how this gene relates to the
+GO terms.
+
+\begin{figure}[ht]
+  \begin{center}
+    \includegraphics[height=5in, width=5in]{heatmap.jpg}
+    \caption{ Heatmap view of gene expression from Chronic Lymphocytic
+      Leukemia (CLL) patients. }
+    \label{Fig:heatmap}
+  \end{center}
+\end{figure}
+
+\begin{figure}[ht]
+  \begin{center}
+    \includegraphics[height=4.3in, width=4.3in]{GOgraph.jpg}
+    \caption{ Gene Ontology (GO) graph that is associated with the genes
+      displayed in the heatmap.  Notice that one gene (PDE8A) is colored blue
+      in the heatmap and all of the corresponding GO terms are colored blue in
+      the graph. }
+    \label{Fig:GOgraph}
+  \end{center}
+\end{figure}
 
@@
 initial heatmap view or not. Thus, we looked for a solution that was
-more based on component software constructed within an object oriented
+based more on component software constructed within an object oriented
 programming environment.
@@
 knew no other languages to be able to extend and enhance the
-system. Therefore, we built our software on top the \Rpackage{RGtk} of
+system. Therefore, we built our software on top the \Rpackage{RGtk} package of
 D. Temple Lang, since it provides such an interface. We also vastly
-preferred Gtk over Tcl/Tk since its widgets are more sophisticated
+preferred Gtk over Tcl/Tk since its widgets are more sophisticated,
 which we found results in a much shorter development time.
@@
 performed in response to events. We also wanted to support a more
-sophisticate developer who might want to take our basic infrastructure
+sophisticated developer who might want to take our basic infrastructure
 and create an interactive system, say for the interactive analysis of
-time series data, using our infrastructure. Hence, all commands are
+time series data, using our design. Hence, all commands are
 available both through the GUI and through command line functions.
@@
 We first give our definitions of linked views, interactive views and
-linked data sets are given. Our model is typically a statistical data
-matrix which represents data on a number of variables, for each of a
+linked data sets. Our model is typically a statistical data
+matrix which represents data on a number of variables, for a
 number of samples.  By \textit{linked view} we mean that if a
@@
 genes. If the entities are the genes then when two scatter plots are
-created as views, where on displays sample one versus sample two and
-the second displays sample three versus sample four, then the points
+created as views, where the first plot displays sample one versus sample two 
+and the second displays sample three versus sample four, then the points
 on each plot that referred to the same gene are corresponding
@@
 
-By \textit{inked data sets} we mean that the data have more than one
+By \textit{linked data sets} we mean that the data have more than one
 conceptual grouping and that there is a relationship between these
@@
 data set A and data set A updates itself.  Finally, in Step 6, data
-set 1 tells its view to update itself.
+set A tells its view to update itself.
 
@@
 %%for the data sets - it was too confusing
+%%EAW: it's changed
 \begin{figure}[ht]
   \begin{center}
-    \scalebox{0.55}{\includegraphics{firstMessagePassing.jpg}}
-    \caption{ Linked views of linked Data Sets.  When a user interacts with
+    \includegraphics[height=3.3in, width=5in]{firstMessagePassing.jpg}
+    \caption{ Linked views of linked data sets.  When a user interacts with
       the y versus x scatter plot, the first step, indicated by a circled one,
@@
 complex.
+
+\begin{figure}[ht]
+  \begin{center}
+    \includegraphics[height=0.9in, width=3.5in]{newDatabaseTables.jpg}
+    \caption{ Example of Relational Database Tables. }
+    \label{Fig:DBTab}
+  \end{center}
+\end{figure}
+
+\begin{figure}[ht]
+  \begin{center}
+    \includegraphics[height=1.2in, width=1.5in]{newOneDatabaseTable.jpg}
+    \caption{ Example of a Combined Table. }
+    \label{Fig:OneDBTab}
+  \end{center}
+\end{figure}
 
@@
 
-\begin{figure}[ht]
-  \begin{center}
-    \scalebox{0.5}{\includegraphics{newDatabaseTables.jpg}}
-    \caption{ Example of Relational Database Tables. }
-    \label{Fig:DBTab}
-  \end{center}
-\end{figure}
-
 % Clearly, this information could be combined into one large table, called
@@
 %\clearpage
-
-\begin{figure}[ht]
-  \begin{center}
-    \scalebox{0.5}{\includegraphics{newOneDatabaseTable.jpg}}
-    \caption{ Example of a Combined Table. }
-    \label{Fig:OneDBTab}
-  \end{center}
-\end{figure}
 
@@
   \begin{center}
-    \scalebox{0.7}{\includegraphics{MultipleMVC.jpg}}
+    \includegraphics[height=5in, width=6in]{MultipleMVC.jpg}
     \caption{ Example of Two \Robject{MVC} Objects.  These two \Robject{MVC}
@@
 
-\section{Design Details An MVC}
+\section{MVC Design Details}
 \label{Sec:OneMVC}
@@
 each type of model. In Figure~\ref{Fig:Model} we show the inheritance
-structure for the three types of models we have implemented, on based
+structure for the three types of models we have implemented, one based
 on data frames, one based on graphs and one based on the expression
@@
   \begin{center}
-    \scalebox{0.7}{\includegraphics{ModelClass.jpg}}
+    \includegraphics[height=3in, width=4.7in]{ModelClass.jpg}
     \caption{ Inheritance for Model Objects. }
@@
 the model.  There are some data that are required by all views and all
-must respond to a certain subset of events so we have implemented this
-class by first creating a single, virtual class which the specialized
-views extend.  As for our MVC classes, this allows us to define
+must respond to a certain subset of events so we have implemented the view
+classes by first creating a single, virtual class which the specialized
+views extend.  As with our model classes, this allows us to define
 methods and data storage once, for the virtual class, and only extend
@@
   \begin{center}
-    \scalebox{0.7}{\includegraphics{newViewClass.jpg}}
+    \includegraphics[height=4in, width=5in]{newViewClass.jpg}
     \caption{ Inheritance for View Objects. }
@@
 that are a two dimensional data structure, such as a matrix or a data
-frame. Other views we have implemented include: scatterplot
-(\Rclass{sPlotView}), graph (\Rclass{graphView}), and heatmaps
+frame. Other views we have implemented include: scatter plots
+(\Rclass{sPlotView}), graphs (\Rclass{graphView}), and heatmaps
 (\Rclass{heatmapView}).
@@
 over a scatter plot view, it is the scatter plot's
-\Rfunction{clickEvent} method that is invoked.  Using the information
-that is passed to it, the method converts the pixel coordinates into
-user coordinates and then finds the closest object in the plot.
+\Rfunction{clickEvent} method that is invoked.  Then within the
+\Rfunction{clickEvent} method, the \Rfunction{identifyView} method is called
+to convert the pixel coordinates into an object in the plot.
 
@@
 model sends them a message that the data have been altered. The
-actions taken depend on the view, for example for a spread sheet view
+actions taken depend on the view, for example for a spreadsheet view
 the method may select or unselect a row on the spreadsheeet while for
-a scatterplot view the method may change the appearance of a point on
+a scatter plot view the method may change the appearance of a point on
 the plot.  The inheritance structure of the view classes is
@@
 \Robject{MVC} is of class \Rclass{dfModel}, then the command line
-functions only allow the user create a scatterplot or a spreadsheet
+functions only allow the user to create a scatter plot or a spreadsheet
 view of the active model, since these are the only ones we have
@@
 %%right, nor that I am following the design decision here.
+%% EAW: When we decided that we wanted callback functions to be
+%% interchangeable, we decided that the callback functions must have the same
+%% signature so that calls to different callback functions would be
+%% identical, as we wouldn't know ahead of time what function we would be
+%% calling.  All callback functions must know which object in the plot to
+%% act on, but there was no other universal information that all callback
+%% functions needed.  Thus, for any other information that a callback function
+%% would need, preprocessing functions were used.  As for that one parameter
+%% that the callback function needed, the parameter type may be different
+%% depending on the plot (and model) type - whatever is considered the object
+%% of interest in a plot (and model).  Thus, for plots of data frames,
+%% the object in the plot is considered a row from the data frame and for
+%% plots of graphs, the object in the plot is considered a node in the graph. 
 In our design a callback function takes only one parameter. Further,
@@
 %%need to be set once, and then used.
+%% EAW: The preprocessing functions are called once when the callback function
+%% is linked to the event - they are not called each time the event occurs.
 Any other information a callback function may need is set using the
-preprocessing functions.  The preprocessing functions take no
+preprocessing functions, which are called once when the callback function is
+linked to the event.  The preprocessing functions take no
 parameters.  The purpose of the preprocessing functions is to set
@@
 example, the preprocessing function opens a color browser to allow the
-user to choose what the new color is.
+user to choose the new color.
 
@@
 %%excluded, and why are we wanting to be the "event police?"
+%% EAW: Any callback function can be linked to any event, but developers may
+%% only want certain callback functions available for certain events.  For
+%% example, a developer may only want the color a point callback function to
+%% be linked to a button click event (i.e. not the mouse over event) or the
+%% tooltips callback function only linked with the mouse over event.  Thus,
+%% I let the developer determine which events the callback function can
+%% be linked to (if the developer wants, the callback function can be linked
+%% to all events).  It is when a callback function is added as a possible
+%% response to an event that the developer gets to decide which events it can
+%% be a response to.  
+%% Otherwise, I would just store a list of all callback functions, which could
+%% be linked to any event.  I think the way it is implemented gives the
+%% developer a little more control over how the callback functions are used -
+%% but the developer is allowed to have his/her callback function be available
+%% for all events. 
 Users can link actions to events either through the GUI or via
-commandline functions.  Note that not all callback functions make
-sense for each event so the controller environment also stores a list,
-for each even,t of appropriate instances that can be linked to that
-event.
+command line functions.  The controller environment for each \Robject{MVC}
+object stores a list, for each event, of appropriate callback functions that
+can be linked to that event and it also stores which callback function is
+currently linked to the event.  When a developer creates a new callback
+function, the developer can decide which events the callback function can
+potentially be linked to.  For example, the developer may only want the color
+a point callback function to be available for a button click event and not a
+mouse over event.
 
@@
   \begin{center}
-    \scalebox{0.6}{\includegraphics{fullMVCClass.jpg}}
+    \includegraphics[height=3.2in, width=1.2in]{fullMVCClass.jpg}
     \caption{ The MVC Class. }
@@
 %%with pictures of what is described below?
+%% EAW: Yes, we could.  My only concern is confusing the idea of views
+%% with models.  I think it's important to show the inheritance
+%% structure of the models and not just views of the models.  
 Thus, the \Rclass{linkedModelMVC} class not only binds the model,
@@
 parent-child (tree) hierarchy, as shown in Figure \ref{Fig:Hier}.  In this
-picture, `model1' is the name of the original \Robject{MVC} object
+picture, `modelA' is the name of the original \Robject{MVC} object
 (model) and from it, two \Robject{MVC} objects were created,
-`model2.1' and `model2.2'. Then from `model2.1', three
-\Robject{MVC} objects were created, `model3.1', `model3.2', and
-`model3.3'.  As an example, model1 could be gene 
-expression data, model2.1 could be gene expression data that is a subset of
-model1 (based on filtering), model2.2 could be derived by performing
-multidimensional scaling (MDS) on model1, model 3.1 could be the gene ontology
-(GO) graph representing the molecular functions of the genes in model2.1,
-model 3.2 could be the KEGG pathways that the genes in model2.1 are involved
-in, and model 3.3 could be the chromosome locations of the genes in model2.1.
+`modelB' and `modelC'. Then from `modelB', three
+\Robject{MVC} objects were created, `modelD', `modelE', and
+`modelF'.  As an example, modelA could be gene 
+expression data, modelB could be gene expression data that is a subset of
+modelA (based on filtering), modelC could be derived by performing
+multidimensional scaling (MDS) on modelA, modelD could be the Gene Ontology
+(GO) graph representing the molecular functions of the genes in modelB,
+modelE could be the KEGG pathways that the genes in modelB are involved
+in, and modelF could be the chromosome locations of the genes in modelB.
 
@@
   \begin{center}
-    \scalebox{0.6}{\includegraphics{Hierarchy2.jpg}}
+    \includegraphics[height=2.5in, width=4.2in]{Hierarchy2.jpg}
     \caption{ The Parent-Child Relationship Between \Robject{MVCs}. }
@@
 %%why are they the only ones we care about?
+%% EAW: I've added some sentences below.
+%% My concerns/questions are:
+%% Obviously the other message class that could be implemented is the delete
+%% message (particularly, deleting a view and deleting a model).  I remember
+%% talking about this a while back and I don't recall exactly why we decided
+%% not to implement them.  At the time, we discussed the delete message we did
+%% not let a model be deleted (at least it wasn't implemented yet).  Message
+%% classes are meant to allow communication between the components - thus,
+%% deleting a view may be considered to take place solely within the
+%% controller (i.e. the response to a delete event for a view is handled by
+%% the contoller) and thus, a message between components is not necessary.
+%% However, deleting a view seems to be the exact opposite operation as adding
+%% a view and we have a message class for that operation.  
+%% Plus, I'm not sure how adding a model is considered a way to communicate
+%% within a MVC object since it creates the MVC object - adding a model
+%% creates the MVC object that is added to the overall environment that stores
+%% information about the system as a whole.
 Currently, there are two types of messages within an MVC: an add
@@
 updating views versus models), it makes sense to have two separate
-message classes for these elements.  These messages can be sent to
-either a model, or a view and their effect is quite different.
+message classes for these elements.  The adding a model message creates the
+\Robject{MVC} object because of the one-to-one relationship between a model
+and a \Robject{MVC}, whereas the adding a view message updates an
+existing \Robject{MVC} object.  The updating a model message is sent by the
+controller in response to an event (such as a button click event over a view)
+and the updating a view message is sent by the model when the model has
+changed.  Notice how these messages allow communication between components
+of a \Robject{MVC} object.
 
@@
 %%mean something.
+%% EAW: It's been changed.
 An example of message passing within a \Robject{MVC} object is shown
-in Figure~\ref{Fig:MPwithin}.  In this example, the response to a user
+in Figure~\ref{Fig:MPwithin}.  In this example, there are two views of the
+model (view 1 and view 2), and the response to a user
 performing a button click event over a point in the scatter plot is to
@@
 in Section~\ref{Ssec:OneEvent}.  Here the user first clicks on a point
-in the scatter plot, which is shown as Step 1 (the circled 1 on the
+in view 1, which is shown as Step 1 (the circled 1 on the
 figure).  In the figure, the point that has been clicked has an arrow
 next to it.  Then, in Step 2, an update data message is sent to the
-model, which updates itself accordingly, ie. by changing the color
+model, which updates itself accordingly, i.e. by changing the color
 associated with the corresponding element.  After the model has been
@@
 Step 3, an update view message is created and sent to all views of
-this model.  In response to this message, the view updates itself in
-Step 4 and now the point is colored red on the view to reflect the
-change in the model. Any other view that contains this point should
-also be altered to reflect the new color.
+this model.  In response to this message, the views update themselves in
+Step 4 and now each view has one point that is colored red to reflect the
+change in the model.
 
@@
   \begin{center}
-    \scalebox{0.6}{\includegraphics{newMPwithin.jpg}}
+    \includegraphics[height=3.5in, width=5in]{newMPwithin2.jpg}
     \caption{ Message Passing within a \Robject{MVC} object.  The first step,
       indicated by a circled one, occurs when a user interacts with the y
-      versus x scatter plot.  Then an update data message is sent to the model
-      in the second step, which alters the model.  Next in the third step all
-      views of this model are sent an update view message, which causes the
-      view to be updated in the fourth step. }
+      versus x scatter plot (view 1).  Then an update data message is sent to
+      the model in the second step, which alters the model.  Next in the third
+      step all views of this model are sent an update view message, which
+      causes the views to be updated in the fourth step. }
     \label{Fig:MPwithin}
@@
 %%describe, succinctly, what needs to happen, and if possible how.
+%% EAW: Not sure if you still want me to respond to this since you've
+%%      commented out the following text.
 
@@
 %%value of the slider, or density smooth.
+%% EAW: Again, not sure if you wanted me to respond to this since the below
+%%      text is commented out.
 
@@
 can be propagated to the city MVC, but some piece of software must
-link the state to the cities. The message is interpreted by the model
-which changes some data and then sends messages to all views that
-there has been an update.
+link the state to the cities. %The message is interpreted by the model
+%which changes some data and then sends messages to all views that
+%there has been an update. - EAW: I commented out the previous sentence
+%because I'm not sure how it fits into the paragraph.
 
@@
 linked in the following manner. The gene identifiers in the
-experimental microarray data are linked to the GO terms.  Thus, to
-pass information between these two data sets requires that these
+experimental microarray data are linked to the GO terms.  Thus,
+passing information between these two data sets requires that these
 mappings be available.
@@
 %%numbers.
+%% EAW: It's been changed.
 Consider Figure~\ref{Fig:Hier}, which was shown previously.  If
-model1's model changes, then this information first gets passed to
-model2.1 and model2.2 and these two \Robject{MVCs} may make changes to
-their models based on this information.  Then model2.1 passes this
-information to model3.1, model3.2, and model3.3 based on the changes
-that model2.1 made to its model.  Thus, information passing between
-model1 and model3.1, for instance, only occurs through model2.1.  So
+modelA's model changes, then this information first gets passed to
+modelB and modelC and these two \Robject{MVCs} may make changes to
+their models based on this information.  Then modelB passes this
+information to modelD, modelE, and modelF based on the changes
+that modelB made to its model.  Thus, information passing between
+modelA and modelD, for instance, only occurs through modelB.  So
 messages can only be sent from parent \Robject{MVC} to child
@@
 %%functions, toParent and fromParent. The rest has been said.
+%% EAW: Not sure if you wanted me to change anything here.
 For messages to pass between \Robject{MVC} objects, the information from one
@@
 converts the parent model's data change into useful information for
-this particular child model.  Thus, the message sent may be specific
-to the child.
-
-In Section~\ref{Ssec:OneMess} we was noted that
-there are two main types of messages: an add message and an update message.
-These message types also are seen in passing messages between different
-\Robject{MVC} objects, as shown in Figure \ref{Fig:BetwMess}.
+this particular child model.  Thus, the message sent to all children is the
+same, but each child converts the message, using its \Rfunction{fromParent}
+function, to make it specific for that child.
+%Thus, the message sent may be specific
+%to the child.
+
+The message classes for creating a child \Robject{MVC} from an existing
+\Robject{MVC} and for sending messages between \Robject{MVC} objects are shown
+in Figure \ref{Fig:BetwMess}.  The message types for passing information
+between \Robject{MVCs} are similar to the messages
+that are passed within \Robject{MVC} objects.  Now the add child message is
+similar to the add model message, and the send parent message and the send
+child message are similar to the update model message.
 
@@
   \begin{center}
-    \scalebox{0.6}{\includegraphics{newMessageClass2.jpg}}
+    \includegraphics[height=3.4in, width=4.5in]{newMessageClass2.jpg}
     \caption{ Inheritance for Message Classes that are Passed Between
@@
   \begin{center}
-    \scalebox{0.55}{\includegraphics{newMP.jpg}}
+    \includegraphics[height=2.8in, width=4.5in]{newMP.jpg}
     \caption{ Message Passing Between and Within \Robject{MVCs}.  When a user
       interacts with the y versus x scatter plot, the first step, indicated by
-      a circled one, is to send an update data message to model 2.  After
-      model 2 has been updated, all views of this model must be notified, which
+      a circled one, is to send an update data message to model B.  After
+      model B has been updated, all views of this model must be notified, which
       is shown in the second step when an update view message is sent.  Next
-      any data that are linked to model 2 must be notified that model 2
+      any data that are linked to model B must be notified that model B
       changed, which is shown in the third step as a send parent message.
-      This message is converted into useful information for model 1 in
+      This message is converted into useful information for model A in
       the fourth step by the \Rfunction{toParent} function, and this
-      information is sent to model 1 in the fifth step when an update data
-      message is sent.  Finally, after model 1 is updated, an update view
-      message is sent in the sixth step to update all views of model 1. }
+      information is sent to model A in the fifth step when an update data
+      message is sent.  Finally, after model A is updated, an update view
+      message is sent in the sixth step to update all views of model A. }
     \label{Fig:MessPass}
@@
 %%FIXME: please give models names, not numbers
+%% EAW: fixed.
 
@@
 that is indicated by the arrow, on the scatter plot.  An update data message is
-sent to Model 2 in response to this user interaction.  As soon as Model 2 is
+sent to Model B in response to this user interaction.  As soon as Model B is
 updated, it sends an update view message to all its views so that the views
@@
 initially clicked is red.  Then a send parent message is sent because MVC
-2, which is the \Robject{MVC} object for Model 2, has a parent \Robject{MVC},
-which is MVC 1.  Next the \Rfunction{toParent} function converts the child's
+B, which is the \Robject{MVC} object for Model B, has a parent \Robject{MVC},
+which is MVC A.  Next the \Rfunction{toParent} function converts the child's
 update data message into an update data message that the parent \Robject{MVC}
 understands.  Then the update data message that was created by the
-\Rfunction{toParent} function is sent to Model 1.  Finally, after Model 1 has
-been updated, an update view message is sent to all of Model 1's views.  It is
-in this last step that the view of Model 1 has a point colored red to indicate
-which data from Model 1 is linked to data from Model 2.  All of the data and
+\Rfunction{toParent} function is sent to Model A.  Finally, after Model A has
+been updated, an update view message is sent to all of Model A's views.  It is
+in this last step that the view of Model A has a point colored red to indicate
+which data from Model A are linked to data from Model B.  All of the data and
 the views are synchronized.
@@
 
-Several examples of linked data sets were given at the beginning of
-this paper, including linked tables from a relational database, and
-meta data that are linked to experimental data.  The MVC paradigm was
-discussed as a useful design for creating linked views of a data set.
-Starting with the MVC paradigm, this paper provided a design for
-implementing linked, interactive views of linked data sets based on
-expanding the MVC paradigm.  For visualizing linked data sets, the MVC
-design, which is a well known and thus, well tested software paradigm,
-was extended to multiple \Robject{MVC} objects where each
-\Robject{MVC} object contains one data set and its views.  This
-solution broke down the problem of linked, interactive views of linked
-data sets into components (\Robject{MVC} objects) that can be reused,
-extended and linked.
-
-Once the design settled on \Robject{MVC} objects as the component of interest
-for creating linked views of linked data sets, the next step was to determine
-how these \Robject{MVC} objects were related.  Because the focus in this paper
-was on experimental data that are linked to meta-data, a hierarchical
-relationship between \Robject{MVC} objects was built into the \Rclass{MVC}
-class definition.  The experimental data can be thought of as determining what
-meta-data are studied and thus, the meta-data are derived from the
-experimental data, which shows a parent-child relationship.
-
-Another consideration in the design was creating a method for communicating
-within and between \Robject{MVC} objects.  This method must pass information
-between the components while keeping the components independent.
-Communication within a \Robject{MVC} object was handled using instances of the
-message classes, which performed one of four functions: adding a model, adding
-a view, updating a model, or updating a view.  Starting with this design, it
-seemed a natural extension to have message classes that passed information
-between \Robject{MVC} objects.  When a \Robject{MVC's} model is updated, it
-notifies its parent and child \Robject{MVCs} that a change has occurred using
-message objects.  Then when a \Robject{MVC} gets a message from a parent or
-child \Robject{MVC}, the \Robject{MVC} can either decide to ignore the message
-or read the message, which results in this \Robject{MVC's} model being
-updated.  However, the \Robject{MVC} object can not decide to only see certain
-types of messages.  It must either accept all messages and update its model
-based on the messages, or ignore all messages.  A future goal for message
-passing is to allow more selectivity in which messages are accepted between
-\Robject{MVC} objects so that the linking between \Robject{MVC} objects is
-more sophisticated.
-
-%%FIXME: this belongs in the introductin. Here you should say
+%Several examples of linked data sets were given at the beginning of
+%this paper, including linked tables from a relational database, and
+%meta data that are linked to experimental data.  The MVC paradigm was
+%discussed as a useful design for creating linked views of a data set.
+%Starting with the MVC paradigm, this paper provided a design for
+%implementing linked, interactive views of linked data sets based on
+%expanding the MVC paradigm.  For visualizing linked data sets, the MVC
+%design, which is a well known and thus, well tested software paradigm,
+%was extended to multiple \Robject{MVC} objects where each
+%\Robject{MVC} object contains one data set and its views.  This
+%solution broke down the problem of linked, interactive views of linked
+%data sets into components (\Robject{MVC} objects) that can be reused,
+%extended and linked.
+
+%Once the design settled on \Robject{MVC} objects as the component of interest
+%for creating linked views of linked data sets, the next step was to determine
+%how these \Robject{MVC} objects were related.  Because the focus in this paper
+%was on experimental data that are linked to meta-data, a hierarchical
+%relationship between \Robject{MVC} objects was built into the \Rclass{MVC}
+%class definition.  The experimental data can be thought of as determining what
+%meta-data are studied and thus, the meta-data are derived from the
+%experimental data, which shows a parent-child relationship.
+
+%Another consideration in the design was creating a method for communicating
+%within and between \Robject{MVC} objects.  This method must pass information
+%between the components while keeping the components independent.
+%Communication within a \Robject{MVC} object was handled using instances of the
+%message classes, which performed one of four functions: adding a model, adding
+%a view, updating a model, or updating a view.  Starting with this design, it
+%seemed a natural extension to have message classes that passed information
+%between \Robject{MVC} objects.  When a \Robject{MVC's} model is updated, it
+%notifies its parent and child \Robject{MVCs} that a change has occurred using
+%message objects.  Then when a \Robject{MVC} gets a message from a parent or
+%child \Robject{MVC}, the \Robject{MVC} can either decide to ignore the message
+%or read the message, which results in this \Robject{MVC's} model being
+%updated.  However, the \Robject{MVC} object can not decide to only see certain
+%types of messages.  It must either accept all messages and update its model
+%based on the messages, or ignore all messages.  A future goal for message
+%passing is to allow more selectivity in which messages are accepted between
+%\Robject{MVC} objects so that the linking between \Robject{MVC} objects is
+%more sophisticated.
+
+%%FIXME: this belongs in the introduction. Here you should say
 %%something more like: RGtk worked well. We could construct the tools
@@
 %% function right?
+%%
+%% EAW: I've tried to change the conclusions to include some of your comments.
 
@@
 The first goal was accomplished by implementing \Robject{MVC} objects as
-components that communicate with each other through messages.  The second goal
-of creating interactive views was met by using the \Rpackage{RGtk} and
-\Rpackage{gtkDevice} packages to create devices that could respond to events.
-In addition the \Rclass{gEventFun} class was designed so that users could
-alter the response to an event.  Finally, the last goal of creating an
-extensible design was achieved through the inheritance structure for classes,
-which allows for extensions by the user, and through command line functions,
-which let the user create new menu items to add to the GUI and also let the
-user create new callback functions that can be linked to events.
-
-Although extensibility was a goal, certain design decisions, such as expecting
-the data sets to be linked through a parent-child hierarchy, may not work with
-some data sets.  As an example, the linked tables from a relational database
-are not related through a tree structure and thus, having parent and child
-\Robject{MVC} objects may not make sense with these data sets.  However, the
-design allows users to add new types of models, new types of views, and new
-message classes so there is still flexibility for additions as long as the
-data sets are connected through parent-child relationships.  These new class
-definitions could be created by the user in a new package that extends the
-\Rpackage{MVCClass} package that was discussed in this paper.  Other additions
-that the user may want to perform, such as adding new menus to the GUI and
-adding new responses to an event can be performed through command line
-functions that are defined in the \Rpackage{iSNetwork} package.
+components that communicate with each other through messages.  Encapsulating
+each data set in a \Robject{MVC} object was particularly useful because it
+allowed the reuse of code for communicating between the components of a
+\Robject{MVC} object, which made generating new model and view classes easier.
+However, communication between the \Robject{MVC} objects was, at times, more
+complicated.
+
+The crux of the communication between \Robject{MVC} objects is the link
+function, which converts information that one model can understand into
+information that a linked model can understand.  If there is a one-to-one
+relationship between the elements of the two linked models, then the link
+function is straightforward:  if element E changes in model A and if element E
+in model A is linked to element F in model B, then the link function will
+tell model B that element F should be updated.  However, if there is a
+one-to-many relationship or a many-to-many relationship, then the link
+function may be more complex.  Suppose that element E changes in model A and
+element E is linked to elements G, H, I, and J in model C, but these elements
+in model C are also linked to other elements in model A (in other words, there
+is a many-to-many relationship between the elements in models A and C).  Now
+the link function must decide whether to tell elements G, H, I, and J in model
+C to be updated in response to a change in element E in model A.  The link
+function may think of these relationships as an `any' versus an `all'
+relationship.  In an `any' relationship, the link function tells elements G,
+H, I, and J in model C to be updated if any element they are linked from model
+A is updated.  In an `all' relationship, the link function tells elements G,
+H, I, and J in model C to be updated only if all elements they are linked to
+in model A are updated.  Deciding how to represent these relationships between
+elements in different models is left to the developer when creating the link
+functions.  While this flexibility lets the developer represent different
+relationships between different models, the developer is then responsible for
+knowing what makes sense (for example, biologically) when creating the link
+functions.
+
+The second goal of creating interactive views was met by using the
+\Rpackage{RGtk} and \Rpackage{gtkDevice} packages to create devices that could
+respond to events.  Using these packages made creating interactive views
+particularly straightforward because any static R plot could be made
+interactive by plotting it on a Gtk device and by having the plotting function
+return information about the location of plot objects on the device.  In
+addition the \Rclass{gEventFun} class was designed so that users could alter
+the response to an event.  The response to an event could be as simple as
+selecting the plot object that was just clicked by a mouse button or the
+response could be as complicated as creating a new model.  This flexibility
+lets the developer create a more varied interactive environment.
+
+Finally, the last goal of creating an extensible design was achieved through
+the inheritance structure for classes, which allows for extensions by the
+user, and through command line functions, which let the user create new menu
+items to add to the GUI and also let the user create new callback functions
+that can be linked to events.  However, in terms of extensibility, certain
+design decisions, such as expecting the data sets to be linked through a
+parent-child hierarchy, may not work with some data sets.  As an example, the
+linked tables from a relational database are not related through a tree
+structure and thus, having parent and child \Robject{MVC} objects may not make
+sense with these data sets.  However, the design allows users to add new types
+of models, new types of views, and new message classes so there is still
+flexibility for additions as long as the data sets are connected through
+parent-child relationships.  These new class definitions could be created by
+the user in a new package that extends the \Rpackage{MVCClass} package that
+was discussed in this paper.  Other additions that the user may want to
+perform, such as adding new menus to the GUI and adding new responses to an
+event can be performed through command line functions that are defined in the
+\Rpackage{iSNetwork} package, which is the front end for the design
+described in this paper.
 

### 2006-07-20 inst/doc/MVCClass.Rnw
Moved some of the class definitions into the BioMVCClass package so that the MVCClass package no longer depends on Biobase, graph, or Rgraphviz.
git-svn-id: file:///home/git/hedgehog.fhcrc.org/bioconductor/trunk/madman/Rpacks/MVCClass@18314 bc3139a8-67e5-0310-9ffc-ced21a209358
Changed the text for the vignette, added a man page, and removed some old pictures that are not being used.
git-svn-id: file:///home/git/hedgehog.fhcrc.org/bioconductor/trunk/madman/Rpacks/MVCClass@18900 bc3139a8-67e5-0310-9ffc-ced21a209358

@@
 \author{Elizabeth Whalen}
-\date{May 5, 2006}
+\date{July 19, 2006}
 
@@
 classes and generic functions that will be used by other packages to create 
-model-view-controller applications.  Currently, the packages that use the 
-\Rpackage{MVCClass} package are \Rpackage{iSPlot} and \Rpackage{iSNetwork}.  
-Both packages use the definitions from \Rpackage{MVCClass} to create linked, 
-interactive views of data.
+model-view-controller applications.  Currently, the \Rpackage{iSPlot} and the
+\Rpackage{iSNetwork} packages use the \Rpackage{MVCClass} package to create
+linked, interactive views of data.  The \Rpackage{BioMVCClass} package extends
+the classes in the \Rpackage{MVCClass} package to include more model and view
+classes.  These definitions were put in a separate R package because the
+\Rpackage{BioMVCClass} packages depends on the following R packages:
+\Rpackage{Biobase}, \Rpackage{graph}, and \Rpackage{Rgraphviz}.
 
@@
 slots, parentMVC and childMVCList.  These two slots also show that a MVC
-object can have at most one parent MVC and can have as many children MVCs as
-it wants.
+object can have at most one parent MVC and can have many children MVCs.
 
@@
 structure for the model classes.  Here the virtual class,
-\Robject{gModel}, has the slots modelData, linkData, virtualData, and
-modelName.  These four slots are common information that all models will
-need.  The modelData slot is the data set for this model, the linkData slot is
+\Robject{gModel}, has the slots: modelData, linkData, virtualData, modelName,
+and modelVar.  These five slots are common information that all models need.
+The modelData slot is the data set for this model, the linkData slot is
 a list of two functions, \Rfunction{toParent} and \Rfunction{fromParent},
@@
 that needs to be stored with the model so that it can be shown in all views of
-this model, and the modelName slot is the name of the model.
+this model, the modelName slot is the name of the model (and the name of the
+MVC), and the modelVar slot is a list of named model variables (for example, a
+statistic for each element in the model).
 
@@
 
-For specific model classes, there are \Robject{dfModel}, \Robject{graphModel},
-and \Robject{exprModel}.  The \Robject{dfModel} class represents a model where
-the modelData has class data frame (or matrix), the \Robject{graphModel} class
-represents a model where the modelData has class graph, and the
-\Robject{exprModel} class represents a model where the modelData has class
-exprSet.
+In this package, the only specific model class is \Robject{dfModel}.
+The \Robject{dfModel} class represents a model where the modelData has class
+data frame (or matrix).  Other model classes are defined in the
+\Rpackage{BioMVCClass} package.
 
@@
 \Rpackage{iSPlot}).
+
+The model classes may also be asked to provide information by a
+\Robject{gAskAncestorMessage} object (which is discussed in Section
+\ref{Sec:Message}) and thus, the generic function \Rfunction{provideInfo} has
+been defined.  Again, the \Rfunction{provideInfo} methods are defined in
+packages that use this package, such as \Rpackage{iSNetwork}.
 
@@
 is not intended to have any objects as it just represents a general plot.  The
-plot classes that can have
-objects are the specific plot classes, \Robject{sPlotView},
-\Robject{graphView}, and \Robject{heatmapView}, which inherit from the
-\Robject{plotView} class.
+plot classes that can have objects are the specific plot classes,
+\Robject{sPlotView} and \Robject{qqPlotView}, which inherit from the
+\Robject{plotView} class.  More view classes are defined in the
+\Rpackage{BioMVCClass} package.
 
@@
 column name if the model is a data frame) from the
-model that is shown as the y variable in the plot.  The \Robject{graphView}
-class represents a graph plot view and it has the extra slot, grLayout, which
-holds a \Robject{Ragraph} object that represents the layout of the graph plot.
-The \Robject{heatmapView} represents a heatmap view of the data and it contains
-the slot, ordering, which is a list that is returned from the
-\Rfunction{heatmap} function to give information about the dendrogram
-ordering.  Note that some of these views are only applicable for certain model
-types.  For example, the \Robject{graphView} class will only make sense for a
-view of a graph object, which would be stored in the \Robject{graphModel}
-class.
+model that is shown as the y variable in the plot.  The \Robject{qqPlotView}
+represents a qq-plot view and it has the extra slots, xval and yval, where
+xval are the numeric values of the points on the x-axis and yval are the
+numeric values of the points on the y-axis.
 
@@
 \Rfunction{clickEvent}, \Rfunction{updateView}, \Rfunction{redrawView} and
-\Rfunction{identifyView}.  Methods for these generic 
-functions will be defined in packages that use this package (in 
-\Rpackage{iSPlot} and \Rpackage{iSNetwork}).  When these methods are defined, 
-the \Rfunction{motionEvent} method will be used to respond to a mouse over 
-event on a view, the \Rfunction{clickEvent} method will be used to respond to 
-a click event over a view, the \Rfunction{updateView} method will be used to 
-update only a portion of the view, the \Rfunction{redrawView} method will 
-be used to completely redraw the view, and the \Rfunction{identifyView} method
-will be used to identify an object (such as a point, a node, etc.) on a view.
+\Rfunction{identifyView}.  Methods for these generic functions are defined in
+packages that use this package, such as \Rpackage{iSPlot} and
+\Rpackage{iSNetwork}.  When these methods are defined, the
+\Rfunction{motionEvent} method responds to a mouse over event on a view, the
+\Rfunction{clickEvent} method responds to a click event over a view, the
+\Rfunction{updateView} method updates only a portion of the view, the
+\Rfunction{redrawView} method completely redraws the view, and the
+\Rfunction{identifyView} method identifies an object (such as a point, a node,
+etc.) on a view.
 
@@
 The figure shows the inheritance structure for message objects that can 
-be passed within one MVC object and a later figure in this section will show 
-the inheritance structure for messages objects that can be passed between MVC 
-objects.
+be passed within one MVC object and Figure \ref{Fig:BetwMess} in this section
+shows the inheritance structure for messages objects that can be passed
+between MVC objects.
 
@@
 that gives the type of addition or alteration to perform.  Both
-\Robject{gAddMessage} and \Robject{gUpdateMessage} are virtual classes.
+\Robject{gAddMessage} and \Robject{gUpdateMessage} are also virtual classes.
 
@@
 from within this MVC object or it may come from a different MVC object that
-has changed its model and is related to this MVC object.  
-Information on passing messages between MVC
-objects will be given later in this section.  Finally, the
-\Robject{gUpdateViewMessage} class represents a message to update a view.
+has changed its model and is related to this MVC object.  Information on
+passing messages between MVC objects is given later in this section.  Finally,
+the \Robject{gUpdateViewMessage} class represents a message to update a view.
 
-Message classes have also been defined to pass information between MVC objects.
-The two main types of messages, the add and update messages, will also be used 
-here.  The inheritance structure for messages passed between MVC objects is 
-shown in the figure below.
+Message classes have also been defined to pass information between MVC
+objects.  The message types for passing information between \Robject{MVCs} are
+similar to the messages that are passed within \Robject{MVC} objects.  Now the
+add child message is similar to the add model message, and the send parent
+message and the send child message are similar to the update model message.
+The inheritance structure for messages passed between MVC objects is shown in
+the Figure \ref{Fig:BetwMess}.
 
@@
 The first message class that will be discussed is \Robject{gAddChildMessage}.
-A \Robject{gAddChildMessage} object will be very similar to a
+A \Robject{gAddChildMessage} object is very similar to a
 \Robject{gAddDataMessage} object because they are both adding a new
@@
 data to fill the modelData and virtualData slots, respectively, of the new
-model, and the type slot gives the type of the model (currently, the options
-are ``exprSet'', ``graph'', or ``data.frame'').
+model, and the type slot gives the type of the model.
 
@@
 
-A generic function, \Rfunction{handleMessage}, is defined that will be used to 
-read message objects.  Methods for this generic function will be defined in 
-packages that use this package (\Rpackage{iSPlot} and \Rpackage{iSNetwork}).
+Finally, the \Robject{gAskAncestorMessage} class, which inherits from the
+\Robject{gMessage} class, is used to to ask an ancestor MVC for information
+about its model.  This class is used when a child MVC is being created and it
+needs information from more than just its immediate parent, i.e. it also needs
+some information from its grandparent or a different ancestor.  The
+\Robject{gAskAncestorMessage} class has three slots, from, type, and mData.
+The from slot is the name of the MVC that asked for the information (i.e. where
+this message came from), the type slot is the type of model to look for in an
+ancestor, and the mData slot is the information being asked for in the model.
+
+A generic function, \Rfunction{handleMessage}, is defined that is used to 
+read message objects.  Methods for this generic function are defined in 
+packages that use this package, such as \Rpackage{iSPlot} and
+\Rpackage{iSNetwork}.
 
@@
 or a key press event, is what allows users to have an interactive
-environment.  Then when an event happens something will occur in response
+environment.  Then when an event happens something occurs in response
 depending on the callback function linked to the event.  Now to create a
@@
 want a point to be hidden.  To allow this flexibility, a class called
-\Robject{gEventFun} was created.  By using this class, users will be able to
+\Robject{gEventFun} was created.  By using this class, users are able to
 change the response to an event.
@@
 callback function.  Having the callback function information stored in an 
-object will allow a user to connect events to callback functions.  The class 
-definition for \Robject{gEventFun} is shown in the figure below.
+object allows a user to connect events to callback functions.  The class 
+definition for \Robject{gEventFun} is shown in Figure \ref{Fig:EventFun}.
 
@@
 are callFun, shortName, and preprocessFun.  The callFun slot is the callback
-function, the shortName slot is a short description of the 
-function, and the preprocessFun slot is a list of all
-preprocessing functions that must be called before the callback function.
+function, the shortName slot is a short description of the function, and the
+preprocessFun slot is a list of all preprocessing functions that must be
+called before the callback function.
 
 The callback function that is stored in the callFun slot must take only one
-parameter and the class of this parameter will depend on the model class in
-the active MVC object.  For example, for a graph model, the
-parameter will be of class \Robject{AgNode}, representing a node in the graph.
-For a data frame model, the parameter will be of class character, representing
-a row in the data frame.
+parameter and the class of this parameter depends on the model class in
+the active MVC object.  For example, for a data frame model, the parameter
+is of class character, representing a row in the data frame.
 
-The preprocessing functions that are stored in the preprocessFun slot will
+The preprocessing functions that are stored in the preprocessFun slot
 take no parameters.  The purpose of the preprocessing functions is to set some
-variables that the callback function will need.  Thus, not all callback
+variables that the callback function needs.  Thus, not all callback
 functions need preprocessing functions and if no preprocessing functions
@@
 color of a node in a graph, then a preprocessing function is needed to
-determine what color this will be.  So in this example, the preprocessing
+determine what color is used.  So in this example, the preprocessing
 function opens a color browser to allow the user to choose what the new
@@
 generic function definitions.  Currently, this package is used by the 
-\Rpackage{iSPlot} and \Rpackage{iSNetwork} packages.
+\Rpackage{iSPlot} and \Rpackage{iSNetwork} packages, and is extended by the
+\Rpackage{BioMVCClass} package.
 

### 2006-07-22 inst/paper/MVC.tex
Changed text size and margins.
git-svn-id: file:///home/git/hedgehog.fhcrc.org/bioconductor/trunk/madman/Rpacks/MVCClass@18926 bc3139a8-67e5-0310-9ffc-ced21a209358

@@
-\documentclass{article}[11pt]
+\documentclass[11pt]{article}
 \setlength{\topmargin}{-2cm}
 \setlength{\textheight}{24cm}
-\setlength{\oddsidemargin}{10mm}
-\setlength{\evensidemargin}{10mm}
-\setlength{\textwidth}{15cm}
+\setlength{\oddsidemargin}{-.5in}
+%\setlength{\evensidemargin}{.5in}
+\setlength{\textwidth}{7.5in}
 \setlength{\parskip}{3mm}
@@
 \usepackage{hyperref}
+\usepackage{times}
 \usepackage[authoryear,round]{natbib}
@@
 
-\abstract{ In this paper we consider a generalization of the model
+\begin{abstract} 
+In this paper we consider a generalization of the model
   view controller (MVC) paradigm to allow for more flexible handling
@@
   MVCs. We provide an implementation of our work in the R package
-\Rpackage{MVCClass} available at \url{http://www.bioconductor.org}. }
+\Rpackage{MVCClass} available at \url{http://www.bioconductor.org}. 
+\end{abstract}
 

### 2006-07-29 inst/paper/MVC.tex
Made changes to paper 2 and also updated the vignette.
git-svn-id: file:///home/git/hedgehog.fhcrc.org/bioconductor/trunk/madman/Rpacks/MVCClass@18994 bc3139a8-67e5-0310-9ffc-ced21a209358

@@
 \usepackage[authoryear,round]{natbib}
+\usepackage{subfigure}
 
@@
 two or three dimensions are needed to view the data.  Many authors
-(citations) have suggested that linked, interactive views of the data
+\cite{intGrUnwin} \cite{iPlots} \cite{MONDRIAN} \cite{GGobi}
+have suggested that linked, interactive views of the data
 provide a valuable and important set of tools for comprehending such
@@
 paradigm, where the data (i.e. the model) are often presumed to be in
-a standard format. We found that adding new views under such
-conditions could be quite problematic and often leads to very complex
-code. Hence we considered an approach that created new instances of
+a single, standard format.  We found that adding new views under such
+conditions could be quite problematic and often led to very complex
+code.  As we shall demonstrate, there are also issues that arise due to
+mappings that are not one-to-one between elements on different views.
+Hence we considered an approach that created new instances of
 the MVC that were linked to the parent that spawned them.
@@
 stored as a tree.  Each parent can have multiple children, but the
-children can have a single parent. Communication between parent and
+children have a single parent.  Communication between parent and
 child is bidirectional, and a variety of different filters can be used
@@
 classes, each with a model appropriate to the specific data of
-interest, that are linked in a natural way. Interacting with the view
+interest, that can be linked in a natural way.  Interacting with the view
 in a child MVC, first sends a message to the child MVC, which will
@@
 GO nodes to be colored blue, one can easily see how this gene relates to the
-GO terms.
+GO terms.  Also by interacting with the node with an arrow next to it in
+Figure \ref{Fig:GOgraph}, which represents the signal transducer activity GO
+term, one can create a heatmap view that shows only the genes annotated at
+that node, as shown in Figure \ref{Fig:heatmapsub}.  Notice how all genes
+annotated at the signal transducer activity GO term have a blue label in
+Figure \ref{Fig:heatmapsub}, which shows that this view was derived from a
+blue colored node in Figure \ref{Fig:GOgraph}.  Compare the heatmaps in
+Figures \ref{Fig:heatmapsub} and \ref{Fig:heatmap}.  The heatmap in Figure
+\ref{Fig:heatmapsub} is just a subset of the heatmap in Figure
+\ref{Fig:heatmap}, however in Figure \ref{Fig:heatmapsub} eleven genes have a
+blue label and in Figure \ref{Fig:heatmap} only one gene has a blue label.
+This difference in these heatmaps indicates that views are synchronized based
+on the parent-child relationships.  Figure \ref{Fig:heatmap} is the parent of
+Figure \ref{Fig:GOgraph}, which is the parent of Figure \ref{Fig:heatmapsub}.
+Figures \ref{Fig:heatmap} and \ref{Fig:GOgraph} are synchronized, and Figures
+\ref{Fig:GOgraph} and \ref{Fig:heatmapsub} are synchronized, but Figures
+\ref{Fig:heatmap} and \ref{Fig:heatmapsub} are not synchronized by color
+because they do not have a parent-child relationship: Figure \ref{Fig:heatmap}
+is the grandparent of Figure \ref{Fig:heatmapsub}.
 
@@
   \begin{center}
-    \includegraphics[height=4.3in, width=4.3in]{GOgraph.jpg}
+    \includegraphics[height=4.3in, width=4.3in]{GOgraph2.jpg}
     \caption{ Gene Ontology (GO) graph that is associated with the genes
-      displayed in the heatmap.  Notice that one gene (PDE8A) is colored blue
-      in the heatmap and all of the corresponding GO terms are colored blue in
-      the graph. }
+      displayed in the heatmap in Figure \ref{Fig:heatmap}.  Notice that one
+      gene (PDE8A) is colored blue in the heatmap and all of the corresponding
+      GO terms are colored blue in the graph.  The arrow next to one of the
+      nodes indicates that this node was used to create the view shown in
+      Figure \ref{Fig:heatmapsub}.  The node with the arrow next to it
+      represents the signal transducer activity GO term. }
     \label{Fig:GOgraph}
+  \end{center}
+\end{figure}
+
+\begin{figure}[ht]
+  \begin{center}
+    \includegraphics[height=5in, width=5in]{heatmapsub.jpg}
+    \caption{ A heatmap view that shows only the genes that are annotated at
+      the node in Figure \ref{Fig:GOgraph} that has an arrow next to it.  This
+      node represents the GO term, signal transducer activity, and as shown in
+      this heatmap, eleven genes are annotated at this node.  Notice that all
+      the labels in this heatmap are colored blue, indicating that the parent
+      node that was used to create this view was also colored blue.  
+      % should I also mention that this is different from the genes that are
+      % selected in the grandparent model, which is shown in Figure
+      % \ref{Fig:heatmap}??  These eleven genes are also shown in the
+      % grandparent view, but only one of them is colored blue showing that
+      % views are synchronized by parent-child relationships.
+     }
+    \label{Fig:heatmapsub}
   \end{center}
@@
 called a key.  In this example, each table can be considered a data set
-and they are linked through the keys.  But the idea is more general
-and we shall discuss other situations where the linked data sets do
-not easily correspond to tables, however, we must have a map that goes
-from the entities in one data set to those in the other as well as the
-inverse map, in order to propagate signals between the MVCs.
+and they are linked through the keys. 
+Linked relational database tables could be combined into one
+large table, but doing so would result in a large amount of redundancy if
+the elements in the different tables are linked in one-to-many or many-to-many
+relationships.  Instead in relational databases, the tables are created such
+that each data element only needs to be stored in one table and then is linked
+to data elements in other tables.  Our design will build on this
+concept, but the idea is more general and we shall discuss other situations
+where the linked data sets do not easily correspond to tables.  As an example
+suppose we have gene expression data and we
+also know the Gene Ontology (GO) molecular function terms for these genes. 
+These data have a many-to-many relationship where each gene can be
+annotated at more than one GO term and each GO term can describe more than one
+gene.  Now these data could be combined in one data structure, but in
+our design we find that creating a MVC for the gene expression data and a MVC
+for the GO graph allows us to create views where we can explore the
+relationships between the genes and the GO terms.  To explore these
+relationships between data elements in different MVCs, we
+must have a map that goes from the entities in one data set to those in the
+other, as well as the inverse map, in order to propagate signals between the
+MVCs.
 
@@
 description of the steps that occur when a user interacts with a
-view. In the cartoon there are only two views, one of data set A and the
-other of data set B, which is linked to data set A.  When a user
-interacts with the scatter plot view of data set B, the response to
-this interaction is for data set B to update itself (this is indicated
+view.  In the cartoon there are only two views, one of model A (note 
+that once a data set is loaded in a MVC, we call it a model) and the
+other of the linked model B.  When a user
+interacts with the scatter plot view of model B, the response to
+this interaction is for model B to update itself (this is indicated
 on the figure as a circled 1).  In Step 2, the scatter plot view is
-told to update itself.  Next, in Step 3, because data set B is linked
-to data set A, data set B sends a message to data set A.  In Step 4,
-the information from data set B is converted into information that
-data set A understands.  In Step 5, this information is received by
-data set A and data set A updates itself.  Finally, in Step 6, data
-set A tells its view to update itself.
+told to update itself.  Next, in Step 3, because model B is linked
+to model A, model B sends a message to model A.  In Step 4,
+the information from model B is converted into information that
+model A understands.  In Step 5, this information is received by
+model A and model A updates itself.  Finally, in Step 6, model
+A tells its view to update itself.
 
@@
   \begin{center}
-    \includegraphics[height=3.3in, width=5in]{firstMessagePassing.jpg}
-    \caption{ Linked views of linked data sets.  When a user interacts with
+    \includegraphics[height=3.3in, width=6in]{firstMessagePassing.jpg}
+    \caption{ Linked views of linked models.  When a user interacts with
       the y versus x scatter plot, the first step, indicated by a circled one,
-      is to tell data set B to update.  Then all views of data set B must be
-      updated.  Next any data that are linked to data set B must be notified
-      that data set B changed.  This information from data set B must be
-      altered so that it can be understood by linked data sets, which is shown
-      in the fourth step.  Finally, data set A is notified that it should
-      update in response to the change in data set B and then all views of
-      data set A are updated. }
+      is to tell model B to update.  Then all views of model B must be
+      updated.  Next any data that are linked to model B must be notified
+      that model B changed.  This information from model B must be
+      altered so that it can be understood by linked models, which is shown
+      in the fourth step.  Finally, model A is notified that it should
+      update in response to the change in model B and then all views of
+      model A are updated. }
     \label{Fig:firstMP}
@@
 the data consist of more than one conceptual grouping.  For example,
-consider data stored in a relational database, where there are
-multiple tables of data and rows in different tables are linked
-through a unique identifier called a key.  Although these tables of
-data could be combined into one large table, there are several reasons
-not to do this.  Keeping the data in separate tables reduces
-redundancy, particularly when the rows are related by a one-to-many or
-a many-to-many relationship.
-
-%  Also separate data tables generally lead
-% to more maintainable data.  By storing only one copy of each piece of
-% information (in a row in one table), it is easier to ensure that the
-% data are correctly updated when a change occurs.  All other tables
-% that are connected with this information can then just reference this
-% row that has been updated.
-
-Consider Figure \ref{Fig:DBTab} shown below.  Here the database
+consider data stored in a relational database, such as the data shown in
+Figure \ref{FigDBTab}.  Here the database
 consists of three tables, State, City and StateCity.  The State table
@@
 tables, in a standard MVC paradigm, it seems that either all data must
-be combined into a single large table (Figure~\ref{Fig:OneDBTab}), or
+be combined into a single large table (Figure \ref{Fig:OneDBTab}), or
 the processing and accessing of the model data must become much more
@@
 
+%where there are
+%multiple tables of data and rows in different tables are linked
+%through a unique identifier called a key.  Although these tables of
+%data could be combined into one large table, there are several reasons
+%not to do this.  Keeping the data in separate tables reduces
+%redundancy, particularly when the rows are related by a one-to-many or
+%a many-to-many relationship.
+
+%  Also separate data tables generally lead
+% to more maintainable data.  By storing only one copy of each piece of
+% information (in a row in one table), it is easier to ensure that the
+% data are correctly updated when a change occurs.  All other tables
+% that are connected with this information can then just reference this
+% row that has been updated.
+
+%Consider Figure \ref{Fig:DBTab} shown below.  Here the database
+%consists of three tables, State, City and StateCity.  The State table
+%has three columns, State (the key), Life Expectancy, and Income.  The
+%City table has three columns, City (the key), Precipitation, and
+%Temperature.  The StateCity table has two columns, State and City.
+%This last table relates which cities are in which states.  In this
+%example, each city can be in only one state, but each state can have
+%more than one city so there is a one-to-many relationship between
+%states and cities.  In order to support interactive views of all data
+%tables, in a standard MVC paradigm, it seems that either all data must
+%be combined into a single large table (Figure \ref{Fig:OneDBTab}), or
+%the processing and accessing of the model data must become much more
+%complex.
+
+%\begin{figure}[ht]
+%  \begin{center}
+%    \includegraphics[height=0.9in, width=3.5in]{newDatabaseTables.jpg}
+%    \caption{ Example of relational database tables. }
+%    \label{Fig:DBTab}
+%  \end{center}
+%\end{figure}
+
+%\begin{figure}[ht]
+%  \begin{center}
+%    \includegraphics[height=1.2in, width=1.5in]{newOneDatabaseTable.jpg}
+%    \caption{ Example of a combined table. }
+%    \label{Fig:OneDBTab}
+%  \end{center}
+%\end{figure}
+
 \begin{figure}[ht]
   \begin{center}
-    \includegraphics[height=0.9in, width=3.5in]{newDatabaseTables.jpg}
-    \caption{ Example of Relational Database Tables. }
+    \subfigure[Example of relational database
+    tables.]{
     \label{Fig:DBTab}
-  \end{center}
-\end{figure}
-
-\begin{figure}[ht]
-  \begin{center}
-    \includegraphics[height=1.2in, width=1.5in]{newOneDatabaseTable.jpg}
-    \caption{ Example of a Combined Table. }
+    \includegraphics[height=1.2in, width=3.5in]{newDatabaseTables.jpg}}
+    \hspace{.3in}
+    \subfigure[Example of a combined table.]{
     \label{Fig:OneDBTab}
+    \includegraphics[height=1.2in, width=1.5in]{newOneDatabaseTable.jpg}}
+    \caption{Relational database tables.}
+%    \label{Fig:DBTab}
   \end{center}
@@
 to deal with ensuring that when, for example, a state is selected all
-cities within it are highlit. The second approach requires writing a
-large amount of code, but is likely to provide little
-extensibility. We should also mention that our main area of
+cities within it are highlighted.  The second approach requires writing a
+large amount of code, but is unlikely to be extensible.
+We should also mention that our main area of
 application is a setting where one representation of the data is as a
@@
 MVCs when the underlying data themselves are derived from the same source.
+We believe that there will often be a hierarchical relationship
+between the models and the MVCs that contain them.  We have found
+that allowing the user to create new MVCs based on a current MVC is an
+important and useful paradigm.
 
@@
 \Robject{MVCs} are encapsulated in a set of messages.  We use one
-\Robject{MVC} object to handle each data set.  Thus, within a
+\Robject{MVC} object to handle each model.  Thus, within a
 \Robject{MVC} object, one can create linked views of that particular
-data set, using the standard MVC design.  Between \Robject{MVC}
+model, using the standard MVC design.  Between \Robject{MVC}
 objects we propose a second messaging system that allows messages to
-be sent between linked data sets.
+be sent between linked models.
 
@@
 based on the color.  Even though there are two \Robject{MVC} objects
-here, one \Robject{MVC} for each data set, the views are linked.
+here, one \Robject{MVC} for each model, the views are linked.
 
@@
 % reflect the relationship between the data sets.
-
-We believe that there will often be a hierarchical relationship
-between the data sets and the MVCs that contain them. We have found
-that allowing the user to create new MVCs based on a current MVC is an
-important and useful paradigm.  The original MVC is the parent
-MVC and the new MVC the child.
 
@@
 implementation we created a single virtual class that represents an
-abstract model and that has storage locations for data that will be
+abstract model and that has storage locations for data that are
 needed by all models.  Then, we created a number of subclasses, one for
-each type of model. In Figure~\ref{Fig:Model} we show the inheritance
-structure for the three types of models we have implemented, one based
-on data frames, one based on graphs and one based on the expression
-set class of data used by the Bioconductor project \citep{BIOC}.
+each type of model.  Currently, we have implemented three types of models, 
+one based on data frames, one based on graphs, and one based on the 
+expression set class of data used by the Bioconductor project \citep{BIOC}.
+
+%In Figure~\ref{Fig:Model} we show the inheritance
+%structure for the three types of models we have implemented, one based
+%on data frames, one based on graphs and one based on the expression
+%set class of data used by the Bioconductor project \citep{BIOC}.
+
 %%FIXME: we need to send the complete paper2 to the BioC tech report
@@
 
-\begin{figure}[ht]
-  \begin{center}
-    \includegraphics[height=3in, width=4.7in]{ModelClass.jpg}
-    \caption{ Inheritance for Model Objects. }
-    \label{Fig:Model}
-  \end{center}
-\end{figure}
+%\begin{figure}[ht]
+%  \begin{center}
+%    \includegraphics[height=3in, width=4.7in]{ModelClass.jpg}
+%    \caption{ Inheritance for Model Objects. }
+%    \label{Fig:Model}
+%  \end{center}
+%\end{figure}
 
@@
 The view classes represent the visual depictions, or renderings, of
-the model.  There are some data that are required by all views and all
+the model.  There are some data that are required by all views, and all views
 must respond to a certain subset of events so we have implemented the view
@@
 
-The controller performs two tasks: the first task is to store
-information and the second task is to allow the user to determine the
-course of actions.  The information that must be stored primarily
-pertains to state. In order for the user to be able to alter the
-effects of different interactions, such as changing the response to a
-key-click event, or the appearance of the views, the controller must
-retain information as to the state.
+The controller performs two tasks: the first task is to maintain state,
+such as what response will occur for a particular user interaction, 
+and the second task is to control message passing.  Messages, which are
+discussed in Sections \ref{Ssec:OneMess} and \ref{Ssec:MultLink}, are
+responsible for creating new components, such as models and views, and for
+passing information between components, such as a view asking a model for
+information.  The controller determines when these messages are created and
+ensures that they are sent to the proper recipient.
 
@@
 to the state information through an application program interface
-(API). In our implementation users can interact either through a GUI
+(API) and thereby lets the user change the response to different events. 
+In our implementation users can interact either through a GUI
 or through command line functions.  Through the API the user can load
 data, create views, set the response to user interaction with a view,
-quit the program and perform other functions.  On the GUI, there are
-menus to let users decide which action is taken next.  If the user
+quit the program and perform other functions.  On the GUI, which we call the
+control window, users can decide which action is taken next by interacting
+with one of the menus: \texttt{File}, \texttt{Edit},
+\texttt{Display}, \texttt{Event}, and \texttt{Window}.  If the user
 would rather not use the GUI, there are command line functions that
@@
 
-Because we support simultaneous linked \Robject{MVC}
-objects, more than one \Robject{MVC} object can be loaded and hence
-more than one model can be present. However, the control window menus
+We support multiple loaded \Robject{MVCs}, but the control window menus
 and the command line functions only perform operations on one
@@
 
-In our implementation we provide a control window, for users that want
-to interact via the GUI.  It has menu items corresponding to
-\texttt{File}, \texttt{Edit}, \texttt{Display}, \texttt{Event}, and
-\texttt{Window}.  These menus always appear on the control window, but
+Although the same menus always appear on the control window,
 the items under each menu may change depending on the type of model
@@
 (\Rclass{spreadView}), but a graph plot (\Rclass{graphView}) does not
-make sense.  Thus, when the model in the ``active'' \Robject{MVC} is
+make sense.  Thus, when the model in the \textit{active} \Robject{MVC} is
 of class \Rclass{dfModel}, then the Display menu has menu items to
-create a scatter plot and a spreadsheet.  Also, if the ``active''
-\Robject{MVC} is of class \Rclass{dfModel}, then the command line
-functions only allow the user to create a scatter plot or a spreadsheet
-view of the active model, since these are the only ones we have
-implemented.
+create a scatter plot and a spreadsheet.
 
 Because one of our goals is to create a software package that is
-flexible and extensible, users are able to add menus and menu items to
-the control window through the command line to access new
+flexible and extensible, users and developers can add menus and menu items 
+to the control window through the command line to access new
 functionality that they have created.  As an example if a future
@@
 a new menu item for that view could be added to the \texttt{Display}
-menu.  By allowing users to change the controller, they can add new
-functionality that is specific to their problem.  Note that when users
-make additions to the API, they should add new command line functions
-in addition to adding menus and menu items to access that
-functionality.
+menu.  When adding to the API, command line functions should be created as
+well as menu items to access the new functionality.
 
@@
 a flexible environment it is important to be able to change the
-response, to an event.  For example, when a left button press event
+response to an event.  For example, when a left button press event
 occurs, a user may sometimes want a point to be colored and at other
 times she may want a point to be hidden.  However, different responses
-to an event may require functions with different signatures.  In our
-example coloring a point requires two parameters, the new color to use
-and the point to color, while hiding a point
-requires just one parameter, the point to hide.  Since the intention is
-to be able to interchange the responses to an event, the signatures of
-different callback functions should be the same.
+to an event may require functions with different signatures, where a function
+signature is the combination of the function name along with its parameters.
+In our example coloring a point requires two parameters, the new color to use
+and the point to color, while hiding a point requires just one parameter, the
+point to hide.
 
@@
 %% plots of graphs, the object in the plot is considered a node in the graph. 
-In our design a callback function takes only one parameter. Further,
-within a specific MVC all callback functions must have the same
-signature so that different callback functions can be interchanged to
-allow different responses to an event.  The class of that parameter
-depends on the type of model represented by the \Robject{MVC}.  For
-example, for a graph model, the parameter corresponds to a node in the
-graph, while for a data frame model, the parameter is of class
-\Rclass{character}, representing a case.
+Since the intention is to be able to interchange the responses to an event,
+the signatures of different callback functions should be the same.
+In our design a callback function takes only one parameter. 
+The class of that parameter depends on the type of model represented by the
+\Robject{MVC}.  For example, for a graph model, the parameter corresponds to
+the name of a node in the graph, while for a data frame model, the parameter
+represents a case.  We expect a one-to-one relationship between the elements
+on a view and the elements in the model.  Continuing the previous example,
+each point on a scatter plot represents one case in the data frame model and
+each node in the graph plot represents one node in the graph model.  Clearly
+it is possible for two points in a scatter plot to overlap if they have the
+same x axis and y axis values, but they are still considered two points
+representing two cases in the model.
 
@@
 Any other information a callback function may need is set using the
-preprocessing functions, which are called once when the callback function is
-linked to the event.  The preprocessing functions take no
+preprocessing functions, which are automatically called once by the controller
+when the callback
+function is linked to the event.  The preprocessing functions take no
 parameters.  The purpose of the preprocessing functions is to set
 variables that the callback function needs.  Thus, not all callback
-functions need preprocessing functions and if no preprocessing
-functions are needed, then this slot is NULL.  If the callback
+functions need preprocessing functions.  If the callback
 function sets the color of a node in a graph, then a preprocessing
@@
 
-Our implementation is shown in Figure \ref{Fig:MVCClass}. It consists
+Our implementation consists
 of a virtual class, \Rclass{MVC}, with three slots, which are
@@
 bind the model, view, and controller into a single unit. This class is
-sufficient for all single MVC activities. To support linked data sets
+sufficient for all single MVC activities. To support linked models
 we extend this class with the \Rclass{linkedModelMVC} class, which has
 two additional slots: \Rslot{parentMVC}, which is the name of the
-parent model and \Rslot{childMVCList}, which is a
-list of the child models.
-
-\begin{figure}[ht]
-  \begin{center}
-    \includegraphics[height=3.2in, width=1.2in]{fullMVCClass.jpg}
-    \caption{ The MVC Class. }
-    \label{Fig:MVCClass}
-  \end{center}
-\end{figure}
+parent model and \Rslot{childMVCList}, which is a list of the child models.
+These are the models that will communicate with each other: parents send
+information to their children and children send information to their parent.
+More information on this communication is given in Section \ref{Sec:MultMVC}.
+
+%\begin{figure}[ht]
+%  \begin{center}
+%    \includegraphics[height=3.2in, width=1.2in]{fullMVCClass.jpg}
+%    \caption{ The MVC Class. }
+%    \label{Fig:MVCClass}
+%  \end{center}
+%\end{figure}
 
@@
 
-New models, and thus new \Robject{MVC} objects, can either be loaded
-or they can be created from an existing \Robject{MVC}. As described in
-Section~\ref{sec:micro} we are interested in data analytic settings
-where experimental data are linked to meta-data and that different
-views of the data based on the meta-data are of substantial interest.
-In this situation, the experimental data can be thought of as
-determining what meta-data are studied.  Thus, there is a hierarchical
-relationship between these data sets where one \Robject{MVC} object
-creates another \Robject{MVC} object.  This focus led to the decision
-that a \Robject{MVC} object can have at most one parent \Robject{MVC},
-from which it was derived, but it can have many child \Robject{MVCs}.
+As shown in Figure \ref{Fig:Hier}, there are many cases where there are
+natural submodels that can be derived from a parent model.  We propose that
+each submodel is itself a \Robject{MVC} instance, and hence can have multiple
+linked views.  Even though each submodel is its own \Robject{MVC} instance,
+there will be times when the submodel \Robject{MVC} needs to communicate with
+its parent \Robject{MVC}.  There are many different ways of communicating
+between parent and child \Robject{MVCs} and we discuss this in Section
+\ref{Sec:MultMVC}.
+
+Deciding when new data should be added to the current model versus generating
+a new submodel in a \Robject{MVC} instance is done by comparing the
+relationship between the new data and the data in the current model.  If the
+new data has a one-to-one relationship with the elements in the current model,
+then the new data can be added to the current model.  For example, if there is
+a test statistic for each row in a data frame, then this could be added to the
+model.  However, if the new data has a one-to-many or a many-to-many
+relationship with the elements in the current model, then a submodel is
+generated in a new \Robject{MVC} instance.  An example here is the
+relationship between genes and GO terms.  If the original model contains gene
+expression levels and a user wants to explore the GO terms for these genes,
+then the GO graph representing the GO terms will be placed in a new
+\Robject{MVC} because the genes and GO terms have a many-to-many relationship.
+
+%New models, and thus new \Robject{MVC} objects, can either be loaded
+%or they can be created from an existing \Robject{MVC}. As described in
+%Section~\ref{sec:micro} we are interested in data analytic settings
+%where experimental data are linked to meta-data and that different
+%views of the data based on the meta-data are of substantial interest.
+%In this situation, the experimental data can be thought of as
+%determining what meta-data are studied.  Thus, there is a hierarchical
+%relationship between these data sets where one \Robject{MVC} object
+%creates another \Robject{MVC} object.  This focus led to the decision
+%that a \Robject{MVC} object can have at most one parent \Robject{MVC},
+%from which it was derived, but it can have many child \Robject{MVCs}.
 
@@
 %% controller (i.e. the response to a delete event for a view is handled by
-%% the contoller) and thus, a message between components is not necessary.
+%% the controller) and thus, a message between components is not necessary.
 %% However, deleting a view seems to be the exact opposite operation as adding
@@
 
-We now describe the support needed for multiple linked data sets.
+We now describe the support needed for multiple linked models.
 Each \Robject{MVC} object represents one model and its views.  In our
-paradigm data sets are linked through some additional meta-data. In
+paradigm models are linked through some additional meta-data. In
 the case of gene expression data the GO view is linked to the heatmap
@@
 microarray data that are linked to the Gene Ontology molecular
-function graph (Section~\ref{sec:micro}, these two data sets are
+function graph (Section~\ref{sec:micro}), these two models are
 linked in the following manner. The gene identifiers in the
 experimental microarray data are linked to the GO terms.  Thus,
-passing information between these two data sets requires that these
+passing information between these two models requires that these
 mappings be available.
@@
 \Robject{MVC} or from child \Robject{MVC} to parent \Robject{MVC}.
-
-\subsection{Linking Data Sets}
+This communication between parent and child \Robject{MVCs} was shown in Figures
+\ref{Fig:heatmap}, \ref{Fig:GOgraph}, and \ref{Fig:heatmapsub} in Section
+\ref{sec:micro}.  Here views of the parent and child models are
+synchronized, but the selected items in a grandparent model may not
+match the selected items in a grandchild model (as shown by Figures
+\ref{Fig:heatmap} and \ref{Fig:heatmapsub} where the selected genes are
+different in these two heatmaps).  Determining how \Robject{MVC} models are
+linked is a difficult task when the relationship between elements in the two
+models is not one-to-one.
+
+\subsection{Linking Models}
 \label{Ssec:MultLink}
@@
 
+The link functions, \Rfunction{toParent} and \Rfunction{fromParent}, can
+become quite complex depending on the relationship between the elements in the
+two models.  Suppose that the data elements in the two models have a
+one-to-one relationship where data element 1 in child model A is linked to
+data element 2 in parent model B.  Now if data element 1 in child model A is
+selected, then the \Rfunction{toParent} function will pass along the
+information that data element 2 in parent model B should be selected.  Here
+the linking is straightforward.  However, if the data elements in the two
+models have a one-to-many or many-to-many relationship, the decision about
+which linked elements to update is more complex.  Now suppose that data
+element 1 in child model A is linked to data elements 10, 11, and 12 in parent
+model B and data element 2 in child model A is linked to data elements 12 and
+13 in parent model B.  Here a many-to-many relationship exists between the
+data elements.  If the user selects element 1 in child model A, the
+\Rfunction{toParent} function must decide if elements 10, 11, and 12 should be
+selected in parent model B or if only elements 10 and 11 should be selected in
+parent model B (because element 12 in model B is also linked to element 2,
+which is not selected, in model A).
+
+These two options for the linking function can be represented as \textit{any}
+versus \textit{all} linking.  In \textit{any} linking, if any element is
+updated, then the corresponding linked element is updated.  This linking could
+also be represented as \textit{or} linking if one is thinking in terms of
+boolean operators.  In \textit{all} linking, the element in the second model
+is updated only if all linked elements are updated in the original model.
+Similarly, this linking could be also be represented as \textit{and} linking
+if one is thinking in terms of boolean operators.
+
+Making the previous example even more complex suppose that child model A also
+had data element 3 which is linked to data elements 12, 13, and 14 in parent
+model B.  Now if data elements 1 and 3 are selected in child model A, the
+\Rfunction{toParent} function can link based on one of these options: i) select
+data elements 10, 11, 12, 13, and 14 in parent model B, which indicates that
+any items in model B that are linked to elements 1 and 3 in model A are
+selected, ii) select data element 12, which indicates that only data elements
+that are linked to both elements 1 and 3 in model A are selected, or iii)
+select no data elements in model B, which indicates that only elements linked
+to both elements 1 and 3 and no other elements in model A are selected.
+Another consideration in the linking functions may be to change the response
+depending on the type of selection being performed.  For example, \textit{any}
+linking between models may be appropriate when an item in one model was
+highlighted, but if an element is hidden, then \textit{all} linking may be
+suitable.  This flexibility lets developers determine the appropriate
+relationships between models, but also requires thoughtful responses and
+potentially complex linking functions.
+
 The message classes for creating a child \Robject{MVC} from an existing
@@
   \begin{center}
-    \includegraphics[height=3.4in, width=4.5in]{newMessageClass2.jpg}
+    \includegraphics[height=4in, width=4.5in]{MVCMessageClass2.jpg}
     \caption{ Inheritance for Message Classes that are Passed Between
@@
 %this paper, including linked tables from a relational database, and
-%meta data that are linked to experimental data.  The MVC paradigm was
+%meta-data that are linked to experimental data.  The MVC paradigm was
 %discussed as a useful design for creating linked views of a data set.
@@
 function, which converts information that one model can understand into
-information that a linked model can understand.  If there is a one-to-one
-relationship between the elements of the two linked models, then the link
-function is straightforward:  if element E changes in model A and if element E
-in model A is linked to element F in model B, then the link function will
-tell model B that element F should be updated.  However, if there is a
-one-to-many relationship or a many-to-many relationship, then the link
-function may be more complex.  Suppose that element E changes in model A and
-element E is linked to elements G, H, I, and J in model C, but these elements
-in model C are also linked to other elements in model A (in other words, there
-is a many-to-many relationship between the elements in models A and C).  Now
-the link function must decide whether to tell elements G, H, I, and J in model
-C to be updated in response to a change in element E in model A.  The link
-function may think of these relationships as an `any' versus an `all'
-relationship.  In an `any' relationship, the link function tells elements G,
-H, I, and J in model C to be updated if any element they are linked from model
-A is updated.  In an `all' relationship, the link function tells elements G,
-H, I, and J in model C to be updated only if all elements they are linked to
-in model A are updated.  Deciding how to represent these relationships between
-elements in different models is left to the developer when creating the link
-functions.  While this flexibility lets the developer represent different
+information that a linked model can understand.  As discussed previously, the
+linking functions, \Rfunction{toParent} and \Rfunction{fromParent}, can be
+quite complex if the relationship between the data elements in the two models
+is one-to-many or many-to-many.  In response to an update in one model, the
+linking function must decide if all linked elements in the second model
+should be updated, if any linked elements should be updated, or if a subset of
+linked elements should be updated.  These decisions must be made by the
+developer to properly represent the relationship between the data elements in
+the two models.  While this flexibility lets the developer represent different
 relationships between different models, the developer is then responsible for
@@
 that can be linked to events.  However, in terms of extensibility, certain
-design decisions, such as expecting the data sets to be linked through a
-parent-child hierarchy, may not work with some data sets.  As an example, the
+design decisions, such as expecting the models to be linked through a
+parent-child hierarchy, may not work with some models.  As an example, the
 linked tables from a relational database are not related through a tree
 structure and thus, having parent and child \Robject{MVC} objects may not make
-sense with these data sets.  However, the design allows users to add new types
+sense with these models.  However, the design allows users to add new types
 of models, new types of views, and new message classes so there is still
-flexibility for additions as long as the data sets are connected through
+flexibility for additions as long as the models are connected through
 parent-child relationships.  These new class definitions could be created by
@@
 mouse movement events), and users can create child \Robject{MVC} objects
-through the GUI or through command line functions.  Using the functionality in
-the \Rpackage{iSNetwork} package, the user can create views of linked data sets
+through the GUI or through command line functions.  With the functionality in
+the \Rpackage{iSNetwork} package, the user can create views of linked models
 and can explore visually the relationships between variables in different data

### 2006-07-29 inst/paper/MVC.tex
fixed up a reference
git-svn-id: file:///home/git/hedgehog.fhcrc.org/bioconductor/trunk/madman/Rpacks/MVCClass@18995 bc3139a8-67e5-0310-9ffc-ced21a209358

@@
 corresponding component in all other views also changes appearance.
-In our example, correspondence would be measured through the sample,
+In our example, correspondence is determined by the sample,
 but there are other possibilities, as we shall explain later.  As an
@@
 
+%%FIXME: need to reference iPlots from Simon, and some of these refs need
+%%to be put up at the front where I have a citations in parens
 Linked views, which have a long history in statistics, are now
@@
 consider data stored in a relational database, such as the data shown in
-Figure \ref{FigDBTab}.  Here the database
+panel (a) of Figure~\ref{Fig:OneDBTab}.  Here the database
 consists of three tables, State, City and StateCity.  The State table
@@
 tables, in a standard MVC paradigm, it seems that either all data must
-be combined into a single large table (Figure \ref{Fig:OneDBTab}), or
+be combined into a single large table (panel (b) of Figure~\ref{Fig:OneDBTab}), or
 the processing and accessing of the model data must become much more

### 2006-07-30 inst/paper/MVC.tex
did some commits
git-svn-id: file:///home/git/hedgehog.fhcrc.org/bioconductor/trunk/madman/Rpacks/MVCClass@18999 bc3139a8-67e5-0310-9ffc-ced21a209358

@@
 
+%%Cite DataDesk as a system that does this, 
+%% also this book:  Andreas Buja, Catherine Hurley, and John Alan McDonald.
+Elements of a viewing pipeline for data analysis.
+In W.
+S.
+Cleveland and M.
+E.
+McGill, editors, Dynamic Graphics for Statistics.
+Wadsworth and Brooks/Cole, Belmont, Ca., 1988.
+
+%% also this paper: Interactive data visualization using focusing and linking
+%%Full text 	pdf formatPdf (541 KB)
+%%Source 	IEEE Visualization archive
+%%Proceedings of the 2nd conference on Visualization '91 table of contents
+%%San Diego, California
+%%SESSION: Multivariate visualization table of contents
+%%Pages: 156 - 163  
+%%Year of Publication: 1991
+%%ISBN:0-8186-2245-8 (PAPER) 
 There are many ways to visualize multivariate data when more than
 two or three dimensions are needed to view the data.  Many authors
-\cite{intGrUnwin} \cite{iPlots} \cite{MONDRIAN} \cite{GGobi}
+\citep{intGrUnwin, iPlots, MONDRIAN, GGobi}
 have suggested that linked, interactive views of the data
@@
 programming environment.
-
+%%FIXME: cite the JSS submission
 We had several objectives, or design criteria. They were that the
@@
     \caption{ Linked views of linked models.  When a user interacts with
-      the y versus x scatter plot, the first step, indicated by a circled one,
+      the scatter plot, the first step, indicated by a circled 1,
       is to tell model B to update.  Then all views of model B must be
@@
 consider data stored in a relational database, such as the data shown in
-panel (a) of Figure~\ref{Fig:OneDBTab}.  Here the database
+Figure~\ref{Fig:DBTab}.  Here the database
 consists of three tables, State, City and StateCity.  The State table
@@
 tables, in a standard MVC paradigm, it seems that either all data must
-be combined into a single large table (panel (b) of Figure~\ref{Fig:OneDBTab}), or
+be combined into a single large table (Figure~\ref{Fig:OneDBTab}), or
 the processing and accessing of the model data must become much more
@@
 application is a setting where one representation of the data is as a
-graph (with nodes and edges) while the other is as a matrix. We can
+graph (with nodes and edges) while the other is as a rectangular array. We can
 envisage lots of different MVCs that operate on graph based data, and
@@
 one view for the city-level data.  The state and city data are linked
-based on the color.  Even though there are two \Robject{MVC} objects
-here, one \Robject{MVC} for each model, the views are linked.
+based on the color.  But there are two distinct MVCs, not one. 
+The city data have one MVC and the state data a second.  Even though 
+there are two \Robject{MVC} objects here the views are linked.
 
@@
 
-In the MVC design, the model stores the data and among other things it
+In our MVC design, the model stores the data and among other things it
 supports all updating and other mechanisms that might be needed.
@@
 To make this addition, the user extends the \Rclass{gModel} class in his or
-her own package.
+her own R software package.
 
@@
 As for specific view classes, the \Rclass{spreadView} class represents
-a spreadsheet view of the data.  This view only makes sense for models
-that are a two dimensional data structure, such as a matrix or a data
-frame. Other views we have implemented include: scatter plots
+a spreadsheet view of the data.  This view requires input data that are
+inherently two-dimensional and for graph data, we make a reduction to the 
+list of nodes and some attributes.  Other views we have implemented 
+include: scatter plots
 (\Rclass{sPlotView}), graphs (\Rclass{graphView}), and heatmaps
@@
 relationship between the \Robject{MVC} instances and model objects.
-Because of this relationship, the \Robject{MVC} object and the model
-object are identified by the same name.
-
+%%FIXME: this seems like an implementation detail
+%%Because of this relationship, the \Robject{MVC} object and the model
+%%object are identified by the same name.
 Our implementation consists
@@
 view, and controller objects together, but it also shows
-that the \Robject{linkedModelMVC} objects are related to each other through a
+that the \Robject{linkedModelMVC} objects can be related to each other 
+through a
 parent-child (tree) hierarchy, as shown in Figure \ref{Fig:Hier}.  In this
-picture, `modelA' is the name of the original \Robject{MVC} object
+picture, modelA is the name of the original \Robject{MVC} object
 (model) and from it, two \Robject{MVC} objects were created,
-`modelB' and `modelC'. Then from `modelB', three
-\Robject{MVC} objects were created, `modelD', `modelE', and
-`modelF'.  As an example, modelA could be gene 
+modelB and modelC. Then from `modelB', three
+\Robject{MVC} objects were created, modelD, modelE, and
+modelF.  As an example, modelA could be gene 
 expression data, modelB could be gene expression data that is a subset of
@@
 new data has a one-to-one relationship with the elements in the current model,
-then the new data can be added to the current model.  For example, if there is
+then we believe that in most cases the new data should be added to the 
+current model.  For example, if there is
 a test statistic for each row in a data frame, then this could be added to the
 model.  However, if the new data has a one-to-many or a many-to-many
-relationship with the elements in the current model, then a submodel is
-generated in a new \Robject{MVC} instance.  An example here is the
+relationship with the elements in the current model, then serious consideration should be given to the creation of a submodel.
+If there is already an appropriate MVC class (e.g.
+if the submodel is a graph) then it can be
+generated by creating a new instance of the appropriate MVC class.  If there
+is no appropriate implementation, then the user will need to create one.  
+As an example, consider the
 relationship between genes and GO terms.  If the original model contains gene
 expression levels and a user wants to explore the GO terms for these genes,
-then the GO graph representing the GO terms will be placed in a new
-\Robject{MVC} because the genes and GO terms have a many-to-many relationship.
+then we encourage the creation of the GO graph representing the GO terms,  and
+creating a new graph \Robject{MVC} object.
+%%FIXME: maybe explicitly name the class in the above line
 
@@
 \ref{Fig:heatmap} and \ref{Fig:heatmapsub} where the selected genes are
-different in these two heatmaps).  Determining how \Robject{MVC} models are
-linked is a difficult task when the relationship between elements in the two
-models is not one-to-one.
+different in these two heatmaps).  Deciding how to link \Robject{MVC} models 
+is a difficult task especially when the relationship between elements in the 
+two models is not one-to-one.
 
@@
 
-The goals for this paper were to create linked views of linked data sets, to
+The goals for this research were to create software implementations and conceptual paradigms to support the use of linked views of linked data sets, to
 create interactive views where the response to an event can be changed, and to
@@
 
-The functionality that is defined in the \Rpackage{iSNetwork} package is
+The functionality that is provided in the \Rpackage{iSNetwork} package is
 discussed in detail in the Vignette for this R package.  Users can load data,

### 2006-08-09 inst/paper/MVC.tex
Added new versions of papers and added paper 2 version 2 (paper2vs.tex) that is my current version to give to the committee.
git-svn-id: file:///home/git/hedgehog.fhcrc.org/bioconductor/trunk/madman/Rpacks/MVCClass@19090 bc3139a8-67e5-0310-9ffc-ced21a209358

@@
 %%ISBN:0-8186-2245-8 (PAPER)
+
+%% EAW: put new citations below
 There are many ways to visualize multivariate data when more than
 two or three dimensions are needed to view the data.  Many authors
-\citep{intGrUnwin, iPlots, MONDRIAN, GGobi}
+\citep{intGrUnwin, DynGraphics, IEEEVisual, GGobi, DataDesk}
 have suggested that linked, interactive views of the data
@@
 %%to be put up at the front where I have a citations in parens
+%%EAW: added a reference to iPlots below
 Linked views, which have a long history in statistics, are now
@@
 software \cite{GGobi}.  Linked views have been implemented in several
-pieces of statistical software, such as XLisp-Stat \cite{Lisp} and
+pieces of statistical software, such as XLisp-Stat \cite{Lisp}, iPlots
+\cite{iPlots}, and
 GGobi \cite{GGobiMan}.  Implementing linked views is based on the
@@
     \includegraphics[height=5in, width=6in]{MultipleMVC.jpg}
-    \caption{ Example of Two \Robject{MVC} Objects.  These two \Robject{MVC}
+    \caption{ Example of two \Robject{MVC} objects.  These two \Robject{MVC}
       objects are linked by the StateCity table and the color of the points
@@
     \includegraphics[height=4in, width=5in]{newViewClass.jpg}
-    \caption{ Inheritance for View Objects. }
+    \caption{ Inheritance for view objects. }
     \label{Fig:View}
@@
 actions taken depend on the view, for example for a spreadsheet view
-the method may select or unselect a row on the spreadsheeet while for
+the method may select or unselect a row on the spreadsheet while for
 a scatter plot view the method may change the appearance of a point on
@@
 model.  However, if the new data has a one-to-many or a many-to-many
-relationship with the elements in the current model, then serious consideration should be given to the creation of a submodel.
-If there is already an appropriate MVC class (e.g.
-if the submodel is a graph) then it can be
-generated by creating a new instance of the appropriate MVC class.  If there
+relationship with the elements in the current model, then serious
+consideration should be given to the creation of a submodel.  If there is
+already an appropriate MVC class (e.g. if the submodel is a graph) then it can
+be generated by creating a new instance of the appropriate MVC class.  If there
 is no appropriate implementation, then the user will need to create one.  
@@
 %%FIXME: maybe explicitly name the class in the above line
+%% EAW: Not sure what you mean about naming the class: creating the GO graph
+%% creates an object of class graphModel that is placed in the model slot of a
+%% linkedModelMVC object.  As for naming the model (MVC), the user must decide
+%% the name.
 
@@
 
-The functionality that is provided in the \Rpackage{iSNetwork} package is
-discussed in detail in the Vignette for this R package.  Users can load data,
+The functionality that is provided in the \Rpackage{iSNetwork} package, which
+is an implementation of the research discussed in this paper, is
+discussed in detail in the vignette for this R package.  Users can load data,
 create views of the data, interact with these views and change the response to
@@
 
+\section*{Acknowledgements}
+We are indebted to Dr. Duncan Temple Lang for the creation of the
+\Rpackage{RGtk} and \Rpackage{gtkDevice} packages as well as helpful advice on
+their use.
+
 \bibliography{paper2}

### 2006-08-09 inst/paper/paper2.tex
Added new versions of papers and added paper 2 version 2 (paper2vs.tex) that is my current version to give to the committee.
git-svn-id: file:///home/git/hedgehog.fhcrc.org/bioconductor/trunk/madman/Rpacks/MVCClass@19090 bc3139a8-67e5-0310-9ffc-ced21a209358

@@
 \usepackage{graphicx}
+\usepackage{hyperref}
+\usepackage{times}
+\usepackage[authoryear,round]{natbib}
+\usepackage{subfigure}
 
@@
 \newcommand{\Rclass}[1]{\texttt{#1}}
+\renewcommand{\baselinestretch}{1.5}
 
@@
 
-\begin{titlepage}
-\begin{center}
-\Large
-\vspace*{1in}
-\textbf{Creating Linked, Interactive Views to \linebreak Explore Linked Data} \\
-\vspace{1in}
-\textbf{Dana-Farber Cancer Institute}\\ 
-\textbf{May 6, 2006} \\
-\vspace{1in} 
-\textbf{Elizabeth Whalen} \\
-\end{center}
-\end{titlepage}
-\normalsize
-\newpage
-
-%\renewcommand{\baselinestretch}{1.5}
+\author{Elizabeth Whalen\\Harvard School of Public Health}
+\title{Generalizing the Model View Controller Paradigm}
+
+\maketitle
+
+\begin{abstract} 
+  In this paper we consider a generalization of the model
+  view controller (MVC) paradigm to allow for more flexible handling
+  of submodels.  Instead of using a single MVC we develop a tree
+  structure of related MVCs, where each MVC has a single parent, but a
+  parent can spawn many different child MVCs.  Communication between
+  parent and child MVCs allows for very flexible linking of the
+  MVCs.  We provide an implementation of our work in the R package
+  \Rpackage{MVCClass} available at \url{http://www.bioconductor.org}. 
+\end{abstract}
+
 \section{Introduction}\label{Sec:Intro}
@@
 
-%\clearpage
-
 \begin{figure}[!h]
@@
 discussed further in Sections \ref{Sec:OneMVC} and \ref{Sec:MultMVC}.
-
-%\clearpage
 
@@
 between states and cities.
-
-%\clearpage
 
@@
 that there are advantages to storing the data in separate structures.
-
-%\clearpage
 
@@
 
-%\clearpage
-
 \begin{figure}[ht]
@@
 represents the data so for a \Robject{spreadView} object the
-\Rfunction{updateView} method may select or unselect a row on the spreadsheeet
+\Rfunction{updateView} method may select or unselect a row on the spreadsheet
 while for a \Robject{sPlotView} object the \Rfunction{updateView} method may
@@
   \begin{center}
-    \scalebox{0.6}{\includegraphics{newMessageClass2.jpg}}
+    \scalebox{0.6}{\includegraphics{MVCMessageClass2.jpg}}
+%    \scalebox{0.6}{\includegraphics{newMessageClass2.jpg}}
     \caption{ Inheritance for Message Classes that are Passed Between

### 2006-08-09 inst/paper/paper2v2.tex
Added new versions of papers and added paper 2 version 2 (paper2vs.tex) that is my current version to give to the committee.
git-svn-id: file:///home/git/hedgehog.fhcrc.org/bioconductor/trunk/madman/Rpacks/MVCClass@19090 bc3139a8-67e5-0310-9ffc-ced21a209358

--- 2006-08-09 inst/paper/paper2.tex
@@
 
-\subsection{Background}\label{Ssec:Backg}
-
-There are many ways to visualize multivariate data where more than
-two or three dimensions are needed to view the data.  Several of these ideas
-were discussed in previous work \cite{EW05}, so the focus in this paper
-is on one specific method of visualizing multivariate data that is
-particularly useful for exploratory data analysis; creating linked,
-interactive views of linked data sets.
+%\subsection{Background}\label{Ssec:Backg}
+
+%There are many ways to visualize multivariate data where more than
+%two or three dimensions are needed to view the data.  Several of these ideas
+%were discussed in previous work \cite{EW05}, so the focus in this paper
+%is on one specific method of visualizing multivariate data that is
+%particularly useful for exploratory data analysis; creating linked,
+%interactive views of linked data sets.
+
+There are many ways to visualize multivariate data when more than
+two or three dimensions are needed to view the data.  Many authors
+\citep{intGrUnwin, DynGraphics, IEEEVisual, GGobi, DataDesk}
+have suggested that linked, interactive views of the data
+provide a valuable and important set of tools for comprehending such
+data.  The model view controller (MVC) paradigm is now a standard
+mechanism for implementing such systems.  The paradigm has proven very
+effective and allows for clear separation of different components.
+In this paper we consider an extension of the MVC paradigm to allow
+for, what we believe to be, an easy expansion to linked data sets.
+
+The sorts of problems we are interested in, are those where a user
+might begin with a large data set, identify some interesting subset
+and want to create interactive views of that subset.  In some cases the
+natural data representation of the subset is in a form other than that
+of the initial data set.  This leads to complications in the usual MVC
+paradigm, where the data (i.e. the model) are often presumed to be in
+a single, standard format.  We found that adding new views under such
+conditions could be quite problematic and often led to very complex
+code.  As we shall demonstrate, there are also issues that arise due to
+mappings that are not one-to-one between elements on different views.
+Hence we considered an approach that created new instances of
+the \Robject{MVC} that were linked to the parent that spawned them.
+
+In particular, our implementation consists of a hierarchy of \Robject{MVCs},
+stored as a tree.  Each parent can have multiple children, but the
+children have a single parent.  Communication between parent and
+child is bidirectional, and a variety of different filters can be used
+to control flow.  Thus, we can devise a number of purpose built \Rclass{MVC}
+classes, each with a model appropriate to the specific data of
+interest, that can be linked in a natural way.  Interacting with the view
+in a child \Robject{MVC}, first sends a message to the child \Robject{MVC},
+which will update all of its views, but the message will also propagate to the
+parent \Robject{MVC} (and further up to the root).  The parent \Robject{MVC}
+can choose whether to react to messages from its children -- all control being
+in the hands of the users, should they want it.
+
+\subsection*{A motivating example}
+\label{sec:micro}
+
+Our development of this model was driven primarily by an interest in
+analyzing high throughput biological data.  Such data tend to be large
+and very complex, and hence ideally suited to interactive
+visualization.  In order to provide the reader with some idea of the
+sorts of interactions we would like to support, we provide here a short
+description of a type of experiment and some of the likely
+interactions that would be of interest to a data analyst.
+
+A gene expression experiment provides data on the expression of mRNA
+abundance on thousands to tens of thousands of genes for some number
+of samples.  Usually there are many fewer samples than genes.  There is
+much that is known about the genes, such as what chromosome they are
+located on and what sorts of activities they are engaged in, and in many
+cases we want to restrict attention to a specific subset of genes, or
+to alter the view so that rather than being gene centric it becomes
+centered on some biological aspect, such as chromosome location.
+
+One particularly rich data source is the Gene Ontology (GO), \cite{GO}
+which provides three controlled vocabularies that are represented as
+directed acyclic graphs, with a root node.  Terms typically have a
+child-parent relationship, with the child terms being more specific
+and the parent terms more general.  The root node represents the most
+general term.  Genes are then annotated at specific nodes in these
+graphs where the node is determined, by a domain expert, to describe a
+property of the gene.  A gene is then also associated with all parents,
+since if the specific term applies so must all more general ones.
+
+There is often substantial interest in relating gene expression data
+to such a graph based model, or to chromosomal locations, or to
+activity in known biological pathways. Data mappings for this latter
+aspect can be obtained from different sources such as \cite{KEGG}.
+Further, users typically want to begin with a heatmap view. Then from
+that derive a GO graph view, and additionally a chromosomal location
+view. In order to interact with the GO graph, the primary unit is the
+node in that graph (and sometimes one expects with the edges), while
+for the heatmap view the basic element is generally the gene.  In Figures
+\ref{Fig:heatmap} and \ref{Fig:GOgraph}, a heatmap view of gene expression
+from Chronic Lymphocytic Leukemia (CLL) patients and the corresponding GO
+graph are shown, respectively.  By interacting with the heatmap to color one
+gene blue (shown as gene PDE8A on the heatmap), which causes the corresponding
+GO nodes to be colored blue, one can easily see how this gene relates to the
+GO terms.  Also by interacting with the node with an arrow next to it in
+Figure \ref{Fig:GOgraph}, which represents the signal transducer activity GO
+term, one can create a heatmap view that shows only the genes annotated at
+that node, as shown in Figure \ref{Fig:heatmapsub}.  Notice how all genes
+annotated at the signal transducer activity GO term have a blue label in
+Figure \ref{Fig:heatmapsub}, which shows that this view was derived from a
+blue colored node in Figure \ref{Fig:GOgraph}.  Compare the heatmaps in
+Figures \ref{Fig:heatmapsub} and \ref{Fig:heatmap}.  The heatmap in Figure
+\ref{Fig:heatmapsub} is just a subset of the heatmap in Figure
+\ref{Fig:heatmap}, however in Figure \ref{Fig:heatmapsub} eleven genes have a
+blue label and in Figure \ref{Fig:heatmap} only one gene has a blue label.
+This difference in these heatmaps indicates that views are synchronized based
+on the parent-child relationships.  Figure \ref{Fig:heatmap} is the parent of
+Figure \ref{Fig:GOgraph}, which is the parent of Figure \ref{Fig:heatmapsub}.
+Figures \ref{Fig:heatmap} and \ref{Fig:GOgraph} are synchronized, and Figures
+\ref{Fig:GOgraph} and \ref{Fig:heatmapsub} are synchronized, but Figures
+\ref{Fig:heatmap} and \ref{Fig:heatmapsub} are not synchronized by color
+because they do not have a parent-child relationship: Figure \ref{Fig:heatmap}
+is the grandparent of Figure \ref{Fig:heatmapsub}.
+
+\begin{figure}[ht]
+  \begin{center}
+    \includegraphics[height=5in, width=5in]{heatmap.jpg}
+    \caption{ Heatmap view of gene expression from Chronic Lymphocytic
+      Leukemia (CLL) patients. }
+    \label{Fig:heatmap}
+  \end{center}
+\end{figure}
+
+\begin{figure}[ht]
+  \begin{center}
+    \includegraphics[height=4.3in, width=4.3in]{GOgraph2.jpg}
+    \caption{ Gene Ontology (GO) graph that is associated with the genes
+      displayed in the heatmap in Figure \ref{Fig:heatmap}.  Notice that one
+      gene (PDE8A) is colored blue in the heatmap and all of the corresponding
+      GO terms are colored blue in the graph.  The arrow next to one of the
+      nodes indicates that this node was used to create the view shown in
+      Figure \ref{Fig:heatmapsub}.  The node with the arrow next to it
+      represents the signal transducer activity GO term. }
+    \label{Fig:GOgraph}
+  \end{center}
+\end{figure}
+
+\begin{figure}[ht]
+  \begin{center}
+    \includegraphics[height=5in, width=5in]{heatmapsub.jpg}
+    \caption{ A heatmap view that shows only the genes that are annotated at
+      the node in Figure \ref{Fig:GOgraph} that has an arrow next to it.  This
+      node represents the GO term, signal transducer activity, and as shown in
+      this heatmap, eleven genes are annotated at this node.  Notice that all
+      the labels in this heatmap are colored blue, indicating that the parent
+      node that was used to create this view was also colored blue.  
+     }
+    \label{Fig:heatmapsub}
+  \end{center}
+\end{figure}
+
+\subsection*{Overview}
+\label{Ssec:Over}
+
+We found that constructing a single MVC that would encompass all
+anticipated views was very cumbersome. It would also be inefficient
+since a MVC model for the GO graph, or for a chromosome view would
+also be of substantial interest, regardless of whether there was an
+initial heatmap view or not. Thus, we looked for a solution that was
+based more on component software constructed within an object oriented
+programming environment.
+%%FIXME: cite the JSS submission
+We had several objectives, or design criteria. They were that the
+software should be extensible in a high level language. We chose R as
+the language for our implementation and wanted to allow users that
+knew no other languages to be able to extend and enhance the
+system.  Therefore, we built our software on top the \Rpackage{RGtk} package of
+D. Temple Lang, since it provides such an interface. We also vastly
+preferred Gtk+ over Tcl/Tk since its widgets are more sophisticated,
+which we found results in a much shorter development time.
+
+We wanted to provide a system that was usable and extensible by
+relatively naive end users and hence have provided a graphical user
+interface (GUI) which they can use to both carry out an analysis, but
+also supports a variety of extensions, such as changing the actions
+performed in response to events. We also wanted to support a more
+sophisticated developer who might want to take our basic infrastructure
+and create an interactive system, say for the interactive analysis of
+time series data, using our design. Hence, all commands are
+available both through the GUI and through command line functions.
+
+\section{Methods}
+\label{Sec:Methods}
 
@@
 have been implemented in several pieces of statistical software, such as
-XLisp-Stat \cite{Lisp} and GGobi \cite{GGobiMan}.  Implementing linked views
+XLisp-Stat \cite{Lisp}, iPlots
+\cite{iPlots}, and GGobi \cite{GGobiMan}.  Implementing linked views
 is based on the model-view-controller (MVC) design \cite{DesignPatterns},
@@
     \scalebox{0.55}{\includegraphics{Event.jpg}}
-    \caption{ Flow Chart of the Response to an Event. }
+    \caption{ Flow chart of the response to an event. }
     \label{Fig:Event}
@@
 considered a data set and they are linked through the keys.  Another situation
-where linked data sets occur is when there is experimental or study data that
+where linked data occur is when there is experimental or study data that
 are linked to meta-data.  An example of this situation is given in Section
@@
 
-As a motivating example of linked views based on linked data sets, consider
-Figure \ref{Fig:firstMP}.  In this figure, there are two views, one view of
-data set 1 and one view of data set 2.  Data set 1 and data set 2 are linked.
-When a user initially interacts with the scatter plot view of data set 2, the
-response to this interaction is to tell data set 2 to update itself (this is
-indicated on the figure as a circled 1).  Then in Step 2, data set 2 tells the
-scatter plot view that is should update itself in response to the change in the
-data.  Next, in Step 3, because data set 2 has changed and because it is
-linked to data set 1, data set 2 sends information to data set 1 about the
-change.  In Step 4, the information from data set 2 is converted into
-information that data set 1 can understand.  In Step 5, this information is
-received by data set 1 and data set 1 updates its data.  Finally, in Step 6,
-data set 1 tells its view to update itself.  This example shows the components
+Consider Figure \ref{Fig:firstMP}, which provides a cartoon
+description of the steps that occur when a user interacts with a
+view.  In the cartoon there are only two views, one of model A (note 
+that once a data set is loaded in a MVC, we call it a model) and the
+other of the linked model B.  When a user interacts with the scatter plot view
+of model B, the response to this interaction is for model B to update itself
+(this is indicated on the figure as a circled 1).  In Step 2, the scatter plot
+view is told to update itself.  Next, in Step 3, because model B is linked
+to model A, model B sends a message to model A.  In Step 4,
+the information from model B is converted into information that
+model A understands.  In Step 5, this information is received by
+model A and model A updates itself.  Finally, in Step 6, model
+A tells its view to update itself.  This example shows the components
 and the communication between the components that are necessary to create
-linked views of underlying linked data sets.
+linked views of underlying linked data.
 
@@
     \scalebox{0.55}{\includegraphics{firstMessagePassing.jpg}}
-    \caption{ Linked views of linked Data Sets.  When a user interacts with
-      the y versus x scatter plot, the first step, indicated by a circled one,
-      is to tell data set 2 to update.  Then all views of data set 2 must be
-      updated.  Next any data that are linked to data set 2 must be notified
-      that data set 2 changed.  This information from data set 2 must be
-      altered so that it can be understood by linked data sets, which is shown
-      in the fourth step.  Finally, data set 1 is notified that it should
-      update in response to the change in data set 2 and then all views of
-      data set 1 are updated. }
+    \caption{ Linked views of linked models.  When a user interacts with
+      the scatter plot, the first step, indicated by a circled 1,
+      is to tell model B to update.  Then all views of model B must be
+      updated.  Next any data that are linked to model B must be notified
+      that model B changed.  This information from model B must be
+      altered so that it can be understood by linked models, which is shown
+      in the fourth step.  Finally, model A is notified that it should
+      update in response to the change in model B and then all views of
+      model A are updated. }
     \label{Fig:firstMP}
@@
 that it expects one data set, which limits the type of data we can visualize.
-There are many situations where multiple data sets occur, particularly any
-time there are experimental or study data linked to meta-data.  Being able to
-visualize data from multiple data sets is helpful to understand the
-relationships between variables.  Thus, after discussing the limitations of
+Thus, after discussing the limitations of
 the MVC design in Section \ref{Ssec:Limit}, an extension of the MVC design is
@@
 linked \Robject{MVC} objects.  The design of individual \Robject{MVC} objects
-is discussed in 
-Section \ref{Sec:OneMVC} while the implementation of linked \Robject{MVC}
-objects is discussed in Section \ref{Sec:MultMVC}.  The goals of the software
-discussed in this paper are to create linked views of linked data sets, to
-create interactive views where the response to an event can be changed, and to
-create an extensible design that allows future users to make additions.
-
-\subsection{Design and Implementation of Linked, Interactive
-  Views}\label{Ssec:Design}
+is discussed in Section \ref{Sec:OneMVC} while the implementation of linked
+\Robject{MVC} objects is discussed in Section \ref{Sec:MultMVC}.  The goals of
+the software discussed in this paper are to create linked views of linked data
+sets, to create interactive views where the response to an event can be
+changed, and to create an extensible design that allows future users to make
+additions.
+
+\subsection{Design and Implementation of Linked, Interactive Views}
+\label{Ssec:Design}
 
@@
 paradigm is powerful because it decouples the views from the model.  Thus,
-only one copy of the data needs to be stored and then all views refer to this
+only one copy of the data need to be stored and then all views refer to this
 data set.  With all views based on the same data, any changes to the data can
@@
     \scalebox{0.5}{\includegraphics{ExMVC.jpg}}
-    \caption{ A Model with its Views. }
+    \caption{ A model with its views. }
     \label{Fig:ExMVC}
@@
 communication between components include the model notifying views that the
-data have changed, a view asking the controller how it should respond to user
-interaction, and a view telling the model that the data should change in
-response to user interaction with the view.  This message passing between
-components in the MVC paradigm is discussed in detail in Section
-\ref{Ssec:OneMess}.
-
-In \cite{EW05}, we discuss the creation of R software to fulfill three goals:
-to create views that are linked, to create views that are interactive, and to
-create a design that is extensible.  To satisfy the overall goals of the
-design, two R packages, called \Rpackage{MVCClass} and \Rpackage{iSPlot},
-were created.  The \Rpackage{MVCClass} package implements the class and
-generic function definitions needed for the MVC design and the
-\Rpackage{iSPlot} package creates the front end to let the user create linked,
-interactive views of two-dimensional data, such as data frames and matrices,
-through a GUI or through command line functions.  The reason for
-separating the software into two packages was that it allowed the class
-definitions in the \Rpackage{MVCClass} package to be used by any other R
-package that wanted to create linked views.
-
-\subsection{Limitations of the MVC Design}\label{Ssec:Limit}
+data have changed, the controller asking the view which object on the plot the
+user just interacted with, and the controller telling the model that the data
+should change in response to user interaction with the view.  This message
+passing between components in the MVC paradigm is discussed in detail in
+Section \ref{Ssec:OneMess}.
+
+In \cite{EW05}, we discuss the creation of \textsf{R} software to fulfill
+three goals: to create views that are linked, to create views that are
+interactive, and to create a design that is extensible.  To implement the
+overall goals of the design, two \textsf{R} packages, called
+\Rpackage{MVCClass} and \Rpackage{iSPlot}, were created.  The
+\Rpackage{MVCClass} package implements the class and generic function
+definitions needed for the MVC design and the \Rpackage{iSPlot} package
+creates the front end to let the user create linked, interactive views of
+two-dimensional data, such as data frames and matrices, through a GUI or
+through command line functions.  The reason for separating the software into
+two packages was that it allowed the class definitions in the
+\Rpackage{MVCClass} package to be used by any other \textsf{R} package that
+wanted to create linked views.
+
+\subsection{Limitations of the MVC Design}
+\label{Ssec:Limit}
 
@@
     \scalebox{0.5}{\includegraphics{newDatabaseTables.jpg}}
-    \caption{ Example of Relational Database Tables. }
+    \caption{ Example of relational database tables. }
     \label{Fig:DBTab}
@@
 prone and requires less memory to store the data.  Another consideration in
-combining data sets is that it may give the appearance of a relationship
+combining data is that it may give the appearance of a relationship
 between variables that does not exist.  For example, it may not make sense to
@@
     \scalebox{0.5}{\includegraphics{newOneDatabaseTable.jpg}}
-    \caption{ Example of a Combined Table. }
+    \caption{ Example of a combined table. }
     \label{Fig:OneDBTab}
@@
 In a relational database all of the data are stored in tables, but consider the
-situation where there are multiple data sets that are of different data
+situation where there are linked data that are of different data
 structures.  For example, suppose that you have experimental microarray data,
@@
 directed acyclic graph, describing the molecular functions carried out by
-these genes.  These two data sets (the experimental data and the meta-data)
+these genes.  The experimental data and the meta-data
 are linked through the genes.  Each gene can have more than one molecular
@@
 that function so there is a many-to-many relationship between the genes and
-the molecular functions.  Though these two data sets are linked, the
+the molecular functions.  Though the data are linked, the
 microarray data and the Gene Ontology data are different data structures.
 
-Now for the original MVC design, where one data set is expected, it is not
-clear how to combine these two data sets into one data set.  The meta-data
-graph could be stored as a matrix, but a graph is a data structure that can be
-well represented as an object with slots for nodes and edges, especially if
-the graph is sparse.  As with relational database tables, combining the data
-sets complicates data storage, alteration, and retrieval.  For example, by
-combining the data sets a structure may be forced on the data that may not fit
-well, especially if the original data sets are instances of different classes.
-In contrast, keeping the data sets stored separately in their original
+Now for the original MVC design, where one grouping of the data is expected,
+it is not clear how to combine the experimental data and meta-data.  The
+meta-data graph could be stored as a matrix, but a graph is a data structure
+that can be well represented as an object with slots for nodes and edges,
+especially if the graph is sparse.  As with relational database tables,
+combining the data complicates data storage, alteration, and retrieval.  For
+example, by combining the data, a structure may be forced on the data that may
+not fit well, especially if the original data are instances of different
+classes.  In contrast, keeping the data stored separately in their original
 structures simplifies the processes of storing, updating, and accessing the
@@
 
-\section{Extending the MVC Design}\label{Sec:Extend}
+\section{Extending the MVC Design}
+\label{Sec:Extend}
 
 As discussed in Section \ref{Sec:Intro}, the MVC design typically uses a single
-conceptual unit, such as a data frame or a matrix, for the model.  In Section
-\ref{Ssec:Limit} we provided several examples where it is advantageous to link
-multiple data sets.  These examples included linked tables in relational
-databases and experimental data linked to meta-data.  Being able to visualize
-variables simultaneously from different data sets that are linked is useful for
-exploring relationships between these variables.  We next describe how the MVC
-design can be revised and extended to accommodate multiple data sets.
-
-\subsection{One Large Data Set}\label{Ssec:OneDS}
-
-One possibility for dealing with multiple data sets within the MVC design is to
-combine the multiple data sets into one large data set.  With this new
-combined data, the original MVC design can be used to create linked views.
-
-Unfortunately, as discussed in Section \ref{Ssec:Limit}, combining
-multiple data sets into one large data set can be problematic.  Even if the
-multiple data sets have the same data structure, combining the data sets often
-results in redundant data and makes updating and representing the
-data more complicated.  In the situation where the data sets do not have the
-same structure, it may be necessary to define a new class to appropriately
-represent the combined data.  Also, combining the data into one data structure
-may give the appearance that some variables have a relationship that does not
-exist.  Just because two data sets are linked does not necessarily mean that
-there is a meaningful relationship between their data.  Thus, another way
-of linking views based on multiple data sets needs to be designed.
-
-\subsection{Multiple MVC Objects}\label{Ssec:MMVC}
-
-A more straightforward approach may be to link \Robject{MVCs}.  Each
-\Robject{MVC} is a coherent object and the interactions between \Robject{MVCs}
-are encapsulated in a set of messages.  We use one \Robject{MVC} object to
-handle each data set.  Thus, within a \Robject{MVC} object, one can create
-linked views of that particular data set, using the standard MVC design.  But
-between \Robject{MVC} objects we design a second messaging system so that a
-change in one data set is sent to the other linked data sets.
+conceptual unit, such as a data frame or a matrix, for the model.  And
+we believe that such an approach is a good one.  MVCs should be
+specialized to certain types of data, as this allows them to provide
+better and more natural interactions.  However, we additionally argue
+that there is great benefit to considering linking these different
+MVCs when the underlying data themselves are derived from the same source.
+We believe that there will often be a hierarchical relationship
+between the models and the MVCs that contain them.  We have found
+that allowing the user to create new MVCs based on a current MVC is an
+important and useful paradigm.
+
+Each \Robject{MVC} is a coherent object and the interactions between
+\Robject{MVCs} are encapsulated in a set of messages.  We use one
+\Robject{MVC} object to handle each model.  Thus, within a
+\Robject{MVC} object, one can create linked views of that particular
+model, using the standard MVC design.  Between \Robject{MVC}
+objects we propose a second messaging system that allows messages to
+be sent between linked models.
+
+%In Section
+%\ref{Ssec:Limit} we provided several examples where it is advantageous to link
+%multiple data sets.  These examples included linked tables in relational
+%databases and experimental data linked to meta-data.  Being able to visualize
+%variables simultaneously from different data sets that are linked is useful
+%for exploring relationships between these variables.  We next describe how
+%the MVC design can be revised and extended to accommodate multiple data sets.
+
+%\subsection{One Large Data Set}\label{Ssec:OneDS}
+%
+%One possibility for dealing with multiple data sets within the MVC design is
+%to combine the multiple data sets into one large data set.  With this new
+%combined data, the original MVC design can be used to create linked views.
+%
+%Unfortunately, as discussed in Section \ref{Ssec:Limit}, combining
+%multiple data sets into one large data set can be problematic.  Even if the
+%multiple data sets have the same data structure, combining the data sets often
+%results in redundant data and makes updating and representing the
+%data more complicated.  In the situation where the data sets do not have the
+%same structure, it may be necessary to define a new class to appropriately
+%represent the combined data.  Also, combining the data into one data structure
+%may give the appearance that some variables have a relationship that does not
+%exist.  Just because two data sets are linked does not necessarily mean that
+%there is a meaningful relationship between their data.  Thus, another way
+%of linking views based on multiple data sets needs to be designed.
+
+%\subsection{Multiple MVC Objects}\label{Ssec:MMVC}
 
@@
     \end{tabular}
-    \caption{State and City Values.}\label{Tab:CityState}
+    \caption{State and city values.}\label{Tab:CityState}
   \end{center}
@@
 \normalsize
-As an example, suppose that one has two data sets: one with information at the
+As an example, suppose that one has two models: one with information at the
 state level and one with information at the city level.  These tables are
@@
 High Temperature in January (in Fahrenheit from 2005)
-data.  These two data sets are linked by the StateCity table that links cities
-with the states they reside in.  Now for these two data sets there are
+data.  These two models are linked by the StateCity table that links cities
+with the states they reside in.  Now for these two models there are
 two \Robject{MVC} objects.  One \Robject{MVC} object for the State data and
@@
     \scalebox{0.7}{\includegraphics{MultipleMVC.jpg}}
-    \caption{ Example of Two \Robject{MVC} Objects.  These two \Robject{MVC}
+    \caption{ Example of two \Robject{MVC} objects.  These two \Robject{MVC}
       objects are linked by the StateCity table and the color of the points
@@
 points referring to Minnesota are green.  Even though there are two
-\Robject{MVC} objects here, with one \Robject{MVC} object for each data set,
-it is still clear that the views are linked through the data sets and that is
+\Robject{MVC} objects here, with one \Robject{MVC} object for each model,
+it is still clear that the views are linked through the models and that is
 shown in this example through the coloring.
@@
 
-\section{Design Details of One MVC Object}\label{Sec:OneMVC}
-
-\subsection{Implementation Details}\label{SSec:OneOver}
+\section{Design Details of One MVC Object}
+\label{Sec:OneMVC}
+
+\subsection{Implementation Details}
+\label{SSec:OneOver}
 
@@
 package implements the design of the MVC classes and generic functions.  The
-\Rpackage{MVCClass} package defines the individual model,
-view, and controller classes as well as the \Rclass{MVC}
-class that binds the individual model, view, and
-controller classes together into one unit.  The package also defines
-the message classes that allow the components to communicate with
-each other.  Thus, the \Rpackage{MVCClass} package contains the classes
-necessary for creating linked views.   The goal of placing these class and
-generic function definitions into one package is that they are available for
-other R packages that implement linked views of data, such as the
-\Rpackage{iSPlot} and the \Rpackage{iSNetwork} packages.
-
-With the class definitions created in the \Rpackage{MVCClass} package, the
-\Rpackage{iSNetwork} package can decide which classes from \Rpackage{MVCClass}
+\Rpackage{MVCClass} package defines the individual model, view, and controller
+classes as well as the \Rclass{MVC} class that binds the individual model,
+view, and controller classes together into one unit.  The \Rpackage{MVCClass}
+package also defines the message classes that allow the components to
+communicate with each other.  Thus, the \Rpackage{MVCClass} package contains
+the classes necessary for creating linked views.  Also, the
+\Rpackage{BioMVCClass} package creates new model and view classes that extend
+the classes available in the \Rpackage{MVCClass} package.  The goal of placing
+these class and generic function definitions in other packages is that they are
+available for other \textsf{R} packages that implement linked views of data,
+such as the \Rpackage{iSPlot} and the \Rpackage{iSNetwork} packages.
+
+With the class definitions created in the \Rpackage{MVCClass} and the
+\Rpackage{BioMVCClass} packages, the
+\Rpackage{iSNetwork} package can decide which classes
 to use based on the overall goals of the combined software.  The
@@
 To create the GUI, \Rpackage{iSNetwork} uses the R packages,
-\Rpackage{RGtk} and \Rpackage{gtkDevice}.  \Rpackage{RGtk} is an R package
-that allows one to interact with Gtk functions from the R interface.  Gtk is
-an open-source X Window toolkit \cite{Gtk} for creating user interfaces.  The
-\Rpackage{gtkDevice} package creates a Gtk device by turning a Gtk drawing
-area into a device that acts like an R device, but still responds to events so
-that the view can be interactive.  For more information about these packages,
-see \cite{EW05}.
-
-The next subsections explain the class definitions and the methods that
+\Rpackage{RGtk} and \Rpackage{gtkDevice}.  \Rpackage{RGtk} is an \textsf{R}
+package that allows one to interact with Gtk+ functions from the \textsf{R}
+interface.  Gtk+ is an open-source window toolkit \cite{Gtk} for creating user
+interfaces.  The \Rpackage{gtkDevice} package creates a Gtk+ device by turning
+a Gtk+ drawing area into a device that acts like an \textsf{R} device, but
+still responds to events so that the view can be interactive.  For more
+information about these packages, see \cite{EW05}.
+
+The next sections explain the class definitions and the methods that
 pertain to a single \Robject{MVC} object.  Thus, the model classes,
@@
 three components in one \Robject{MVC} object are discussed in the following
-subsections.  All of the class definitions are implemented in the
-\Rpackage{MVCClass} package and all of the methods are defined in the
-\Rpackage{iSNetwork} package.  Message class definitions that pertain
-to communication between the \Robject{MVC} objects are discussed in Section
-\ref{Sec:MultMVC}.
-
-\subsection{Model}\label{Ssec:OneModel}
+sections.  All of the class definitions are implemented in the
+\Rpackage{MVCClass} and \Rpackage{BioMVCClass} packages and all of the methods
+are defined in the \Rpackage{iSNetwork} package.  Message class definitions
+that pertain to communication between the \Robject{MVC} objects are discussed
+in Section \ref{Sec:MultMVC}.
+
+\subsection{Model}
+\label{Ssec:OneModel}
 
@@
 classes.  Here the virtual class, \Rclass{gModel}, has the slots
-\Rslot{modelData}, \Rslot{linkData}, \Rslot{virtualData}, and
-\Rslot{modelName}.  These four slots are common information that all models
+\Rslot{modelData}, \Rslot{linkData}, \Rslot{virtualData}, \Rslot{modelName},
+and \Rslot{modelVar}.  These five slots are common information that all models
 need to store.  The \Rslot{modelData} slot is the data set for this model, the
@@
 pertaining to the views that needs to be stored with the model so that it can
-be shown in all views of this model, and the \Rslot{modelName} slot is the
-name of the model.
+be shown in all views of this model, the \Rslot{modelName} slot is the
+name of the model, and \Rslot{modelVar} is a list of any model variables,
+which are extra data that are stored with the model, such as test statistics
+calculated from the model data.
 
@@
     \scalebox{0.7}{\includegraphics{ModelClass.jpg}}
-    \caption{ Inheritance for Model Objects. }
+    \caption{ Inheritance for model classes. }
     \label{Fig:Model}
@@
 
-\subsection{Views}\label{Ssec:OneViews}
-
-The view classes represent the visual depictions of the model.  All
-views need to store some common information and respond to certain events
-through methods.  This consideration leads to an object model where the
-different view classes inherit from a view virtual class,
-called \Rclass{genView}.  The object model for the view classes is
-shown in Figure \ref{Fig:View}.
+\subsection{Views}
+\label{Ssec:OneViews}
+
+The view classes represent the visual depictions, or renderings, of
+the model.  There are some data that are required by all views, and all views
+must respond to a certain subset of events so we have implemented the view
+classes by first creating a single, virtual class which the specialized
+views extend.  As with our model classes, this allows us to define
+methods and data storage once, for the virtual class, and only extend
+it for views that specifically require additional data storage, or
+specialized handling.  The object model for the view classes is shown
+in Figure \ref{Fig:View}.
 
@@
     \scalebox{0.7}{\includegraphics{newViewClass.jpg}}
-    \caption{ Inheritance for View Objects. }
+    \caption{ Inheritance for view classes. }
     \label{Fig:View}
@@
 and \Rslot{winNum}.  \Rslot{dataName} is the name of the model that the view
-displays, \Rslot{win} is the Gtk window object that holds the view, and
+displays, \Rslot{win} is the Gtk+ window object that holds the view, and
 \Rslot{winNum} is the number of the window so that the window can be
@@
 frame.  The information needed for this view is stored in the slot,
-\Rslot{clist}, which is the Gtk spreadsheet object.
+\Rslot{clist}, which is the Gtk+ spreadsheet object.
 
@@
 and \Rslot{drArea}, which store the device number of the plot, the plotting
-parameters, and the Gtk drawing area object, respectively.  This class is a
+parameters, and the Gtk+ drawing area object, respectively.  This class is a
 virtual class as it represents a general plot.  The instantiable plot classes
@@
 holds a \Robject{Ragraph} object that represents the layout of the graph plot.
-The \Rclass{heatmapView} represents a heatmap
-view of the data and it contains the slot, \Rslot{ordering}, which is a list
-that is returned from the \Rfunction{heatmap} function to give information
-about the ordering in the row and column dendrograms.  Note that some of these
+The \Rclass{heatmapView} represents a heatmap view of the data and it contains
+the slots, \Rslot{ordering} and \Rslot{rNames}, where \Rslot{ordering} is a
+list that is returned from the \Rfunction{heatmap} function to give
+information about the ordering in the row and column dendrograms and
+\Rslot{rNames} are the row names that were used to create the heatmap in case
+a subset of the data was used to create the view.  Note that some of these
 views are only applicable for certain model types.  For example, the
@@
 
-Since all views are expected to be interactive, all view classes have methods
-that correspond to user interaction.  Interactive views mean a response
-occurs when the view has focus and an event occurs.  The events we are most
-interested in responding to include key press
-events, where a certain configuration of keys are pressed; button click
-events, where one of the mouse buttons is clicked over the view; and mouse
-movement events, where the cursor moves over the view.  Recall from Section
-\ref{Ssec:Backg} that when an event occurs Gtk emits a signal that is caught
-by the signal handler.  The signal handler then calls the callback function
-and it is in the callback function that the corresponding method is called.
-For example, when the button click event occurs on a scatter plot view, the
-callback function calls the scatter plot's \Rfunction{clickEvent} method.
-Using the information that is passed to the \Rfunction{clickEvent} method in
-the \Rslot{where} parameter, the method converts the Gtk pixel coordinates
-that indicate where the event occurred into user coordinates and then into an
-object from the plot, such as a point on the plot.
-
-For \Rclass{sPlotView}, the \Rfunction{clickEvent} method is called when Gtk
-emits a signal that a mouse button has been pressed and the
-\Rfunction{motionEvent} method is called when Gtk emits a signal that
-the cursor has been moved over the view.  Both the \Rfunction{clickEvent} and
-\Rfunction{motionEvent} methods for a \Robject{sPlotView} object
-determine which point on the plot was clicked or under the cursor,
-respectively, using the data in the \Rslot{where} parameter.  For
-\Rclass{spreadView}, the \Rfunction{clickEvent} method is called when Gtk
-emits a signal that a row in the spreadsheet has been selected or unselected
-and the \Rfunction{clickEvent} method determines which row was changed.
-For \Rclass{graphView}, the \Rfunction{clickEvent} method is called when Gtk
-emits a signal that a mouse button has been pressed and the
-\Rfunction{motionEvent} is called when Gtk emits a signal that the cursor has
-been moved over the view.  Here, though, the \Rfunction{clickEvent} and
-\Rfunction{motionEvent} methods determine which node was clicked or under
-the cursor, respectively, using the data in the \Rslot{where} parameter.
-
-The \Rclass{heatmapView} class is different from the other views in that it
-has neither a \Rfunction{clickEvent} or a \Rfunction{motionEvent} method.  The
-current implementation of the \Rfunction{heatmap} function does not return
-information about where plot objects are placed on the device.  As a result,
-it is not possible to convert the user coordinates of where the event occurred
-into an object (such as the row and column coordinates of the heatmap) on
-the view.  Thus, the \Rclass{heatmapView} class is the only view that is not
-currently interactive.
+Since all views are expected to be interactive they have methods that
+correspond to the different supported user interactions.  The events
+we are most interested in responding to include key press events,
+where a certain configuration of keys are pressed; button click
+events, where one of the mouse buttons is clicked over the view; and
+mouse movement events, where the cursor moves over the view.  When an
+event occurs it creates a signal, that is caught by the signal
+handler.  The signal handler then calls the callback function and it
+is in the callback function that the corresponding method is called.
+For example, when the user clicks the mouse button and the cursor is
+over a scatter plot view, it is the scatter plot's
+\Rfunction{clickEvent} method that is invoked.  Then within the
+\Rfunction{clickEvent} method, the \Rfunction{identifyView} method is called
+to convert the pixel coordinates into an object in the plot.
+
+As can be seen in Figure \ref{Fig:View}, \Robject{sPlotView},
+\Robject{graphView}, \Robject{heatmapView}, and \Robject{spreadView} all have
+a \Rfunction{clickEvent} method, which is called when the button click event
+occurs or in the case of \Robject{spreadView} when a row is selected on the
+spreadsheet. Currently, only \Robject{sPlotView} and \Robject{graphView} have
+\Rfunction{motionEvent} methods, which is called when the cursor moves over
+the view.  The \Rfunction{clickEvent} and \Rfunction{motionEvent} methods call
+the \Rfunction{identifyView} method to determine which object on the view was
+under the cursor.
 
@@
 
-Note that when updating a view in R we do not have access to objects on a
-view.  Thus, a point on a plot is not an object that allows us to update its
-attributes and then have that point redrawn to reflect its new attributes.
-Instead in R we must currently draw over a portion of the view to update the
-view.  However, if the method of creating R plots changed, the
-\Rfunction{updateView} methods could be modified.
+Note that when updating a view in \textsf{R} we do not have access to objects
+on a view.  Thus, a point on a plot is not an object that allows us to update
+its attributes and then have that point redrawn to reflect its new attributes.
+Instead in \textsf{R} we must currently draw over a portion of the view to
+update the view.  However, if the method of creating \textsf{R} plots changed,
+the \Rfunction{updateView} methods could be modified.
 
@@
 
-\subsection{Controller}\label{Ssec:OneCont}
-
-The controller must perform two tasks: the first task is to store information
-and the second task is to allow the user to determine the course of actions.
-For the first function, which is implemented in the \Rpackage{MVCClass}
+\subsection{Controller}
+\label{Ssec:OneCont}
+
+The controller performs two tasks: the first task is to maintain state,
+such as what response will occur for a particular user interaction, 
+and the second task is to control message passing.  Messages, which are
+discussed in Sections \ref{Ssec:OneMess} and \ref{Ssec:MultLink}, are
+responsible for creating new components, such as models and views, and for
+passing information between components, such as a view asking a model for
+information.  The controller determines when these messages are created and
+ensures that they are sent to the proper recipient.
+
+For the first task, which is implemented in the \Rpackage{MVCClass}
 package, the controller is implemented as an environment that is stored in the
@@
 storing data, such as a list or a database, an environment is nice because in
-R an environment is not copied each time a value within the environment
-changes.  The controller in this first role needs to store information
-pertaining to state, such as information about the appearance of the views in
-a particular model-view-controller entity.
-
-For the second function, which is implemented in the \Rpackage{iSNetwork}
-package, the controller gives the user access to the application program
-interface (API), either through the GUI or through command line functions.
-Through the API the user can load data, create views, set the response to user
-interaction with a view, quit the program and perform other functions.  On the
-GUI, there are menus to let users decide which action is taken next.  If
-the user would rather not use the GUI, there are command line functions that
-provide access to the same functionality that is available through the GUI.
-The GUI is shown in Figure \ref{Fig:ContWin} and is referred to as the
-control window.
+\textsf{R} an environment is not copied each time a value within the
+environment changes.
+
+For the second task, the controller gives the user restricted access
+to the state information through an application program interface
+(API) and thereby lets the user change the response to different events. 
+In our implementation users can interact either through a GUI
+or through command line functions.  Through the API the user can load
+data, create views, set the response to user interaction with a view,
+quit the program and perform other functions.  On the GUI, which is shown in
+Figure \ref{Fig:ContWin} and is called the control window, users can decide
+which action is taken next by interacting with one of the menus.  If the user
+would rather not use the GUI, there are command line functions that
+provide access to the same functionality that is available through the
+GUI.
 
@@
     \scalebox{0.35}{\includegraphics{ControlWindow.jpg}}
-    \caption{ The Control Window. }
+    \caption{ The control window. }
     \label{Fig:ContWin}
@@
 As shown in Figure \ref{Fig:ContWin}, the menus on the control window are
-File, Edit, Display, Event, and Window.  These menus always appear on the
-control window, but the menu items under each menu change depending on
-the type of model that is active.  Because the goal is to have multiple
-linked \Robject{MVC} objects, more than one \Robject{MVC} object is loaded and
-thus, more than one model. However, the control window menus and the command
-line functions only perform operations on one \Robject{MVC} object at a time.
-Thus, in the \Rpackage{iSNetwork} package there is the concept of the
-``active'' \Robject{MVC} object, where the ``active'' \Robject{MVC} object
-(and thus, the ``active'' model) means that operations performed through the
-control window menus or through the command line functions affect that
-\Robject{MVC} object.  For example, the Display menu allows users to create
-views of the active model, but only certain views are appropriate for a type
-of model, as mentioned in 
-Section \ref{Ssec:OneViews}.  If the active model was a data frame model
-(\Rclass{dfModel}), then the views that the user could create are a
-scatter plot (\Rclass{sPlotView}) and a spreadsheet (\Rclass{spreadView}),
-but a graph plot (\Rclass{graphView}) does not make sense.  Thus, when the
-model in the ``active'' \Robject{MVC} is of class \Rclass{dfModel}, then the
-Display 
-menu has menu items to create a scatter plot and a spreadsheet.  Also, if
-the ``active'' \Robject{MVC} is of class \Rclass{dfModel}, then the command
-line functions only let the user create a scatter plot and a spreadsheet view
-of the active model.
-
-Because one of the goals of creating linked views based on multiple linked
-data sets is to create a software package that is flexible and extensible,
-users are able to add menus and menu items to the control window through
-the command line to access new functionality that they have created.  As an
-example if a future programmer wants to add a new type of view, such as a
-histogram, then a new menu item for that view could be added to the Display
-menu, as long as that view makes sense for the active \Robject{MVC's} model.
-By allowing users to change the controller, they can add new functionality that
-is specific to their problem.  Note that when users make additions to the API,
-they should add new command line functions in addition to adding menus and
-menu items to access that functionality.
-
-\subsubsection{Linking Events to Functions}\label{Ssec:OneEvent}
+\texttt{File}, \texttt{Edit}, \texttt{Display}, \texttt{Event}, and
+\texttt{Window}.  These menus always appear on the control window, but the
+menu items under each menu change depending on the type of model that is
+active.  Because the goal is to have multiple linked \Robject{MVC} objects,
+more than one \Robject{MVC} object is loaded and thus, more than one model.
+However, the control window menus and the command line functions only perform
+operations on one \Robject{MVC} object at a time.  Thus, in the
+\Rpackage{iSNetwork} package there is the concept of the \textit{active}
+\Robject{MVC} object, where the \textit{active} \Robject{MVC} object (and
+thus, the active model) means that operations performed through the control
+window menus or through the command line functions affect that \Robject{MVC}
+object.  For example, the \texttt{Display} menu allows users to create views
+of the active model, but only certain views are appropriate for a type of
+model, as mentioned in Section \ref{Ssec:OneViews}.  If the active model was a
+data frame model (\Rclass{dfModel}), then the views that the user could create
+are a scatter plot (\Rclass{sPlotView}) and a spreadsheet
+(\Rclass{spreadView}), but a graph plot (\Rclass{graphView}) does not make
+sense.  Thus, when the model in the \textit{active} \Robject{MVC} is of class
+\Rclass{dfModel}, then the \texttt{Display} menu has menu items to create a
+scatter plot and a spreadsheet.  Also, if the \textit{active} \Robject{MVC} is
+of class \Rclass{dfModel}, then the command line functions only let the user
+create a scatter plot and a spreadsheet view of the active model.
+
+Because one of our goals is to create a software package that is
+flexible and extensible, users and developers can add menus and menu items 
+to the control window through the command line to access new
+functionality that they have created.  As an example if a future
+programmer wants to add a new type of view, such as a histogram, then
+a new menu item for that view could be added to the \texttt{Display}
+menu.  When adding to the API, command line functions should be created as
+well as menu items to access the new functionality.
+
+\subsubsection{Linking Events to Functions}
+\label{Ssec:OneEvent}
 
@@
 which is called the callback function, as discussed in Section
-\ref{Ssec:Backg}.  However, different responses to an event may require
-functions with different signatures.  Using the example in the previous
-paragraph, a callback function that colored a point needs two parameters,
-the new color to use and the point to color, while a callback function that
-hid a point needs just one parameter, the point to hide.  Since the
-intention is to be able to interchange the responses to an event, the
+\ref{Sec:Intro}.  However, different responses to an event may require
+functions with different signatures, where a function signature is the
+combination of the function name along with its parameters.  Using the example
+in the previous paragraph, a callback function that colored a point needs two
+parameters, the new color to use and the point to color, while a callback
+function that hid a point needs just one parameter, the point to hide.  Since
+the intention is to be able to interchange the responses to an event, the
 signatures of different callback functions should be the same.  These design
@@
 class are \Rslot{callFun}, \Rslot{shortName}, and \Rslot{preprocessFun}.  The
-\Rslot{callFun} slot is the callback function, the
-\Rslot{shortName} slot is a short description of the function that
-identifies the callback function, and the
-\Rslot{preprocessFun} slot is a list of all preprocessing
+\Rslot{callFun} slot is the callback function, the \Rslot{shortName} slot is a
+short description of the function that identifies the callback function, and
+the \Rslot{preprocessFun} slot is a list of all preprocessing
 functions that must be called before the callback function.
@@
     \scalebox{0.6}{\includegraphics{EventClass.jpg}}
-    \caption{ The gEventFun Class. }
+    \caption{ The gEventFun class. }
     \label{Fig:EventFun}
@@
 
-The callback function that is stored in the \Rslot{callFun} slot must take
-only one parameter.  As discussed previously, all callback functions must have
-the same signature so that different callback functions can be interchanged in
-the code to allow different responses to an event.  All callback functions
-need to know what object they should act on so this is the one
-parameter that the callback functions take.  Any other information a callback
-function may need is set using the preprocessing functions.  The class of
-the one parameter for the callback function depends on the model class in
-the active \Robject{MVC} object.  For example, for a graph model, the
-parameter is of class \Rclass{AgNode}, representing a node in the graph.  For
-a data frame model, the parameter is of class \Rclass{character}, representing
-a row in the data frame.
-
-The preprocessing functions that are stored in the preprocessFun slot
-take no parameters.  The purpose of the preprocessing functions is to set
-variables that the callback function needs.  Thus, not all callback
-functions need preprocessing functions and if no preprocessing functions
-are needed, then this slot is NULL.  If the callback function sets the
-color of a node in a graph, then a preprocessing function is needed to
-determine what color is used.  So in this example, the preprocessing
-function opens a color browser to allow the user to choose what the new
-color is.
-
-When a user wants to link a callback function stored in an \Robject{gEventFun}
-instance to an event, this can be performed either through the GUI or through
-the command line functions.  The controller environment in each \Robject{MVC}
-object, which is discussed in detail in Section \ref{Ssec:OneMVC},
-stores which \Robject{gEventFun} object (representing the callback
-function) is linked with each event.  Note that not all callback functions
-make sense for each event so the controller environment also stores a list
-for each event of appropriate \Robject{gEventFun} instances that can be linked
-to that event.
-
-\subsection{MVC}\label{Ssec:OneMVC}
+Since the intention is to be able to interchange the responses to an event,
+the signatures of different callback functions should be the same.
+In our design a callback function takes only one parameter. 
+The class of that parameter depends on the type of model represented by the
+\Robject{MVC}.  For example, for a graph model, the parameter corresponds to
+the name of a node in the graph, while for a data frame model, the parameter
+represents a case.  We expect a one-to-one relationship between the elements
+on a view and the elements in the model.  Continuing the previous example,
+each point on a scatter plot represents one case in the data frame model and
+each node in the graph plot represents one node in the graph model.  Clearly
+it is possible for two points in a scatter plot to overlap if they have the
+same x axis and y axis values, but they are still considered two points
+representing two cases in the model.
+
+Any other information a callback function may need is set using the
+preprocessing functions, which are automatically called once by the controller
+when the callback function is linked to the event.  The preprocessing
+functions take no parameters.  The purpose of the preprocessing functions is
+to set variables that the callback function needs.  Thus, not all callback
+functions need preprocessing functions.  If the callback
+function sets the color of a node in a graph, then a preprocessing
+function is needed to determine what color is used.  So in this
+example, the preprocessing function opens a color browser to allow the
+user to choose the new color.
+
+Users can link actions to events either through the GUI or via
+command line functions.  The controller environment for each \Robject{MVC}
+object stores a list, for each event, of appropriate callback functions that
+can be linked to that event and it also stores which callback function is
+currently linked to the event.  When a developer creates a new callback
+function, the developer can decide which events the callback function can
+potentially be linked to.  For example, the developer may only want the color
+a point callback function to be available for a button click event and not a
+mouse over event.
+
+\subsection{MVC}
+\label{Ssec:OneMVC}
 
@@
     \scalebox{0.6}{\includegraphics{fullMVCClass.jpg}}
-    \caption{ The MVC Class. }
+    \caption{ The MVC class. }
     \label{Fig:MVCClass}
@@
 view, and controller objects together, but it also shows
-that the \Robject{linkedModelMVC} objects are related to each other through a
-parent-child (tree) hierarchy, as shown in Figure \ref{Fig:Hier}.  In this
-picture, `model1' is the name of the original \Robject{MVC} object
+that the \Robject{linkedModelMVC} objects can be related to each other 
+through a parent-child (tree) hierarchy, as shown in Figure \ref{Fig:Hier}.
+In this picture, modelA is the name of the original \Robject{MVC} object
 (model) and from it, two \Robject{MVC} objects were created,
-`model2.1' and `model2.2'. Then from `model2.1', three
-\Robject{MVC} objects were created, `model3.1', `model3.2', and
-`model3.3'.  As an example, model1 could be gene 
-expression data, model2.1 could be gene expression data that is a subset of
-model1 (based on filtering), model2.2 could be derived by performing
-multidimensional scaling (MDS) on model1, model 3.1 could be the gene ontology
-(GO) graph representing the molecular functions of the genes in model2.1,
-model 3.2 could be the KEGG pathways that the genes in model2.1 are involved
-in, and model 3.3 could be the chromosome locations of the genes in model2.1.
+modelB and modelC. Then from `modelB', three
+\Robject{MVC} objects were created, modelD, modelE, and
+modelF.  As an example, modelA could be gene 
+expression data, modelB could be gene expression data that is a subset of
+modelA (based on filtering), modelC could be derived by performing
+multidimensional scaling (MDS) on modelA, modelD could be the Gene Ontology
+(GO) graph representing the molecular functions of the genes in modelB,
+modelE could be the KEGG pathways that the genes in modelB are involved
+in, and modelF could be the chromosome locations of the genes in modelB.
 
@@
     \scalebox{0.6}{\includegraphics{Hierarchy2.jpg}}
-    \caption{ The Parent-Child Relationship Between \Robject{MVCs}. }
+    \caption{ The parent-child relationship between \Robject{MVCs}. }
     \label{Fig:Hier}
@@
 
-A design decision was that a new model (and thus, a new
-\Robject{MVC} object) could either be loaded or could be created from an
-already loaded \Robject{MVC} object.  As mentioned in Section
-\ref{Ssec:MMVC}, the focus in this paper is on experimental data that are
-linked to meta-data, as the linked data sets that we are most interested in
-visualizing.  In this situation, the experimental data (and its
-\Robject{MVC} object) can be thought of as determining what
-meta-data are studied.  Thus, there is a hierarchical relationship between
-these data sets where one \Robject{MVC} object creates another
-\Robject{MVC} object.  This focus led to the decision that a
-\Robject{MVC} object can have at most one parent
-\Robject{MVC}, from which it was derived, but it can have many
-child \Robject{MVCs}.  An easy example of a child model derived
-from a parent model is to start with a data frame as the parent model and then
-take a subset of this data frame as the child model.
-
-\subsection{Messages Within a MVC Object}\label{Ssec:OneMess}
+As shown in Figure \ref{Fig:Hier}, there are many cases where there are
+natural submodels that can be derived from a parent model.  We propose that
+each submodel is itself a \Robject{MVC} instance, and hence can have multiple
+linked views.  Even though each submodel is its own \Robject{MVC} instance,
+there will be times when the submodel \Robject{MVC} needs to communicate with
+its parent \Robject{MVC}.  There are many different ways of communicating
+between parent and child \Robject{MVCs} and we discuss this in Section
+\ref{Sec:MultMVC}.
+
+Deciding when new data should be added to the current model versus generating
+a new submodel in a \Robject{MVC} instance is done by comparing the
+relationship between the new data and the data in the current model.  If the
+new data has a one-to-one relationship with the elements in the current model,
+then we believe that in most cases the new data should be added to the 
+current model.  For example, if there is
+a test statistic for each row in a data frame, then this could be added to the
+model.  However, if the new data has a one-to-many or a many-to-many
+relationship with the elements in the current model, then serious
+consideration should be given to the creation of a submodel.  If there is
+already an appropriate MVC class (e.g. if the submodel is a graph) then it can
+be generated by creating a new instance of the appropriate MVC class.  If there
+is no appropriate implementation, then the user will need to create one.  
+As an example, consider the
+relationship between genes and GO terms.  If the original model contains gene
+expression levels and a user wants to explore the GO terms for these genes,
+then we encourage the creation of the GO graph representing the GO terms, and
+creating a new graph \Robject{MVC} object.
+
+\subsection{Messages Within a MVC Object}
+\label{Ssec:OneMess}
 
@@
 about messages that are passed between \Robject{MVC} objects, please see
-Section\ref{Ssec:MultMess}.
+Section \ref{Ssec:MultMess}.
 
@@
 the \Rfunction{handleMessage} method, which determines where the information
-should be sent and hence is the dispatcher.   Also, since messages
+should be sent and hence, is the dispatcher.   Also, since messages
 are sent when something has changed, either through addition, alteration, or
@@
     \scalebox{0.6}{\includegraphics{newMessageClass.jpg}}
-    \caption{ Inheritance for Message Objects. }
+    \caption{ Inheritance for message classes. }
     \label{Fig:Mess}
@@
 \Rslot{mData}, and \Rslot{type}.  \Rslot{dataName} is the character string
-that gives the name of the model, 
-\Rslot{mData} is a list of data needed to perform the addition or alteration
-operation, and \Rslot{type} is a character string that gives the type of
-addition or alteration to perform.  Both \Rclass{gAddMessage} and
-\Rclass{gUpdateMessage} are virtual classes.
+that gives the name of the model, \Rslot{mData} is a list of data needed to
+perform the addition or alteration operation, and \Rslot{type} is a character
+string that gives the type of addition or alteration to perform.  Both
+\Rclass{gAddMessage} and \Rclass{gUpdateMessage} are virtual classes.
 
@@
 
-%\subsubsection*{Example of Message Passing}
-
-An example of message passing within a \Robject{MVC} object is shown in Figure
-\ref{Fig:MPwithin}.  In this example, the response to a user performing a
-button click event over a point in the scatter plot is to color the point red.
-Setting the response to an event was discussed in Section
-\ref{Ssec:OneEvent}.  Here the user first clicks on a point in the
-scatter plot, which is shown as Step 1 (the circled 1 on the figure).  In the
-figure, the point that has been clicked has an arrow next to it.  Then, in
-Step 2, an update data message is sent to the model, which updates
-the \Rslot{virtualData} slot of the model object.  After the
-model has been updated, the views must be notified that a
-change has occurred so in Step 3, an update view message is created and sent
-to all views of this model.  In response to this
-message, the view updates itself in Step 4 and now the
-point is colored red on the view to reflect the change in the
-model.  In this example, the \Rslot{from} slot in the
-\Robject{gUpdateDataMessage} object is the name of this \Robject{MVC} object
-since the message came from within the \Robject{MVC}.
+An example of message passing within a \Robject{MVC} object is shown
+in Figure~\ref{Fig:MPwithin}.  In this example, there are two views of the
+model (view 1 and view 2), and the response to a user
+performing a button click event over a point in the scatter plot is to
+color the point red.  Setting the response to an event was discussed
+in Section~\ref{Ssec:OneEvent}.  Here the user first clicks on a point
+in view 1, which is shown as Step 1 (the circled 1 on the
+figure).  In the figure, the point that has been clicked has an arrow
+next to it.  Then, in Step 2, an update data message is sent to the
+model, which updates itself accordingly, i.e. by changing the color
+associated with the corresponding element.  After the model has been
+updated, the views must be notified that a change has occurred so in
+Step 3, an update view message is created and sent to all views of
+this model.  In response to this message, the views update themselves in
+Step 4 and now each view has one point that is colored red to reflect the
+change in the model.
 
@@
     \scalebox{0.6}{\includegraphics{newMPwithin.jpg}}
-    \caption{ Message Passing within a \Robject{MVC} object.  The first step,
+    \caption{ Message passing within a \Robject{MVC} object.  The first step,
       indicated by a circled one, occurs when a user interacts with the y
-      versus x scatter plot.  Then an update data message is sent to the model
-      in the second step, which alters the model.  Next in the third step all
-      views of this model are sent an update view message, which causes the
-      view to be updated in the fourth step. }
+      versus x scatter plot (view 1).  Then an update data message is sent to
+      the model in the second step, which alters the model.  Next in the third
+      step all views of this model are sent an update view message, which
+      causes the views to be updated in the fourth step. }
     \label{Fig:MPwithin}
@@
 
-\section{Details of Multiple MVCs}\label{Sec:MultMVC}
+\section{Details of Multiple MVCs}
+\label{Sec:MultMVC}
 
 We now describe the implementation of multiple \Robject{MVC} objects for
-linking views of multiple linked data sets.  Each \Robject{MVC} object
-represents one model and its views.  Now for the
-views of different data sets to be linked, they must link through
-the data sets so this linking information must be stored somewhere.  Also,
-there must be message classes to pass information between
-\Robject{MVC} objects.
+linking views of multiple linked models.  Each \Robject{MVC} object
+represents one model and its views.  In our
+paradigm models are linked through some additional meta-data. In
+the case of gene expression data the GO view is linked to the heatmap
+view through the mapping of genes to GO categories. In the city and
+state example, the cities are linked via the states in which they are
+contained. If, in the state MVC, a state is selected that information
+can be propagated to the city MVC, but some piece of software must
+link the state to the cities.
 
@@
 the Gene Ontology molecular function graph that was first mentioned in Section
-\ref{Ssec:Limit}, these two data sets are linked in the following manner: the
+\ref{Ssec:Limit}, these two models are linked in the following manner: the
 Affymetrix identifiers in the experimental microarray data are linked to Locus
 Link identifiers, which are in turn linked to the Gene Ontology molecular
-function terms.  Thus, to pass information between these two data sets a table
+function terms.  Thus, to pass information between these two models a table
 that links Affymetrix identifiers to Gene Ontology molecular function terms
@@
 Affymetrix identifiers to Gene Ontology molecular function terms and vice
-versa.  Once it is possible to link these data sets, then information must be
+versa.  Once it is possible to link these models, then information must be
 passed between them through messages to keep the views synchronized.
@@
 that \Robject{MVC} objects are related through a tree hierarchy where a child
-model is derived from its parent model.  Consider Figure
-\ref{Fig:Hier}, which was shown previously.  If model1's model changes, then
-this information first gets passed to model2.1 and model2.2 and these two
-\Robject{MVCs} may make changes to their models based on this
-information.  Then model2.1 passes this information to model3.1, model3.2, and
-model3.3 based on the changes that model2.1 made to its model.  Thus,
-information passing between model1 and model3.1, for instance, only occurs
-through model2.1.  So messages can only be sent from parent \Robject{MVC} to
-child \Robject{MVC} or from child \Robject{MVC} to parent \Robject{MVC}.
-
-Next we consider how the data sets are linked (and thus, how \Robject{MVC}
+model is derived from its parent model.  
+Consider Figure~\ref{Fig:Hier}, which was shown previously.  If
+modelA's model changes, then this information first gets passed to
+modelB and modelC and these two \Robject{MVCs} may make changes to
+their models based on this information.  Then modelB passes this
+information to modelD, modelE, and modelF based on the changes
+that modelB made to its model.  Thus, information passing between
+modelA and modelD, for instance, only occurs through modelB.  So
+messages can only be sent from parent \Robject{MVC} to child
+\Robject{MVC} or from child \Robject{MVC} to parent \Robject{MVC}.
+This communication between parent and child \Robject{MVCs} was shown in Figures
+\ref{Fig:heatmap}, \ref{Fig:GOgraph}, and \ref{Fig:heatmapsub} in Section
+\ref{sec:micro}.  Here views of the parent and child models are
+synchronized, but the selected items in a grandparent model may not
+match the selected items in a grandchild model (as shown by Figures
+\ref{Fig:heatmap} and \ref{Fig:heatmapsub} where the selected genes are
+different in these two heatmaps).  Deciding how to link \Robject{MVC} models 
+is a difficult task especially when the relationship between elements in the 
+two models is not one-to-one.
+
+Next we consider how the models are linked (and thus, how \Robject{MVC}
 objects are linked) followed by how messages are passed between these
@@
 
-\subsection{Linking Data Sets}\label{Ssec:MultLink}
+\subsection{Linking Models}
+\label{Ssec:MultLink}
 
@@
 model's data change into useful information for this
-particular child model.  Thus, the parent model's data
-change may be converted into different information for different child models
-and this can happen because each child model has
-its own \Rfunction{fromParent} function.  When a child \Robject{MVC} object is
-created and its model's \Rslot{linkData} slot is filled, the
-\Rfunction{toParent} function converts information from this child
-model into useful information for its parent \Robject{MVC} and the
-\Rfunction{fromParent} function converts a message from its parent
-\Robject{MVC} into useful information for this child \Robject{MVC}.
+particular child model.  Thus, the message sent to all children is the
+same, but each child converts the message, using its \Rfunction{fromParent}
+function, to make it specific for that child.
+
+The link functions, \Rfunction{toParent} and \Rfunction{fromParent}, can
+become quite complex depending on the relationship between the elements in the
+two models.  Suppose that the data elements in the two models have a
+one-to-one relationship where data element 1 in child model A is linked to
+data element 2 in parent model B.  Now if data element 1 in child model A is
+selected, then the \Rfunction{toParent} function will pass along the
+information that data element 2 in parent model B should be selected.  Here
+the linking is straightforward.  However, if the data elements in the two
+models have a one-to-many or many-to-many relationship, the decision about
+which linked elements to update is more complex.  Now suppose that data
+element 1 in child model A is linked to data elements 10, 11, and 12 in parent
+model B and data element 2 in child model A is linked to data elements 12 and
+13 in parent model B.  Here a many-to-many relationship exists between the
+data elements.  If the user selects element 1 in child model A, the
+\Rfunction{toParent} function must decide if elements 10, 11, and 12 should be
+selected in parent model B or if only elements 10 and 11 should be selected in
+parent model B (because element 12 in model B is also linked to element 2,
+which is not selected, in model A).
+
+These two options for the linking function can be represented as \textit{any}
+versus \textit{all} linking.  In \textit{any} linking, if any element is
+updated, then the corresponding linked element is updated.  This linking could
+also be represented as \textit{or} linking if one is thinking in terms of
+boolean operators.  In \textit{all} linking, the element in the second model
+is updated only if all linked elements are updated in the original model.
+Similarly, this linking could be also be represented as \textit{and} linking
+if one is thinking in terms of boolean operators.
+
+Making the previous example even more complex suppose that child model A also
+had data element 3 which is linked to data elements 12, 13, and 14 in parent
+model B.  Now if data elements 1 and 3 are selected in child model A, the
+\Rfunction{toParent} function can link based on one of these options: i) select
+data elements 10, 11, 12, 13, and 14 in parent model B, which indicates that
+any items in model B that are linked to elements 1 and 3 in model A are
+selected, ii) select data element 12, which indicates that only data elements
+that are linked to both elements 1 and 3 in model A are selected, or iii)
+select no data elements in model B, which indicates that only elements linked
+to both elements 1 and 3 and no other elements in model A are selected.
+Another consideration in the linking functions may be to change the response
+depending on the type of selection being performed.  For example, \textit{any}
+linking between models may be appropriate when an item in one model was
+highlighted, but if an element is hidden, then \textit{all} linking may be
+suitable.  This flexibility lets developers determine the appropriate
+relationships between models, but also requires thoughtful responses and
+potentially complex linking functions.
 
@@
 
-\subsection{Messages Between MVC Objects}\label{Ssec:MultMess}
-
-In Section \ref{Ssec:OneMess}, where the message classes to pass
-information within one \Robject{MVC} object were discussed, it was noted that
-there are two main types of messages: an add message and an update message.
-These message types also are seen in passing messages between different
-\Robject{MVC} objects, as shown in Figure \ref{Fig:BetwMess}.
+\subsection{Messages Between MVC Objects}
+\label{Ssec:MultMess}
+
+The message classes for creating a child \Robject{MVC} from an existing
+\Robject{MVC} and for sending messages between \Robject{MVC} objects are shown
+in Figure \ref{Fig:BetwMess}.  The message types for passing information
+between \Robject{MVCs} are similar to the messages
+that are passed within \Robject{MVC} objects.  Now the add child message is
+similar to the add model message, and the send parent message and the send
+child message are similar to the update model message.
 
@@
 %    \scalebox{0.6}{\includegraphics{newMessageClass2.jpg}}
-    \caption{ Inheritance for Message Classes that are Passed Between
+    \caption{ Inheritance for message classes that are passed between
       \Robject{MVCs}. }
@@
 
-The first message class that is discussed is
-\Rclass{gAddChildMessage}.  A \Robject{gAddChildMessage} object is very
+A \Robject{gAddChildMessage} object is very 
 similar to a \Robject{gAddDataMessage} object because they are both adding a
@@
 new model (and new \Robject{MVC}), the \Rslot{mData} slot is a list that
-contains the
-model data and virtual data to fill the \Rslot{modelData} and
+contains the model data and virtual data to fill the \Rslot{modelData} and
 \Rslot{virtualData} slots, respectively, of the new model, and the
@@
 \Rfunction{initialize} method properly fills the slots of the object when it
-is created and the \Rfunction{handleMessage} method ensures that the new
+is created.  The \Rfunction{handleMessage} method ensures that the new
 model and \Robject{MVC} objects are created (which is what the
-\Rfunction{handleMessage} method for the \Rclass{gAddDataMessage} class does)
-and then it fills the \Rslot{linkData} slot of the new child model
+\Rfunction{handleMessage} method for the \Rclass{gAddDataMessage} class does),
+then it fills the \Rslot{linkData} slot of the new child model
 so that the \Rfunction{toParent} and \Rfunction{fromParent} functions are
-available for messaging between \Robject{MVC} objects and finally it stores
+available for messaging between \Robject{MVC} objects, and finally it stores
 information in the child \Robject{MVC} object's controller that determines if
@@
 These message classes ensure that when a model is updated, its parent and
-child models are notified of the change and most importantly are
+child models are notified of the change and most importantly, are
 notified of the change in a way that each model can properly act upon that
@@
     \scalebox{0.55}{\includegraphics{newMP.jpg}}
-    \caption{ Message Passing Between and Within \Robject{MVCs}.  When a user
+    \caption{ Message passing between and within \Robject{MVCs}.  When a user
       interacts with the y versus x scatter plot, the first step, indicated by
-      a circled one, is to send an update data message to model 2.  After
-      model 2 has been updated, all views of this model must be notified, which
+      a circled one, is to send an update data message to model B.  After
+      model B has been updated, all views of this model must be notified, which
       is shown in the second step when an update view message is sent.  Next
-      any data that are linked to model 2 must be notified that model 2
+      any data that are linked to model B must be notified that model B
       changed, which is shown in the third step as a send parent message.
-      This message is converted into useful information for model 1 in
+      This message is converted into useful information for model A in
       the fourth step by the \Rfunction{toParent} function, and this
-      information is sent to model 1 in the fifth step when an update data
-      message is sent.  Finally, after model 1 is updated, an update view
-      message is sent in the sixth step to update all views of model 1. }
+      information is sent to model A in the fifth step when an update data
+      message is sent.  Finally, after model A is updated, an update view
+      message is sent in the sixth step to update all views of model A. }
     \label{Fig:MessPass}
@@
 Figure \ref{Fig:MessPass} is very similar to Figure \ref{Fig:firstMP} shown in
-Section \ref{Ssec:Backg}.  Now, however, Figure \ref{Fig:MessPass} shows how
+Section \ref{Sec:Methods}.  Now, however, Figure \ref{Fig:MessPass} shows how
 \Robject{MVC} objects are defined and how messages are passed within one
@@
 initially clicked is red.  Then a send parent message is sent because MVC
-2, which is the \Robject{MVC} object for Model 2, has a parent \Robject{MVC},
-which is MVC 1.  Next the \Rfunction{toParent} function converts the child's
+B, which is the \Robject{MVC} object for Model B, has a parent \Robject{MVC},
+which is MVC A.  Next the \Rfunction{toParent} function converts the child's
 update data message into an update data message that the parent \Robject{MVC}
 understands.  Then the update data message that was created by the
-\Rfunction{toParent} function is sent to Model 1.  Finally, after Model 1 has
-been updated, an update view message is sent to all of Model 1's views.  It is
-in this last step that the view of Model 1 has a point colored red to indicate
-which data from Model 1 is linked to data from Model 2.  All of the data and
+\Rfunction{toParent} function is sent to Model A.  Finally, after Model A has
+been updated, an update view message is sent to all of Model A's views.  It is
+in this last step that the view of Model A has a point colored red to indicate
+which data from Model A are linked to data from Model B.  All of the data and
 the views are synchronized.
 
-\section{Conclusions}\label{Sec:Conc}
-
-Several examples of linked data sets were given at the beginning of this
-paper, including linked tables from a relational database, and meta data that
-are linked to experimental data.  The MVC paradigm was
-discussed as a useful design for creating linked views of a data set.
-Starting with the MVC paradigm, this paper provided a design for implementing
-linked, interactive views of linked data sets based on expanding the MVC
-paradigm.  For visualizing linked data sets, the MVC design, which is a well
-known and thus, well tested software paradigm, was extended to multiple
-\Robject{MVC} objects where each \Robject{MVC} object contains one data set
-and its views.  This solution broke down the problem of linked, interactive
-views of linked data sets into components (\Robject{MVC} objects) that can be
-reused, extended and linked.
-
-Once the design settled on \Robject{MVC} objects as the component of interest
-for creating linked views of linked data sets, the next step was to determine
-how these \Robject{MVC} objects were related.  Because the focus in this paper
-was on experimental data that are linked to meta-data, a hierarchical
-relationship between \Robject{MVC} objects was built into the \Rclass{MVC}
-class definition.  The experimental data can be thought of as determining what
-meta-data are studied and thus, the meta-data are derived from the
-experimental data, which shows a parent-child relationship.
-
-Another consideration in the design was creating a method for communicating
-within and between \Robject{MVC} objects.  This method must pass information
-between the components while keeping the components independent.
-Communication within a \Robject{MVC} object was handled using instances of the
-message classes, which performed one of four functions: adding a model, adding
-a view, updating a model, or updating a view.  Starting with this design, it
-seemed a natural extension to have message classes that passed information
-between \Robject{MVC} objects.  When a \Robject{MVC's} model is updated, it
-notifies its parent and child \Robject{MVCs} that a change has occurred using
-message objects.  Then when a \Robject{MVC} gets a message from a parent or
-child \Robject{MVC}, the \Robject{MVC} can either decide to ignore the message
-or read the message, which results in this \Robject{MVC's} model being
-updated.  However, the \Robject{MVC} object can not decide to only see certain
-types of messages.  It must either accept all messages and update its model
-based on the messages, or ignore all messages.  A future goal for message
-passing is to allow more selectivity in which messages are accepted between
-\Robject{MVC} objects so that the linking between \Robject{MVC} objects is
-more sophisticated.
-
-The goals for this paper were to create linked views of linked data sets, to
-create interactive views where the response to an event can be changed, and to
-create an extensible design that allows future users to make additions.
+\section{Conclusions}
+\label{Sec:Conc}
+
+The goals for this research were to create software implementations and
+conceptual paradigms to support the use of linked views of linked data sets,
+to create interactive views where the response to an event can be changed, and
+to create an extensible design that allows future users to make additions.
 The first goal was accomplished by implementing \Robject{MVC} objects as
-components that communicate with each other through messages.  The second goal
-of creating interactive views was met by using the \Rpackage{RGtk} and
-\Rpackage{gtkDevice} packages to create devices that could respond to events.
-In addition the \Rclass{gEventFun} class was designed so that users could
-alter the response to an event.  Finally, the last goal of creating an
-extensible design was achieved through the inheritance structure for classes,
-which allows for extensions by the user, and through command line functions,
-which let the user create new menu items to add to the GUI and also let the
-user create new callback functions that can be linked to events.
-
-Although extensibility was a goal, certain design decisions, such as expecting
-the data sets to be linked through a parent-child hierarchy, may not work with
-some data sets.  As an example, the linked tables from a relational database
-are not related through a tree structure and thus, having parent and child
-\Robject{MVC} objects may not make sense with these data sets.  However, the
-design allows users to add new types of models, new types of views, and new
-message classes so there is still flexibility for additions as long as the
-data sets are connected through parent-child relationships.  These new class
-definitions could be created by the user in a new package that extends the
-\Rpackage{MVCClass} package that was discussed in this paper.  Other additions
-that the user may want to perform, such as adding new menus to the GUI and
-adding new responses to an event can be performed through command line
-functions that are defined in the \Rpackage{iSNetwork} package.
-
-The functionality that is defined in the \Rpackage{iSNetwork} package is
-discussed in detail in the Vignette for this R package.  Users can load data,
+components that communicate with each other through messages.  Encapsulating
+each data set in a \Robject{MVC} object was particularly useful because it
+allowed the reuse of code for communicating between the components of a
+\Robject{MVC} object, which made generating new model and view classes easier.
+However, communication between the \Robject{MVC} objects was, at times, more
+complicated.
+
+The crux of the communication between \Robject{MVC} objects is the link
+function, which converts information that one model can understand into
+information that a linked model can understand.  As discussed previously, the
+linking functions, \Rfunction{toParent} and \Rfunction{fromParent}, can be
+quite complex if the relationship between the data elements in the two models
+is one-to-many or many-to-many.  In response to an update in one model, the
+linking function must decide if all linked elements in the second model
+should be updated, if any linked elements should be updated, or if a subset of
+linked elements should be updated.  These decisions must be made by the
+developer to properly represent the relationship between the data elements in
+the two models.  While this flexibility lets the developer represent different
+relationships between different models, the developer is then responsible for
+knowing what makes sense (for example, biologically) when creating the link
+functions.
+
+The second goal of creating interactive views was met by using the
+\Rpackage{RGtk} and \Rpackage{gtkDevice} packages to create devices that could
+respond to events.  Using these packages made creating interactive views
+particularly straightforward because any static R plot could be made
+interactive by plotting it on a Gtk device and by having the plotting function
+return information about the location of plot objects on the device.  In
+addition the \Rclass{gEventFun} class was designed so that users could alter
+the response to an event.  The response to an event could be as simple as
+selecting the plot object that was just clicked by a mouse button or the
+response could be as complicated as creating a new model.  This flexibility
+lets the developer create a more varied interactive environment.
+
+Finally, the last goal of creating an extensible design was achieved through
+the inheritance structure for classes, which allows for extensions by the
+user, and through command line functions, which let the user create new menu
+items to add to the GUI and also let the user create new callback functions
+that can be linked to events.  However, in terms of extensibility, certain
+design decisions, such as expecting the models to be linked through a
+parent-child hierarchy, may not work with some models.  As an example, the
+linked tables from a relational database are not related through a tree
+structure and thus, having parent and child \Robject{MVC} objects may not make
+sense with these models.  However, the design allows users to add new types
+of models, new types of views, and new message classes so there is still
+flexibility for additions as long as the models are connected through
+parent-child relationships.  These new class definitions could be created by
+the user in a new package that extends the \Rpackage{MVCClass} package that
+was discussed in this paper.  Other additions that the user may want to
+perform, such as adding new menus to the GUI and adding new responses to an
+event can be performed through command line functions that are defined in the
+\Rpackage{iSNetwork} package, which is the front end for the design
+described in this paper.
+
+The functionality that is provided in the \Rpackage{iSNetwork} package, which
+is an implementation of the research discussed in this paper, is
+discussed in detail in the vignette for this R package.  Users can load data,
 create views of the data, interact with these views and change the response to

### 2006-08-10 inst/paper/paper2v2.tex
Made changes to paper 2.
git-svn-id: file:///home/git/hedgehog.fhcrc.org/bioconductor/trunk/madman/Rpacks/MVCClass@19115 bc3139a8-67e5-0310-9ffc-ced21a209358

@@
 graphical user interface (GUI), to pressing certain keys.  Events are
-interpreted based on which window has focus.  For example, in the XWindows
+interpreted based on which window has focus.  For example, in the X Windows
 system, if a user types C on the keyboard and the window that has focus is a
@@
 In the design presented in this paper, the components consist of \Robject{MVC}
-objects and communication within and between \Robject{MVC} objects is
+objects, and communication within and between \Robject{MVC} objects is
 performed by message objects.  This solution uses the power and simplicity of
@@
 must respond to a certain subset of events so we have implemented the view
-classes by first creating a single, virtual class which the specialized
+classes by first creating a single, virtual class, which the specialized
 views extend.  As with our model classes, this allows us to define
@@
 (model) and from it, two \Robject{MVC} objects were created,
-modelB and modelC. Then from `modelB', three
+modelB and modelC. Then from modelB, three
 \Robject{MVC} objects were created, modelD, modelE, and
@@
   \begin{center}
-    \scalebox{0.6}{\includegraphics{newMPwithin.jpg}}
+    \scalebox{0.6}{\includegraphics{newMPwithin2.jpg}}
     \caption{ Message passing within a \Robject{MVC} object.  The first step,
@@
 \ref{Ssec:Limit}, these two models are linked in the following manner: the
-Affymetrix identifiers in the experimental microarray data are linked to Locus
-Link identifiers, which are in turn linked to the Gene Ontology molecular
+Affymetrix identifiers in the experimental microarray data are linked to Entrez
+Gene identifiers, which are in turn linked to the Gene Ontology molecular
 function terms.  Thus, to pass information between these two models a table
@@
 change into useful information for the parent model.  
-
 Now suppose that the parent model changed instead.  This information

### 2006-10-26 inst/paper/MVC.tex
Changed the MVC paper.
git-svn-id: file:///home/git/hedgehog.fhcrc.org/bioconductor/trunk/madman/Rpacks/MVCClass@20825 bc3139a8-67e5-0310-9ffc-ced21a209358

@@
 \author{Elizabeth Whalen\\Harvard School of Public Health \\ 
+        Duncan Temple Lang\\University of California at Davis \\
         Robert Gentleman\\Fred Hutchinson Cancer Research Center}
@@
 provide a valuable and important set of tools for comprehending such
-data.  The model view controller (MVC) paradigm is now a standard
+data.  The model view controller (MVC) paradigm, as introduced in the
+Smalltalk system \cite{Smalltalk}, is now a standard
 mechanism for implementing such systems.  The paradigm has proven very
@@
 paradigm, where the data (i.e. the model) are often presumed to be in
-a single, standard format.  We found that adding new views under such
-conditions could be quite problematic and often led to very complex
-code.  As we shall demonstrate, there are also issues that arise due to
-mappings that are not one-to-one between elements on different views.
-Hence we considered an approach that created new instances of
-the MVC that were linked to the parent that spawned them.
+a single, standard format when implementing the design.  We found that 
+adding new views under such conditions could be quite problematic and 
+often led to very complex code.  As we shall demonstrate, there are 
+also issues that arise due to mappings that are not one-to-one between 
+elements on different views.  Hence we considered an approach that 
+created new instances of the MVC that were linked to the parent that 
+spawned them.
 
@@
 update all of its views, but the message will also propagate to the
-parent MVC (and further up to the root). The parent MVC can choose to
-react, or not, to messages from its children -- all control being in
+parent MVC (and further up to the root).  Similarly, messages propagate 
+to the child MVCs if the user initially interacts with a view in the 
+parent MVC.  The MVC can choose to react, or not, to messages from its 
+parent or children -- all control being in
 the hands of the users, should they want it.
@@
 view. In order to interact with the GO graph, the primary unit is the
-node in that graph (and sometimes one expects with the edges), while
+node in that graph (though edges could also be considered the primary 
+unit), while
 for the heatmap view the basic element is generally the gene.  In Figures
@@
 system. Therefore, we built our software on top the \Rpackage{RGtk} package of
-D. Temple Lang, since it provides such an interface. We also vastly
+D. Temple Lang \cite{RGtk}, since it provides such an interface. We also vastly
 preferred Gtk over Tcl/Tk since its widgets are more sophisticated,
@@
 software \cite{GGobi}.  Linked views have been implemented in several
-pieces of statistical software, such as XLisp-Stat \cite{Lisp}, iPlots
-\cite{iPlots}, and
+pieces of statistical software, such as DataDesk \cite{DataDesk},
+XLisp-Stat \cite{Lisp}, iPlots \cite{iPlots}, and
 GGobi \cite{GGobiMan}.  Implementing linked views is based on the
@@
 a pointing device, such as a mouse, or via key strokes, or any of many
-other types of input devices.  Interactivity means that there is a
-response to some user action.  It is a powerful paradigm and allows
+other types of input devices.  Interactivity means that there is an
+immediate response to some user action.  It is a powerful paradigm and allows
 users to quickly navigate and explore complex documents or complex
@@
 interacts with the scatter plot view of model B, the response to
-this interaction is for model B to update itself (this is indicated
+this interaction is for the controller to tell model B to update itself 
+(this is indicated
 on the figure as a circled 1).  In Step 2, the scatter plot view is
@@
       the scatter plot, the first step, indicated by a circled 1,
-      is to tell model B to update.  Then all views of model B must be
-      updated.  Next any data that are linked to model B must be notified
-      that model B changed.  This information from model B must be
+      is for the controller to tell model B to update.  Then all views of
+      model B must be updated.  Next any data that are linked to model B must
+      be notified that model B changed.  This information from model B must be
       altered so that it can be understood by linked models, which is shown
@@
 application is a setting where one representation of the data is as a
-graph (with nodes and edges) while the other is as a rectangular array. We can
+graph (with nodes and edges) while the other is as a rectangular array. 
+We can
 envisage lots of different MVCs that operate on graph based data, and
@@
 implement these as yet another separate, specialized MVC, that could
-be linked to any one of a variety of other MVCs.
+be linked to any one of a variety of other MVCs.  In our design, each 
+MVC will contain one model and thus, linked models also indicate linked 
+MVCs.
 
@@
 The city data have one MVC and the state data a second.  Even though 
-there are two \Robject{MVC} objects here the views are linked.
+there are two \Robject{MVC} objects here, the views are linked.
 
@@
 To make this addition, the user extends the \Rclass{gModel} class in his or
-her own R software package.
+her own R software package.  More information on extending the classes can be
+found in the \Rpackage{MVCClass} and \Rpackage{iSNetwork} Vignettes.
 
@@
 switch between active models, but since the interactions could be
-quite different, there must always be an active MVC.
+quite different, there must always be an active \Robject{MVC}.
 
@@
 movement or a key press event, is what allows users to have an
-interactive environment.  When an event occurs, the callback function
-linked to the event determines the response to that event.  To create
+interactive environment.  When an event occurs, the callback function linked
+to the event (this information is stored in the controller in the MVC)
+determines the response to that event.  To create
 a flexible environment it is important to be able to change the
@@
 %% plots of graphs, the object in the plot is considered a node in the graph.
+
 Since the intention is to be able to interchange the responses to an event,
@@
 figure).  In the figure, the point that has been clicked has an arrow
-next to it.  Then, in Step 2, an update data message is sent to the
+next to it.  Then, in Step 2, an update data message is sent by the 
+controller to the
 model, which updates itself accordingly, i.e. by changing the color
@@
       indicated by a circled one, occurs when a user interacts with the y
-      versus x scatter plot (view 1).  Then an update data message is sent to
+      versus x scatter plot (view 1).  Then an update data message is sent by
+      the controller to
       the model in the second step, which alters the model.  Next in the third
@@
 microarray data that are linked to the Gene Ontology molecular
-function graph (Section~\ref{sec:micro}), these two models are
-linked in the following manner. The gene identifiers in the
+function graph (as discussed in Section~\ref{sec:micro}), these two models 
+are linked in the following manner. The gene identifiers in the
 experimental microarray data are linked to the GO terms.  Thus,
@@
 \Rslot{linkData} slot is used to convert the child model's data
-change into useful information for the parent model.
+change into useful information for the parent model if there are any linked
+elements in the parent model.
 
@@
       interacts with the y versus x scatter plot, the first step, indicated by
-      a circled one, is to send an update data message to model B.  After
-      model B has been updated, all views of this model must be notified, which
-      is shown in the second step when an update view message is sent.  Next
-      any data that are linked to model B must be notified that model B
-      changed, which is shown in the third step as a send parent message.
-      This message is converted into useful information for model A in
-      the fourth step by the \Rfunction{toParent} function, and this
-      information is sent to model A in the fifth step when an update data
+      a circled one, is for the controller to send an update data message to
+      model B.  After model B has been updated, all views of this model must
+      be notified, which is shown in the second step when an update view
+      message is sent.  Next any data that are linked to model B must be
+      notified that model B changed, which is shown in the third step as a
+      send parent message.  This message is converted into useful information
+      for model A in the fourth step by the \Rfunction{toParent} function, and
+      this information is sent to model A in the fifth step when an update data
       message is sent.  Finally, after model A is updated, an update view
@@
 
-The goals for this research were to create software implementations and conceptual paradigms to support the use of linked views of linked data sets, to
-create interactive views where the response to an event can be changed, and to
-create an extensible design that allows future users to make additions.
+The goals for this research were to create software implementations and
+conceptual paradigms to support the use of linked views of linked data sets,
+to create interactive views where the response to an event can be changed, and
+to create an extensible design that allows future users to make additions.
 The first goal was accomplished by implementing \Robject{MVC} objects as
@@
 The second goal of creating interactive views was met by using the
-\Rpackage{RGtk} and \Rpackage{gtkDevice} packages to create devices that could
-respond to events.  Using these packages made creating interactive views
-particularly straightforward because any static R plot could be made
-interactive by plotting it on a Gtk device and by having the plotting function
-return information about the location of plot objects on the device.  In
-addition the \Rclass{gEventFun} class was designed so that users could alter
+\Rpackage{RGtk} and \Rpackage{gtkDevice} packages \cite{RGtk} to create
+devices that could respond to events.  Using these packages made creating
+interactive views particularly straightforward because any static R plot could
+be made interactive by plotting it on a Gtk device and by having the plotting
+function return information about the location of plot objects on the device.
+In addition the \Rclass{gEventFun} class was designed so that users could alter
 the response to an event.  The response to an event could be as simple as
@@
 
-\section*{Acknowledgements}
-We are indebted to Dr. Duncan Temple Lang for the creation of the
-\Rpackage{RGtk} and \Rpackage{gtkDevice} packages as well as helpful advice on
-their use.
+%\section*{Acknowledgements}
+%We are indebted to Dr. Duncan Temple Lang for the creation of the
+%\Rpackage{RGtk} and \Rpackage{gtkDevice} packages as well as helpful advice on
+%their use.
 

### 2006-11-21 inst/paper/MVC.tex
Made changes to paper 2.  Still needs some cutting.
git-svn-id: file:///home/git/hedgehog.fhcrc.org/bioconductor/trunk/madman/Rpacks/MVCClass@21240 bc3139a8-67e5-0310-9ffc-ced21a209358

@@
 \end{abstract}
+
+\setlength{\baselineskip}{20pt}
 
@@
 elements on different views.  Hence we considered an approach that 
-created new instances of the MVC that were linked to the parent that 
+created new MVC instances that were linked to the parent MVC that 
 spawned them.
@@
 In particular, our implementation consists of a hierarchy of MVCs,
-stored as a tree.  Each parent can have multiple children, but the
+stored as a tree, as shown in Figure \ref{Fig:Hier}.  Each parent can 
+have multiple children, but the
 children have a single parent.  Communication between parent and
@@
 the hands of the users, should they want it.
+
+\begin{figure}[ht]
+  \begin{center}
+    \includegraphics[height=2.5in, width=4.2in]{Hierarchy2.jpg}
+    \caption{ The Parent-Child Relationship Between \Robject{MVCs}. }
+    \label{Fig:Hier}
+  \end{center}
+\end{figure}
 
@@
 blue label and in Figure \ref{Fig:heatmap} only one gene has a blue label.
-This difference in these heatmaps indicates that views are synchronized based
+The difference in these heatmaps indicates that views are synchronized based
 on the parent-child relationships.  Figure \ref{Fig:heatmap} is the parent of
@@
 
-\section{Methods}
-
-We first give our definitions of linked views, interactive views and
+In this paper we describe an extensible design for creating linked,
+interactive views of linked data sets using an extension of the MVC 
+paradigm.  Section \ref{Sec:Back} gives definitions of the terms used in this
+paper and describes the MVC paradigm along with its
+limitations.  Then Section \ref{Sec:OneMVC} describes our implementation of
+the MVC paradigm and how it was extended to accommodate linked data sets.
+
+\section{Background}
+\label{Sec:Back}
+
+\subsection{Definitions}
+\label{Ssec:Def}
+
+We first give our definitions of linked views, and
 linked data sets.  Our model is typically a statistical data
@@
 pieces of statistical software, such as DataDesk \cite{DataDesk},
-XLisp-Stat \cite{Lisp}, iPlots \cite{iPlots}, and
-GGobi \cite{GGobiMan}.  Implementing linked views is based on the
-model-view-controller (MVC) design \cite{DesignPatterns}, which is a
-widely used and well understood paradigm.
-
-We take a fairly general view of interactivity.  Users can interact via
-a pointing device, such as a mouse, or via key strokes, or any of many
-other types of input devices.  Interactivity means that there is an
-immediate response to some user action.  It is a powerful paradigm and allows
-users to quickly navigate and explore complex documents or complex
-data sets.  As an example, suppose that a user has two views: one view
-is a graph where each node represents a particular set of genes while
-the other view is a heatmap of gene expression data.  If the user
-clicks on a node then the response to this action could be to
-highlight all genes in the heatmap view that correspond to the
-selected node.
+XLisp-Stat \cite{Lisp}, iPlots \cite{iPlots}, and GGobi \cite{GGobiMan}.  
+Implementing linked views is generally based on the MVC paradigm
+\cite{DesignPatterns}, which is widely used and well understood.  We describe
+it in detail in Section \ref{Ssec:MVC}.
+
+%We take a fairly general view of interactivity.  Users can interact via
+%a pointing device, such as a mouse, or via key strokes, or any of many
+%other types of input devices.  Interactivity means that there is an
+%immediate response to some user action.  It is a powerful paradigm and allows
+%users to quickly navigate and explore complex documents or complex
+%data sets.  As an example, suppose that a user has two views: one view
+%is a graph where each node represents a particular set of genes while
+%the other view is a heatmap of gene expression data.  If the user
+%clicks on a node then the response to this action could be to
+%highlight all genes in the heatmap view that correspond to the
+%selected node.
 
@@
 
-Consider Figure \ref{Fig:firstMP}, which provides a cartoon
-description of the steps that occur when a user interacts with a
-view.  In the cartoon there are only two views, one of model A (note 
+\subsection{Model View Controller (MVC) Paradigm}
+\label{Ssec:MVC}
+
+The MVC paradigm \citep{DesignPatterns} is a design that consists of
+three types of objects: the controller, which defines what actions
+occur in response to user input; the view, which consists of displays
+of the data; and the model, which manages the data.  The power of the
+MVC design is that it decouples the views of the model from the model
+itself by creating a subscribe/notify procedure between them. In other
+words, the views subscribe to a particular model and the model must
+notify the views when a change occurs so that the views can be
+updated.  This allows multiple views of the model that are linked, but
+are not aware of each other.  And similarly, the model has no
+particular knowledge of any of the views, just how to notify them of
+changes to the contents of the model.
+
+In the MVC paradigm the components are separated, but they need a 
+method to communicate with each other, which can be implemented using 
+a messaging system.  If we have linked data sets, then Figure
+\ref{Fig:firstMP} displays the steps that are needed to synchronize 
+each model with its views as well as synchronizing the linked data sets 
+when a user interacts with a view.
+In the figure there are only two views, one of model A (note 
 that once a data set is loaded in a MVC, we call it a model) and the
@@
 % are discussed further in Sections \ref{Sec:OneMVC} and
-% \ref{Sec:MultMVC}.
+% \ref{Ssec:MultMVC}.
 
@@
 % Section \ref{Sec:OneMVC} while the implementation of linked \Robject{MVC}
-% objects is discussed in Section \ref{Sec:MultMVC}.  The goals of the software
+% objects is discussed in Section \ref{Ssec:MultMVC}.  The goals of the software
 % discussed in this paper are to create linked views of linked data sets, to
@@
 
-\subsection{Limitations of the MVC Design}\label{Ssec:Limit}
-
-In essence the MVC design assumes that there is one data set, which is
-called the model, that all views reference.  But problems arise when
-the data consist of more than one conceptual grouping.  For example,
-consider data stored in a relational database, such as the data shown in
-Figure~\ref{Fig:DBTab}.  Here the database
+\subsection{Limitations of the MVC Paradigm}\label{Ssec:Limit}
+
+When implementing the MVC paradigm, it is often assumed that there is 
+one data set, which is called the model, that all views reference.  
+But problems arise when the data consist of more than one conceptual 
+grouping.  For example, consider data stored in a relational database, 
+such as the data shown in Figure \ref{Fig:MultMVC}.
+%Figure~\ref{Fig:DBTab}.  
+Here the database
 consists of three tables, State, City and StateCity.  The State table
@@
 states and cities.  In order to support interactive views of all data
-tables, in a standard MVC paradigm, it seems that either all data must
-be combined into a single large table (Figure~\ref{Fig:OneDBTab}), or
+tables, in a standard MVC implementation, it seems that either all 
+data must be combined into a single large table, or
 the processing and accessing of the model data must become much more
-complex.
+complex.  Notice that the MVC paradigm allows a model that is complex, but
+accessing the model data is more straightforward when the data is a single
+conceptual unit.  This single data structure as a model creates a well 
+defined message passing system between the model and its views as well as
+between the model and the controller.
 
@@
 
-\begin{figure}[ht]
-  \begin{center}
-    \subfigure[Example of relational database
-    tables.]{
-    \label{Fig:DBTab}
-    \includegraphics[height=1.2in, width=3.5in]{newDatabaseTables.jpg}}
-    \hspace{.3in}
-    \subfigure[Example of a combined table.]{
-    \label{Fig:OneDBTab}
-    \includegraphics[height=1.2in, width=1.5in]{newOneDatabaseTable.jpg}}
-    \caption{Relational database tables.}
+%\begin{figure}[ht]
+%  \begin{center}
+%    \subfigure[Example of relational database
+%    tables.]{
 %    \label{Fig:DBTab}
-  \end{center}
-\end{figure}
-
-We felt that neither approach had much appeal.  A single large table
-suggests that all data are similar, and still leaves issues about how
+%    \includegraphics[height=1.2in, width=3.5in]{newDatabaseTables.jpg}}
+%    \hspace{.3in}
+%    \subfigure[Example of a combined table.]{
+%    \label{Fig:OneDBTab}
+%    \includegraphics[height=1.2in, width=1.5in]{newOneDatabaseTable.jpg}}
+%    \caption{Relational database tables.}
+%%    \label{Fig:DBTab}
+%  \end{center}
+%\end{figure}
+
+We felt that neither approach of combining linked data into one data 
+structure or of having a complex model had much appeal.  A single large 
+table suggests that all data are similar, and still leaves issues about how
 to deal with ensuring that when, for example, a state is selected all
 cities within it are highlighted.  The second approach requires writing a
-large amount of code, but is unlikely to be extensible.
-We should also mention that our main area of
-application is a setting where one representation of the data is as a
+large amount of code, but is unlikely to be extensible.  One area of 
+interest is a setting where one representation of the data is as a
 graph (with nodes and edges) while the other is as a rectangular array. 
-We can
-envisage lots of different MVCs that operate on graph based data, and
+We can envisage lots of different MVCs that operate on graph based data, and
 equally MVCs that operate on matrix based data.  Specifying a mechanism
@@
 appropriately linked instances of the two classes.  We also argue that
-there are surely other specific types of models that represent
+there are other specific types of models that represent
 specific types of data (e.g. time series) and it would make sense to
-implement these as yet another separate, specialized MVC, that could
-be linked to any one of a variety of other MVCs.  In our design, each 
-MVC will contain one model and thus, linked models also indicate linked 
-MVCs.
+implement these as other specialized MVCs.
 
@@
 
-\section{Extending the MVC Design}
-\label{Sec:Extend}
-
-As discussed in Section \ref{Sec:Intro}, the MVC design typically uses a single
-conceptual unit, such as a data frame or a matrix, for the model.  And
-we believe that such an approach is a good one.  MVCs should be
-specialized to certain types of data, as this allows them to provide
-better and more natural interactions.  However, we additionally argue
-that there is great benefit to considering linking these different
-MVCs when the underlying data themselves are derived from the same source.
-We believe that there will often be a hierarchical relationship
-between the models and the MVCs that contain them.  We have found
-that allowing the user to create new MVCs based on a current MVC is an
-important and useful paradigm.
-
-Each \Robject{MVC} is a coherent object and the interactions between
-\Robject{MVCs} are encapsulated in a set of messages.  We use one
-\Robject{MVC} object to handle each model.  Thus, within a
-\Robject{MVC} object, one can create linked views of that particular
-model, using the standard MVC design.  Between \Robject{MVC}
-objects we propose a second messaging system that allows messages to
-be sent between linked models.
+%\subsection{Extending the MVC Design}
+%\label{Ssec:Extend}
+
+%As discussed in Section \ref{Sec:Intro}, the MVC design typically uses a 
+%single
+%conceptual unit, such as a data frame or a matrix, for the model.  And
+%we believe that such an approach is a good one.  
+%We believe that MVCs should be specialized to certain types of data, such 
+%as a graph or a data frame, as this allows them to provide
+%better and more natural interactions.  However, we additionally argue
+%that there is great benefit to considering linking these different
+%MVCs when the underlying data themselves are derived from the same source.
+%We believe that there will often be a hierarchical relationship
+%between the models and the MVCs that contain them.  We have found
+%that allowing the user to create new MVCs based on a current MVC is an
+%important and useful paradigm.
+
+%Each \Robject{MVC} is a coherent object and the interactions between
+%\Robject{MVCs} are encapsulated in a set of messages.  We use one
+%\Robject{MVC} object to handle each model.  Thus, within a
+%\Robject{MVC} object, one can create linked views of that particular
+%model, using the standard MVC design.  Between \Robject{MVC}
+%objects we propose a second messaging system that allows messages to
+%be sent between linked models.  In Figure \ref{Fig:MultMVC} we show one 
+%view for state-level data and one view for the city-level data.  The state 
+%and city data are linked based on the color.  But there are two distinct 
+%MVCs, not one.  The city data have one MVC and the state data a second.  
+%Even though there are two \Robject{MVC} objects here, the views are linked.
 
@@
 
+%\begin{figure}[ht]
+%  \begin{center}
+%    \includegraphics[height=5in, width=6in]{MultipleMVC.jpg}
+%    \caption{ Example of two \Robject{MVC} objects.  These two \Robject{MVC}
+%      objects are linked by the StateCity table and the color of the points
+%    indicates the linking between the data.  The points referring to the state
+%    of Pennsylvania are red, the points referring to the state of California
+%    are blue, and the points referring to Minnesota are green. }
+%    \label{Fig:MultMVC}
+%  \end{center}
+%\end{figure}
+
+%In Figure \ref{Fig:MultMVC} we show one view for state-level data and
+%one view for the city-level data.  The state and city data are linked
+%based on the color.  But there are two distinct MVCs, not one. 
+%The city data have one MVC and the state data a second.  Even though 
+%there are two \Robject{MVC} objects here, the views are linked.
+
+% Not only does this solution use the power and simplicity of the MVC design, but
+% this multiple MVC design is structured to fit the linking seen in the multiple
+% data sets.  Clearly, the design should be built around the problem and
+% multiple \Robject{MVC} objects with messaging between the \Robject{MVC}
+% objects fits the problem of creating linked views of multiple linked data
+% sets.  In designing an appropriate messaging system between \Robject{MVC}
+% objects, consideration is given to what information should pass between
+% \Robject{MVC} objects and to how messages propagate between \Robject{MVC}
+% objects.  As for the relationship between the \Robject{MVC} objects, it should
+% reflect the relationship between the data sets.
+
+\section{MVC Implementation Details}
+\label{Sec:OneMVC}
+
+We believe that MVCs should be specialized to certain types of data, such 
+as a graph or a data frame, as this allows them to provide
+better and more natural interactions.  However, we additionally argue
+that there is great benefit to considering linking these different
+MVCs when the underlying data themselves are derived from the same source.
+We believe that there will often be a hierarchical relationship
+between the models and the MVCs that contain them.  We have found
+that allowing the user to create new MVCs based on a current MVC is an
+important and useful paradigm.
+
+Each \Robject{MVC} is a coherent object and the interactions between
+\Robject{MVCs} are encapsulated in a set of messages.  We use one
+\Robject{MVC} object to handle each model.  Thus, within a
+\Robject{MVC} object, one can create linked views of that particular
+model, using the standard MVC design.  Between \Robject{MVC}
+objects we propose a second messaging system that allows messages to
+be sent between linked models.  In Figure \ref{Fig:MultMVC} we show one 
+view for state-level data and one view for the city-level data.  The state 
+and city data are linked based on the color.  But there are two distinct 
+MVCs, not one.  The city data have one MVC and the state data a second.  
+Even though there are two \Robject{MVC} objects here, the views are linked.
+
 \begin{figure}[ht]
@@
 
-In Figure \ref{Fig:MultMVC} we show one view for state-level data and
-one view for the city-level data.  The state and city data are linked
-based on the color.  But there are two distinct MVCs, not one. 
-The city data have one MVC and the state data a second.  Even though 
-there are two \Robject{MVC} objects here, the views are linked.
-
-% Not only does this solution use the power and simplicity of the MVC design, but
-% this multiple MVC design is structured to fit the linking seen in the multiple
-% data sets.  Clearly, the design should be built around the problem and
-% multiple \Robject{MVC} objects with messaging between the \Robject{MVC}
-% objects fits the problem of creating linked views of multiple linked data
-% sets.  In designing an appropriate messaging system between \Robject{MVC}
-% objects, consideration is given to what information should pass between
-% \Robject{MVC} objects and to how messages propagate between \Robject{MVC}
-% objects.  As for the relationship between the \Robject{MVC} objects, it should
-% reflect the relationship between the data sets.
-
-\section{MVC Design Details}
-\label{Sec:OneMVC}
-
-\subsection{Implementation Details}
-\label{SSec:OneOver}
-
-We have implemented linked, interactive views of linked data sets
-using an object-oriented programming design which allows for
-extensibility.  We provide this implementation in the
-\Rpackage{MVCClass} package, written in R, which implements the basic
-classes and generic functions.  The \Rpackage{MVCClass} package
-defines the individual model, view, and controller classes as well as
-the \Rclass{MVC} class that binds the individual model, view, and
-controller classes together into one unit.  The package also defines
-the message classes that allow the components to communicate with each
-other.  Thus, the \Rpackage{MVCClass} package contains the classes
-necessary for creating linked views.
+%\subsection{Implementation Details}
+%\label{SSec:OneOver}
+%
+%We have implemented linked, interactive views of linked data sets
+%using an object-oriented programming design which allows for
+%extensibility.  We provide this implementation in the
+%\Rpackage{MVCClass} package, written in R, which implements the basic
+%classes and generic functions.  The \Rpackage{MVCClass} package
+%defines the individual model, view, and controller classes as well as
+%the \Rclass{MVC} class that binds the individual model, view, and
+%controller classes together into one unit.  The package also defines
+%the message classes that allow the components to communicate with each
+%other.  Thus, the \Rpackage{MVCClass} package contains the classes
+%necessary for creating linked views.
 
@@
 % model and receive information from its parent model, respectively (see
-% Sections \ref{Ssec:OneMVC} and \ref{Ssec:MultLink} for more information on how
+% Sections \ref{Ssec:OneMVC} and \ref{Sssec:MultLink} for more information on how
 % parent and child models are linked), the \Rslot{virtualData} slot is data
@@
 model class called \Rclass{tsModel} to represent time series data.
-To make this addition, the user extends the \Rclass{gModel} class in his or
+To make this addition, the user extends the \Rclass{gModel} class, which is
+the general model class, in his or
 her own R software package.  More information on extending the classes can be
-found in the \Rpackage{MVCClass} and \Rpackage{iSNetwork} Vignettes.
+found in the \Rpackage{MVCClass} and \Rpackage{iSNetwork} vignettes, which are
+supplied with the software packages.
 
@@
 and the second task is to control message passing.  Messages, which are
-discussed in Sections \ref{Ssec:OneMess} and \ref{Ssec:MultLink}, are
+discussed in Sections \ref{Ssec:OneMess} and \ref{Sssec:MultLink}, are
 responsible for creating new components, such as models and views, and for
@@
 
-\subsection{MVC}
+\subsection{Putting it all Together in a MVC Class}
 \label{Ssec:OneMVC}
@@
 information to their children and children send information to their parent.
-More information on this communication is given in Section \ref{Sec:MultMVC}.
+More information on this communication is given in Section \ref{Ssec:MultMVC}.
 
@@
 
-\begin{figure}[ht]
-  \begin{center}
-    \includegraphics[height=2.5in, width=4.2in]{Hierarchy2.jpg}
-    \caption{ The Parent-Child Relationship Between \Robject{MVCs}. }
-    \label{Fig:Hier}
-  \end{center}
-\end{figure}
+%\begin{figure}[ht]
+%  \begin{center}
+%    \includegraphics[height=2.5in, width=4.2in]{Hierarchy2.jpg}
+%    \caption{ The Parent-Child Relationship Between \Robject{MVCs}. }
+%    \label{Fig:Hier}
+%  \end{center}
+%\end{figure}
 
@@
 between parent and child \Robject{MVCs} and we discuss this in Section
-\ref{Sec:MultMVC}.
+\ref{Ssec:MultMVC}.
 
 Deciding when new data should be added to the current model versus generating
-a new submodel in a \Robject{MVC} instance is done by comparing the
+a new submodel in a \Robject{MVC} instance should be done by comparing the
 relationship between the new data and the data in the current model.  If the
@@
 %% information about the system as a whole.
-Currently, there are two types of messages within an MVC: an add
+Currently, two types of messages that are passed within an MVC have been
+implemented : an add
 message and an update message.  Both messages request that some
@@
 
-\section{Details of Multiple MVCs}
-\label{Sec:MultMVC}
+\subsection{Details of Multiple MVCs}
+\label{Ssec:MultMVC}
 
@@
 
-\subsection{Linking Models}
-\label{Ssec:MultLink}
+\subsubsection{Linking Models}
+\label{Sssec:MultLink}
 
@@
 
-The message classes for creating a child \Robject{MVC} from an existing
-\Robject{MVC} and for sending messages between \Robject{MVC} objects are shown
-in Figure \ref{Fig:BetwMess}.  The message types for passing information
-between \Robject{MVCs} are similar to the messages
-that are passed within \Robject{MVC} objects.  Now the add child message is
-similar to the add model message, and the send parent message and the send
-child message are similar to the update model message.
-
-\begin{figure}[ht]
-  \begin{center}
-    \includegraphics[height=4in, width=4.5in]{MVCMessageClass2.jpg}
-    \caption{ Inheritance for Message Classes that are Passed Between
-      \Robject{MVCs}. }
-    \label{Fig:BetwMess}
-  \end{center}
-\end{figure}
+%The message classes for creating a child \Robject{MVC} from an existing
+%\Robject{MVC} and for sending messages between \Robject{MVC} objects are shown
+%in Figure \ref{Fig:BetwMess}.  The message types for passing information
+%between \Robject{MVCs} are similar to the messages
+%that are passed within \Robject{MVC} objects.  Now the add child message is
+%similar to the add model message, and the send parent message and the send
+%child message are similar to the update model message.
+
+%\begin{figure}[ht]
+%  \begin{center}
+%    \includegraphics[height=4in, width=4.5in]{MVCMessageClass2.jpg}
+%    \caption{ Inheritance for Message Classes that are Passed Between
+%      \Robject{MVCs}. }
+%    \label{Fig:BetwMess}
+%  \end{center}
+%\end{figure}
 
@@
 
-An example of how these messages are passed and converted is shown in
-Figure \ref{Fig:MessPass}.  These message classes ensure that when a
-model is updated, its parent and child models are notified of the
-change and most importantly are notified of the change in a way that
+An example of how messages are passed and converted between MVCs 
+is shown in Figure \ref{Fig:MessPass}.  The message classes ensure 
+that when a model is updated, its parent and child models are notified 
+of the change and most importantly are notified of the change in a way that
 each model can properly act upon that information.  Because each

### 2008-02-27 inst/paper/paper2.tex
Replaced reference to soon to be defunct class exprSet with ExpressionSet.
git-svn-id: file:///home/git/hedgehog.fhcrc.org/bioconductor/trunk/madman/Rpacks/MVCClass@30240 bc3139a8-67e5-0310-9ffc-ced21a209358

@@
 \Rslot{modelData} slot has class graph, and the \Rclass{exprModel} class
-represents a model where the \Rslot{modelData} slot has class exprSet, a class
-used to represent gene expression data \cite{BioC}.
+represents a model where the \Rslot{modelData} slot has class ExpressionSet,
+a class used to represent gene expression data \cite{BioC}.
 
@@
 \Rslot{type} slot gives the type of the model (currently, the options are
-``exprSet'', ``graph'', or ``data.frame'').  As with the previous
+``ExpressionSet'', ``graph'', or ``data.frame'').  As with the previous
 message classes, the \Rclass{gAddChildMessage} class has two methods:
